### 2019-03-22 icepack.tex
First commit

\documentclass{article}

\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amsthm}
%\usepackage{amssymb}
%\usepackage{mathrsfs}
%\usepackage{fullpage}
%\usepackage{mathptmx}
%\usepackage[varg]{txfonts}
\usepackage{natbib}
\usepackage{color}
\usepackage[charter]{mathdesign}
\usepackage[pdftex]{graphicx}
%\usepackage{float}
%\usepackage{hyperref}
%\usepackage[modulo, displaymath, mathlines]{lineno}
%\usepackage{setspace}
%\usepackage[titletoc,toc,title]{appendix}

%\linenumbers
%\doublespacing

\theoremstyle{definition}
\newtheorem*{defn}{Definition}
\newtheorem*{exm}{Example}

\theoremstyle{plain}
\newtheorem*{thm}{Theorem}
\newtheorem*{lem}{Lemma}
\newtheorem*{prop}{Proposition}
\newtheorem*{cor}{Corollary}

\newcommand{\argmin}{\text{argmin}}
\newcommand{\ud}{\hspace{2pt}\mathrm{d}}
\newcommand{\bs}{\boldsymbol}
\newcommand{\PP}{\mathsf{P}}

\title{\emph{icepack}: a novel glacier flow modeling package}
\author{Daniel Shapero}
\date{}

\begin{document}

\tableofcontents
\newpage

\maketitle

\begin{abstract}
In this paper we introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
Icepack is built on the finite element modeling library firedrake, which implements the domain-specific language UFL for the specification of weak forms of partial differential equations.
The diagnostic models implemented in icepack use variational formulations that are specified in UFL.
Individual components of each model's action functional can be easily substituted for components of the user's choosing, facilitating experimentation with the model physics.
Additionally, many post-processing and analysis tasks on simulation results amount to the evaluation of some functional.
By using a variational formulation of the model physics, the specification of a problem and the analysis of the solution employ the same conceptual vocabulary.
A third advantage of variational principles is that the action functional itself can be used to define a solver convergence criterion that is independent of the mesh and requires little tuning on the part of the user.
Icepack features a 3D diagnostic model based on terrain-following coordinates and vertical spectral discretization.
This model resolves both plug- and shear-flow components of horizontal ice flow with a minimum of computational expense over 2D, depth-averaged models.
Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
The overall design philosophy of icepack is to enable usability for as wide a swathe of the glaciological community as possible, including both experts and novices in numerical PDE solvers and high-performance computing.
\end{abstract}

\section{Introduction}

Numerical modeling has become an essential part of the workflow of glaciologists across all disciplines.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item predicting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
    \item exploring poorly-understood aspects of glacier physics, such as hydrology and calving,
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like.
\end{enumerate}
Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several glacier flow models already exist and are effective in the hands of experts.
These models are usually written in compiled programming languages such as C, C++, and Fortran, which afford an unsurpassed degree of computational speed.
Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for using numerical models to glaciologists who are not experts in modeling as such.
We wanted to make a tool that would be accessible also to non-experts.

\textcolor{red}{We focus in this paper only on novel aspects of icepack that differ substantially from existing packages.}

The glacier flow modeling package closest in spirit to icepack is VarGlaS \citep{brinkerhoff2013data}.
VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}.
The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for specifying weak forms of PDE.
Icepack improves upon the groundwork laid in VarGlaS through its use of tensor product finite elements on extruded meshes, both of which are only available in firedrake \citep{bercea2016structure, mcrae2016automated}.
These two features enable the definition of a much simpler 3D flow model that we will describe in the next section.

The two main components of a glacier flow model are a \emph{diagnostic} and a \emph{prognostic} equation.
The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation.
The inputs to the diagnostic equation are the ice thickness, surface elevation, rheology, ice velocity at the inflow boundary, and the coefficient of friction for contact with the bed and side walls.
The output of the diagnostic equation is the ice velocity throughout the entire region of interest.
The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
The inputs to the prognostic equation are the current value of the ice thickness, the ice velocity, and the surface and basal mass balance.
The output is the ice thickness at a later time.
Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation.

The rheology and friction coefficient may also be described in terms of other fields.
For example, the rheology can be parameterized in terms of the ice temperature, englacial water content, and damage state.
Likewise, the friction coefficient can be described in terms of a geological roughness factor and the subglacial water pressure.
The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.

\section{Diagnostic models}

There are two diagnostic models implemented in icepack.
The shallow stream approximation (SSA) is a 2D model describing the depth-averaged velocity of a fast-flowing grounded ice stream or floating ice shelf \citep{macayeal1989large}.
The SSA model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
Plug flow is a good approximation near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
The first-order or Blatter-Pattyn (BP) approximation, by contrast, is a 3D model describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
The only approximation in the BP model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.

Both the SSA and BP models in icepack are described through \emph{action principles} \citep{dukowicz2010consistent}.
Rather than describe the velocity as the weak solution of a nonlinear partial differential equation, an action principle instead states that the velocity is the minimizer of a certain nonlinear functional, the action.
Most importantly for our application is that the action functional is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge.
The two formulations are mathematically equivalent, but minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation.

Many dissipative physical systems have convex action principles.
In the theory of non-equilibrium thermodynamics, the existence of a convex action functional follows from the Onsager reciprocity relations \citep{de2013non}.
The applicability of this theory, however, hinges on a number of assumptions that are hard to justify without question.
In any case, the existence of an action functional is an immense mathematical convenience.
Depending on one's viewpoint, it may also be a much deeper consequence of thermodynamics.

\subsection{Implementation with firedrake}

\subsection{Substitution of model components}

\subsection{Defining objective convergence criteria}

\section{Prognostic model}

\section{Data assimilation}

\section{Demonstrations}

\section{Testing}

\section{Usability}

\section{Discussion}

\bibliographystyle{plainnat}
\bibliography{icepack.bib}

\end{document}

### 2019-03-22 icepack.tex
Started writing about model substitution

@@
 There are two diagnostic models implemented in icepack.
-The shallow stream approximation (SSA) is a 2D model describing the depth-averaged velocity of a fast-flowing grounded ice stream or floating ice shelf \citep{macayeal1989large}.
+The \emph{shallow stream approximation} (SSA) is a 2D model describing the depth-averaged velocity of a fast-flowing grounded ice stream or floating ice shelf \citep{macayeal1989large}.
 The SSA model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
-Plug flow is a good approximation near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
-The first-order or Blatter-Pattyn (BP) approximation, by contrast, is a 3D model describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
+Plug flow is a good approximation in fast-flowing ice streams and outlet glaciers near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
+The \emph{first-order} or \emph{Blatter-Pattyn} (BP) approximation, by contrast, is a 3D model describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
 The only approximation in the BP model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
@@
 
-\subsection{Substitution of model components}
+\subsection{Substituting model components}
+
+Many aspects of glacier physics are not completely understood.
+For example, the most common description of basal friction assumes that glacier sliding occurs through regelation, in which case the basal shear stress can be written
+\begin{equation}
+    \tau_b = -C|u|^{\frac{1}{m} - 1}u
+\end{equation}
+with $m = 3$ \citep{weertman1957sliding}.
+Many authors have argued that, in fast-flowing regions of the ice sheet, glacier sliding occurs instead by plastic deformation within the subglacial sediments \citep{tulaczyk2000basal}.
+For plastic sliding, the shear stress is dependent on the yield strength of the subglacial sediments and not on the sliding speed, in which case $m = \infty$ would be most appropriate.
+Finally, the Schoof sliding law aims to reconcile the two modes of sliding \citep{schoof2005effect}.
+Below a critical sliding speed, the shear stress is aymptotic to $|u|^{1/m}$ as in the Weertman sliding law, but above the critical speed the shear stress is independent of the sliding speed.
+
+The Weertman and plastic sliding laws possess the same functional form but differ only in the value of a single scalar parameter $m$.
+The Schoof sliding law, on the other hand, has a totally different functional dependence on the velocity.
+Several authors, including Schoof, have proposed that the basal shear stress is also a function of the effective pressure $N = \rho gh - p_w$ within the subglacial hydrological system \citep{budd1979empirical, schoof2005effect}.
+Implementing these more sophisticated mathematical models would require adding an extra argument to the procedure for solving the diagnostic equation.
+
+One of our goals with icepack is to facilitate experimentation with the model physics.
+To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
+For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
+In programming terms this amounts to changing the number of arguments to the function that calculates basal shear stress.
+For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
+In C++ one could obviate the need to modify the diagnostic solve routine by using variadic templates.
+But variadic templates are a more advanced language feature and using them would arguably present a steep learning curve for novices.
+
+In icepack, any given model physics component can be substituted for a parameterization of the user's choosing.
+Every flow model in icepack is represented by a Python object, and each model object is initialized with a set of functions to calculate the various terms of the action functional.
+For example, the ice stream model is initialized with functions to calculate the contributions due to viscosity, basal friction, side wall drag, and gravity.
+Sensible parameterizations are selected by default.
+If the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
+When the diagnostic solve procedure is invoked, all of its arguments are then passed on by keyword to the method that calculates the action functional.
+Any fields that are unnecessary for calculating a given component are simply ignored.
+For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function simply makes no use of these arguments.
+The functional form can be changed entirely and new arguments can be added, with the only restriction that they are passed as keyword and not as positional arguments.
+
+This approach to making the physics models more extensible restricts us to using keyword arguments instead of positional arguments.
+In this case, passing arguments by keyword enhances readability and comprehensibility.
+The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used for each ($C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.).
+The order of the arguments is arbitrary and immaterial.
+For defining, say, the bilinear form that represents a non-symmetric PDE, the order of the arguments does have some inherent mathematical significance.
 

### 2019-03-22 icepack.tex
Wrote subsection on action principle + convergence criteria

@@
 Both the SSA and BP models in icepack are described through \emph{action principles} \citep{dukowicz2010consistent}.
-Rather than describe the velocity as the weak solution of a nonlinear partial differential equation, an action principle instead states that the velocity is the minimizer of a certain nonlinear functional, the action.
-Most importantly for our application is that the action functional is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
-Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge.
-The two formulations are mathematically equivalent, but minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation.
+Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a certain nonlinear functional, called the action.
+The action consists of four terms:
+\begin{align}
+    & \text{action} = \iint\text{stress} \times \text{strain rate}\ud z\ud x - \int\text{basal friction} \times \text{sliding velocity}\ud x \nonumber \\
+    & \quad - \iint\text{surface slope}\times\text{velocity}\ud z\ud x - \iint\text{ocean pressure}\times\text{velocity}\ud z\ud \gamma
+    \label{action-functional}
+\end{align}
+where $\ud z\ud x$ denotes integration over the entire glacier, $\ud x$ denotes integration over the glacier footprint, and $\ud z\ud\gamma$ over the side wall boundary.
+The action has units of power (energy/time) and can be related to the rate of decrease of the thermodynamic free energy.
+Moreover, the energy lost to viscous and frictional heating can be calculated from the action and its Legendre transform.
+The action principle can be viewed as a consequence of the Onsager reciprocity relations for systems near to equilibrium \citep{de2013non}.
 
-Many dissipative physical systems have convex action principles.
-In the theory of non-equilibrium thermodynamics, the existence of a convex action functional follows from the Onsager reciprocity relations \citep{de2013non}.
-The applicability of this theory, however, hinges on a number of assumptions that are hard to justify without question.
-In any case, the existence of an action functional is an immense mathematical convenience.
-Depending on one's viewpoint, it may also be a much deeper consequence of thermodynamics.
+Irrespective of any deep connections with thermodynamics, the existence of an action principle is a tremendous convenience for numerical analysis.
+For viscous flow problems near to steady-state, the action is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
+Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge from any initial guess.
+Minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation, although the two formulations are equivalent.
 
@@
 
-\subsection{Defining objective convergence criteria}
+\subsection{Convergence criteria}
+
+Calculating the ice velocity from the geometry and other input fields involves minimizing a convex functional.
+The minimizer can only be approximated, so we employ iterative procedures based on Newton's method with a line search.
+We must then choose some metric to decide when the current guess for the velocity is good enough to stop the iteration.
+
+A convergence criterion that works equally well independent of the mesh, finite element discretization, and the quality of the initial guess can be defined based on the idea of the \emph{Newton decrement} \citep{nocedal2006numerical}.
+The search direction $v_k$ at step $k$ for Newton's method is defined as
+\begin{equation}
+    v_k = -\ud^2J(u_k)^{-1}\ud J(u_k).
+\end{equation}
+Since the second derivative operator $\ud^2J(u_k)$ is positive-definite, this is a descent direction for the action $J$:
+\begin{equation}
+    \ud J(u_k)\cdot v_k < 0.
+\end{equation}
+The absolute value of the quantity in the last equation is defined as the Newton decrement.
+For $u_k$ sufficiently close to the true solution $u$, the Newton decrement roughly tells us how much we can expect the action to decrease:
+\begin{equation}
+    J(u_k) - J(u) \approx \frac{1}{2}|\ud J(u_k)\cdot v_k|.
+\end{equation}
+We can then use the Newton decrement to decide when to stop the iteration.
+
+As shown in equation \eqref{action-functional}, the action has units of power and is the sum of the dissipation due to viscosity, friction, gravitational driving, and restrain by ocean pressure at the terminus.
+The viscous and frictional terms are convex and positive functions of the velocity, while the gravitational and terminus stress terms are linear in the velocity and can be of either sign.
+If we define the \emph{scale functional}
+\begin{equation}
+    K(u) = \text{viscous dissipation} + \text{frictional dissipation}
+\end{equation}
+as only the positive parts of the action, then the convergence criterion
+\begin{equation}
+    |\ud J(u_k)\cdot v_k| < \epsilon K(u_k)
+\end{equation}
+is independent of the discretization.
+The intuition behind this criterion is that iteration is halted when the expected decrease in the action functional is much smaller than the positive part of the action itself.
+
+We have found empirically that, with this criterion and the Newton solver implementation in icepack, the iteration usually converges to machine precision in around 8 steps.
+For exceptionally bad initial guesses the iteration count can reach as high as 20 but rarely more.
+We also observe the expected doubling of the number of accurate digits in the value of the action once the velocity guesses are within the convergence basin of the true solution.
+Other convergence criteria, such as using relative change in the velocity guesses, can terminate prematurely when the initial guess is very far outside the quadratic convergence basin.
+
+The numerical solvers in icepack have been designed so that users who are not familiar with numerical optimization need not be confronted with a possibly bewildering array of algorithmic parameters.
+Consequently, sensible defaults have been chosen for the Armijo and Wolfe criteria \citep{nocedal2006numerical}, and the tolerance for the line search is chosen based on that of the outer-level Newton iteration.
+The Newton search direction is calculated using a direct factorization solver rather than, say, the conjugate gradient algorithm, as the use of another iterative method would introduce yet another algorithmic parameter.
+The existing algorithmic parameters can nonetheless be changed from their defaults by passing keyword arguments for advanced users who are interested in performance optimization.
 

### 2019-03-23 icepack.tex
Bash on mixed-language projects

@@
 
-This approach to making the physics models more extensible restricts us to using keyword arguments instead of positional arguments.
-In this case, passing arguments by keyword enhances readability and comprehensibility.
-The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used for each ($C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.).
+We are restricted to using keyword arguments instead of positional arguments by adopting this approach.
+For the particular case of passing several fields to a physics solver routine, using only keyword arguments enhances readability and comprehensibility.
+The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used in the literature -- $C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.
 The order of the arguments is arbitrary and immaterial.
-For defining, say, the bilinear form that represents a non-symmetric PDE, the order of the arguments does have some inherent mathematical significance.
+For defining, say, the bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance and our design would be inappropriate.
+
+One of our main design goals for icepack is to enable users to ``plug and play'' their own physics.
+Since the physics of glacier flow are not understood completely, this focus on extensibility is essential for glaciological modeling \textcolor{red}{reword somehow}.
+Writing the entire package in Python was necessary to meet this goal.
+Many scientific software packages consist of a highly-optimized core library in C, C++, or Fortran, together with a wrapper that exposes this library to an interpreted language like Python.
+This approach in principle offers the best of both worlds -- the runtime efficiency of a compiled language with the convenience and interactivity of an interpreted one.
+Many practitioners, however, will use such a library entirely through the interpreted surface layer and have no familiarity with the language in which the fast core was written.
+These users then have a steep learning curve to overcome when they need to modify functionality that lives in the core library, as they now need to learn an entirely new programming language.
+Some low-level kernel code is always necessary for runtime efficiency of course.
+With Firedrake, this responsibility is deferred to a code generator of which the average user remains blissfully unaware.
 

### 2019-03-24 icepack.tex
Wrote some about prognostic solver

@@
 One of our main design goals for icepack is to enable users to ``plug and play'' their own physics.
-Since the physics of glacier flow are not understood completely, this focus on extensibility is essential for glaciological modeling \textcolor{red}{reword somehow}.
+Since many aspects of glacier flow are not understood completely, we view extensibility as an essential requirement for any software package in this domain.
 Writing the entire package in Python was necessary to meet this goal.
@@
 
+The PDE describing how the ice thickness changes in time is a simple conservative advection equation:
+\begin{equation}
+    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
+\end{equation}
+where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
+The thickness is represented using continuous Galerkin basis functions.
+The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
+The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.
+
+The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
+Other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \textcolor{red}{cite ISSM, VarGlaS}.
+The SUPG method is explicit in time and conditionally stable \textcolor{red}{cite Donea and Huerta?}.
+This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
+The implicit Euler scheme requires solving a more complicated linear system, but the extra computational cost comes with the advantage of unconditional stability.
+Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
+The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
+The extra cost of using an implicit time discretization for the mass transport equation is dwarfed by the cost of the diagnostic solve in any case.
+Advanced users who are interested in maximizing performance can subclass the mass transport solver to implement a faster explicit scheme.
+
+The implicit Euler scheme introduces a large degree of artificial diffusion \textcolor{red}{cite Donea and Huerta}.
+Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
+The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.
+Other problems within glaciology have more of a hyperbolic character and there the argument for using shock-capturing methods is much stronger.
+For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \textcolor{red}{cite Albrecht and Levermann}.
+The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge Kutta method would be more appropriate.
+
 \section{Data assimilation}

### 2019-03-25 icepack.tex
Add description of hybrid model

@@
 
+At the highest level, each physics model in icepack is represented by a Python class.
+So far, the models available are \texttt{IceShelf}, \texttt{IceStream}, and \texttt{HybridModel}.
+Each class has three main responsibilities.
+First, the model class has to have a method that takes in the various fields (ice velocity, thickness, surface elevation, etc.) and returns a symbolic representation of the action functional for that particular model.
+Second, the model class must have a method that takes in the input fields and returns an approximate minimizer of the action functional.
+The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the first method calculates.
+The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
+Finally the model must have a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
+The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.
+
+\textcolor{red}{Some sample code.}
+
+The Unified Form Language for specifying weak forms of PDEs contains all of the primitives necessary to express individual terms of the action functional.
+These primitives largely consist of the basic vector calculus operators like the gradient of a field, tensor calculus operations like taking the dot product of two vectors or tensors, scalar functions like the square root or exponential, and symbolic integration over the mesh or its boundary.
+For example, the strain rate for a given velocity field $u$ can be written as \texttt{sym(grad(u))}, where the function \texttt{grad} represents the symbolic gradient of a field and \texttt{sym} represents the symmetrization of a rank-2 tensor.
+
+\textcolor{red}{Brief discussion of SSA model.}
+
+The 3D flow model, as represented in the \texttt{HybridModel} class, uses several features that are only available in Firedrake to better exploit the underlying mathematics.
+First, rather than represent the 3D physical domain itself, we instead use \emph{terrain-following coordinates} and \emph{extruded meshes}.
+The terrain-following vertical coordinate $\zeta$ is defined as
+\begin{equation}
+    \zeta = \frac{z - b}{h}
+\end{equation}
+where $b$ is the ice base.
+We can then think of the computational domain as the Cartesian product of a 2D footprint domain $\Omega$ and the unit interval $[0, 1]$.
+
+Both the bed elevation and thickness depend on $x$ and $y$.
+As a result, the formula for the horizontal gradient of a field in terrain-following coordinates includes an additional geometric correction factor.
+Letting $\nabla_z$ and $\nabla_\zeta$ denote the horizontal gradient with respectively $z$ and $\zeta$ held constant, the chain rule gives us that
+\begin{equation}
+    \nabla_zq = \nabla_\zeta q + \frac{\partial q}{\partial\zeta}\nabla\zeta,
+\end{equation}
+where we can calculate the spatial gradient of $\zeta$ as
+\begin{equation}
+    \nabla\zeta = -h^{-1}\left\{(1 - \zeta)\nabla b + \zeta\nabla s\right\}.
+\end{equation}
+Likewise, the strain rate of a vector field can be expressed as
+\begin{equation}
+    \dot\varepsilon_z(u) = \dot\varepsilon_\zeta(u) + \frac{1}{2}\left(u \otimes\nabla\zeta + \nabla\zeta\otimes u\right)
+\end{equation}
+where $\otimes$ is the tensor product of two vectors.
+The geometric correction factor can easily be represented in UFL.
+By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is no more complex than in Cartesian coordinates.
+
+Terrain-following coordinates are especially advantageous on extruded meshes.
+An extruded mesh takes a 2D footprint mesh and lifts it into 3D by taking the Cartesian product with the unit interval.
+The cells of an extruded mesh are triangular prisms instead of the more common tetrahedra used for general 3D meshes.
+In Firedrake, an extruded mesh is created by calling the function \texttt{ExtrudedMesh} on the 2D footprint.
+Not every 3D domain can be described by extruding a 2D domain, but the geometry of most glacier flow problems can.
+
+A 2D mesh can be extruded to any number of vertical layers.
+In Elmer/Ice, for example, the user chooses the number of vertical layers in order to set the resolution along this axis \citep{gagliardini2013capabilities}.
+The horizontal velocity is expected to be very smooth as a function of depth.
+For example, under the plug flow approximation, the horizontal velocity is constant with depth.
+Under the shallow ice approximation, the horizontal velocity varies with depth as $1 - (1 - \zeta)^{n + 1}$ where $n = 3$ is the Glen flow law exponent.
+This extra information about our solution suggests a different discretization strategy.
+
+Rather than extrude the footprint mesh into many vertical layers, we might instead use only one vertical layer and increase the polynomial degree in the vertical direction only to obtain higher resolution.
+To support this use case, Firedrake includes the ability to define \emph{tensor product} elements on extruded meshes.
+Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
+\begin{equation}
+    \Phi_{kl}(x, y, \zeta) = \phi_k(x, y)\psi_l(\zeta).
+\end{equation}
+For example, we can use piecewise linear or quadratic elements on triangles and use quintic or higher degree polynomials in the vertical.
+Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
+Rather than use the usual Lagrange interpolating polynomial basis in the vertical dimension, we can instead use the Legendre polynomial basis.
+The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.
+To select the Legendre polynomial basis, the user passes the keyword argument \texttt{vfamily=`Gauss-Legendre'} or \texttt{`GL'} for short to the constructor for the function space.
+
+The combination of using extruded meshes and tensor product elements in the vertical direction is referred to as \emph{semi-discretization}.
+This approach can be thought of merely as a way to discretize a PDE that has special structure.
+Alternatively, we can view vertical semi-discretization as defining a family of models indexed by the number of vertical basis functions.
+The order-$d$ model defines a coupled system of PDEs for $d$ vector fields.
+Each vector field represents one mode of vertical variability, similar to the distinction between barotropic and baroclinic modes in atmospheric physics and oceanography.
+The system is then discretized in the horizontal and solved numerically.
+In any case, the code is the same regardless of how one wishes to think about the underlying mathematics.
+
+We are then faced with the choice of how many vertical basis functions to use.
+Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
+The degree 1 model is the most minimalistic model that still exhibits vertical shear.
+Going to higher and higher degree gives a more accurate approximation at the expense of greater computational effort.
+
+This approach for implementing a 3D flow model works completely seamlessly but for one important detail.
+The backpressure from ocean water at the calving front of a marine-terminating glacier is not a smooth function of depth.
+The pressure is 0 above the water line and linearly increasing below it:
+\begin{equation}
+    \text{backpressure power} = \int_\Gamma\int_0^1 \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+\ud\zeta\ud\gamma,
+    \label{backpressure}
+\end{equation}
+where $\zeta_{\text{sl}}$ denotes the relative depth to the water line and the subscript $+$ denotes the positive part of a real number.
+Were we to use the standard asssembly procedure in Firedrake to evaluate this integral, we would get an inaccurate result due to an insufficient number of integration points.
+The resulting velocity solutions are then wildly inaccurate due to the mis-specification of the Neumann boundary condition.
+A blunt solution to this problem would be to pass an extra argument to the Firedrake form compiler that specifies a much greater integration accuracy in the vertical for this term.
+This fix reduces, but does not eliminate, the errors in the velocities, and it incurs an appreciable computational cost.
+
+We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
+The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
+Since the result of this symbolic Legendre polynomial expansion is the same upon every invocation, we can save the effort of recalculating it every time by caching the results.
+The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts \textcolor{red}{figure}.
+While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property.
+Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the expansion limit.
+
+The exact symbolic integration approach is both faster and more accurate than using a large number of quadrature points.
+The same technique could be used to exactly calculate the ocean backpressure for any model, say the full Stokes equations, using terrain-following coordinates together with a Legendre polynomial expansion in the vertical.
+
+For the Stokes equations, transforming the vertical coordinate also circumvents the issue of how to enforce the partial Dirichlet boundary condition $u\cdot \nu = 0$ at the ice base, where $\nu$ is the unit outward normal vector.
+This boundary condition is difficult to impose exactly because the unit outward normal vector $\nu$ is defined on mesh faces while the velocity is defined at mesh vertices.
+Elmer/Ice uses an ad-hoc procedure to define the unit normal vectors at mesh nodes \citep{gagliardini2013capabilities}.
+With a transformation to terrain-following coordinates, we can just set the terrain-following vertical velocity $\omega$ to be 0 at the ice base to impose this boundary condition.
+
+We also argue that answers expressed in terrain-following coordinates are more intuitive in some respects than in Cartesian coordinates.
+At the bed of a grounded glacier, the vertical velocity in Cartesian coordinates is
+\begin{equation}
+    w = -\dot a_b + u\cdot\nabla b.
+\end{equation}
+Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
+In other words it is not immediately clear whether the vertical velocity is more a result of basal mass balance or of geometry without more information.
+By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.
+
+\textcolor{red}{Speed of hybrid model.}
+One of the main advantages of this hybrid model is the ease with which a more accurate solution can be bootstrapped from a less accurate solution, since the two fields are defined on exactly the same geometry.
+For example, if we only have a very crude initial guess for the solution, we can obtain a degree-1 solution relatively cheaply.
+The resulting velocity field is then used as a much more informed initial guess for a degree-5 solution.
+\textcolor{red}{Benchmark.}
+Since the vertical basis functions are orthogonal, the projection of the degree-1 solution into the degree-5 function space can be calculated by simply copying array elements.
+By contrast, if we tried to use the same strategy to bootstrap a solution with a small number of vertical layers to a solution with a larger number of vertical layers, we would also have to interpolate values to the fine layers.
+
+Finally, we note that this approach was first described in \citet{bassis2010hamilton}.
+This work focused solely on flowband models and did not consider the problem of how to calculate the power dissipated at the calving front.
+Moreover, this work used only the two vertical basis functions $\psi_1(\zeta) = 1$, $\psi_2(\zeta) = 1 - (1 - \zeta)^{n + 1}$, the idea being that the vertical profile can be written as a direct superposition of the plug flow and shallow ice approximation modes.
+By contrast, we leave the number of vertical basis functions up to the user.
+This flexibility enables the multigrid-type approach described above.
+
 \subsection{Substituting model components}
@@
 
-\subsection{Convergence criteria}
+\subsection{Convergence criteria} \label{sec:convergence-criteria}
 
@@
 
-\section{Prognostic model}
+\section{Prognostic model} \label{sec:prognostic-model}
 

### 2019-03-26 icepack.tex
Started writing about inverse problems

@@
 
+Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
+The class \texttt{InverseProblem} represents the specification of the inverse problem.
+To specify an inverse problem, the user must provide:
+\begin{itemize}
+    \item the model object and the method that solves the diagnostic equation,
+    \item the objective and regularization functionals,
+    \item the observed field and the name of the argument to the diagnostic solver,
+    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver,
+    \item extra data passed to the diagnostic solver such as boundary conditions.
+\end{itemize}
+At present, this class assumes that the observed state is always the ice velocity, but in principle the same design would suffice for more complicated inverse problems.
+More importantly, the inverse problem is flexible enough to account for the fact that we might have substituted in our own parameterization for the rheology or friction coefficient.
+For example, nearly all studies in the literature use an alternate parameterization of these coefficients in terms of some auxiliary field in order to guarantee positivity \textcolor{red}{citations}.
+This information is passed to the inverse problem by specifying the names of the keyword arguments for the observed state and the field we wish to observe.
+
+The class \texttt{InverseSolver} is an abstract interface for classes that actually solve the inverse problem.
+There are three implementations of this interface in icepack: \texttt{GradientDescentSolver}, \texttt{BFGSSolver}, and \texttt{GaussNewtonSolver}.
+All of these classes are based around the general idea of first computing a search direction and then performing a line search in every iteration.
+They differ in how the search direction is computed.
+The gradient descent solver uses the search direction
+\begin{equation}
+    \phi_k = -M^{-1}\ud J(\theta_k)
+\end{equation}
+where $M$ is the finite element mass matrix.
+Gradient descent is a popular choice because the objective functional is always decreasing along this search direction.
+However, the search direction can be poorly scaled to the physical dimensions of the problem at hand.
+This method can be very expensive and brittle in the initial iterations and often takes many steps to converge.
+The BFGS method uses the past $m$ iterations of the algorithm to compute a low-rank approximation to the inverse of the objective functional's Hessian matrix; see \citet{nocedal2006numerical} for a more in-depth discussion.
+The BFGS method converges faster than gradient descent but suffers from many of the same brittleness issues in the initial iterations before it has accumulated enough history.
+Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian.
+Each iteration of Gauss-Newton is more expensive than that of BFGS or gradient descent because it requires the solution of a more complex linear system than just the mass matrix.
+The Gauss-Newton method converges fastest by far in virtually every test case we have found, in some instances by up to factor of 50.
+
+The derivative of the objective functional with respect to the unknown parameter is calculated using the symbolic differentiation features of Firedrake.
+The user does not need to provide any routines for the derivatives, only the symbolic form of the error metric and the regularization functional.
+The model object is responsible for providing the symbolic form of the action functional.
+
+The bootstrapping approach used to calculate higher-order solutions to the hybrid model from lower-order solutions rather than from a cold initial guess can also be used for inverse problems.
+First, we estimate the unknown parameter using only degree-1 vertical basis functions for the velocity field, which captures most of the spatial variability of the unknown parameter.
+The resulting coefficient is then used as the initial guess with higher degree vertical basis functions for the velocity \textcolor{red}{benchmark}.
+
 \section{Demonstrations}

### 2019-03-27 icepack.tex
References for section on prognostic solver

@@
 The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for using numerical models to glaciologists who are not experts in modeling as such.
-We wanted to make a tool that would be accessible also to non-experts.
+We wanted to make a tool that would be both accessible to non-experts and useful for experts.
 
@@
 The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
-Other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \textcolor{red}{cite ISSM, VarGlaS}.
-The SUPG method is explicit in time and conditionally stable \textcolor{red}{cite Donea and Huerta?}.
+Other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
+The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
 This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
-The implicit Euler scheme requires solving a more complicated linear system, but the extra computational cost comes with the advantage of unconditional stability.
+The implicit Euler scheme requires solving a non-symmetric linear system, but the extra computational cost comes with the advantage of unconditional stability.
 Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
@@
 
-The implicit Euler scheme introduces a large degree of artificial diffusion \textcolor{red}{cite Donea and Huerta}.
+The implicit Euler scheme tends to diffuse out sharp discontinuities that may be present in the true solution \citep{donea2003finite}.
 Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
@@
 Other problems within glaciology have more of a hyperbolic character and there the argument for using shock-capturing methods is much stronger.
-For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \textcolor{red}{cite Albrecht and Levermann}.
+For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \citep{albrecht2014fracture}.
 The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge Kutta method would be more appropriate.
@@
 
+One of the main goals for icepack was to create a tool that is accessible to researchers who might not be experts in scientific computing.
+Previous work on numerical modeling of glacier flow has focused largely on technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory, does it scale to large numbers of processors, can models accurately predict grounding line retreat, etc.
+The fallible human learning to use the model is largely absent from the discussion.
+
+The field of \emph{human-computer interaction} (HCI) asks how we can design software that is easiest to learn and use effectively.
+In the following, we will describe some of the design choices in icepack and how they relate to what HCI researchers call the \emph{cognitive dimensions of notations}.
+\citet{green1996usability} introduced this concept to assess the usability of visual programming languages, but the criteria they laid out in this paper have been used to analyze software systems across many disciplines.
+
+\textbf{Consistency: After a user learns part of the software, can they guess the remaining parts?}
+Each of the model objects in icepack is a class with a method ending in \texttt{solve} that takes in keyword arguments for the various input fields and options for things like boundary conditions.
+Users already familiar with, say, the \texttt{IceStream} class can then use the \texttt{HeatTransport} class under the assumption that the input fields -- the current temperature $T$, ice velocity $u$, and basal melt rate $m$ -- are passed as keyword arguments with the same name as the fields themselves.
+This obviates the need for consulting the documentation or examples.
+
+\textbf{Progressive evaluation: How easily can users get feedback during their use of the software?}
+Progressive evaluation is the main advantage of having a user interface in Python or another interpreted language, as opposed to using compiled programs that must run all at once.
+Icepack includes a suite of demos that demonstrate all of the main functions of the package.
+These demos are formatted as Jupyter notebooks, a document format that includes code, figures, and explanatory text with typeset mathematics that can be run interactively in a web browser \citep{kluyver2016jupyter}.
+A notebook is a sequence of cells, each of which consists of either code or text.
+The code cells can be executed one-at-a-time and this enables users to insert diagnostics in each cell as they go.
+Most importantly, the notebook format also integrates the plotting library matplotlib so that figures created in each cell appear directly beneath that cell as it is executed.
+The ability to visualize the current state of the simulation at every step is a powerful tool for debugging and sanity-checking.
+\textcolor{red}{Picture of an executed jupyter notebook.}
+Finally, a Jupyter notebook and the accompanying plots can be rendered into HTML and published using the tool \texttt{nbconvert}.
+
+\textbf{Abstraction gradient: What are the levels of abstraction exposed by the library?
+Can irrelevant details be hidden?}
+The API for icepack has been designed so that the user only needs to make decisions about what problem they wish to solve and not how to solve it.
+Where a choice concerns more the ``how'' than the ``what'', a sensible default has been chosen.
+Moreover, the default behavior we choose always biases for correctness rather than speed.
+For example, the Newton solver uses a direct factorization method to solve the linear system for the search direction because factorization requires no tuning whereas iterative methods do.
+A user interested in achieving greater runtime performance can pass additional keyword arguments instead specifying, say, the conjugate gradient method.
+This choice is of interest mostly to advanced users so we keep the linear solver method as a default argument.
+In so doing, we avoid confronting novice users with an option that they might not understand.
+
+Advanced users who wish to tune solver performance for large simulations will need some way to make choices about algorithms.
+For example, a more advanced user might choose the GMRES iterative solver together with an incomplete LU preconditioner to solve linear systems.
+The solver classes, as opposed to the model classes, provide the interface for making these kinds of algorithmic choices.
+While alternative solvers might offer faster runtime performance than direct factorization, they also requires making additional choices choices -- how often to restart GMRES?
+How much fill-in to allow in the incomplete factorization?
+Many glaciologists might not have the background in computational science to know that adjusting these parameters could make the difference between solver convergence or breakdown.
+\textcolor{red}{Also stabilization of the Stokes equations.}
+
 \section{Discussion}

### 2019-04-06 icepack.tex
More subsections

@@
 
+\subsection{Terrain-following coordinates and extruded meshes}
+
 The 3D flow model, as represented in the \texttt{HybridModel} class, uses several features that are only available in Firedrake to better exploit the underlying mathematics.
-First, rather than represent the 3D physical domain itself, we instead use \emph{terrain-following coordinates} and \emph{extruded meshes}.
+Rather than represent the 3D physical domain itself, we instead use \emph{terrain-following coordinates} and \emph{extruded meshes}.
 The terrain-following vertical coordinate $\zeta$ is defined as
@@
 
-A 2D mesh can be extruded to any number of vertical layers.
-In Elmer/Ice, for example, the user chooses the number of vertical layers in order to set the resolution along this axis \citep{gagliardini2013capabilities}.
+For the Stokes equations, this alternative coordinate system also helps avoid the problem of how to enforce the condition $u\cdot \nu = -\dot a_b$ at the ice base, where $\nu$ is the unit outward normal vector and $\dot a_b$ is the basal mass balance.
+This boundary condition is difficult to impose exactly because the unit outward normal vector $\nu$ is defined on mesh faces while the velocity is defined at mesh vertices.
+Elmer/Ice uses an ad-hoc procedure to define the unit normal vectors at mesh nodes \citep{gagliardini2013capabilities}.
+With a transformation to terrain-following coordinates, we can just set the terrain-following vertical velocity $\omega$ to be 0 at the ice base to impose this boundary condition.
+
+We also argue that answers expressed in terrain-following coordinates are more intuitive in some respects than in Cartesian coordinates.
+At the bed of a grounded glacier, the vertical velocity in Cartesian coordinates is
+\begin{equation}
+    w = -\dot a_b + u\cdot\nabla b.
+\end{equation}
+Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
+In other words it is not immediately clear whether the vertical velocity is more a result of basal mass balance or of geometry without more information.
+By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.
+
+\subsection{Vertical spectral elements}
+
+Having chosen to use an extruded mesh, we must then decide how many vertical layers to use.
+In Elmer/Ice, for example, the number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.
 The horizontal velocity is expected to be very smooth as a function of depth.
@@
 
-Rather than extrude the footprint mesh into many vertical layers, we might instead use only one vertical layer and increase the polynomial degree in the vertical direction only to obtain higher resolution.
+Rather than extrude the footprint mesh into many vertical layers, we might instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
 To support this use case, Firedrake includes the ability to define \emph{tensor product} elements on extruded meshes.
@@
 
-We are then faced with the choice of how many vertical basis functions to use.
+The user then has to decide how many vertical modes are enough.
 Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
@@
 The same technique could be used to exactly calculate the ocean backpressure for any model, say the full Stokes equations, using terrain-following coordinates together with a Legendre polynomial expansion in the vertical.
-
-For the Stokes equations, transforming the vertical coordinate also circumvents the issue of how to enforce the partial Dirichlet boundary condition $u\cdot \nu = 0$ at the ice base, where $\nu$ is the unit outward normal vector.
-This boundary condition is difficult to impose exactly because the unit outward normal vector $\nu$ is defined on mesh faces while the velocity is defined at mesh vertices.
-Elmer/Ice uses an ad-hoc procedure to define the unit normal vectors at mesh nodes \citep{gagliardini2013capabilities}.
-With a transformation to terrain-following coordinates, we can just set the terrain-following vertical velocity $\omega$ to be 0 at the ice base to impose this boundary condition.
-
-We also argue that answers expressed in terrain-following coordinates are more intuitive in some respects than in Cartesian coordinates.
-At the bed of a grounded glacier, the vertical velocity in Cartesian coordinates is
-\begin{equation}
-    w = -\dot a_b + u\cdot\nabla b.
-\end{equation}
-Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
-In other words it is not immediately clear whether the vertical velocity is more a result of basal mass balance or of geometry without more information.
-By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.
 
@@
 In C++ one could obviate the need to modify the diagnostic solve routine by using variadic templates.
-But variadic templates are a more advanced language feature and using them would arguably present a steep learning curve for novices.
+But variadic templates are a more advanced language feature and relying on them would arguably present a steep learning curve for novices.
 
@@
 
-We are restricted to using keyword arguments instead of positional arguments by adopting this approach.
+We are restricted to using keyword arguments instead of positional arguments in adopting this approach.
 For the particular case of passing several fields to a physics solver routine, using only keyword arguments enhances readability and comprehensibility.
@@
 is independent of the discretization.
-The intuition behind this criterion is that iteration is halted when the expected decrease in the action functional is much smaller than the positive part of the action itself.
+The intuition behind this criterion is that the iteration is halted when the expected decrease in the action functional is much smaller than the positive part of the action itself.
 
@@
 The Newton search direction is calculated using a direct factorization solver rather than, say, the conjugate gradient algorithm, as the use of another iterative method would introduce yet another algorithmic parameter.
-The existing algorithmic parameters can nonetheless be changed from their defaults by passing keyword arguments for advanced users who are interested in performance optimization.
+Advanced users who are interested in performance optimization can change these algorithmic parameters by passing extra arguments to the solve procedure.
 
@@
 This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
-The implicit Euler scheme requires solving a non-symmetric linear system, but the extra computational cost comes with the advantage of unconditional stability.
+The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.
+
 Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.

### 2019-04-06 icepack.tex
Some references

@@
 This flexibility enables the multigrid-type approach described above.
+\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
+\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem via finite differences in the vertical and finite elements in the horizontal, but did not use terrain-following coordinates.
+This work also adopts the viewpoint that semi-discretization defines a family of models, rather than merely a problem-specific discretization strategy.
 

### 2019-08-15 icepack.tex
Assorted corrections

@@
 In this paper we introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
-Icepack is built on the finite element modeling library firedrake, which implements the domain-specific language UFL for the specification of weak forms of partial differential equations.
-The diagnostic models implemented in icepack use variational formulations that are specified in UFL.
-Individual components of each model's action functional can be easily substituted for components of the user's choosing, facilitating experimentation with the model physics.
-Additionally, many post-processing and analysis tasks on simulation results amount to the evaluation of some functional.
+Icepack is built on the finite element modeling library firedrake, which uses the domain-specific language UFL for the describing of weak forms of partial differential equations.
+The diagnostic models in icepack are formulated through action principles that are specified in UFL.
+The components of each action functional can be substituted for different forms of the user's choosing, which makes it easy to experiment with the model physics.
+Many post-processing and analysis tasks on simulation results also amount to the evaluation of some functional.
 By using a variational formulation of the model physics, the specification of a problem and the analysis of the solution employ the same conceptual vocabulary.
@@
 Icepack features a 3D diagnostic model based on terrain-following coordinates and vertical spectral discretization.
-This model resolves both plug- and shear-flow components of horizontal ice flow with a minimum of computational expense over 2D, depth-averaged models.
+This model resolves both plug- and shear-flow components of horizontal ice flow with a minimum of extra computational expense over 2D, depth-averaged models.
 Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
-The overall design philosophy of icepack is to enable usability for as wide a swathe of the glaciological community as possible, including both experts and novices in numerical PDE solvers and high-performance computing.
+The overall design philosophy of icepack is to be as usable as possible for a wide a swathe of the glaciological community, including both experts and novices in computational science.
 \end{abstract}
@@
     \item predicting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
-    \item exploring poorly-understood aspects of glacier physics, such as hydrology and calving,
+    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood,
     \item estimating unobservable quantities, such as bed friction or rheology, from observational data, and
@@
 Several glacier flow models already exist and are effective in the hands of experts.
-These models are usually written in compiled programming languages such as C, C++, and Fortran, which afford an unsurpassed degree of computational speed.
+These models are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
 Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
-The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for using numerical models to glaciologists who are not experts in modeling as such.
-We wanted to make a tool that would be both accessible to non-experts and useful for experts.
+The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in numerical modeling.
+Our goal is to make a tool that will be both accessible to non-experts and useful for experts.
 
@@
 The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation.
-The inputs to the diagnostic equation are the ice thickness, surface elevation, rheology, ice velocity at the inflow boundary, and the coefficient of friction for contact with the bed and side walls.
-The output of the diagnostic equation is the ice velocity throughout the entire region of interest.
+The inputs to the diagnostic equation are the ice thickness, surface elevation, velocity at the inflow boundary, rheology, and bed friction coefficient.
+The output of the diagnostic equation is the ice velocity throughout the entire domain.
 The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
-The inputs to the prognostic equation are the current value of the ice thickness, the ice velocity, and the surface and basal mass balance.
-The output is the ice thickness at a later time.
+The inputs to the prognostic equation are the current value of the ice thickness, velocity, and the surface and basal mass balance.
+The output is the ice thickness at the subsequent timestep.
 Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation.
 
-The rheology and friction coefficient may also be described in terms of other fields.
-For example, the rheology can be parameterized in terms of the ice temperature, englacial water content, and damage state.
-Likewise, the friction coefficient can be described in terms of a geological roughness factor and the subglacial water pressure.
+The rheology and friction coefficient are functions of other fields that have their own evolution equations.
+For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
+Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock.
 The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.
@@
 
-Irrespective of any deep connections with thermodynamics, the existence of an action principle is a tremendous convenience for numerical analysis.
+The action principle is especially useful for designing a robust numerical solver.
 For viscous flow problems near to steady-state, the action is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
@@
 Elmer/Ice uses an ad-hoc procedure to define the unit normal vectors at mesh nodes \citep{gagliardini2013capabilities}.
-With a transformation to terrain-following coordinates, we can just set the terrain-following vertical velocity $\omega$ to be 0 at the ice base to impose this boundary condition.
+With a transformation to terrain-following coordinates, we can set the terrain-following vertical velocity $\omega$ to be $-\dot a_b/h$ at the ice base to impose this boundary condition.
 
@@
 In Elmer/Ice, for example, the number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.
-The horizontal velocity is expected to be very smooth as a function of depth.
+The horizontal velocity for many realistic flows is very smooth as a function of depth.
 For example, under the plug flow approximation, the horizontal velocity is constant with depth.
 Under the shallow ice approximation, the horizontal velocity varies with depth as $1 - (1 - \zeta)^{n + 1}$ where $n = 3$ is the Glen flow law exponent.
-This extra information about our solution suggests a different discretization strategy.
+This extra information about our solution suggests a modal rather than a nodal discretization strategy.
 
@@
 The degree 1 model is the most minimalistic model that still exhibits vertical shear.
-Going to higher and higher degree gives a more accurate approximation at the expense of greater computational effort.
+Going to higher degree gives a more accurate approximation at the expense of greater computational effort.
 
@@
 A blunt solution to this problem would be to pass an extra argument to the Firedrake form compiler that specifies a much greater integration accuracy in the vertical for this term.
-This fix reduces, but does not eliminate, the errors in the velocities, and it incurs an appreciable computational cost.
+This fix reduces the errors in the velocities, but it does not eliminate them completely and it incurs a large computational cost.
 
@@
 The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
-Since the result of this symbolic Legendre polynomial expansion is the same upon every invocation, we can save the effort of recalculating it every time by caching the results.
 The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts \textcolor{red}{figure}.
 While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property.
-Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the expansion limit.
+Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the number of vertical modes.
 
@@
 For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
-In C++ one could obviate the need to modify the diagnostic solve routine by using variadic templates.
-But variadic templates are a more advanced language feature and relying on them would arguably present a steep learning curve for novices.
+In C++ one could avoid modifying the diagnostic solve routine by using variadic templates.
+Variadic templates are a more advanced language feature and relying on them would present a steep learning curve for novices.
 
-In icepack, any given model physics component can be substituted for a parameterization of the user's choosing.
-Every flow model in icepack is represented by a Python object, and each model object is initialized with a set of functions to calculate the various terms of the action functional.
+Any model physics component in icepack can be substituted for a parameterization of the user's choosing.
+Each flow model object is initialized with a set of functions to calculate the various terms of the action functional.
 For example, the ice stream model is initialized with functions to calculate the contributions due to viscosity, basal friction, side wall drag, and gravity.
-Sensible parameterizations are selected by default.
-If the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
+The default parameterizations are set to the most common choices in the literature, but if the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
 When the diagnostic solve procedure is invoked, all of its arguments are then passed on by keyword to the method that calculates the action functional.
-Any fields that are unnecessary for calculating a given component are simply ignored.
-For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function simply makes no use of these arguments.
+Any fields that are unnecessary for calculating a given component are ignored.
+For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function makes no use of these arguments.
 The functional form can be changed entirely and new arguments can be added, with the only restriction that they are passed as keyword and not as positional arguments.
@@
 These users then have a steep learning curve to overcome when they need to modify functionality that lives in the core library, as they now need to learn an entirely new programming language.
-Some low-level kernel code is always necessary for runtime efficiency of course.
-With Firedrake, this responsibility is deferred to a code generator of which the average user remains blissfully unaware.
+Some low-level kernel code is, of course, always necessary for runtime efficiency.
+With Firedrake, this responsibility is deferred to a code generator.
+The average user can remain largely unaware of how the code generator works even when solving very advanced problems.
 
@@
 
-As shown in equation \eqref{action-functional}, the action has units of power and is the sum of the dissipation due to viscosity, friction, gravitational driving, and restrain by ocean pressure at the terminus.
-The viscous and frictional terms are convex and positive functions of the velocity, while the gravitational and terminus stress terms are linear in the velocity and can be of either sign.
+As shown in equation \eqref{action-functional}, the action has units of power and is the sum of the dissipation due to viscosity, friction, gravitational driving, and ocean back-pressure at the terminus.
+The viscous and frictional terms are convex, positive functions of the velocity.
+The gravitational and terminus stress terms are linear in the velocity and can be of either sign.
 If we define the \emph{scale functional}
@@
 
-One of the main goals for icepack was to create a tool that is accessible to researchers who might not be experts in scientific computing.
+One of the main goals for icepack is to create a tool that is accessible to researchers who might not be experts in scientific computing.
 Previous work on numerical modeling of glacier flow has focused largely on technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory, does it scale to large numbers of processors, can models accurately predict grounding line retreat, etc.
@@
 Progressive evaluation is the main advantage of having a user interface in Python or another interpreted language, as opposed to using compiled programs that must run all at once.
-Icepack includes a suite of demos that demonstrate all of the main functions of the package.
-These demos are formatted as Jupyter notebooks, a document format that includes code, figures, and explanatory text with typeset mathematics that can be run interactively in a web browser \citep{kluyver2016jupyter}.
+Icepack includes a suite of example programs that demonstrate all of the main functions of the package.
+These demos are formatted as Jupyter notebooks, a document format that includes code, figures, and explanatory text with typeset mathematics that runs interactively in a web browser \citep{kluyver2016jupyter}.
 A notebook is a sequence of cells, each of which consists of either code or text.
@@
 This choice is of interest mostly to advanced users so we keep the linear solver method as a default argument.
-In so doing, we avoid confronting novice users with an option that they might not understand.
+In so doing, we avoid confronting novice users with options that they might not understand.
 

### 2019-08-15 icepack.tex
Add MISMIP+ demo

@@
 In this paper we introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
-Icepack is built on the finite element modeling library firedrake, which uses the domain-specific language UFL for the describing of weak forms of partial differential equations.
+Icepack is built on the finite element modeling library firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
 The diagnostic models in icepack are formulated through action principles that are specified in UFL.
@@
 Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
-The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in numerical modeling.
+The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
 Our goal is to make a tool that will be both accessible to non-experts and useful for experts.
-
-\textcolor{red}{We focus in this paper only on novel aspects of icepack that differ substantially from existing packages.}
 
@@
 VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}.
-The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for specifying weak forms of PDE.
-Icepack improves upon the groundwork laid in VarGlaS through its use of tensor product finite elements on extruded meshes, both of which are only available in firedrake \citep{bercea2016structure, mcrae2016automated}.
-These two features enable the definition of a much simpler 3D flow model that we will describe in the next section.
+The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for describing weak forms of PDE.
+Icepack improves upon the groundwork laid in VarGlaS in three main respects.
+First, icepack includes a simple 3D flow model that uses several features only available in firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
+This model will be described in section \S\ref{sec:physics-hybrid-model}.
+Second, the architecture of icepack is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of the models included.
+The software design that made this possible is discussed in section \S\ref{sec:physics-substitution}.
+Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.
 
@@
 
-\section{Diagnostic models}
+\section{Physics}
 
-There are two diagnostic models implemented in icepack.
-The \emph{shallow stream approximation} (SSA) is a 2D model describing the depth-averaged velocity of a fast-flowing grounded ice stream or floating ice shelf \citep{macayeal1989large}.
-The SSA model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
+\subsection{Mass transport}
+
+The PDE describing how the ice thickness changes in time is a conservative advection equation:
+\begin{equation}
+    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
+\end{equation}
+where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
+The thickness is represented using continuous Galerkin basis functions.
+The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
+The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.
+
+The surface elevation is calculated at each timestep of a simulation as
+\begin{equation}
+    s = \max\{b + h, (1 - \rho_I / \rho_W)h\},
+\end{equation}
+the first case corresponding to grounded ice and the second case corresponding to floating ice.
+In the immediate vicinity of the grounding line, the assumption that the floating ice is in hydrostatic equilibrium with the ocean fails.
+Most models assume hydrostasy and icepack does as well.
+Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.
+
+\subsection{Momentum transport}
+
+There are \textcolor{green}{three} diagnostic models implemented in icepack.
+The \emph{shallow ice approximation} (SIA) and shallow stream approximation (SSA) are 2D models describing the depth-averaged velocity of a glacier.
+\textcolor{green}{The SIA model describes the deep interior of ice sheets well, where flow velocities are on the order of 0.1 - 10 m/year.
+This model assumes that nearly all flow is by vertical shear, which happens when the ice is frozen to the bed or sliding at a small fraction of the surface speed.}
+The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.
+This model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
 Plug flow is a good approximation in fast-flowing ice streams and outlet glaciers near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
@@
 
-Both the SSA and BP models in icepack are described through \emph{action principles} \citep{dukowicz2010consistent}.
-Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a certain nonlinear functional, called the action.
+\subsubsection{Variational principles}
+
+All of the diagnostic models in icepack are described through \emph{variational} or \emph{action} principles \citep{dukowicz2010consistent}.
+Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a functional called the action.
 The action consists of four terms:
@@
 
-\subsection{Implementation with firedrake}
+\subsubsection{\textcolor{green}{Shallow ice approximation}}
 
-At the highest level, each physics model in icepack is represented by a Python class.
-So far, the models available are \texttt{IceShelf}, \texttt{IceStream}, and \texttt{HybridModel}.
-Each class has three main responsibilities.
-First, the model class has to have a method that takes in the various fields (ice velocity, thickness, surface elevation, etc.) and returns a symbolic representation of the action functional for that particular model.
-Second, the model class must have a method that takes in the input fields and returns an approximate minimizer of the action functional.
-The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the first method calculates.
-The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
-Finally the model must have a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
-The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.
+\subsubsection{Shallow stream approximation}
 
-\textcolor{red}{Some sample code.}
+\subsection{Hybrid model} \label{sec:physics-hybrid-model}
 
-The Unified Form Language for specifying weak forms of PDEs contains all of the primitives necessary to express individual terms of the action functional.
-These primitives largely consist of the basic vector calculus operators like the gradient of a field, tensor calculus operations like taking the dot product of two vectors or tensors, scalar functions like the square root or exponential, and symbolic integration over the mesh or its boundary.
-For example, the strain rate for a given velocity field $u$ can be written as \texttt{sym(grad(u))}, where the function \texttt{grad} represents the symbolic gradient of a field and \texttt{sym} represents the symmetrization of a rank-2 tensor.
+The shallow ice and shallow stream models in icepack are completely standard.
+The hybrid flow model that we describe here is based on variational principles, terrain-following coordinates, and vertical spectral methods.
+Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
+\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
+\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
+\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem, but this model used a finite difference discretization in the vertical direction.
 
-\textcolor{red}{Brief discussion of SSA model.}
+\subsubsection{Terrain-following coordinates}
 
-\subsection{Terrain-following coordinates and extruded meshes}
-
-The 3D flow model, as represented in the \texttt{HybridModel} class, uses several features that are only available in Firedrake to better exploit the underlying mathematics.
-Rather than represent the 3D physical domain itself, we instead use \emph{terrain-following coordinates} and \emph{extruded meshes}.
-The terrain-following vertical coordinate $\zeta$ is defined as
+Rather than use the usual Cartesian coordinate system, the hybrid flow model uses terrain-following coordinates.
+The terrain-following vertical coordinate $\zeta$ is
 \begin{equation}
@@
     \dot\varepsilon_z(u) = \dot\varepsilon_\zeta(u) + \frac{1}{2}\left(u \otimes\nabla\zeta + \nabla\zeta\otimes u\right)
+    \label{eq:geometric-correction}
 \end{equation}
 where $\otimes$ is the tensor product of two vectors.
-The geometric correction factor can easily be represented in UFL.
-By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is no more complex than in Cartesian coordinates.
-
-Terrain-following coordinates are especially advantageous on extruded meshes.
-An extruded mesh takes a 2D footprint mesh and lifts it into 3D by taking the Cartesian product with the unit interval.
-The cells of an extruded mesh are triangular prisms instead of the more common tetrahedra used for general 3D meshes.
-In Firedrake, an extruded mesh is created by calling the function \texttt{ExtrudedMesh} on the 2D footprint.
-Not every 3D domain can be described by extruding a 2D domain, but the geometry of most glacier flow problems can.
 
@@
 Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
-In other words it is not immediately clear whether the vertical velocity is more a result of basal mass balance or of geometry without more information.
+In other words, it is not immediately clear whether the vertical velocity is a result of basal mass balance or of geometry without additional information.
 By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.
 
-\subsection{Vertical spectral elements}
+\subsubsection{Spectral discretization}
 
-Having chosen to use an extruded mesh, we must then decide how many vertical layers to use.
-In Elmer/Ice, for example, the number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.
-The horizontal velocity for many realistic flows is very smooth as a function of depth.
+Terrain-following coordinates open up several choices for how to describe the vertical variation of the velocity field.
+In Elmer/Ice, for example, the user can extend the finite element discretization into a number of vertical layers.
+The number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.
+
+The horizontal velocity for many realistic flows is very smooth as a function of depth and this suggests a different approach.
 For example, under the plug flow approximation, the horizontal velocity is constant with depth.
@@
 
-Rather than extrude the footprint mesh into many vertical layers, we might instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
-To support this use case, Firedrake includes the ability to define \emph{tensor product} elements on extruded meshes.
+Rather than divide the spatial domain into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
+This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element.
 Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
@@
 For example, we can use piecewise linear or quadratic elements on triangles and use quintic or higher degree polynomials in the vertical.
-Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
 Rather than use the usual Lagrange interpolating polynomial basis in the vertical dimension, we can instead use the Legendre polynomial basis.
 The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.
-To select the Legendre polynomial basis, the user passes the keyword argument \texttt{vfamily=`Gauss-Legendre'} or \texttt{`GL'} for short to the constructor for the function space.
 
@@
 The system is then discretized in the horizontal and solved numerically.
-In any case, the code is the same regardless of how one wishes to think about the underlying mathematics.
+In any case, the code is the same regardless of how one views the underlying mathematics.
 
@@
 
-This approach for implementing a 3D flow model works completely seamlessly but for one important detail.
-The backpressure from ocean water at the calving front of a marine-terminating glacier is not a smooth function of depth.
-The pressure is 0 above the water line and linearly increasing below it:
-\begin{equation}
-    \text{backpressure power} = \int_\Gamma\int_0^1 \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+\ud\zeta\ud\gamma,
-    \label{backpressure}
-\end{equation}
-where $\zeta_{\text{sl}}$ denotes the relative depth to the water line and the subscript $+$ denotes the positive part of a real number.
-Were we to use the standard asssembly procedure in Firedrake to evaluate this integral, we would get an inaccurate result due to an insufficient number of integration points.
-The resulting velocity solutions are then wildly inaccurate due to the mis-specification of the Neumann boundary condition.
-A blunt solution to this problem would be to pass an extra argument to the Firedrake form compiler that specifies a much greater integration accuracy in the vertical for this term.
-This fix reduces the errors in the velocities, but it does not eliminate them completely and it incurs a large computational cost.
-
-We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
-The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
-The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts \textcolor{red}{figure}.
-While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property.
-Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the number of vertical modes.
-
-The exact symbolic integration approach is both faster and more accurate than using a large number of quadrature points.
-The same technique could be used to exactly calculate the ocean backpressure for any model, say the full Stokes equations, using terrain-following coordinates together with a Legendre polynomial expansion in the vertical.
-
-\textcolor{red}{Speed of hybrid model.}
-One of the main advantages of this hybrid model is the ease with which a more accurate solution can be bootstrapped from a less accurate solution, since the two fields are defined on exactly the same geometry.
-For example, if we only have a very crude initial guess for the solution, we can obtain a degree-1 solution relatively cheaply.
-The resulting velocity field is then used as a much more informed initial guess for a degree-5 solution.
-\textcolor{red}{Benchmark.}
-Since the vertical basis functions are orthogonal, the projection of the degree-1 solution into the degree-5 function space can be calculated by simply copying array elements.
-By contrast, if we tried to use the same strategy to bootstrap a solution with a small number of vertical layers to a solution with a larger number of vertical layers, we would also have to interpolate values to the fine layers.
-
-Finally, we note that this approach was first described in \citet{bassis2010hamilton}.
-This work focused solely on flowband models and did not consider the problem of how to calculate the power dissipated at the calving front.
-Moreover, this work used only the two vertical basis functions $\psi_1(\zeta) = 1$, $\psi_2(\zeta) = 1 - (1 - \zeta)^{n + 1}$, the idea being that the vertical profile can be written as a direct superposition of the plug flow and shallow ice approximation modes.
-By contrast, we leave the number of vertical basis functions up to the user.
-This flexibility enables the multigrid-type approach described above.
-\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
-\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem via finite differences in the vertical and finite elements in the horizontal, but did not use terrain-following coordinates.
-This work also adopts the viewpoint that semi-discretization defines a family of models, rather than merely a problem-specific discretization strategy.
-
-\subsection{Substituting model components}
+\subsection{Substituting model components} \label{sec:physics-substitution}
 
@@
 For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
-In programming terms this amounts to changing the number of arguments to the function that calculates basal shear stress.
+In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
 For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
@@
 For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function makes no use of these arguments.
-The functional form can be changed entirely and new arguments can be added, with the only restriction that they are passed as keyword and not as positional arguments.
+The user can completely change the functional form and add new arguments, with the restriction that arguments are passed by keyword and not positionally.
 
-We are restricted to using keyword arguments instead of positional arguments in adopting this approach.
-For the particular case of passing several fields to a physics solver routine, using only keyword arguments enhances readability and comprehensibility.
+In adopting this approach, we are restricted to using keyword arguments instead of positional arguments.
+Using only keyword arguments enhances readability and comprehensibility for this particular case of invoking a physics solver.
 The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used in the literature -- $C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.
 The order of the arguments is arbitrary and immaterial.
-For defining, say, the bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance and our design would be inappropriate.
+The preference for argument passing by name is specific to this use case and not universal.
+For example, in defining a bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance.
+In that case using positional arguments would better correspond with the underlying problem.
 
@@
 
-\subsection{Convergence criteria} \label{sec:convergence-criteria}
+\subsection{Data assimilation}
+
+Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
+The class \texttt{InverseProblem} represents the specification of the inverse problem.
+To specify an inverse problem, the user must provide:
+\begin{itemize}
+    \item the model object and the method that solves the diagnostic equation,
+    \item the objective and regularization functionals,
+    \item the observed field and the name of the argument to the diagnostic solver,
+    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver,
+    \item extra data passed to the diagnostic solver such as boundary conditions.
+\end{itemize}
+At present, this class assumes that the observed state is always the ice velocity, but in principle the same design would suffice for more complicated inverse problems.
+More importantly, the inverse problem is flexible enough to account for the fact that we might have substituted in our own parameterization for the rheology or friction coefficient.
+For example, nearly all studies in the literature use an alternate parameterization of the friction coefficient in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
+This information is passed to the inverse problem by specifying the names of the keyword arguments for the observed state and the field we wish to observe.
+
+The \texttt{InverseSolver} class is responsible for actually solving the inverse problem.
+This class will be described further in section \S\ref{sec:numerics-inverse-solvers}.
+The inverse solver is completely agnostic to the particular parameterization of the physics in terms of the inferred field as long as it guarantees positivity.
+This feature interacts seamlessly with the ability to modify individual model physics components as described in \S\ref{sec:physics-substitution}.
+
+\subsection{Damage transport}
+
+\subsection{Heat transport}
+
+\subsection{\textcolor{green}{Plumes}}
+
+\section{Numerics}
+
+In this section we describe how various parts of the model physics were implemented in icepack, including both software design and the numerical solution procedures.
+From a design perspective, each flow model is represented by a Python class that has three responsibilities.
+First, the model class has to have a method that takes in the various fields (ice velocity, thickness, surface elevation, etc.) and returns a symbolic representation of the action functional for that particular model.
+Second, the model class must have a method that takes in the input fields and returns an approximate minimizer of the action functional.
+The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the first method calculates.
+The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
+Finally the model must have a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
+The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.
+
+The Unified Form Language for specifying weak forms of PDEs contains all of the primitives necessary to express individual terms of the action functional.
+These primitives consist of the basic vector calculus operators like the gradient of a field, tensor calculus operations like taking the dot product of two vectors or tensors, scalar functions like the square root or exponential, and symbolic integration over the mesh or its boundary.
+For example, the strain rate for a given velocity field $u$ can be written as \texttt{sym(grad(u))}, where the function \texttt{grad} represents the symbolic gradient of a field and \texttt{sym} represents the symmetrization of a rank-2 tensor.
+
+\subsection{Advective transport} \label{sec:prognostic-model}
+
+The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
+To solve the mass transport equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
+The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
+This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
+The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.
+
+Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
+The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
+The extra cost of using an implicit time discretization for the mass transport equation is dwarfed by the cost of the diagnostic solve in any case.
+Advanced users who are interested in maximizing performance can subclass the mass transport solver to implement a faster explicit scheme.
+
+The implicit Euler scheme tends to diffuse out sharp discontinuities that may be present in the true solution \citep{donea2003finite}.
+Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
+The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.
+Other problems within glaciology have more of a hyperbolic character and there the argument for using shock-capturing methods is much stronger.
+For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \citep{albrecht2014fracture}.
+The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge Kutta method would be more appropriate.
+
+\subsection{Convex optimization}
+
+The action functional for each diagnostic model is convex, i.e. the second derivative is positive-definite.
+\textcolor{red}{Why it's great...}
+
+Other software packages that treat diagnostic models as nonlinear systems of equations tend to rely on ad-hoc procedures for initializing the numerical solution process.
+For example, without a damping procedure in Newton's method, the iteration can prove unstable if initialized far away from the true solution.
+Some packages combat this problem by using a few iterations of the more robust but slower Picard method first \citep{gagliardini2013capabilities}.
+While this approach can be effective it requires tuning the number of Picard iterations and there is no guarantee that an adequate amount for one problem will work well on another problem.
+This issue rarely appears on realistic input data.
+When solving inverse problems, however, the intermediate guesses for the inferred field can be wildly unrealistic before converging.
+A forward model solver that is not sufficiently robust can crash in these extreme scenarios.
+By contrast, a damped Newton procedure using a line search that satisfies the Armijo-Wolfe criteria is guaranteed to converge on non-degenerate, if unrealistic, input data.
+
+\subsection{Convergence metrics} \label{sec:convergence-criteria}
 
@@
 We have found empirically that, with this criterion and the Newton solver implementation in icepack, the iteration usually converges to machine precision in around 8 steps.
-For exceptionally bad initial guesses the iteration count can reach as high as 20 but rarely more.
+The iteration count can reach as high as 20 for exceptionally bad initial guesses for the velocity or with unphysical fluidity or friction values.
 We also observe the expected doubling of the number of accurate digits in the value of the action once the velocity guesses are within the convergence basin of the true solution.
@@
 
-\section{Prognostic model} \label{sec:prognostic-model}
+\subsection{Inverse solvers} \label{sec:numerics-inverse-solvers}
 
-The PDE describing how the ice thickness changes in time is a simple conservative advection equation:
-\begin{equation}
-    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
-\end{equation}
-where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
-The thickness is represented using continuous Galerkin basis functions.
-The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
-The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.
+The \texttt{InverseProblem} class describes what problem is being solved, while the \texttt{InverseSolver} class is responsible for carrying out the numerical optimization.
+There are three inverse solvers in icepack: a simple gradient descent solver, a quasi-Newton solver based on the BFGS approximation to the Hessian, and a Gauss-Newton solver.
+All of these classes are based around the general idea of first computing a search direction and then performing a line search.
+They differ in how the search direction is computed.
 
-The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
-Other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
-The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
-This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
-The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.
-
-Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
-The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
-The extra cost of using an implicit time discretization for the mass transport equation is dwarfed by the cost of the diagnostic solve in any case.
-Advanced users who are interested in maximizing performance can subclass the mass transport solver to implement a faster explicit scheme.
-
-The implicit Euler scheme tends to diffuse out sharp discontinuities that may be present in the true solution \citep{donea2003finite}.
-Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
-The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.
-Other problems within glaciology have more of a hyperbolic character and there the argument for using shock-capturing methods is much stronger.
-For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \citep{albrecht2014fracture}.
-The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge Kutta method would be more appropriate.
-
-\section{Data assimilation}
-
-Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
-The class \texttt{InverseProblem} represents the specification of the inverse problem.
-To specify an inverse problem, the user must provide:
-\begin{itemize}
-    \item the model object and the method that solves the diagnostic equation,
-    \item the objective and regularization functionals,
-    \item the observed field and the name of the argument to the diagnostic solver,
-    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver,
-    \item extra data passed to the diagnostic solver such as boundary conditions.
-\end{itemize}
-At present, this class assumes that the observed state is always the ice velocity, but in principle the same design would suffice for more complicated inverse problems.
-More importantly, the inverse problem is flexible enough to account for the fact that we might have substituted in our own parameterization for the rheology or friction coefficient.
-For example, nearly all studies in the literature use an alternate parameterization of these coefficients in terms of some auxiliary field in order to guarantee positivity \textcolor{red}{citations}.
-This information is passed to the inverse problem by specifying the names of the keyword arguments for the observed state and the field we wish to observe.
-
-The class \texttt{InverseSolver} is an abstract interface for classes that actually solve the inverse problem.
-There are three implementations of this interface in icepack: \texttt{GradientDescentSolver}, \texttt{BFGSSolver}, and \texttt{GaussNewtonSolver}.
-All of these classes are based around the general idea of first computing a search direction and then performing a line search in every iteration.
-They differ in how the search direction is computed.
 The gradient descent solver uses the search direction
@@
 This method can be very expensive and brittle in the initial iterations and often takes many steps to converge.
+
 The BFGS method uses the past $m$ iterations of the algorithm to compute a low-rank approximation to the inverse of the objective functional's Hessian matrix; see \citet{nocedal2006numerical} for a more in-depth discussion.
-The BFGS method converges faster than gradient descent but suffers from many of the same brittleness issues in the initial iterations before it has accumulated enough history.
+The BFGS method converges faster than gradient descent.
+However, it suffers from many of the same brittleness issues in the initial iterations before it has built up enough history to approximate the Hessian inverse.
+
 Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian.
@@
 
-The bootstrapping approach used to calculate higher-order solutions to the hybrid model from lower-order solutions rather than from a cold initial guess can also be used for inverse problems.
+\subsection{Hybrid model}
+
+The hybrid flow uses several features that are only available in firedrake to better exploit the special structure of the problem.
+Implementing this model also required some mathematical sleight-of-hand related to the terminus boundary condition that has not appeared in the literature before.
+Additionally, the hierarchical structure of spectral basis functions presents an opportunity for developing fast algorithms.
+In all other respects the implementation of the hybrid flow model using convex optimization follows the techniques described above.
+
+\subsubsection{Discretization}
+
+In order to use terrain-following coordinates, the hybrid model assumes that the geometry of the domain is an extruded mesh, where a 2D footprint mesh is lifted into 3D.
+Firedrake includes support for creating extruded meshes by calling the function \texttt{ExtrudedMesh} on the 2D footprint \citep{bercea2016structure}.
+The cells of an extruded mesh are triangular prisms instead of the more common tetrahedra used for general 3D meshes.
+Not every 3D domain can be described by extruding a 2D domain, but the geometry of most glacier flow problems can.
+
+The geometric correction factor in equation \eqref{eq:geometric-correction} for gradients in terrain-following coordinates can easily be represented in UFL.
+By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is only slightly more complex than in Cartesian coordinates.
+
+For problems defined on extruded geometries, firedrake includes support for tensor product elements, which includes using different bases in the horizontal and vertical directions \citep{mcrae2016automated}.
+Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
+In our case, we used the usual continuous Galerkin basis in the horizontal and Gauss-Legendre elements in the vertical.
+To select the Legendre polynomial basis, the user passes the keyword argument \texttt{vfamily=`Gauss-Legendre'} or \texttt{`GL'} for short to the constructor for the function space.
+
+Extruded meshes and tensor product elements are available in firedrake but not in FEniCS.
+Other general-purpose finite element modeling packages that support tensor product elements include deal.II and nektar++ \citep{bangerth2007deal, cantwell2015nektar++}.
+Like most other packages in this domain, deal.II and nektar++ are written in C++, whereas our goal for icepack was to have both the core and the user interface in Python.
+
+\subsubsection{Ocean boundary condition}
+
+Our approach for implementing a hybrid flow model works completely seamlessly but for one important detail.
+The backpressure from ocean water at the calving front of a marine-terminating glacier is not a smooth function of depth.
+The pressure is 0 above the water line and linearly increasing below it:
+\begin{equation}
+    \text{backpressure power} = \int_\Gamma\int_0^1 \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+\ud\zeta\ud\gamma,
+    \label{backpressure}
+\end{equation}
+where $\zeta_{\text{sl}}$ denotes the relative depth to the water line and the subscript $+$ denotes the positive part of a real number.
+Were we to use the standard asssembly procedure in Firedrake to evaluate this integral, we would get an inaccurate result due to an insufficient number of integration points.
+The resulting velocity solutions are then wildly inaccurate due to the mis-specification of the Neumann boundary condition.
+A blunt solution to this problem would be to pass an extra argument to the Firedrake form compiler that specifies a much greater integration accuracy in the vertical for this term.
+This fix reduces the errors in the velocities, but it does not eliminate them completely and it incurs a large computational cost.
+
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/legendre/pressure.png}
+    \caption{The normalized ocean pressure ($p_w / \rho_Wg$) and Legendre polynomial approximations of several degrees (left), and the residuals of the approximation (right).
+    For this particular example, the waterline is at $\zeta = 1/3$.
+    The moments of each of the residuals up to the approximation degree are all zero.}
+    \label{fig:legendre}
+\end{figure}
+
+We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
+The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
+The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts at high degree.
+While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property of Legendre polynomials.
+Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the number of vertical modes.
+An example of the pressure approximations using linear, quadratic, and cubic Legendre polynomials are shown in figure \ref{fig:legendre}.
+
+The exact symbolic integration approach is both faster and more accurate than using a large number of quadrature points.
+The same technique could be used to exactly calculate the ocean backpressure for any model, say the full Stokes equations, using terrain-following coordinates together with a Legendre polynomial expansion in the vertical.
+
+\subsubsection{Multigrid-type algorithm}
+
+One of the main advantages of this hybrid model is the ease with which a more accurate solution can be bootstrapped from a less accurate solution, since the two fields are defined on exactly the same geometry.
+For example, if we only have a very crude initial guess for the solution, we can obtain a degree-1 solution relatively cheaply.
+The resulting velocity field is then used as a much more informed initial guess for a degree-5 solution.
+\textcolor{red}{Benchmark.}
+Since the vertical basis functions are orthogonal, the projection of the degree-1 solution into the degree-5 function space can be calculated by simply copying array elements.
+By contrast, if we tried to use the same strategy to bootstrap a solution with a small number of vertical layers to a solution with a larger number of vertical layers, we would also have to interpolate values to the fine layers.
+
+Other hybrid models in the literature use only a fixed number of vertical basis functions.
+For example, \citet{bassis2010hamilton} use only the two vertical basis functions $\psi_1(\zeta) = 1$, $\psi_2(\zeta) = 1 - (1 - \zeta)^{n + 1}$, the idea being that the vertical profile can be written as a direct superposition of the plug flow and shallow ice approximation modes.
+By contrast, we leave the number of vertical basis functions up to the user.
+This flexibility enables the multigrid-type approach described above.
+
+The bootstrapping approach described here can be used both for the forward model and for inverse problems.
 First, we estimate the unknown parameter using only degree-1 vertical basis functions for the velocity field, which captures most of the spatial variability of the unknown parameter.
@@
 
-\section{Testing}
+\subsection{MISMIP}
+
+\subsection{Larsen C Ice Shelf}
+
+\subsection{Jakobshavn Isbrae}
 
@@
 How much fill-in to allow in the incomplete factorization?
-Many glaciologists might not have the background in computational science to know that adjusting these parameters could make the difference between solver convergence or breakdown.
-\textcolor{red}{Also stabilization of the Stokes equations.}
+Many glaciologists do not have the background in numerical linear algebra to know that adjusting these parameters could make the difference between solver convergence or breakdown.
+Similarly, users might want to select between different discretization strategies for the Stokes equations, which must be chosen carefully in order to satisfy the Ladyzhenskaya-Babu\v{s}ka-Brezzi (LBB) conditions \citep{boffi2013mixed}.
+When using Galerkin least-squares \textcolor{red}{check that's actually what it is} stabilization of the weak form, the user has to pick a value of the stabilization parameter.
+Determining exactly what value of this parameter is necessary to guarantee stability is a subtle problem, even more so for the kinds of highly anisotropic meshes that are commonly encountered in 3D glacier flow modeling.
+If the solver fails to converge, it might not be obvious even to an expert whether the problem lies with the stabilization or the aforementioned parameters of the linear solver.
 

### 2019-08-15 demos/mismip/mismip+.py

@@ -1,0 +1,125 @@
+import os
+import argparse
+import numpy as np
+import matplotlib.pyplot as plt
+import firedrake
+from firedrake import (inner, dx, Constant, interpolate, as_vector, exp, sqrt,
+                       max_value)
+import icepack, icepack.models, icepack.plot
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--verbose', action='store_true')
+parser.add_argument('--input')
+parser.add_argument('--output', default='result')
+parser.add_argument('--melt', action='store_true')
+parser.add_argument('--time', type=float, default=4e3)
+parser.add_argument('--timestep', type=float, default=1.0)
+
+args = parser.parse_args()
+
+# Create the mesh and function spaces
+Lx, Ly = 640e3, 80e3
+ny = 40
+nx = int(Lx/Ly) * ny
+mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
+area = firedrake.assemble(Constant(1) * dx(mesh))
+
+Q = firedrake.FunctionSpace(mesh, family='CG', degree=1)
+V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=1)
+
+# Create the bed topography
+x, y = firedrake.SpatialCoordinate(mesh)
+
+x_c = 300e3
+X = x / x_c
+
+B_0 = -150
+B_2 = -728.8
+B_4 = 343.91
+B_6 = -50.57
+B_x = B_0 + B_2 * X**2 + B_4 * X**4 + B_6 * X**6
+
+f_c = 4e3
+d_c = 500
+w_c = 24e3
+
+B_y = d_c * (1 / (1 + exp(-2 * (y - Ly/2 - w_c) / f_c)) +
+             1 / (1 + exp(2 * (y - Ly / 2 + w_c) / f_c)))
+
+z_deep = -720
+z_b = interpolate(max_value(B_x + B_y, z_deep), Q)
+
+# We use units of megapascals-meters-years, whereas in the paper the parameters
+# are reported in units of pascals, so these values include some conversion
+# factors.
+A = Constant(20)
+C = Constant(1e-2)
+
+# Create the bed friction dissipation functional
+from icepack.constants import (ice_density as ρ_I, water_density as ρ_W,
+                               gravity as g, weertman_sliding_law as m)
+def friction(u, h, s, C):
+    u_b = sqrt(inner(u, u))
[…]

### 2019-08-15 demos/mismip/plot.py

@@ -1,0 +1,36 @@
+import os
+import argparse
+import firedrake
+import icepack.plot
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--input')
+parser.add_argument('--output')
+
+args = parser.parse_args()
+
+Lx, Ly = 640e3, 80e3
+ny = 40
+nx = int(Lx/Ly) * ny
+mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
+
+Q = firedrake.FunctionSpace(mesh, family='CG', degree=1)
+V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=1)
+
+h = firedrake.Function(Q)
+u = firedrake.Function(V)
+
+input_name = os.path.splitext(args.input)[0]
+chk = firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ)
+chk.load(h, name='h')
+chk.load(u, name='u')
+
+fig, axes = icepack.plot.subplots(nrows=2, sharex=True, sharey=True,
+                                  figsize=(6.4, 2.8))
+contours_h = icepack.plot.tricontourf(h, 40, axes=axes[0])
+fig.colorbar(contours_h, ax=axes[0], fraction=0.0075, pad=0.04, label='m')
+axes[0].set_title('Thickness')
+contours_u = icepack.plot.tricontourf(u, 40, axes=axes[1])
+fig.colorbar(contours_u, ax=axes[1], fraction=0.0075, pad=0.04, label='m/year')
+axes[1].set_title('Velocity')
+fig.savefig(args.output, dpi=300, bbox_inches='tight')

### 2019-08-15 icepack.tex
Gigantic reorganization

@@
 Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
-The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
+The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in numerical modeling.
 Our goal is to make a tool that will be both accessible to non-experts and useful for experts.
@@
 VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}.
-The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for describing weak forms of PDE.
-Icepack improves upon the groundwork laid in VarGlaS in three main respects.
-First, icepack includes a simple 3D flow model that uses several features only available in firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
-This model will be described in section \S\ref{sec:physics-hybrid-model}.
-Second, the architecture of icepack is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of the models included.
-The software design that made this possible is discussed in section \S\ref{sec:physics-substitution}.
-Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.
+The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for specifying weak forms of PDE.
+Icepack improves upon the groundwork laid in VarGlaS through its use of tensor product finite elements on extruded meshes, both of which are only available in firedrake \citep{bercea2016structure, mcrae2016automated}.
+These two features enable the definition of a much simpler 3D flow model that we will describe in \S\ref{sec:physics-hybrid-model}.
 
@@
 Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
-\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
+\citet{bassis2010hamilton} \textcolor{red}{and Charlie} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
 \citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
@@
 Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
-In other words, it is not immediately clear whether the vertical velocity is a result of basal mass balance or of geometry without additional information.
+In other words it is not immediately clear whether the vertical velocity is more a result of basal mass balance or of geometry without more information.
 By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.
@@
 
-Rather than divide the spatial domain into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
+Rather than divide the geometry into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
 This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element.
@@
 The system is then discretized in the horizontal and solved numerically.
-In any case, the code is the same regardless of how one views the underlying mathematics.
+In any case, the code is the same regardless of how one wishes to think about the underlying mathematics.
 
@@
 
-\subsection{Substituting model components} \label{sec:physics-substitution}
+\subsection{Substituting model components}
 
@@
 For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
-In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
+In programming terms this amounts to changing the number of arguments to the function that calculates basal shear stress.
 For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
@@
 For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function makes no use of these arguments.
-The user can completely change the functional form and add new arguments, with the restriction that arguments are passed by keyword and not positionally.
+The functional form can be changed entirely and new arguments can be added, with the only restriction that they are passed as keyword and not as positional arguments.
 
-In adopting this approach, we are restricted to using keyword arguments instead of positional arguments.
-Using only keyword arguments enhances readability and comprehensibility for this particular case of invoking a physics solver.
+We are restricted to using keyword arguments instead of positional arguments in adopting this approach.
+For the particular case of passing several fields to a physics solver routine, using only keyword arguments enhances readability and comprehensibility.
 The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used in the literature -- $C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.
 The order of the arguments is arbitrary and immaterial.
-The preference for argument passing by name is specific to this use case and not universal.
-For example, in defining a bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance.
-In that case using positional arguments would better correspond with the underlying problem.
+For defining, say, the bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance and our design would be inappropriate.
 
@@
 More importantly, the inverse problem is flexible enough to account for the fact that we might have substituted in our own parameterization for the rheology or friction coefficient.
-For example, nearly all studies in the literature use an alternate parameterization of the friction coefficient in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
+For example, nearly all studies in the literature use an alternate parameterization of these coefficients in terms of some auxiliary field in order to guarantee positivity \textcolor{red}{citations}.
 This information is passed to the inverse problem by specifying the names of the keyword arguments for the observed state and the field we wish to observe.
@@
 This class will be described further in section \S\ref{sec:numerics-inverse-solvers}.
-The inverse solver is completely agnostic to the particular parameterization of the physics in terms of the inferred field as long as it guarantees positivity.
-This feature interacts seamlessly with the ability to modify individual model physics components as described in \S\ref{sec:physics-substitution}.
+We reiterate the inverse solver is completely agnostic to the particular parameterization of the physics in terms of the field that is being inferred, so long as it guarantees positivity.
 
@@
 For example, without a damping procedure in Newton's method, the iteration can prove unstable if initialized far away from the true solution.
-Some packages combat this problem by using a few iterations of the more robust but slower Picard method first \citep{gagliardini2013capabilities}.
+Some packages combat this problem by using a few iterations of the more robust but slower Picard method first \textcolor{red}{cite}.
 While this approach can be effective it requires tuning the number of Picard iterations and there is no guarantee that an adequate amount for one problem will work well on another problem.
-This issue rarely appears on realistic input data.
-When solving inverse problems, however, the intermediate guesses for the inferred field can be wildly unrealistic before converging.
-A forward model solver that is not sufficiently robust can crash in these extreme scenarios.
-By contrast, a damped Newton procedure using a line search that satisfies the Armijo-Wolfe criteria is guaranteed to converge on non-degenerate, if unrealistic, input data.
+By contrast, a damped Newton procedure using a line search that satisfies the Armijo-Wolfe criteria is guaranteed to converge on non-degenerate input data.
 
@@
 We have found empirically that, with this criterion and the Newton solver implementation in icepack, the iteration usually converges to machine precision in around 8 steps.
-The iteration count can reach as high as 20 for exceptionally bad initial guesses for the velocity or with unphysical fluidity or friction values.
+For exceptionally bad initial guesses the iteration count can reach as high as 20 but rarely more.
 We also observe the expected doubling of the number of accurate digits in the value of the action once the velocity guesses are within the convergence basin of the true solution.
@@
 The BFGS method uses the past $m$ iterations of the algorithm to compute a low-rank approximation to the inverse of the objective functional's Hessian matrix; see \citet{nocedal2006numerical} for a more in-depth discussion.
-The BFGS method converges faster than gradient descent.
-However, it suffers from many of the same brittleness issues in the initial iterations before it has built up enough history to approximate the Hessian inverse.
+The BFGS method converges faster than gradient descent but suffers from many of the same brittleness issues in the initial iterations before it has accumulated enough history.
 
@@
 The geometric correction factor in equation \eqref{eq:geometric-correction} for gradients in terrain-following coordinates can easily be represented in UFL.
-By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is only slightly more complex than in Cartesian coordinates.
+By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is no more complex than in Cartesian coordinates.
 
@@
 Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
-In our case, we used the usual continuous Galerkin basis in the horizontal and Gauss-Legendre elements in the vertical.
+In our case, we used the traditional continuous Galerkin basis in the horizontal and Gauss-Legendre elements in the vertical.
 To select the Legendre polynomial basis, the user passes the keyword argument \texttt{vfamily=`Gauss-Legendre'} or \texttt{`GL'} for short to the constructor for the function space.
@@
 Extruded meshes and tensor product elements are available in firedrake but not in FEniCS.
-Other general-purpose finite element modeling packages that support tensor product elements include deal.II and nektar++ \citep{bangerth2007deal, cantwell2015nektar++}.
+Other general-purpose finite element modeling packages that support tensor product elements include deal.II and nektar++ \textcolor{red}{cite}.
 Like most other packages in this domain, deal.II and nektar++ are written in C++, whereas our goal for icepack was to have both the core and the user interface in Python.
@@
 
-\begin{figure}[h]
-    \includegraphics[width=0.95\linewidth]{demos/legendre/pressure.png}
-    \caption{The normalized ocean pressure ($p_w / \rho_Wg$) and Legendre polynomial approximations of several degrees (left), and the residuals of the approximation (right).
-    For this particular example, the waterline is at $\zeta = 1/3$.
-    The moments of each of the residuals up to the approximation degree are all zero.}
-    \label{fig:legendre}
-\end{figure}
-
 We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
 The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
-The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts at high degree.
-While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property of Legendre polynomials.
+The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts \textcolor{red}{figure}.
+While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property.
 Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the number of vertical modes.
-An example of the pressure approximations using linear, quadratic, and cubic Legendre polynomials are shown in figure \ref{fig:legendre}.
 
@@
 
-\subsection{MISMIP}
+\subsection{ISMIP-HOM}
 
@@
 How much fill-in to allow in the incomplete factorization?
-Many glaciologists do not have the background in numerical linear algebra to know that adjusting these parameters could make the difference between solver convergence or breakdown.
-Similarly, users might want to select between different discretization strategies for the Stokes equations, which must be chosen carefully in order to satisfy the Ladyzhenskaya-Babu\v{s}ka-Brezzi (LBB) conditions \citep{boffi2013mixed}.
-When using Galerkin least-squares \textcolor{red}{check that's actually what it is} stabilization of the weak form, the user has to pick a value of the stabilization parameter.
-Determining exactly what value of this parameter is necessary to guarantee stability is a subtle problem, even more so for the kinds of highly anisotropic meshes that are commonly encountered in 3D glacier flow modeling.
-If the solver fails to converge, it might not be obvious even to an expert whether the problem lies with the stabilization or the aforementioned parameters of the linear solver.
+Many glaciologists might not have the background in computational science to know that adjusting these parameters could make the difference between solver convergence or breakdown.
+\textcolor{red}{Also stabilization of the Stokes equations.}
 

### 2019-08-15 icepack.tex
More contrast with varglas

@@
 VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}.
-The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for specifying weak forms of PDE.
-Icepack improves upon the groundwork laid in VarGlaS through its use of tensor product finite elements on extruded meshes, both of which are only available in firedrake \citep{bercea2016structure, mcrae2016automated}.
-These two features enable the definition of a much simpler 3D flow model that we will describe in \S\ref{sec:physics-hybrid-model}.
+The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for describing weak forms of PDE.
+Icepack improves upon the groundwork laid in VarGlaS in three main respects.
+First, icepack includes a simple 3D flow model that uses several features only available in firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
+This model will be described in section \S\ref{sec:physics-hybrid-model}.
+Second, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.
+Finally, the architecture of icepack is designed to make it easy for users to modify the components of each model, such as the rheology and basal friction.
+The software design that made this possible is discussed in section \S\ref{sec:physics-substitution}.
 
@@
 
-\subsection{Substituting model components}
+\subsection{Substituting model components} \label{sec:physics-substitution}
 

### 2019-08-16 icepack.tex
Added citation to Langdon and Raymond 1978

@@
 Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
-\citet{bassis2010hamilton} \textcolor{red}{and Charlie} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
+\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
 \citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.

### 2019-08-16 icepack.tex
Add Legendre polynomial figure

@@
 Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
-The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in numerical modeling.
+The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
 Our goal is to make a tool that will be both accessible to non-experts and useful for experts.
@@
 This model will be described in section \S\ref{sec:physics-hybrid-model}.
-Second, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.
-Finally, the architecture of icepack is designed to make it easy for users to modify the components of each model, such as the rheology and basal friction.
+Second, the architecture of icepack is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of the models included.
 The software design that made this possible is discussed in section \S\ref{sec:physics-substitution}.
+Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.
 
@@
 Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
-In other words it is not immediately clear whether the vertical velocity is more a result of basal mass balance or of geometry without more information.
+In other words, it is not immediately clear whether the vertical velocity is a result of basal mass balance or of geometry without additional information.
 By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.
@@
 
-Rather than divide the geometry into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
+Rather than divide the spatial domain into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
 This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element.
@@
 This class will be described further in section \S\ref{sec:numerics-inverse-solvers}.
-We reiterate the inverse solver is completely agnostic to the particular parameterization of the physics in terms of the field that is being inferred, so long as it guarantees positivity.
+The inverse solver is completely agnostic to the particular parameterization of the physics in terms of the inferred field as long as it guarantees positivity.
+This feature interacts seamlessly with the ability to modify individual model physics components as described in \S\ref{sec:physics-substitution}.
 
@@
 While this approach can be effective it requires tuning the number of Picard iterations and there is no guarantee that an adequate amount for one problem will work well on another problem.
-By contrast, a damped Newton procedure using a line search that satisfies the Armijo-Wolfe criteria is guaranteed to converge on non-degenerate input data.
+This issue rarely appears on realistic input data.
+When solving inverse problems, however, the intermediate guesses for the inferred field can be wildly unrealistic before converging.
+A forward model solver that is not sufficiently robust can crash in these extreme scenarios.
+By contrast, a damped Newton procedure using a line search that satisfies the Armijo-Wolfe criteria is guaranteed to converge on non-degenerate, if unrealistic, input data.
 
@@
 The BFGS method uses the past $m$ iterations of the algorithm to compute a low-rank approximation to the inverse of the objective functional's Hessian matrix; see \citet{nocedal2006numerical} for a more in-depth discussion.
-The BFGS method converges faster than gradient descent but suffers from many of the same brittleness issues in the initial iterations before it has accumulated enough history.
+The BFGS method converges faster than gradient descent.
+However, it suffers from many of the same brittleness issues in the initial iterations before it has built up enough history to approximate the Hessian inverse.
 
@@
 The geometric correction factor in equation \eqref{eq:geometric-correction} for gradients in terrain-following coordinates can easily be represented in UFL.
-By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is no more complex than in Cartesian coordinates.
+By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is only slightly more complex than in Cartesian coordinates.
 
@@
 Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
-In our case, we used the traditional continuous Galerkin basis in the horizontal and Gauss-Legendre elements in the vertical.
+In our case, we used the usual continuous Galerkin basis in the horizontal and Gauss-Legendre elements in the vertical.
 To select the Legendre polynomial basis, the user passes the keyword argument \texttt{vfamily=`Gauss-Legendre'} or \texttt{`GL'} for short to the constructor for the function space.
@@
 
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/legendre/pressure.png}
+    \caption{The normalized ocean pressure ($p_w / \rho_Wg$) and Legendre polynomial approximations of several degrees (left), and the residuals of the approximation (right).
+    For this particular example, the waterline is at $\zeta = 1/3$.
+    The moments of each of the residuals up to the approximation degree are all zero.}
+    \label{fig:legendre}
+\end{figure}
+
 We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
 The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
-The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts \textcolor{red}{figure}.
-While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property.
+The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts at high degree.
+While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property of Legendre polynomials.
 Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the number of vertical modes.
+An example of the pressure approximations using linear, quadratic, and cubic Legendre polynomials are shown in figure \ref{fig:legendre}.
 

### 2019-09-04 icepack.tex
Assorted corrections

@@
 
-\begin{figure}[h]
-    \includegraphics[width=0.95\linewidth]{demos/legendre/pressure.png}
-    \caption{The normalized ocean pressure ($p_w / \rho_Wg$) and Legendre polynomial approximations of several degrees (left), and the residuals of the approximation (right).
-    For this particular example, the waterline is at $\zeta = 1/3$.
-    The moments of each of the residuals up to the approximation degree are all zero.}
-    \label{fig:legendre}
-\end{figure}
-
 We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
 The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
-The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts at high degree.
+The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts \textcolor{red}{figure}.
 While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property of Legendre polynomials.
 Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the number of vertical modes.
-An example of the pressure approximations using linear, quadratic, and cubic Legendre polynomials are shown in figure \ref{fig:legendre}.
 

### 2019-09-06 icepack.tex
Citations

--- 2019-08-16 icepack.tex
@@
 More importantly, the inverse problem is flexible enough to account for the fact that we might have substituted in our own parameterization for the rheology or friction coefficient.
-For example, nearly all studies in the literature use an alternate parameterization of these coefficients in terms of some auxiliary field in order to guarantee positivity \textcolor{red}{citations}.
+For example, nearly all studies in the literature use an alternate parameterization of the friction coefficient in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
 This information is passed to the inverse problem by specifying the names of the keyword arguments for the observed state and the field we wish to observe.
@@
 For example, without a damping procedure in Newton's method, the iteration can prove unstable if initialized far away from the true solution.
-Some packages combat this problem by using a few iterations of the more robust but slower Picard method first \textcolor{red}{cite}.
+Some packages combat this problem by using a few iterations of the more robust but slower Picard method first \citep{gagliardini2013capabilities}.
 While this approach can be effective it requires tuning the number of Picard iterations and there is no guarantee that an adequate amount for one problem will work well on another problem.
@@
 Extruded meshes and tensor product elements are available in firedrake but not in FEniCS.
-Other general-purpose finite element modeling packages that support tensor product elements include deal.II and nektar++ \textcolor{red}{cite}.
+Other general-purpose finite element modeling packages that support tensor product elements include deal.II and nektar++ \citep{bangerth2007deal, cantwell2015nektar++}.
 Like most other packages in this domain, deal.II and nektar++ are written in C++, whereas our goal for icepack was to have both the core and the user interface in Python.

### 2019-09-13 icepack.tex
Some explanation of MISMIP+, added figure

@@
     \tau_b = -C|u|^{\frac{1}{m} - 1}u
+    \label{eq:weertman-sliding}
 \end{equation}
@@
 
-\subsection{ISMIP-HOM}
+\subsection{MISMIP+}
+
+As a first test case for icepack, we ran the first experiment from the Marine Ice Sheet Model Intercomparison Project version 3 (MISMIP+).
+The parameters and geometry for this experiment can be found in \citet{asay2016experimental}.
+The MISMIP+ experiment has three phases.
+First, the model must find a steady state marine ice sheet with a fixed accumulation rate and no submarine melting.
+Next, submarine melting with a given depth-dependent parameterization is applied for 100 years.
+The increased melt thins the ice ice shelf and initiates a retreat of the glacier grounding line.
+Finally, submarine melting is turned off for at least 200 years, optionally longer.
+The grounding line then readvances, but not as far as its original position.
+
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/mismip/steady-state.png}
+    \caption{The steady state thickness and velocity of the MISMIP+ experimental setup with the Weertman sliding law.}
+    \label{fig:mismip}
+\end{figure}
+
+The original intercomparison project specified that participants could use the Weertman sliding law (equation \eqref{eq:weertman-sliding}) as well as two other sliding laws that transition to a more plastic rheology at high sliding speeds.
+The first alternative sliding law consists of Weertman sliding until the stress reaches a critical value, at which point the constitutive relation transitions to perfect plasticity.
+The second alternative is the Schoof sliding law \citep{schoof2005effect}:
+\begin{equation}
+    \tau_b = -\frac{C|u|^{1/m}\cdot \tau_c}{(C^m|u| + \tau_c^m)^{1/m}}\frac{u}{|u|},
+    \label{eq:schoof-sliding}
+\end{equation}
+where the critical stress $\tau_c$ is a certain specified fraction of the water pressure in the subglacial hydrological system.
+Rather than specify the sliding law as such, icepack expresses the sliding law as the derivative of some action functional.
+In other words, to implement the Schoof sliding law in icepack, we need to know the antiderivative of equation \eqref{eq:schoof-sliding}.
+Using a computer algebra system, we found that the antiderivative of this expression in terms of hypergeometric functions.
+The Unified Form Language has several transcendental functions (sine, cosine, exponential, etc.), but it does not currently support hypergeometric functions.
+Instead, we implemented a sliding relation that exhibits the important features of the Schoof law, i.e.
+it behaves like an $m = 3$ power law at low sliding speeds and $m = \infty$ at high sliding speeds, but which is more tractable algebraically.
+The power dissipation density for this sliding law is
+\begin{equation}
+    P = \tau_c\left(\left(u_c^{\frac{1}{m} + 1} + |u|^{\frac{1}{m} + 1}\right)^{\frac{m}{m + 1}} - u_c\right).
+\end{equation}
+A sample of the source code implementing both of these approaches is shown in \textcolor{red}{Figure X}.
+To change the sliding law, users only need to pass one function or the other to the model object at initialization.
+Users do not need to implement a subclass that overrides some parent method.
+This approach would be idiomatic in C++, but it requires some facility with object-oriented programming that a non-expert might lack.
+
+Figure \ref{fig:mismip} shows the outcome of the MISMIP+ experiment performed using icepack.
 

### 2019-10-06 icepack.tex
More on convex opt

@@
 For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
-In C++ one could avoid modifying the diagnostic solve routine by using variadic templates.
-Variadic templates are a more advanced language feature and relying on them would present a steep learning curve for novices.
+In C++ one could avoid modifying the diagnostic solve routine by using variadic templates, but this is an advanced language feature and relying on it would present a steep learning curve for novices.
 
@@
 
-The action functional for each diagnostic model is convex, i.e. the second derivative is positive-definite.
-\textcolor{red}{Why it's great...}
+The action functional for each diagnostic model is convex, i.e. the second derivative is strictly positive-definite.
+From a theoretical persepctive, convexity guarantees that the problem has a unique solution.
+This property is also especially advantageous for implementing numerical solvers.
+We use a damped Newton method to solve the diagnostic equations.
+Starting from a guess $u_k$ for the velocity, one first computed a \emph{search direction} $v_k$ that solves a linear equation:
+\begin{equation}
+    \ud^2J(u_k)\cdot v_k = -\ud J(u_k).
+    \label{eq:newton-search-direction}
+\end{equation}
+The next value of the velocity is then computed by searching along $v_k$:
+\begin{equation}
+    u_{k + 1} = u_k + \alpha_k\cdot v_k,
+\end{equation}
+where $\alpha$ is chosen to minimize $J$ along the search direction.
+For an initial guess sufficiently close to the exact solution, the undamped Newton method ($\alpha_k = 1$ at every step) converges quadratically.
+The line search step ensures that the method can converge even from a poor initial guess, provided that the line search method satisfies the Armijo-Wolfe criteria \citep{nocedal2006numerical}.
+
+For a convex problem, $\ud^2J$ is a symmetric and positive-definite matrix.
+This has two advantages.
+First, the search direction is always a descent direction for $J$.
+For non-convex problems this is not always the case and some adaptation strategy must be employed in the vicinity of saddle points.
+Second, symmetry and positivity enable the use of specialized linear solvers, such as the Cholesky decomposition or the conjugate gradient algorithm, that are superior to their more general counterparts in many respects.
 
@@
 
-Calculating the ice velocity from the geometry and other input fields involves minimizing a convex functional.
-The minimizer can only be approximated, so we employ iterative procedures based on Newton's method with a line search.
-We must then choose some metric to decide when the current guess for the velocity is good enough to stop the iteration.
+Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation, but few if any have considered the problem of when to stop iterating.
+The most common stopping criteria are when (1) the 2-norm of the residual is sufficiently small or (2) the relative change in the iterates is sufficiently small.
+Each of these approaches has problems.
+The residual norm depends on the discretization and does not weight all degrees of freedom proportionally, e.g. vertex and edge degrees of freedom in higher-order finite element methods.
+The relative change criterion, on the other hand, can suggest convergence when in fact the method has stagnated.
 
 A convergence criterion that works equally well independent of the mesh, finite element discretization, and the quality of the initial guess can be defined based on the idea of the \emph{Newton decrement} \citep{nocedal2006numerical}.
-The search direction $v_k$ at step $k$ for Newton's method is defined as
-\begin{equation}
-    v_k = -\ud^2J(u_k)^{-1}\ud J(u_k).
-\end{equation}
-Since the second derivative operator $\ud^2J(u_k)$ is positive-definite, this is a descent direction for the action $J$:
+Since the second derivative operator $\ud^2J(u_k)$ is positive-definite, the Newton search direction $v_k$ computed from equation \eqref{eq:newton-search-direction} is a descent direction for $J$:
 \begin{equation}

### 2019-10-08 icepack.tex
Started writing Larsen C section

@@
 
+To demonstrate the forward and inverse modeling capabilities of icepack, we will estimate the rheology of the Larsen C Ice Shelf in the Antarctic Peninsula from observational data and project the model forward by 10 years.
+Several recent studies have focused on Larsen C because it may be unstable in the warmer climate of the coming decades.
+From January to March of 2002, the neighboring Larsen B Ice Shelf disintegrated due to surface melt pond-induced fracture \textcolor{red}{check that this is the currently-accepted explanation and cite}, and this mechanism could also lead to the breakup of Larsen C.
+One of the key factors affecting the stability of ice shelves is the presence of marine ice -- seawater that freezes onto the base of an ice shelf -- in the suture zones where two flow units meet \citep{kulessa2014marine}.
+Marine ice is warmer than meteoric ice and usually includes brine pockets, which are discernible in radio echo sounding measurements as the absence of reflection from the ice shelf base \citep{holland2009marine}.
+This warmer and impurity-laden ice is more ductile and thus less prone to fracture than cold and brittle meteoric ice.
+Ocean models predict that marine ice forms under Larsen C as well \citep{holland2009marine}.
+By estimating the material hardness or rheology of an ice shelf, we can constrain where marine ice may be forming.
+
+\subsubsection{Data}
+
+\subsubsection{Parameterization}
+
+The rheology parameter of an ice shelf is strictly positive.
+The optimization algorithm, however, can explore unphysical regions of parameter space without some a priori constraint on what values are reasonable.
+Rather than try to solve an inequality-constrained problem, most studies in glaciology instead re-parameterize the problem in terms of some auxiliary field in such a way that the rheology is manifestly positive.
+In this case we will use the parameterization
+\begin{equation}
+    B = B_0e^{\theta}
+\end{equation}
+and estimate $\theta$.
+The inverse solver calculates the derivative of the objective functional symbolically and is thus agnostic to the particular parameterization.
+The only information that the user needs to pass is the name of the arguments to the forward model representing the parameter and the observed field so that these can be passed by keyword.
+
+\subsubsection{Results}
+
 \subsection{Jakobshavn Isbrae}

### 2019-10-09 icepack.tex
Added figure comparing sliding laws

@@
 where the critical stress $\tau_c$ is a certain specified fraction of the water pressure in the subglacial hydrological system.
-Rather than specify the sliding law as such, icepack expresses the sliding law as the derivative of some action functional.
-In other words, to implement the Schoof sliding law in icepack, we need to know the antiderivative of equation \eqref{eq:schoof-sliding}.
+In icepack sliding laws are not expressed directly but rather as the the derivative of an acction functional.
+To implement the Schoof sliding law we need to know the antiderivative of equation \eqref{eq:schoof-sliding}.
 Using a computer algebra system, we found that the antiderivative of this expression in terms of hypergeometric functions.
@@
 it behaves like an $m = 3$ power law at low sliding speeds and $m = \infty$ at high sliding speeds, but which is more tractable algebraically.
-The power dissipation density for this sliding law is
+Knowing the critical stress $\tau_c$ and the friction coefficient $C$, which has units of stress $\times$ speed${}^{-1/m}$, we can define a \emph{critical speed} $u_c$ as
 \begin{equation}
-    P = \tau_c\left(\left(u_c^{\frac{1}{m} + 1} + |u|^{\frac{1}{m} + 1}\right)^{\frac{m}{m + 1}} - u_c\right).
+    u_c = C^{-m}\tau_c^m.
 \end{equation}
+The power dissipation density for the Schoof-type sliding law that we use is
+\begin{equation}
+    P = \tau_c\left\{\left(u_c^{\frac{1}{m} + 1} + |u|^{\frac{1}{m} + 1}\right)^{\frac{m}{m + 1}} - u_c\right\}
+    \label{eq:power-dissipation-modified-schoof-sliding}
+\end{equation}
+Figure \ref{fig:sliding-laws} shows a comparison of the original Schoof sliding law and the sliding law that arises as the derivative of the functional in equation \eqref{eq:power-dissipation-modified-schoof-sliding}.
+The two have the same asymptotic behavior when the speed is much smaller or much larger than the critical speed; they differ in a relatively small range around the critical speed.
+The relative difference in basal shear stress between the two sliding laws is less than 10\% throughout the entire range.
+
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/sliding/sliding-law.png}
+    \caption{The Weertman, Schoof, and modified Schoof sliding law of equation \eqref{eq:power-dissipation-modified-schoof-sliding}.
+    The critical speed and shear stress are $u_c = 250$ m/year and $\tau_c = 100$ kPa.}
+    \label{fig:sliding-laws}
+\end{figure}
+
 A sample of the source code implementing both of these approaches is shown in \textcolor{red}{Figure X}.

### 2019-10-09 demos/sliding/sliding-law.py

@@ -1,0 +1,39 @@
+import argparse
+import matplotlib.pyplot as plt
+import numpy as np
+
+m = 3
+
+def τ_schoof(u, u_c, τ_c):
+    return τ_c * u**(1/m) / (u + u_c)**(1/m)
+
+def τ_modified(u, u_c, τ_c):
+    return τ_c * u**(1/m) / (u**(1/m + 1) + u_c**(1/m + 1))**(1/(m + 1))
+
+τ_c = 1e2
+u_c = 2.5e2
+
+u = np.geomspace(u_c / 100, 10 * u_c, 201)
+
+fig, ax = plt.subplots(figsize=(8, 3))
+
+τ_s = τ_schoof(u, u_c, τ_c)
+τ_m = τ_modified(u, u_c, τ_c)
+
+ax.plot([u_c, u_c], [0.95 * τ_s[0], 1.05 * τ_c], 'k--', linewidth=0.75, label='critical speed')
+
+ax.plot(u, τ_c / u_c**(1/m) * u**(1/m), 'C1', label='Weertman')
+ax.plot(u, τ_s, 'C2', label='Schoof')
+ax.plot(u, τ_m, 'C0', label='Modified Schoof')
+
+ax.set_xscale('log')
+ax.set_ylim(0.95 * τ_s[0], 1.05 * τ_c)
+ax.set_xlabel('speed (m / year)')
+ax.set_ylabel('stress (kPa)')
+ax.legend()
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--output')
+args = parser.parse_args()
+
+fig.savefig(args.output, dpi=300, bbox_inches='tight')

### 2019-10-09 icepack.tex
Explaining tradeoffs of variational principles

@@
 
-\begin{figure}[h]
-    \includegraphics[width=0.95\linewidth]{demos/mismip/steady-state.png}
-    \caption{The steady state thickness and velocity of the MISMIP+ experimental setup with the Weertman sliding law.}
-    \label{fig:mismip}
-\end{figure}
-
 The original intercomparison project specified that participants could use the Weertman sliding law (equation \eqref{eq:weertman-sliding}) as well as two other sliding laws that transition to a more plastic rheology at high sliding speeds.
@@
 Figure \ref{fig:mismip} shows the outcome of the MISMIP+ experiment performed using icepack.
+
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/mismip/steady-state.png}
+    \caption{The steady state thickness and velocity of the MISMIP+ experimental setup with the Weertman sliding law.}
+    \label{fig:mismip}
+\end{figure}
+
+Using variational principles to express all constitutive laws is less flexible than specifying the sliding law directly and this is a distinct disadvantage.
+We were able to implement a sliding law that exhibits many of the characteristics of the Schoof sliding law.
+\citet{asay2016experimental} also suggest using a sliding law that transitions sharply to exact plasticity above the critical speed.
+Expressing this sliding law in UFL requires a conditional in the velocity and is thus no longer differentiable.
+This lack of differentiability makes the forward solver crash.
+The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.
 

### 2019-10-09 demos/mismip/mismip+.py
Using Schoof sliding law in MISMIP demo

@@ -59,12 +59,15 @@ C = Constant(1e-2)
 from icepack.constants import (ice_density as ρ_I, water_density as ρ_W,
                                gravity as g, weertman_sliding_law as m)
 def friction(u, h, s, C):
-    u_b = sqrt(inner(u, u))
     p_W = ρ_W * g * max_value(0, -(s - h))
     p_I = ρ_I * g * h
     N = p_I - p_W
-    ϕ = N / p_I
-    return m / (m + 1) * C * ϕ * u_b**(1/m + 1) * dx
+    τ_c = N / 2
+
+    u_c = (τ_c / C)**m
+    u_b = sqrt(inner(u, u))
+
+    return τ_c * ((u_c**(1/m + 1) + u_b**(1/m + 1))**(m / (m + 1)) - u_c) * dx
 
 model = icepack.models.IceStream(friction=friction)
 opts = {'dirichlet_ids': [1], 'side_wall_ids': [3, 4], 'tol': 1e-8}

### 2019-10-10 demos/mismip/mismip+.py
MISMIP+ demo works at more than one resolution
To reach steady state, we start from a coarse mesh and successively
refine until we reach the (much higher) resolution necessary for the
retreat part of the MISMIP+ scenario.

@@ -10,20 +10,26 @@ import icepack, icepack.models, icepack.plot
 parser = argparse.ArgumentParser()
 parser.add_argument('--verbose', action='store_true')
 parser.add_argument('--input')
-parser.add_argument('--output', default='result')
+parser.add_argument('--output')
+parser.add_argument('--input-level', type=int, default=0, dest='input_level')
+parser.add_argument('--output-level', type=int, default=0, dest='output_level')
 parser.add_argument('--melt', action='store_true')
 parser.add_argument('--time', type=float, default=4e3)
 parser.add_argument('--timestep', type=float, default=1.0)
 
 args = parser.parse_args()
+if args.output_level < args.input_level:
+    raise ValueError('Output level must be >= input level!')
 
 # Create the mesh and function spaces
 Lx, Ly = 640e3, 80e3
-ny = 40
+ny = 20
 nx = int(Lx/Ly) * ny
-mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
-area = firedrake.assemble(Constant(1) * dx(mesh))
+area = Lx * Ly
 
+coarse_mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
+mesh_hierarchy = firedrake.MeshHierarchy(coarse_mesh, args.output_level)
+mesh = mesh_hierarchy[args.output_level]
 Q = firedrake.FunctionSpace(mesh, family='CG', degree=1)
 V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=1)
 
@@ -69,17 +75,39 @@ def friction(u, h, s, C):
 
     return τ_c * ((u_c**(1/m + 1) + u_b**(1/m + 1))**(m / (m + 1)) - u_c) * dx
 
+# Create the model object and extra options
 model = icepack.models.IceStream(friction=friction)
 opts = {'dirichlet_ids': [1], 'side_wall_ids': [3, 4], 'tol': 1e-8}
 
-# Read in the input data if there is any
+# Read the input data if there is any
 if args.input:
-    h0 = firedrake.Function(Q)
-    u0 = firedrake.Function(V)
+    if args.input_level < args.output_level:
+        input_mesh = mesh_hierarchy[args.input_level]
+        Q_in = firedrake.FunctionSpace(input_mesh, family='CG', degree=1)
+        V_in = firedrake.VectorFunctionSpace(input_mesh, family='CG', degree=1)
+    else:
+        Q_in = Q
+        V_in = V
+
[…]

### 2019-10-10 demos/mismip/plot.py

@@ -5,14 +5,17 @@ import icepack.plot
 
 parser = argparse.ArgumentParser()
 parser.add_argument('--input')
+parser.add_argument('--level', type=int)
 parser.add_argument('--output')
 
 args = parser.parse_args()
 
 Lx, Ly = 640e3, 80e3
-ny = 40
+ny = 20
 nx = int(Lx/Ly) * ny
-mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
+coarse_mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
+mesh_hierarchy = firedrake.MeshHierarchy(coarse_mesh, args.level)
+mesh = mesh_hierarchy[args.level]
 
 Q = firedrake.FunctionSpace(mesh, family='CG', degree=1)
 V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=1)

### 2019-10-14 icepack.tex
Symbol table + description of SSA

@@
 
+\begin{table}[h]
+    \begin{tabular}{l|l}
+        Symbol & Meaning \\
+        \hline
+        $h$ & thickness \\
+        $b$ & bed elevation  \\
+        $s$ & surface elevation  \\
+        $d$ & water depth \\
+        $u$ & velocity \\
+        $\dot a_s$ & surface mass balance \\
+        $\dot a_b$ & basal mass balance \\
+        $\dot\varepsilon$ & strain rate \\
+        $C$ & bed friction coefficient \\
+        $A$ & rheology coefficient \\
+        $\nu$ & unit outward normal \\
+        $J$ & action functional
+    \end{tabular}
+    \caption{Mathematical symbols}
+\end{table}
+
+\begin{table}[h]
+    \begin{tabular}{l|l}
+        Symbol & Meaning \\
+        \hline
+        $n$ & Glen's flow law exponent \\
+        $m$ & Weertman sliding law exponent \\
+        $\rho_I$ & ice density \\
+        $\rho_W$ & seawater density \\
+        $g$ & gravitational acceleration
+    \end{tabular}
+    \caption{Physical constants}
+\end{table}
+
 \subsection{Mass transport}
@@
 \subsubsection{Shallow stream approximation}
+
+The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
+In this case, the momentum equation can be vertically integrated to obtain a 2D system of equations.
+The terms in the action functional are:
+\begin{align}
+    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}\ud x \\
+    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}\ud x \\
+    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\ud x \\
+    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\ud\gamma
+\end{align}
+and the action functional is
+\begin{equation}
+    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}.
+\end{equation}
+When the ice is floating, there are two simplifications: the friction coefficient $C$ is 0, and the surface elevation $s$ can be written in terms of the thickness $h$ as
+\begin{equation}
+    s = (1 - \rho_I / \rho_W)h.
+\end{equation}
+Using integration by parts, the terminus component can be eliminated completely.
+
+Several studies have compared the shallow stream approximation to higher-order models, such as Blatter-Pattyn or full Stokes \textcolor{red}{cite the MIPs}.
+The most appreciable difference between lower- and higher-order models occurs near the glacier grounding line, where the full Stokes equations can represent bridging stresses \textcolor{red}{cite van der Veen}.
+The shallow stream approximation is warranted for studying individual fast-flowing glaciers where the concern is more with studying processes and responses than with complete accuracy.
 
@@
 For example, in defining a bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance.
-In that case using positional arguments would better correspond with the underlying problem.
+In this case, positional arguments correspond better with the semantics of the problem.
 
@@
 We use a damped Newton method to solve the diagnostic equations.
-Starting from a guess $u_k$ for the velocity, one first computed a \emph{search direction} $v_k$ that solves a linear equation:
+Starting from a guess $u_k$ for the velocity, the \emph{search direction} $v_k$ is the unique solution of the linear system
 \begin{equation}
@@
 \end{equation}
-The next value of the velocity is then computed by searching along $v_k$:
-\begin{equation}
-    u_{k + 1} = u_k + \alpha_k\cdot v_k,
-\end{equation}
-where $\alpha$ is chosen to minimize $J$ along the search direction.
+The next approximation for the velocity minimizes $J$ along the search direction $v_k$ starting from $u_k$, i.e.
+\begin{align}
+    u_{k + 1} & = u_k + \alpha_k\cdot v_k, \nonumber\\
+    \alpha_k & = \text{argmin}_\alpha \hspace{2pt}J(u_k + \alpha v_k).
+\end{align}
 For an initial guess sufficiently close to the exact solution, the undamped Newton method ($\alpha_k = 1$ at every step) converges quadratically.
@@
 
-Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation, but few if any have considered the problem of when to stop iterating.
+Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation \textcolor{red}{cite -- I think Mathieu and Dan did this}, but few if any have considered the problem of when to stop iterating.
 The most common stopping criteria are when (1) the 2-norm of the residual is sufficiently small or (2) the relative change in the iterates is sufficiently small.
@@
 The increased melt thins the ice ice shelf and initiates a retreat of the glacier grounding line.
-Finally, submarine melting is turned off for at least 200 years, optionally longer.
+Finally, submarine melting is turned off for at least 100 years, optionally longer.
 The grounding line then readvances, but not as far as its original position.

### 2019-11-12 icepack.tex
Removing some planned sections, add one for ice sheet
Added Larsen results to paper

@@
 
-\subsection{\textcolor{green}{Plumes}}
-
 \section{Numerics}
@@
 
+\subsection{Synthetic ice sheet}
+
 \subsection{Larsen C Ice Shelf}
 
-To demonstrate the forward and inverse modeling capabilities of icepack, we will estimate the rheology of the Larsen C Ice Shelf in the Antarctic Peninsula from observational data and project the model forward by 10 years.
+To demonstrate the inverse solver, we will estimate the rheology of the Larsen C Ice Shelf in the Antarctic Peninsula from observational data.
 Several recent studies have focused on Larsen C because it may be unstable in the warmer climate of the coming decades.
@@
 Ocean models predict that marine ice forms under Larsen C as well \citep{holland2009marine}.
-By estimating the material hardness or rheology of an ice shelf, we can constrain where marine ice may be forming.
+By estimating the material rigidity of an ice shelf, we can constrain where marine ice may be forming.
 
 \subsubsection{Data}
+
+We used the InSAR phase-based ice velocity map from \citet{mouginot2019continent}.
+This dataset has nominal errors over the Larsen Ice Shelf between \textcolor{red}{find it}.
+We used the recently-released BedMachine map of the thickness of Antarctica \citep{morlighem2019deep}.
+This data product improves on existing thickness maps for the continent by using the mass conservation law, together with measurements of the velocity, surface elevation, and surface mass balance, to improve thickness estimates where few measurements are available.
+
+Existing work on glaciological inverse problems uses the mismatch between the computed and observed velocity fields as part of the objective functional.
+The effect of thickness errors is studied largely through a posteriori validation \citep{joughin2004basal, larour2005rheology}.
+Errors in thickness or surface slope can be large enough that it might be impossible to fit the velocity measurements to the degree that statistical theory predicts \citep{macayeal1995basal}.
+The velocity measurements themselves might have significant outliers, in which case using the usual weighted sum of squared misfits as an error metric will give poor results.
+For these reasons, some studies have explored alternative objective functionals \citep{morlighem2010spatial}.
+We have opted to use the regularized $L^1$-type error metric
+\begin{equation}
+    E(u - u^o) = \int_\Omega\left(\sqrt{\frac{|u - u^o|^2}{\sigma^2} + \gamma^2} - \gamma\right)\ud x.
+    \label{eq:l1-error-metric}
+\end{equation}
+This error metric approaches the usual weighted sum of squared errors as $\gamma \to \infty$, and approaches the sparsity-promoting $L^1$ error metric as $\gamma \to 0$.
+For finite, positive values of $\gamma$ this error metric is robust to non-normality or a small fraction of outliers \citep{barron2019general}.
+
+\citet{adusumilli2018variable} used surface elevation measurements from multiple satellite altimeters, together with firn density estimates from a climate model and tide height estimates, to compile a time series of ice thickness change over the Larsen Ice Shelf for the past 23 years.
+This same approach can be used to derive an improved estimate the thickness of the ice shelf from, e.g. the REMA surface elevation dataset.
+Future work will focus on this and other methods to assimilate more accurate thickness data and to explicitly account for the statistical errors in this field.
 
@@
 \begin{equation}
-    B = B_0e^{\theta}
+    A = A_0e^{\theta}
 \end{equation}
@@
 
-\subsection{Jakobshavn Isbrae}
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/larsen/parameter.png}
+    \caption{Left: Inferred fluidity parameter $\theta$.
+    Lower values indicate more deformable ice.
+    Right: stream plot of ice velocity computed from this fluidity parameter.}
+    \label{fig:larsen}
+\end{figure}
+
+The inferred parameter field $\theta$ is shown in figure \ref{fig:larsen}.
+The algorithm detects areas of much lower ice rigidity around highly damaged ice.
+This feature is especially apparent around the large rift emanating from the Gipps Ice Rise, as well as the crevassed areas upstream.
+We find other areas of low rigidity in the suture zones where two flow units converge and where marine ice tends to form, releasing heat to the ice shelf, exactly as observed in \citet{holland2009marine}.
+Finally, the inferred rheology field reproduces features that have already been found in previous studies of Larsen C Ice Shelf \citep{khazendar2011acceleration}.
+
+The final value of the model-data misfit \eqref{eq:l1-error-metric} matches the value we would expect if the velocity errors actually came from this probability distribution.
+In a separate computational experiment, we used the older bedmap2 ice thickness map instead of BedMachine \citep{fretwell2013bedmap2}.
+We were unable to achieve the same model-data misfit using bedmap2 at the same grid resolution with any regularization parameter.
 
@@
 The fallible human learning to use the model is largely absent from the discussion.
+For graduate students or other researchers who are not experts in computational physics, the difficulty of learning to use a particular modeling tool may be more of a rate-limiting factor than the speed or efficiency of the model.
 
@@
 \textcolor{red}{Picture of an executed jupyter notebook.}
-Finally, a Jupyter notebook and the accompanying plots can be rendered into HTML and published using the tool \texttt{nbconvert}.
+The interface for icepack also supports inspection of partial results in other ways.
+Running a simulation amounts to writing a loop that alternates prognostic and diagnostic solves.
+Users can add arbitrary code to this loop for sanity checking or feedback on progress, or they can manually step through the simulation one iteration at a time.
+Other packages only support a more coarse-grained view.
+The user specifies the starting and ending time of a simulation, and the degree of feedback is hard-coded into the package itself.
 
@@
 Can irrelevant details be hidden?}
-The API for icepack has been designed so that the user only needs to make decisions about what problem they wish to solve and not how to solve it.
-Where a choice concerns more the ``how'' than the ``what'', a sensible default has been chosen.
-Moreover, the default behavior we choose always biases for correctness rather than speed.
+The API for icepack has been designed so that the users only need to decide what problem to solve and not how to solve it.
+Where a choice does concern more the ``how'' than the ``what'', we use a sensible default that biases for correctness rather than speed.
 For example, the Newton solver uses a direct factorization method to solve the linear system for the search direction because factorization requires no tuning whereas iterative methods do.
-A user interested in achieving greater runtime performance can pass additional keyword arguments instead specifying, say, the conjugate gradient method.
+A user interested in achieving greater runtime performance can pass additional keyword arguments instead specifying, say, the preconditioned conjugate gradient method.
 This choice is of interest mostly to advanced users so we keep the linear solver method as a default argument.

### 2020-01-26 icepack.tex
Fixes to usability section

--- 2019-10-14 icepack.tex
@@
 
+\subsection{Synthetic ice sheet}
+
 \subsection{Larsen C Ice Shelf}
@@
 
-\subsection{Jakobshavn Isbrae}
-
 \section{Usability}

### 2020-01-29 demos/gibbous/ice-shelf.py
Add ice shelf damage example

@@ -1,0 +1,120 @@
+import os
+import argparse
+import tqdm
+import firedrake
+from firedrake import as_vector, inner, ds
+import icepack
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--mesh')
+parser.add_argument('--input')
+parser.add_argument('--output')
+parser.add_argument('--damage', action='store_true')
+parser.add_argument('--final-time', type=float)
+parser.add_argument('--num-steps', type=int)
+args = parser.parse_args()
+
+# Load the mesh and create some function spaces
+mesh = firedrake.Mesh(args.mesh)
+Q = firedrake.FunctionSpace(mesh, family='CG', degree=2)
+V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=2)
+Δ = firedrake.FunctionSpace(mesh, family='DG', degree=1)
+
+# Load the initial data from an HDF5 file if available
+if args.input:
+    h0 = firedrake.Function(Q)
+    u0 = firedrake.Function(V)
+
+    input_name = os.path.splitext(args.input)[0]
+    with firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ) as chk:
+        chk.load(h0, name='h')
+        chk.load(u0, name='u')
+
+# Otherwise, create some synthetic initial data symbolically
+else:
+    import numpy as np
+    from numpy import pi as π
+
+    inlet_angles = π * np.array([-3/4, -1/2, -1/3, -1/6])
+    inlet_widths = π * np.array([1/8, 1/12, 1/24, 1/12])
+
+    x = firedrake.SpatialCoordinate(mesh)
+
+    R = 200e3
+    u_in = 300
+    h_in = 350
+    hb = 100
+    dh, du = 400, 250
+
+    hs, us = [], []
+    for θ, ϕ in zip(inlet_angles, inlet_widths):
+        x0 = R * as_vector((np.cos(θ), np.sin(θ)))
+        v = -as_vector((np.cos(θ), np.sin(θ)))
+        L = inner(x - x0, v)
+        W = x - x0 - L * v
+        Rn = 2 * ϕ / π * R
+        q = firedrake.max_value(1 - (W / Rn)**2, 0)
+        hs.append(hb + q * ((h_in - hb) - dh * L /R))
+        us.append(firedrake.exp(-4 * (W/R)**2) * (u_in + du * L / R) * v)
+
+    h_expr = firedrake.Constant(hb)
+    for h in hs:
[…]

### 2020-01-29 demos/gibbous/make_mesh.py

@@ -1,0 +1,29 @@
+import argparse
+import numpy as np
+import pygmsh
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--output')
+args = parser.parse_args()
+
+R = 200e3
+δx = 5e3
+geometry = pygmsh.built_in.Geometry()
+
+x1 = geometry.add_point([-R, 0, 0], lcar=δx)
+x2 = geometry.add_point([+R, 0, 0], lcar=δx)
+
+center1 = geometry.add_point([0, 0, 0,], lcar=δx)
+center2 = geometry.add_point([0, -4 * R, 0], lcar=δx)
+
+arcs = [geometry.add_circle_arc(x1, center1, x2),
+        geometry.add_circle_arc(x2, center2, x1)]
+
+line_loop = geometry.add_line_loop(arcs)
+plane_surface = geometry.add_plane_surface(line_loop)
+
+physical_lines = [geometry.add_physical(arc) for arc in arcs]
+physical_surface = geometry.add_physical(plane_surface)
+
+with open(args.output, 'w') as geo_file:
+    geo_file.write(geometry.get_code())

### 2020-01-29 demos/gibbous/make_plots.py

@@ -1,0 +1,72 @@
+import numpy as np
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+import firedrake
+from firedrake import sqrt, inner
+import icepack.plot
+
+def colorbar(figure, axes, mappable, *args, **kwargs):
+    divider = make_axes_locatable(axes)
+    cax = divider.append_axes('right', size='5%', pad=0.05)
+    return figure.colorbar(mappable, *args, cax=cax, **kwargs)
+
+mesh = firedrake.Mesh('ice-shelf.msh')
+Q = firedrake.FunctionSpace(mesh, family='CG', degree=2)
+V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=2)
+Δ = firedrake.FunctionSpace(mesh, family='DG', degree=1)
+
+filename = 'steady-state-undamaged'
+with firedrake.DumbCheckpoint(filename, mode=firedrake.FILE_READ) as chk:
+    h_undamaged = firedrake.Function(Q)
+    u_undamaged = firedrake.Function(V)
+    chk.load(h_undamaged, name='h')
+    chk.load(u_undamaged, name='u')
+
+filename = 'steady-state-damaged'
+with firedrake.DumbCheckpoint(filename, mode=firedrake.FILE_READ) as chk:
+    h_damaged = firedrake.Function(Q)
+    u_damaged = firedrake.Function(V)
+    D = firedrake.Function(Δ)
+    chk.load(h_damaged, name='h')
+    chk.load(u_damaged, name='u')
+    chk.load(D, name='D')
+
+δh = firedrake.interpolate(h_damaged - h_undamaged, Q)
+
+speed = sqrt(inner(u_undamaged, u_undamaged))
+v = u_undamaged / speed
+δu = firedrake.interpolate(inner(u_damaged - u_undamaged, v), Q)
+
+fig, axes = icepack.plot.subplots(nrows=1, ncols=2, sharex=True, sharey=True)
+axes[0].set_ylabel('distance (m)')
+
+umax = icepack.norm(u_undamaged, norm_type='Linfty')
+umax = (int(umax // 50) + 1) * 50
+streamlines = icepack.plot.streamplot(u_undamaged, precision=1e3, density=5e3, axes=axes[0])
+colorbar(fig, axes[0], streamlines, label='velocity (m/a)')
+
+dumax = icepack.norm(δu, norm_type='Linfty')
[…]

### 2020-09-03 icepack.tex
Add more on usability

--- 2019-10-14 icepack.tex
@@
 In this paper we introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
-Icepack is built on the finite element modeling library firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
-The diagnostic models in icepack are formulated through action principles that are specified in UFL.
-The components of each action functional can be substituted for different forms of the user's choosing, which makes it easy to experiment with the model physics.
-Many post-processing and analysis tasks on simulation results also amount to the evaluation of some functional.
+Icepack is built on the finite element modeling library firedrake, which implements the domain-specific language UFL for the specification of weak forms of partial differential equations.
+The diagnostic models implemented in icepack use variational formulations that are specified in UFL.
+Individual components of each model's action functional can be easily substituted for components of the user's choosing, facilitating experimentation with the model physics.
+Additionally, many post-processing and analysis tasks on simulation results amount to the evaluation of some functional.
 By using a variational formulation of the model physics, the specification of a problem and the analysis of the solution employ the same conceptual vocabulary.
@@
 Icepack features a 3D diagnostic model based on terrain-following coordinates and vertical spectral discretization.
-This model resolves both plug- and shear-flow components of horizontal ice flow with a minimum of extra computational expense over 2D, depth-averaged models.
+This model resolves both plug- and shear-flow components of horizontal ice flow with a minimum of computational expense over 2D, depth-averaged models.
 Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
-The overall design philosophy of icepack is to be as usable as possible for a wide a swathe of the glaciological community, including both experts and novices in computational science.
+The overall design philosophy of icepack is to enable usability for as wide a swathe of the glaciological community as possible, including both experts and novices in numerical PDE solvers and high-performance computing.
 \end{abstract}
@@
     \item predicting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
-    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood,
+    \item exploring poorly-understood aspects of glacier physics, such as hydrology and calving,
     \item estimating unobservable quantities, such as bed friction or rheology, from observational data, and
@@
 Several glacier flow models already exist and are effective in the hands of experts.
-These models are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
+These models are usually written in compiled programming languages such as C, C++, and Fortran, which afford an unsurpassed degree of computational speed.
 Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
-The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
-Our goal is to make a tool that will be both accessible to non-experts and useful for experts.
+The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for using numerical models to glaciologists who are not experts in modeling as such.
+We wanted to make a tool that would be both accessible to non-experts and useful for experts.
+
+\textcolor{red}{We focus in this paper only on novel aspects of icepack that differ substantially from existing packages.}
 
@@
 VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}.
-The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for describing weak forms of PDE.
-Icepack improves upon the groundwork laid in VarGlaS in three main respects.
-First, icepack includes a simple 3D flow model that uses several features only available in firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
-This model will be described in section \S\ref{sec:physics-hybrid-model}.
-Second, the architecture of icepack is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of the models included.
-The software design that made this possible is discussed in section \S\ref{sec:physics-substitution}.
-Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.
+The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for specifying weak forms of PDE.
+Icepack improves upon the groundwork laid in VarGlaS through its use of tensor product finite elements on extruded meshes, both of which are only available in firedrake \citep{bercea2016structure, mcrae2016automated}.
+These two features enable the definition of a much simpler 3D flow model that we will describe in the next section.
 
@@
 The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation.
-The inputs to the diagnostic equation are the ice thickness, surface elevation, velocity at the inflow boundary, rheology, and bed friction coefficient.
-The output of the diagnostic equation is the ice velocity throughout the entire domain.
+The inputs to the diagnostic equation are the ice thickness, surface elevation, rheology, ice velocity at the inflow boundary, and the coefficient of friction for contact with the bed and side walls.
+The output of the diagnostic equation is the ice velocity throughout the entire region of interest.
 The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
-The inputs to the prognostic equation are the current value of the ice thickness, velocity, and the surface and basal mass balance.
-The output is the ice thickness at the subsequent timestep.
+The inputs to the prognostic equation are the current value of the ice thickness, the ice velocity, and the surface and basal mass balance.
+The output is the ice thickness at a later time.
 Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation.
 
-The rheology and friction coefficient are functions of other fields that have their own evolution equations.
-For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
-Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock.
+The rheology and friction coefficient may also be described in terms of other fields.
+For example, the rheology can be parameterized in terms of the ice temperature, englacial water content, and damage state.
+Likewise, the friction coefficient can be described in terms of a geological roughness factor and the subglacial water pressure.
 The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.
 
-\section{Physics}
+\section{Diagnostic models}
 
-\begin{table}[h]
-    \begin{tabular}{l|l}
-        Symbol & Meaning \\
-        \hline
-        $h$ & thickness \\
-        $b$ & bed elevation  \\
-        $s$ & surface elevation  \\
-        $d$ & water depth \\
-        $u$ & velocity \\
-        $\dot a_s$ & surface mass balance \\
-        $\dot a_b$ & basal mass balance \\
-        $\dot\varepsilon$ & strain rate \\
-        $C$ & bed friction coefficient \\
-        $A$ & rheology coefficient \\
-        $\nu$ & unit outward normal \\
-        $J$ & action functional
-    \end{tabular}
-    \caption{Mathematical symbols}
-\end{table}
-
-\begin{table}[h]
-    \begin{tabular}{l|l}
-        Symbol & Meaning \\
-        \hline
-        $n$ & Glen's flow law exponent \\
-        $m$ & Weertman sliding law exponent \\
-        $\rho_I$ & ice density \\
-        $\rho_W$ & seawater density \\
-        $g$ & gravitational acceleration
-    \end{tabular}
-    \caption{Physical constants}
-\end{table}
-
-\subsection{Mass transport}
-
-The PDE describing how the ice thickness changes in time is a conservative advection equation:
-\begin{equation}
-    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
-\end{equation}
-where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
-The thickness is represented using continuous Galerkin basis functions.
-The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
-The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.
-
-The surface elevation is calculated at each timestep of a simulation as
-\begin{equation}
-    s = \max\{b + h, (1 - \rho_I / \rho_W)h\},
-\end{equation}
-the first case corresponding to grounded ice and the second case corresponding to floating ice.
-In the immediate vicinity of the grounding line, the assumption that the floating ice is in hydrostatic equilibrium with the ocean fails.
-Most models assume hydrostasy and icepack does as well.
-Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.
-
-\subsection{Momentum transport}
-
-There are \textcolor{green}{three} diagnostic models implemented in icepack.
-The \emph{shallow ice approximation} (SIA) and shallow stream approximation (SSA) are 2D models describing the depth-averaged velocity of a glacier.
-\textcolor{green}{The SIA model describes the deep interior of ice sheets well, where flow velocities are on the order of 0.1 - 10 m/year.
-This model assumes that nearly all flow is by vertical shear, which happens when the ice is frozen to the bed or sliding at a small fraction of the surface speed.}
-The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.
-This model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
+There are two diagnostic models implemented in icepack.
+The \emph{shallow stream approximation} (SSA) is a 2D model describing the depth-averaged velocity of a fast-flowing grounded ice stream or floating ice shelf \citep{macayeal1989large}.
+The SSA model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
 Plug flow is a good approximation in fast-flowing ice streams and outlet glaciers near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
@@
 
-\subsubsection{Variational principles}
-
-All of the diagnostic models in icepack are described through \emph{variational} or \emph{action} principles \citep{dukowicz2010consistent}.
-Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a functional called the action.
+Both the SSA and BP models in icepack are described through \emph{action principles} \citep{dukowicz2010consistent}.
+Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a certain nonlinear functional, called the action.
 The action consists of four terms:
@@
 
-The action principle is especially useful for designing a robust numerical solver.
+Irrespective of any deep connections with thermodynamics, the existence of an action principle is a tremendous convenience for numerical analysis.
 For viscous flow problems near to steady-state, the action is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
@@
 
-\subsubsection{\textcolor{green}{Shallow ice approximation}}
+\subsection{Implementation with firedrake}
 
-\subsubsection{Shallow stream approximation}
+At the highest level, each physics model in icepack is represented by a Python class.
+So far, the models available are \texttt{IceShelf}, \texttt{IceStream}, and \texttt{HybridModel}.
+Each class has three main responsibilities.
+First, the model class has to have a method that takes in the various fields (ice velocity, thickness, surface elevation, etc.) and returns a symbolic representation of the action functional for that particular model.
+Second, the model class must have a method that takes in the input fields and returns an approximate minimizer of the action functional.
+The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the first method calculates.
+The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
+Finally the model must have a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
+The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.
 
-The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
-In this case, the momentum equation can be vertically integrated to obtain a 2D system of equations.
-The terms in the action functional are:
-\begin{align}
-    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}\ud x \\
-    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}\ud x \\
-    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\ud x \\
-    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\ud\gamma
-\end{align}
-and the action functional is
-\begin{equation}
-    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}.
-\end{equation}
-When the ice is floating, there are two simplifications: the friction coefficient $C$ is 0, and the surface elevation $s$ can be written in terms of the thickness $h$ as
-\begin{equation}
-    s = (1 - \rho_I / \rho_W)h.
-\end{equation}
-Using integration by parts, the terminus component can be eliminated completely.
+\textcolor{red}{Some sample code.}
 
-Several studies have compared the shallow stream approximation to higher-order models, such as Blatter-Pattyn or full Stokes \textcolor{red}{cite the MIPs}.
-The most appreciable difference between lower- and higher-order models occurs near the glacier grounding line, where the full Stokes equations can represent bridging stresses \textcolor{red}{cite van der Veen}.
-The shallow stream approximation is warranted for studying individual fast-flowing glaciers where the concern is more with studying processes and responses than with complete accuracy.
+The Unified Form Language for specifying weak forms of PDEs contains all of the primitives necessary to express individual terms of the action functional.
+These primitives largely consist of the basic vector calculus operators like the gradient of a field, tensor calculus operations like taking the dot product of two vectors or tensors, scalar functions like the square root or exponential, and symbolic integration over the mesh or its boundary.
+For example, the strain rate for a given velocity field $u$ can be written as \texttt{sym(grad(u))}, where the function \texttt{grad} represents the symbolic gradient of a field and \texttt{sym} represents the symmetrization of a rank-2 tensor.
 
-\subsection{Hybrid model} \label{sec:physics-hybrid-model}
+\textcolor{red}{Brief discussion of SSA model.}
 
-The shallow ice and shallow stream models in icepack are completely standard.
-The hybrid flow model that we describe here is based on variational principles, terrain-following coordinates, and vertical spectral methods.
-Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
-\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
-\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
-\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem, but this model used a finite difference discretization in the vertical direction.
-
-\subsubsection{Terrain-following coordinates}
-
-Rather than use the usual Cartesian coordinate system, the hybrid flow model uses terrain-following coordinates.
-The terrain-following vertical coordinate $\zeta$ is
+The 3D flow model, as represented in the \texttt{HybridModel} class, uses several features that are only available in Firedrake to better exploit the underlying mathematics.
+First, rather than represent the 3D physical domain itself, we instead use \emph{terrain-following coordinates} and \emph{extruded meshes}.
+The terrain-following vertical coordinate $\zeta$ is defined as
 \begin{equation}
@@
     \dot\varepsilon_z(u) = \dot\varepsilon_\zeta(u) + \frac{1}{2}\left(u \otimes\nabla\zeta + \nabla\zeta\otimes u\right)
-    \label{eq:geometric-correction}
 \end{equation}
 where $\otimes$ is the tensor product of two vectors.
+The geometric correction factor can easily be represented in UFL.
+By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is no more complex than in Cartesian coordinates.
 
-For the Stokes equations, this alternative coordinate system also helps avoid the problem of how to enforce the condition $u\cdot \nu = -\dot a_b$ at the ice base, where $\nu$ is the unit outward normal vector and $\dot a_b$ is the basal mass balance.
+Terrain-following coordinates are especially advantageous on extruded meshes.
+An extruded mesh takes a 2D footprint mesh and lifts it into 3D by taking the Cartesian product with the unit interval.
+The cells of an extruded mesh are triangular prisms instead of the more common tetrahedra used for general 3D meshes.
+In Firedrake, an extruded mesh is created by calling the function \texttt{ExtrudedMesh} on the 2D footprint.
+Not every 3D domain can be described by extruding a 2D domain, but the geometry of most glacier flow problems can.
+
+A 2D mesh can be extruded to any number of vertical layers.
+In Elmer/Ice, for example, the user chooses the number of vertical layers in order to set the resolution along this axis \citep{gagliardini2013capabilities}.
+The horizontal velocity is expected to be very smooth as a function of depth.
+For example, under the plug flow approximation, the horizontal velocity is constant with depth.
+Under the shallow ice approximation, the horizontal velocity varies with depth as $1 - (1 - \zeta)^{n + 1}$ where $n = 3$ is the Glen flow law exponent.
+This extra information about our solution suggests a different discretization strategy.
+
+Rather than extrude the footprint mesh into many vertical layers, we might instead use only one vertical layer and increase the polynomial degree in the vertical direction only to obtain higher resolution.
+To support this use case, Firedrake includes the ability to define \emph{tensor product} elements on extruded meshes.
+Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
+\begin{equation}
+    \Phi_{kl}(x, y, \zeta) = \phi_k(x, y)\psi_l(\zeta).
+\end{equation}
+For example, we can use piecewise linear or quadratic elements on triangles and use quintic or higher degree polynomials in the vertical.
+Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
+Rather than use the usual Lagrange interpolating polynomial basis in the vertical dimension, we can instead use the Legendre polynomial basis.
+The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.
+To select the Legendre polynomial basis, the user passes the keyword argument \texttt{vfamily=`Gauss-Legendre'} or \texttt{`GL'} for short to the constructor for the function space.
+
+The combination of using extruded meshes and tensor product elements in the vertical direction is referred to as \emph{semi-discretization}.
+This approach can be thought of merely as a way to discretize a PDE that has special structure.
+Alternatively, we can view vertical semi-discretization as defining a family of models indexed by the number of vertical basis functions.
+The order-$d$ model defines a coupled system of PDEs for $d$ vector fields.
+Each vector field represents one mode of vertical variability, similar to the distinction between barotropic and baroclinic modes in atmospheric physics and oceanography.
+The system is then discretized in the horizontal and solved numerically.
+In any case, the code is the same regardless of how one wishes to think about the underlying mathematics.
+
+We are then faced with the choice of how many vertical basis functions to use.
+Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
+The degree 1 model is the most minimalistic model that still exhibits vertical shear.
+Going to higher and higher degree gives a more accurate approximation at the expense of greater computational effort.
+
+This approach for implementing a 3D flow model works completely seamlessly but for one important detail.
+The backpressure from ocean water at the calving front of a marine-terminating glacier is not a smooth function of depth.
+The pressure is 0 above the water line and linearly increasing below it:
+\begin{equation}
+    \text{backpressure power} = \int_\Gamma\int_0^1 \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+\ud\zeta\ud\gamma,
+    \label{backpressure}
+\end{equation}
+where $\zeta_{\text{sl}}$ denotes the relative depth to the water line and the subscript $+$ denotes the positive part of a real number.
+Were we to use the standard asssembly procedure in Firedrake to evaluate this integral, we would get an inaccurate result due to an insufficient number of integration points.
+The resulting velocity solutions are then wildly inaccurate due to the mis-specification of the Neumann boundary condition.
+A blunt solution to this problem would be to pass an extra argument to the Firedrake form compiler that specifies a much greater integration accuracy in the vertical for this term.
+This fix reduces, but does not eliminate, the errors in the velocities, and it incurs an appreciable computational cost.
+
+We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
+The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
+Since the result of this symbolic Legendre polynomial expansion is the same upon every invocation, we can save the effort of recalculating it every time by caching the results.
+The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts \textcolor{red}{figure}.
+While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property.
+Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the expansion limit.
+
+The exact symbolic integration approach is both faster and more accurate than using a large number of quadrature points.
+The same technique could be used to exactly calculate the ocean backpressure for any model, say the full Stokes equations, using terrain-following coordinates together with a Legendre polynomial expansion in the vertical.
+
+For the Stokes equations, transforming the vertical coordinate also circumvents the issue of how to enforce the partial Dirichlet boundary condition $u\cdot \nu = 0$ at the ice base, where $\nu$ is the unit outward normal vector.
 This boundary condition is difficult to impose exactly because the unit outward normal vector $\nu$ is defined on mesh faces while the velocity is defined at mesh vertices.
 Elmer/Ice uses an ad-hoc procedure to define the unit normal vectors at mesh nodes \citep{gagliardini2013capabilities}.
-With a transformation to terrain-following coordinates, we can set the terrain-following vertical velocity $\omega$ to be $-\dot a_b/h$ at the ice base to impose this boundary condition.
+With a transformation to terrain-following coordinates, we can just set the terrain-following vertical velocity $\omega$ to be 0 at the ice base to impose this boundary condition.
 
@@
 Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
-In other words, it is not immediately clear whether the vertical velocity is a result of basal mass balance or of geometry without additional information.
+In other words it is not immediately clear whether the vertical velocity is more a result of basal mass balance or of geometry without more information.
 By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.
 
-\subsubsection{Spectral discretization}
+\textcolor{red}{Speed of hybrid model.}
+One of the main advantages of this hybrid model is the ease with which a more accurate solution can be bootstrapped from a less accurate solution, since the two fields are defined on exactly the same geometry.
+For example, if we only have a very crude initial guess for the solution, we can obtain a degree-1 solution relatively cheaply.
+The resulting velocity field is then used as a much more informed initial guess for a degree-5 solution.
+\textcolor{red}{Benchmark.}
+Since the vertical basis functions are orthogonal, the projection of the degree-1 solution into the degree-5 function space can be calculated by simply copying array elements.
+By contrast, if we tried to use the same strategy to bootstrap a solution with a small number of vertical layers to a solution with a larger number of vertical layers, we would also have to interpolate values to the fine layers.
 
-Terrain-following coordinates open up several choices for how to describe the vertical variation of the velocity field.
-In Elmer/Ice, for example, the user can extend the finite element discretization into a number of vertical layers.
-The number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.
+Finally, we note that this approach was first described in \citet{bassis2010hamilton}.
+This work focused solely on flowband models and did not consider the problem of how to calculate the power dissipated at the calving front.
+Moreover, this work used only the two vertical basis functions $\psi_1(\zeta) = 1$, $\psi_2(\zeta) = 1 - (1 - \zeta)^{n + 1}$, the idea being that the vertical profile can be written as a direct superposition of the plug flow and shallow ice approximation modes.
+By contrast, we leave the number of vertical basis functions up to the user.
+This flexibility enables the multigrid-type approach described above.
 
-The horizontal velocity for many realistic flows is very smooth as a function of depth and this suggests a different approach.
-For example, under the plug flow approximation, the horizontal velocity is constant with depth.
-Under the shallow ice approximation, the horizontal velocity varies with depth as $1 - (1 - \zeta)^{n + 1}$ where $n = 3$ is the Glen flow law exponent.
-This extra information about our solution suggests a modal rather than a nodal discretization strategy.
-
-Rather than divide the spatial domain into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
-This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element.
-Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
-\begin{equation}
-    \Phi_{kl}(x, y, \zeta) = \phi_k(x, y)\psi_l(\zeta).
-\end{equation}
-For example, we can use piecewise linear or quadratic elements on triangles and use quintic or higher degree polynomials in the vertical.
-Rather than use the usual Lagrange interpolating polynomial basis in the vertical dimension, we can instead use the Legendre polynomial basis.
-The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.
-
-The combination of using extruded meshes and tensor product elements in the vertical direction is referred to as \emph{semi-discretization}.
-This approach can be thought of merely as a way to discretize a PDE that has special structure.
-Alternatively, we can view vertical semi-discretization as defining a family of models indexed by the number of vertical basis functions.
-The order-$d$ model defines a coupled system of PDEs for $d$ vector fields.
-Each vector field represents one mode of vertical variability, similar to the distinction between barotropic and baroclinic modes in atmospheric physics and oceanography.
-The system is then discretized in the horizontal and solved numerically.
-In any case, the code is the same regardless of how one views the underlying mathematics.
-
-The user then has to decide how many vertical modes are enough.
-Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
-The degree 1 model is the most minimalistic model that still exhibits vertical shear.
-Going to higher degree gives a more accurate approximation at the expense of greater computational effort.
-
-\subsection{Substituting model components} \label{sec:physics-substitution}
+\subsection{Substituting model components}
 
@@
     \tau_b = -C|u|^{\frac{1}{m} - 1}u
-    \label{eq:weertman-sliding}
 \end{equation}
@@
 For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
-In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
+In programming terms this amounts to changing the number of arguments to the function that calculates basal shear stress.
 For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
-In C++ one could avoid modifying the diagnostic solve routine by using variadic templates, but this is an advanced language feature and relying on it would present a steep learning curve for novices.
+In C++ one could obviate the need to modify the diagnostic solve routine by using variadic templates.
+But variadic templates are a more advanced language feature and using them would arguably present a steep learning curve for novices.
 
-Any model physics component in icepack can be substituted for a parameterization of the user's choosing.
-Each flow model object is initialized with a set of functions to calculate the various terms of the action functional.
+In icepack, any given model physics component can be substituted for a parameterization of the user's choosing.
+Every flow model in icepack is represented by a Python object, and each model object is initialized with a set of functions to calculate the various terms of the action functional.
 For example, the ice stream model is initialized with functions to calculate the contributions due to viscosity, basal friction, side wall drag, and gravity.
-The default parameterizations are set to the most common choices in the literature, but if the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
+Sensible parameterizations are selected by default.
+If the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
 When the diagnostic solve procedure is invoked, all of its arguments are then passed on by keyword to the method that calculates the action functional.
-Any fields that are unnecessary for calculating a given component are ignored.
-For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function makes no use of these arguments.
-The user can completely change the functional form and add new arguments, with the restriction that arguments are passed by keyword and not positionally.
+Any fields that are unnecessary for calculating a given component are simply ignored.
+For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function simply makes no use of these arguments.
+The functional form can be changed entirely and new arguments can be added, with the only restriction that they are passed as keyword and not as positional arguments.
 
-In adopting this approach, we are restricted to using keyword arguments instead of positional arguments.
-Using only keyword arguments enhances readability and comprehensibility for this particular case of invoking a physics solver.
+We are restricted to using keyword arguments instead of positional arguments by adopting this approach.
+For the particular case of passing several fields to a physics solver routine, using only keyword arguments enhances readability and comprehensibility.
 The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used in the literature -- $C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.
 The order of the arguments is arbitrary and immaterial.
-The preference for argument passing by name is specific to this use case and not universal.
-For example, in defining a bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance.
-In this case, positional arguments correspond better with the semantics of the problem.
+For defining, say, the bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance and our design would be inappropriate.
 
@@
 These users then have a steep learning curve to overcome when they need to modify functionality that lives in the core library, as they now need to learn an entirely new programming language.
-Some low-level kernel code is, of course, always necessary for runtime efficiency.
-With Firedrake, this responsibility is deferred to a code generator.
-The average user can remain largely unaware of how the code generator works even when solving very advanced problems.
+Some low-level kernel code is always necessary for runtime efficiency of course.
+With Firedrake, this responsibility is deferred to a code generator of which the average user remains blissfully unaware.
 
-\subsection{Data assimilation}
+\subsection{Convergence criteria} \label{sec:convergence-criteria}
+
+Calculating the ice velocity from the geometry and other input fields involves minimizing a convex functional.
+The minimizer can only be approximated, so we employ iterative procedures based on Newton's method with a line search.
+We must then choose some metric to decide when the current guess for the velocity is good enough to stop the iteration.
+
+A convergence criterion that works equally well independent of the mesh, finite element discretization, and the quality of the initial guess can be defined based on the idea of the \emph{Newton decrement} \citep{nocedal2006numerical}.
+The search direction $v_k$ at step $k$ for Newton's method is defined as
+\begin{equation}
+    v_k = -\ud^2J(u_k)^{-1}\ud J(u_k).
+\end{equation}
+Since the second derivative operator $\ud^2J(u_k)$ is positive-definite, this is a descent direction for the action $J$:
+\begin{equation}
+    \ud J(u_k)\cdot v_k < 0.
+\end{equation}
+The absolute value of the quantity in the last equation is defined as the Newton decrement.
+For $u_k$ sufficiently close to the true solution $u$, the Newton decrement roughly tells us how much we can expect the action to decrease:
+\begin{equation}
+    J(u_k) - J(u) \approx \frac{1}{2}|\ud J(u_k)\cdot v_k|.
+\end{equation}
+We can then use the Newton decrement to decide when to stop the iteration.
+
+As shown in equation \eqref{action-functional}, the action has units of power and is the sum of the dissipation due to viscosity, friction, gravitational driving, and restrain by ocean pressure at the terminus.
+The viscous and frictional terms are convex and positive functions of the velocity, while the gravitational and terminus stress terms are linear in the velocity and can be of either sign.
+If we define the \emph{scale functional}
+\begin{equation}
+    K(u) = \text{viscous dissipation} + \text{frictional dissipation}
+\end{equation}
+as only the positive parts of the action, then the convergence criterion
+\begin{equation}
+    |\ud J(u_k)\cdot v_k| < \epsilon K(u_k)
+\end{equation}
+is independent of the discretization.
+The intuition behind this criterion is that iteration is halted when the expected decrease in the action functional is much smaller than the positive part of the action itself.
+
+We have found empirically that, with this criterion and the Newton solver implementation in icepack, the iteration usually converges to machine precision in around 8 steps.
+For exceptionally bad initial guesses the iteration count can reach as high as 20 but rarely more.
+We also observe the expected doubling of the number of accurate digits in the value of the action once the velocity guesses are within the convergence basin of the true solution.
+Other convergence criteria, such as using relative change in the velocity guesses, can terminate prematurely when the initial guess is very far outside the quadratic convergence basin.
+
+The numerical solvers in icepack have been designed so that users who are not familiar with numerical optimization need not be confronted with a possibly bewildering array of algorithmic parameters.
+Consequently, sensible defaults have been chosen for the Armijo and Wolfe criteria \citep{nocedal2006numerical}, and the tolerance for the line search is chosen based on that of the outer-level Newton iteration.
+The Newton search direction is calculated using a direct factorization solver rather than, say, the conjugate gradient algorithm, as the use of another iterative method would introduce yet another algorithmic parameter.
+The existing algorithmic parameters can nonetheless be changed from their defaults by passing keyword arguments for advanced users who are interested in performance optimization.
+
+\section{Prognostic model} \label{sec:prognostic-model}
+
+The PDE describing how the ice thickness changes in time is a simple conservative advection equation:
+\begin{equation}
+    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
+\end{equation}
+where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
+The thickness is represented using continuous Galerkin basis functions.
+The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
+The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.
+
+The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
+Other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \textcolor{red}{cite ISSM, VarGlaS}.
+The SUPG method is explicit in time and conditionally stable \textcolor{red}{cite Donea and Huerta?}.
+This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
+The implicit Euler scheme requires solving a more complicated linear system, but the extra computational cost comes with the advantage of unconditional stability.
+Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
+The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
+The extra cost of using an implicit time discretization for the mass transport equation is dwarfed by the cost of the diagnostic solve in any case.
+Advanced users who are interested in maximizing performance can subclass the mass transport solver to implement a faster explicit scheme.
+
+The implicit Euler scheme introduces a large degree of artificial diffusion \textcolor{red}{cite Donea and Huerta}.
+Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
+The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.
+Other problems within glaciology have more of a hyperbolic character and there the argument for using shock-capturing methods is much stronger.
+For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \textcolor{red}{cite Albrecht and Levermann}.
+The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge Kutta method would be more appropriate.
+
+\section{Data assimilation}
 
@@
 More importantly, the inverse problem is flexible enough to account for the fact that we might have substituted in our own parameterization for the rheology or friction coefficient.
-For example, nearly all studies in the literature use an alternate parameterization of the friction coefficient in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
+For example, nearly all studies in the literature use an alternate parameterization of these coefficients in terms of some auxiliary field in order to guarantee positivity \textcolor{red}{citations}.
 This information is passed to the inverse problem by specifying the names of the keyword arguments for the observed state and the field we wish to observe.
 
-The \texttt{InverseSolver} class is responsible for actually solving the inverse problem.
-This class will be described further in section \S\ref{sec:numerics-inverse-solvers}.
-The inverse solver is completely agnostic to the particular parameterization of the physics in terms of the inferred field as long as it guarantees positivity.
-This feature interacts seamlessly with the ability to modify individual model physics components as described in \S\ref{sec:physics-substitution}.
-
-\subsection{Damage transport}
-
-\subsection{Heat transport}
-
-\subsection{\textcolor{green}{Plumes}}
-
-\section{Numerics}
-
-In this section we describe how various parts of the model physics were implemented in icepack, including both software design and the numerical solution procedures.
-From a design perspective, each flow model is represented by a Python class that has three responsibilities.
-First, the model class has to have a method that takes in the various fields (ice velocity, thickness, surface elevation, etc.) and returns a symbolic representation of the action functional for that particular model.
-Second, the model class must have a method that takes in the input fields and returns an approximate minimizer of the action functional.
-The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the first method calculates.
-The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
-Finally the model must have a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
-The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.
-
-The Unified Form Language for specifying weak forms of PDEs contains all of the primitives necessary to express individual terms of the action functional.
-These primitives consist of the basic vector calculus operators like the gradient of a field, tensor calculus operations like taking the dot product of two vectors or tensors, scalar functions like the square root or exponential, and symbolic integration over the mesh or its boundary.
-For example, the strain rate for a given velocity field $u$ can be written as \texttt{sym(grad(u))}, where the function \texttt{grad} represents the symbolic gradient of a field and \texttt{sym} represents the symmetrization of a rank-2 tensor.
-
-\subsection{Advective transport} \label{sec:prognostic-model}
-
-The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
-To solve the mass transport equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
-The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
-This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
-The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.
-
-Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
-The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
-The extra cost of using an implicit time discretization for the mass transport equation is dwarfed by the cost of the diagnostic solve in any case.
-Advanced users who are interested in maximizing performance can subclass the mass transport solver to implement a faster explicit scheme.
-
-The implicit Euler scheme tends to diffuse out sharp discontinuities that may be present in the true solution \citep{donea2003finite}.
-Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
-The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.
-Other problems within glaciology have more of a hyperbolic character and there the argument for using shock-capturing methods is much stronger.
-For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \citep{albrecht2014fracture}.
-The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge Kutta method would be more appropriate.
-
-\subsection{Convex optimization}
-
-The action functional for each diagnostic model is convex, i.e. the second derivative is strictly positive-definite.
-From a theoretical persepctive, convexity guarantees that the problem has a unique solution.
-This property is also especially advantageous for implementing numerical solvers.
-We use a damped Newton method to solve the diagnostic equations.
-Starting from a guess $u_k$ for the velocity, the \emph{search direction} $v_k$ is the unique solution of the linear system
-\begin{equation}
-    \ud^2J(u_k)\cdot v_k = -\ud J(u_k).
-    \label{eq:newton-search-direction}
-\end{equation}
-The next approximation for the velocity minimizes $J$ along the search direction $v_k$ starting from $u_k$, i.e.
-\begin{align}
-    u_{k + 1} & = u_k + \alpha_k\cdot v_k, \nonumber\\
-    \alpha_k & = \text{argmin}_\alpha \hspace{2pt}J(u_k + \alpha v_k).
-\end{align}
-For an initial guess sufficiently close to the exact solution, the undamped Newton method ($\alpha_k = 1$ at every step) converges quadratically.
-The line search step ensures that the method can converge even from a poor initial guess, provided that the line search method satisfies the Armijo-Wolfe criteria \citep{nocedal2006numerical}.
-
-For a convex problem, $\ud^2J$ is a symmetric and positive-definite matrix.
-This has two advantages.
-First, the search direction is always a descent direction for $J$.
-For non-convex problems this is not always the case and some adaptation strategy must be employed in the vicinity of saddle points.
-Second, symmetry and positivity enable the use of specialized linear solvers, such as the Cholesky decomposition or the conjugate gradient algorithm, that are superior to their more general counterparts in many respects.
-
-Other software packages that treat diagnostic models as nonlinear systems of equations tend to rely on ad-hoc procedures for initializing the numerical solution process.
-For example, without a damping procedure in Newton's method, the iteration can prove unstable if initialized far away from the true solution.
-Some packages combat this problem by using a few iterations of the more robust but slower Picard method first \citep{gagliardini2013capabilities}.
-While this approach can be effective it requires tuning the number of Picard iterations and there is no guarantee that an adequate amount for one problem will work well on another problem.
-This issue rarely appears on realistic input data.
-When solving inverse problems, however, the intermediate guesses for the inferred field can be wildly unrealistic before converging.
-A forward model solver that is not sufficiently robust can crash in these extreme scenarios.
-By contrast, a damped Newton procedure using a line search that satisfies the Armijo-Wolfe criteria is guaranteed to converge on non-degenerate, if unrealistic, input data.
-
-\subsection{Convergence metrics} \label{sec:convergence-criteria}
-
-Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation \textcolor{red}{cite -- I think Mathieu and Dan did this}, but few if any have considered the problem of when to stop iterating.
-The most common stopping criteria are when (1) the 2-norm of the residual is sufficiently small or (2) the relative change in the iterates is sufficiently small.
-Each of these approaches has problems.
-The residual norm depends on the discretization and does not weight all degrees of freedom proportionally, e.g. vertex and edge degrees of freedom in higher-order finite element methods.
-The relative change criterion, on the other hand, can suggest convergence when in fact the method has stagnated.
-
-A convergence criterion that works equally well independent of the mesh, finite element discretization, and the quality of the initial guess can be defined based on the idea of the \emph{Newton decrement} \citep{nocedal2006numerical}.
-Since the second derivative operator $\ud^2J(u_k)$ is positive-definite, the Newton search direction $v_k$ computed from equation \eqref{eq:newton-search-direction} is a descent direction for $J$:
-\begin{equation}
-    \ud J(u_k)\cdot v_k < 0.
-\end{equation}
-The absolute value of the quantity in the last equation is defined as the Newton decrement.
-For $u_k$ sufficiently close to the true solution $u$, the Newton decrement roughly tells us how much we can expect the action to decrease:
-\begin{equation}
-    J(u_k) - J(u) \approx \frac{1}{2}|\ud J(u_k)\cdot v_k|.
-\end{equation}
-We can then use the Newton decrement to decide when to stop the iteration.
-
-As shown in equation \eqref{action-functional}, the action has units of power and is the sum of the dissipation due to viscosity, friction, gravitational driving, and ocean back-pressure at the terminus.
-The viscous and frictional terms are convex, positive functions of the velocity.
-The gravitational and terminus stress terms are linear in the velocity and can be of either sign.
-If we define the \emph{scale functional}
-\begin{equation}
-    K(u) = \text{viscous dissipation} + \text{frictional dissipation}
-\end{equation}
-as only the positive parts of the action, then the convergence criterion
-\begin{equation}
-    |\ud J(u_k)\cdot v_k| < \epsilon K(u_k)
-\end{equation}
-is independent of the discretization.
-The intuition behind this criterion is that the iteration is halted when the expected decrease in the action functional is much smaller than the positive part of the action itself.
-
-We have found empirically that, with this criterion and the Newton solver implementation in icepack, the iteration usually converges to machine precision in around 8 steps.
-The iteration count can reach as high as 20 for exceptionally bad initial guesses for the velocity or with unphysical fluidity or friction values.
-We also observe the expected doubling of the number of accurate digits in the value of the action once the velocity guesses are within the convergence basin of the true solution.
-Other convergence criteria, such as using relative change in the velocity guesses, can terminate prematurely when the initial guess is very far outside the quadratic convergence basin.
-
-The numerical solvers in icepack have been designed so that users who are not familiar with numerical optimization need not be confronted with a possibly bewildering array of algorithmic parameters.
-Consequently, sensible defaults have been chosen for the Armijo and Wolfe criteria \citep{nocedal2006numerical}, and the tolerance for the line search is chosen based on that of the outer-level Newton iteration.
-The Newton search direction is calculated using a direct factorization solver rather than, say, the conjugate gradient algorithm, as the use of another iterative method would introduce yet another algorithmic parameter.
-Advanced users who are interested in performance optimization can change these algorithmic parameters by passing extra arguments to the solve procedure.
-
-\subsection{Inverse solvers} \label{sec:numerics-inverse-solvers}
-
-The \texttt{InverseProblem} class describes what problem is being solved, while the \texttt{InverseSolver} class is responsible for carrying out the numerical optimization.
-There are three inverse solvers in icepack: a simple gradient descent solver, a quasi-Newton solver based on the BFGS approximation to the Hessian, and a Gauss-Newton solver.
-All of these classes are based around the general idea of first computing a search direction and then performing a line search.
+The class \texttt{InverseSolver} is an abstract interface for classes that actually solve the inverse problem.
+There are three implementations of this interface in icepack: \texttt{GradientDescentSolver}, \texttt{BFGSSolver}, and \texttt{GaussNewtonSolver}.
+All of these classes are based around the general idea of first computing a search direction and then performing a line search in every iteration.
 They differ in how the search direction is computed.
-
 The gradient descent solver uses the search direction
@@
 This method can be very expensive and brittle in the initial iterations and often takes many steps to converge.
-
 The BFGS method uses the past $m$ iterations of the algorithm to compute a low-rank approximation to the inverse of the objective functional's Hessian matrix; see \citet{nocedal2006numerical} for a more in-depth discussion.
-The BFGS method converges faster than gradient descent.
-However, it suffers from many of the same brittleness issues in the initial iterations before it has built up enough history to approximate the Hessian inverse.
-
+The BFGS method converges faster than gradient descent but suffers from many of the same brittleness issues in the initial iterations before it has accumulated enough history.
 Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian.
@@
 
-\subsection{Hybrid model}
-
-The hybrid flow uses several features that are only available in firedrake to better exploit the special structure of the problem.
-Implementing this model also required some mathematical sleight-of-hand related to the terminus boundary condition that has not appeared in the literature before.
-Additionally, the hierarchical structure of spectral basis functions presents an opportunity for developing fast algorithms.
-In all other respects the implementation of the hybrid flow model using convex optimization follows the techniques described above.
-
-\subsubsection{Discretization}
-
-In order to use terrain-following coordinates, the hybrid model assumes that the geometry of the domain is an extruded mesh, where a 2D footprint mesh is lifted into 3D.
-Firedrake includes support for creating extruded meshes by calling the function \texttt{ExtrudedMesh} on the 2D footprint \citep{bercea2016structure}.
-The cells of an extruded mesh are triangular prisms instead of the more common tetrahedra used for general 3D meshes.
-Not every 3D domain can be described by extruding a 2D domain, but the geometry of most glacier flow problems can.
-
-The geometric correction factor in equation \eqref{eq:geometric-correction} for gradients in terrain-following coordinates can easily be represented in UFL.
-By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is only slightly more complex than in Cartesian coordinates.
-
-For problems defined on extruded geometries, firedrake includes support for tensor product elements, which includes using different bases in the horizontal and vertical directions \citep{mcrae2016automated}.
-Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
-In our case, we used the usual continuous Galerkin basis in the horizontal and Gauss-Legendre elements in the vertical.
-To select the Legendre polynomial basis, the user passes the keyword argument \texttt{vfamily=`Gauss-Legendre'} or \texttt{`GL'} for short to the constructor for the function space.
-
-Extruded meshes and tensor product elements are available in firedrake but not in FEniCS.
-Other general-purpose finite element modeling packages that support tensor product elements include deal.II and nektar++ \citep{bangerth2007deal, cantwell2015nektar++}.
-Like most other packages in this domain, deal.II and nektar++ are written in C++, whereas our goal for icepack was to have both the core and the user interface in Python.
-
-\subsubsection{Ocean boundary condition}
-
-Our approach for implementing a hybrid flow model works completely seamlessly but for one important detail.
-The backpressure from ocean water at the calving front of a marine-terminating glacier is not a smooth function of depth.
-The pressure is 0 above the water line and linearly increasing below it:
-\begin{equation}
-    \text{backpressure power} = \int_\Gamma\int_0^1 \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+\ud\zeta\ud\gamma,
-    \label{backpressure}
-\end{equation}
-where $\zeta_{\text{sl}}$ denotes the relative depth to the water line and the subscript $+$ denotes the positive part of a real number.
-Were we to use the standard asssembly procedure in Firedrake to evaluate this integral, we would get an inaccurate result due to an insufficient number of integration points.
-The resulting velocity solutions are then wildly inaccurate due to the mis-specification of the Neumann boundary condition.
-A blunt solution to this problem would be to pass an extra argument to the Firedrake form compiler that specifies a much greater integration accuracy in the vertical for this term.
-This fix reduces the errors in the velocities, but it does not eliminate them completely and it incurs a large computational cost.
-
-\begin{figure}[h]
-    \includegraphics[width=0.95\linewidth]{demos/legendre/pressure.png}
-    \caption{The normalized ocean pressure ($p_w / \rho_Wg$) and Legendre polynomial approximations of several degrees (left), and the residuals of the approximation (right).
-    For this particular example, the waterline is at $\zeta = 1/3$.
-    The moments of each of the residuals up to the approximation degree are all zero.}
-    \label{fig:legendre}
-\end{figure}
-
-We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
-The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
-The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts at high degree.
-While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property of Legendre polynomials.
-Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the number of vertical modes.
-An example of the pressure approximations using linear, quadratic, and cubic Legendre polynomials are shown in figure \ref{fig:legendre}.
-
-The exact symbolic integration approach is both faster and more accurate than using a large number of quadrature points.
-The same technique could be used to exactly calculate the ocean backpressure for any model, say the full Stokes equations, using terrain-following coordinates together with a Legendre polynomial expansion in the vertical.
-
-\subsubsection{Multigrid-type algorithm}
-
-One of the main advantages of this hybrid model is the ease with which a more accurate solution can be bootstrapped from a less accurate solution, since the two fields are defined on exactly the same geometry.
-For example, if we only have a very crude initial guess for the solution, we can obtain a degree-1 solution relatively cheaply.
-The resulting velocity field is then used as a much more informed initial guess for a degree-5 solution.
-\textcolor{red}{Benchmark.}
-Since the vertical basis functions are orthogonal, the projection of the degree-1 solution into the degree-5 function space can be calculated by simply copying array elements.
-By contrast, if we tried to use the same strategy to bootstrap a solution with a small number of vertical layers to a solution with a larger number of vertical layers, we would also have to interpolate values to the fine layers.
-
-Other hybrid models in the literature use only a fixed number of vertical basis functions.
-For example, \citet{bassis2010hamilton} use only the two vertical basis functions $\psi_1(\zeta) = 1$, $\psi_2(\zeta) = 1 - (1 - \zeta)^{n + 1}$, the idea being that the vertical profile can be written as a direct superposition of the plug flow and shallow ice approximation modes.
-By contrast, we leave the number of vertical basis functions up to the user.
-This flexibility enables the multigrid-type approach described above.
-
-The bootstrapping approach described here can be used both for the forward model and for inverse problems.
+The bootstrapping approach used to calculate higher-order solutions to the hybrid model from lower-order solutions rather than from a cold initial guess can also be used for inverse problems.
 First, we estimate the unknown parameter using only degree-1 vertical basis functions for the velocity field, which captures most of the spatial variability of the unknown parameter.
@@
 
-\subsection{MISMIP+}
-
-As a first test case for icepack, we ran the first experiment from the Marine Ice Sheet Model Intercomparison Project version 3 (MISMIP+).
-The parameters and geometry for this experiment can be found in \citet{asay2016experimental}.
-The MISMIP+ experiment has three phases.
-First, the model must find a steady state marine ice sheet with a fixed accumulation rate and no submarine melting.
-Next, submarine melting with a given depth-dependent parameterization is applied for 100 years.
-The increased melt thins the ice ice shelf and initiates a retreat of the glacier grounding line.
-Finally, submarine melting is turned off for at least 100 years, optionally longer.
-The grounding line then readvances, but not as far as its original position.
-
-The original intercomparison project specified that participants could use the Weertman sliding law (equation \eqref{eq:weertman-sliding}) as well as two other sliding laws that transition to a more plastic rheology at high sliding speeds.
-The first alternative sliding law consists of Weertman sliding until the stress reaches a critical value, at which point the constitutive relation transitions to perfect plasticity.
-The second alternative is the Schoof sliding law \citep{schoof2005effect}:
-\begin{equation}
-    \tau_b = -\frac{C|u|^{1/m}\cdot \tau_c}{(C^m|u| + \tau_c^m)^{1/m}}\frac{u}{|u|},
-    \label{eq:schoof-sliding}
-\end{equation}
-where the critical stress $\tau_c$ is a certain specified fraction of the water pressure in the subglacial hydrological system.
-In icepack sliding laws are not expressed directly but rather as the the derivative of an acction functional.
-To implement the Schoof sliding law we need to know the antiderivative of equation \eqref{eq:schoof-sliding}.
-Using a computer algebra system, we found that the antiderivative of this expression in terms of hypergeometric functions.
-The Unified Form Language has several transcendental functions (sine, cosine, exponential, etc.), but it does not currently support hypergeometric functions.
-Instead, we implemented a sliding relation that exhibits the important features of the Schoof law, i.e.
-it behaves like an $m = 3$ power law at low sliding speeds and $m = \infty$ at high sliding speeds, but which is more tractable algebraically.
-Knowing the critical stress $\tau_c$ and the friction coefficient $C$, which has units of stress $\times$ speed${}^{-1/m}$, we can define a \emph{critical speed} $u_c$ as
-\begin{equation}
-    u_c = C^{-m}\tau_c^m.
-\end{equation}
-The power dissipation density for the Schoof-type sliding law that we use is
-\begin{equation}
-    P = \tau_c\left\{\left(u_c^{\frac{1}{m} + 1} + |u|^{\frac{1}{m} + 1}\right)^{\frac{m}{m + 1}} - u_c\right\}
-    \label{eq:power-dissipation-modified-schoof-sliding}
-\end{equation}
-Figure \ref{fig:sliding-laws} shows a comparison of the original Schoof sliding law and the sliding law that arises as the derivative of the functional in equation \eqref{eq:power-dissipation-modified-schoof-sliding}.
-The two have the same asymptotic behavior when the speed is much smaller or much larger than the critical speed; they differ in a relatively small range around the critical speed.
-The relative difference in basal shear stress between the two sliding laws is less than 10\% throughout the entire range.
-
-\begin{figure}[h]
-    \includegraphics[width=0.95\linewidth]{demos/sliding/sliding-law.png}
-    \caption{The Weertman, Schoof, and modified Schoof sliding law of equation \eqref{eq:power-dissipation-modified-schoof-sliding}.
-    The critical speed and shear stress are $u_c = 250$ m/year and $\tau_c = 100$ kPa.}
-    \label{fig:sliding-laws}
-\end{figure}
-
-A sample of the source code implementing both of these approaches is shown in \textcolor{red}{Figure X}.
-To change the sliding law, users only need to pass one function or the other to the model object at initialization.
-Users do not need to implement a subclass that overrides some parent method.
-This approach would be idiomatic in C++, but it requires some facility with object-oriented programming that a non-expert might lack.
-
-Figure \ref{fig:mismip} shows the outcome of the MISMIP+ experiment performed using icepack.
-
-\begin{figure}[h]
-    \includegraphics[width=0.95\linewidth]{demos/mismip/steady-state.png}
-    \caption{The steady state thickness and velocity of the MISMIP+ experimental setup with the Weertman sliding law.}
-    \label{fig:mismip}
-\end{figure}
-
-Using variational principles to express all constitutive laws is less flexible than specifying the sliding law directly and this is a distinct disadvantage.
-We were able to implement a sliding law that exhibits many of the characteristics of the Schoof sliding law.
-\citet{asay2016experimental} also suggest using a sliding law that transitions sharply to exact plasticity above the critical speed.
-Expressing this sliding law in UFL requires a conditional in the velocity and is thus no longer differentiable.
-This lack of differentiability makes the forward solver crash.
-The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.
-
-\subsection{Larsen C Ice Shelf}
-
-To demonstrate the forward and inverse modeling capabilities of icepack, we will estimate the rheology of the Larsen C Ice Shelf in the Antarctic Peninsula from observational data and project the model forward by 10 years.
-Several recent studies have focused on Larsen C because it may be unstable in the warmer climate of the coming decades.
-From January to March of 2002, the neighboring Larsen B Ice Shelf disintegrated due to surface melt pond-induced fracture \textcolor{red}{check that this is the currently-accepted explanation and cite}, and this mechanism could also lead to the breakup of Larsen C.
-One of the key factors affecting the stability of ice shelves is the presence of marine ice -- seawater that freezes onto the base of an ice shelf -- in the suture zones where two flow units meet \citep{kulessa2014marine}.
-Marine ice is warmer than meteoric ice and usually includes brine pockets, which are discernible in radio echo sounding measurements as the absence of reflection from the ice shelf base \citep{holland2009marine}.
-This warmer and impurity-laden ice is more ductile and thus less prone to fracture than cold and brittle meteoric ice.
-Ocean models predict that marine ice forms under Larsen C as well \citep{holland2009marine}.
-By estimating the material hardness or rheology of an ice shelf, we can constrain where marine ice may be forming.
-
-\subsubsection{Data}
-
-\subsubsection{Parameterization}
-
-The rheology parameter of an ice shelf is strictly positive.
-The optimization algorithm, however, can explore unphysical regions of parameter space without some a priori constraint on what values are reasonable.
-Rather than try to solve an inequality-constrained problem, most studies in glaciology instead re-parameterize the problem in terms of some auxiliary field in such a way that the rheology is manifestly positive.
-In this case we will use the parameterization
-\begin{equation}
-    B = B_0e^{\theta}
-\end{equation}
-and estimate $\theta$.
-The inverse solver calculates the derivative of the objective functional symbolically and is thus agnostic to the particular parameterization.
-The only information that the user needs to pass is the name of the arguments to the forward model representing the parameter and the observed field so that these can be passed by keyword.
-
-\subsubsection{Results}
-
-\subsection{Jakobshavn Isbrae}
+\section{Testing}
 
@@
 
-One of the main goals for icepack is to create a tool that is accessible to researchers who might not be experts in scientific computing.
+One of the main goals for icepack was to create a tool that is accessible to researchers who might not be experts in scientific computing.
 Previous work on numerical modeling of glacier flow has focused largely on technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory, does it scale to large numbers of processors, can models accurately predict grounding line retreat, etc.
@@
 Progressive evaluation is the main advantage of having a user interface in Python or another interpreted language, as opposed to using compiled programs that must run all at once.
-Icepack includes a suite of example programs that demonstrate all of the main functions of the package.
-These demos are formatted as Jupyter notebooks, a document format that includes code, figures, and explanatory text with typeset mathematics that runs interactively in a web browser \citep{kluyver2016jupyter}.
+Icepack includes a suite of demos that demonstrate all of the main functions of the package.
+These demos are formatted as Jupyter notebooks, a document format that includes code, figures, and explanatory text with typeset mathematics that can be run interactively in a web browser \citep{kluyver2016jupyter}.
 A notebook is a sequence of cells, each of which consists of either code or text.
@@
 This choice is of interest mostly to advanced users so we keep the linear solver method as a default argument.
-In so doing, we avoid confronting novice users with options that they might not understand.
+In so doing, we avoid confronting novice users with an option that they might not understand.
 

### 2020-09-03 demos/mismip/mismip+.py
Coarsen some early stages of MISMIP+ spin-up

@@ -13,11 +13,11 @@ import icepack, icepack.plot
 parser = argparse.ArgumentParser()
 parser.add_argument('--input')
 parser.add_argument('--output')
-parser.add_argument('--input-level', type=int, default=0, dest='input_level')
-parser.add_argument('--output-level', type=int, default=0, dest='output_level')
+parser.add_argument('--input-level', type=int, default=0)
+parser.add_argument('--output-level', type=int)
 parser.add_argument('--melt', action='store_true')
-parser.add_argument('--time', type=float, default=4e3)
-parser.add_argument('--timestep', type=float, default=1.0)
+parser.add_argument('--time', type=float)
+parser.add_argument('--timestep', type=float)
 
 args = parser.parse_args()
 if args.output_level < args.input_level:

### 2020-11-03 icepack.tex

--- 2019-11-12 icepack.tex
@@
 
-There are \textcolor{green}{three} diagnostic models implemented in icepack.
-The \emph{shallow ice approximation} (SIA) and shallow stream approximation (SSA) are 2D models describing the depth-averaged velocity of a glacier.
-\textcolor{green}{The SIA model describes the deep interior of ice sheets well, where flow velocities are on the order of 0.1 - 10 m/year.
-This model assumes that nearly all flow is by vertical shear, which happens when the ice is frozen to the bed or sliding at a small fraction of the surface speed.}
+There are three diagnostic models implemented in icepack.
+The \emph{shallow ice approximation} (SIA) and \emph{shallow stream approximation} (SSA) are 2D models describing the depth-averaged velocity of ice flow, while the \emph{first-order} or \emph{Blatter-Pattyn} approximation is a 3D model describing the horizontal velocity (termed the \emph{hybrid model} in icepack) \citep{blatter1995velocity, pattyn2003new}.
+The SIA model \citep{hutter1981effect} describes the interior of ice sheets well, where flow can be simply described as bed-parallel shear and surface and basal slopes are small.
+This approximation does not apply well near ice divides, where vertical velocities dominate, and near ice margins, where basal sliding can be a large fraction of the surface speed.
 The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.
@@
 Plug flow is a good approximation in fast-flowing ice streams and outlet glaciers near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
-The \emph{first-order} or \emph{Blatter-Pattyn} (BP) approximation, by contrast, is a 3D model describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
-The only approximation in the BP model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
+The only approximation in the hybrid model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
 This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
 Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.
+
+\textcolor{red}{Should Glen's flow law be mentioned or written somewhere?}
 
@@
 
-\subsubsection{\textcolor{green}{Shallow ice approximation}}
+\subsubsection{Shallow ice approximation}
+
+The shallow ice model can be derived from the Blatter-Pattyn approximation by assuming the ice flow is dominated by bed-parallel shear and that surface and basal slopes are small, i.e.
+$\partial u/\partial x \ll \partial u/\partial z$.
+The result is a 2D system of equations for depth-averaged horizontal velocity.
+The terms in the action functional are:
+
+\begin{align}
+    \text{mass} & =\frac{1}{2}\int_\Omega u\cdot u\ud x\\\
+    \text{gravity} & = \int_\Omega\frac{2A(\varrho_I g)^n}{n+2} (\nabla h^2\cdot u) h^{n+1} \nabla s^{n-1}\ud x\\
+    \text{penalty} & = \frac{1}{2}\int_\Omega l^2\nabla u\cdot \nabla u\ud x 
+\end{align}
+
+where
+
+\begin{equation}
+l = 2 \text{ max}(\text{cell diameter}, 5h)
+\end{equation}
+
+and the action fuctional is
+
+\begin{equation}
+    J = \text{mass} + \text{gravity} + \text{penalty}.
+\end{equation}
+
+We have included an additional penalty term, which helps alleviate numerical artifacts, especially near the ice margins and termini where the shallow ice approximation is less applicable.
+
+The shallow ice approximation applies well in ice-sheet interiors.
+For this reason, and because the equations are particularly simple to solve, this approximation has been a common choice for ice-sheet modeling \textcolor{red}{cite}.
+Caution must be taken, however, because areas of the ice sheet that experience high vertical velocities (e.g., ice divides) or plug flow (e.g., ice streams) are not well represented by this approximation.
 
@@
 The shallow ice and shallow stream models in icepack are completely standard.
+\textcolor{red}{JB: Aren't the models in icepack based on variational principles?
+This sentence and the next suggest that only the hybrid model is based on variational principles.}
 The hybrid flow model that we describe here is based on variational principles, terrain-following coordinates, and vertical spectral methods.
@@
 \subsection{Synthetic ice sheet}
+As a second test case for icepack, we ran an experiment inspired by \cite{kessler2008fjord}.
+In the original work, the authors couple an ice-flow model with simple models for bed-erosion, calving, and \textcolor{red}{glacial isostatic adjustment (GIA)} to investigate the formation of deep fjords that are characteristic of coastlines on Baffin Island, Greenland, and British Columbia, among others.
+Their first experiment is set in an idealized circular domain that contains a bedrock plateau rimmed by a ridge that is studded with four valleys.
+They find that fjords become well-developed after about one million years and that ice velocities are highest in these fjords.
+
+Similar to \cite{kessler2008fjord}, we use the shallow ice model; however, we do not couple it to bed-erosion, calving, and GIA models.
+We initialize the experiment with a simple, unrealistic ice sheet placed on the bed topography \textcolor{red}{(figure X)}.
+We let the ice sheet evolve without climate forcing for 500 years until it is near equilibrium.
+The time it takes to reach equilibrium is much shorter than the \cite{kessler2008fjord} experiment because the bedrock is static.
+The first part of our simulation is an exponential relaxation towards equilibrium.
+A velocity pattern emerges that is similar to results from \cite{kessler2008fjord}, with the highest velocities where ice is constricted through the deepest valleys \textcolor{red}{(figure X)}.
+In and upstream of the valleys, the surface is drawn down due to the elevated export of ice \textcolor{red}{as shown in figure X}.
+
+The ice-flow model chosen for this simulation is computationally cheap compared to the other ice-flow models, such that it can be used to simulate an ice sheet over thousands of years.
+By simply switching between ice-flow models, icepack is able to simulate a range of glaciological conditions over a range of timescales and spatial scales.
+For example, icepack can be used to simulate a small domain dominated by an ice stream, an ice catchment that includes slow-moving interior ice and a floating ice shelf, or, as in this example, an entire ice sheet.
 

### 2020-11-03 icepack.tex
Mention possibility of DG for mass transport

@@
 The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
-The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.
+The current mass transport implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
+Implementing a solver that also works for discontinuous basis functions involves is more challenging but still possible.
 
@@
 The SIA model \citep{hutter1981effect} describes the interior of ice sheets well, where flow can be simply described as bed-parallel shear and surface and basal slopes are small.
-This approximation does not apply well near ice divides, where vertical velocities dominate, and near ice margins, where basal sliding can be a large fraction of the surface speed.
+This approximation breaks down near ice divides, where vertical velocities dominate, and near ice margins, where basal sliding can be a large fraction of the surface speed.
 The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.
@@
 The terms in the action functional are:
-
 \begin{align}
     \text{mass} & =\frac{1}{2}\int_\Omega u\cdot u\ud x\\\
-    \text{gravity} & = \int_\Omega\frac{2A(\varrho_I g)^n}{n+2} (\nabla h^2\cdot u) h^{n+1} \nabla s^{n-1}\ud x\\
-    \text{penalty} & = \frac{1}{2}\int_\Omega l^2\nabla u\cdot \nabla u\ud x 
+    \text{gravity} & = \int_\Omega\frac{2A(\rho_I g)^n}{n+2} h^{n + 1}|\nabla s|^{n - 1}\nabla s\cdot u\ud x\\
+    \text{penalty} & = \frac{1}{2}\int_\Omega \ell^2\nabla u\cdot \nabla u\ud x
 \end{align}
-
-where
-
-\begin{equation}
-l = 2 \text{ max}(\text{cell diameter}, 5h)
-\end{equation}
-
 and the action fuctional is
-
 \begin{equation}
@@
 \end{equation}
-
-We have included an additional penalty term, which helps alleviate numerical artifacts, especially near the ice margins and termini where the shallow ice approximation is less applicable.
+The default value of the length scale $\ell$ in the penalty term is defined as:
+\begin{equation}
+    \ell = 2\max\{\text{cell diameter}, 5h\}
+\end{equation}
+but users can adjust this to the value of their choice.
+This penalty term smooths over numerical artifacts, especially near the ice margins and termini.
+In these regions the shallow ice approximation is less applicable, so the error in solving a different set of equations is small compared to the inherent modeling error in using these equations in the first place.
 
@@
 
-The shallow ice and shallow stream models in icepack are completely standard.
-\textcolor{red}{JB: Aren't the models in icepack based on variational principles?
-This sentence and the next suggest that only the hybrid model is based on variational principles.}
-The hybrid flow model that we describe here is based on variational principles, terrain-following coordinates, and vertical spectral methods.
+The shallow ice and shallow stream models in icepack follow directly from the variational principles described above.
+The hybrid flow model, while also based on a variational principle, uses two more tricks: terrain-following coordinates and vertical spectral methods.
 Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
@@
 Elmer/Ice uses an ad-hoc procedure to define the unit normal vectors at mesh nodes \citep{gagliardini2013capabilities}.
-With a transformation to terrain-following coordinates, we can set the terrain-following vertical velocity $\omega$ to be $-\dot a_b/h$ at the ice base to impose this boundary condition.
+This procedure is nearly always effective in practice.
+But with a transformation to terrain-following coordinates, we can set the terrain-following vertical velocity $\omega$ to be $-\dot a_b/h$ at the ice base to impose this boundary condition with no additional intervention.
 

### 2020-11-03 icepack.tex
Expand C++ screed + update for API change

@@
 Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
-The degree 1 model is the most minimalistic model that still exhibits vertical shear.
+The degree 2 model is the most minimalistic model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
 Going to higher degree gives a more accurate approximation at the expense of greater computational effort.
@@
 For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
-In C++ one could avoid modifying the diagnostic solve routine by using variadic templates, but this is an advanced language feature and relying on it would present a steep learning curve for novices.
+In C++ one could avoid modifying the diagnostic solve routine by several different means.
+The function signature could be expressed using variadic templates.
+This is an advanced language feature and relying on it would present a steep learning curve for novices.
+The inputs could also be wrapped up in a class.
+Users can then define a subclass if they need to change the input fields.
+This would be a fairly idiomatic approach for the language but requires more knowledge on the part of users of object-oriented programming and its conventions than the approach we'll describe below.
+Finally, the inputs could be wrapped in a dictionary (a \texttt{std::map} from the C++ standard library).
+This approach sacrifices the static type checking guarantees of C++, one of the main advantages of the language.
+At this rate there is no marginal benefit over using a dynamically-typed language such as Python.
 
@@
 Using only keyword arguments enhances readability and comprehensibility for this particular case of invoking a physics solver.
-The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used in the literature -- $C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.
+The user only needs to know the argument names, which are chosen to agree with the English name most commonly used in the literature -- ``friction'', ``rheology'', ``velocity'', etc.
 The order of the arguments is arbitrary and immaterial.
@@
 These users then have a steep learning curve to overcome when they need to modify functionality that lives in the core library, as they now need to learn an entirely new programming language.
-Some low-level kernel code is, of course, always necessary for runtime efficiency.
-With Firedrake, this responsibility is deferred to a code generator.
+Some low-level kernel code is always necessary for runtime efficiency but with Firedrake this responsibility is deferred to a code generator.
 The average user can remain largely unaware of how the code generator works even when solving very advanced problems.
@@
 In this section we describe how various parts of the model physics were implemented in icepack, including both software design and the numerical solution procedures.
-From a design perspective, each flow model is represented by a Python class that has three responsibilities.
-First, the model class has to have a method that takes in the various fields (ice velocity, thickness, surface elevation, etc.) and returns a symbolic representation of the action functional for that particular model.
-Second, the model class must have a method that takes in the input fields and returns an approximate minimizer of the action functional.
-The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the first method calculates.
+The key classes that users interact with are flow models and solvers.
+The role of the model classes is to describe what problem is being solved.
+These classes describe the diagnostic model by taking in the input fields -- ice velocity, thickness, surface elevation, etc. -- and returning a symbolic representation of the action functional.
+There are several model classes, one for each set of physics equations: \texttt{ShallowIce}, \texttt{IceShelf}, \texttt{IceStream}, \texttt{HybridModel}.
+The model classes do not dictate how that problem should be solved numerically; this is the realm of a separate class called \texttt{FlowSolver}.
+This flow solver class has methods for computing the solutions of the diagnostic and prognostic equations and works the same regardless of which model is being solved.
+The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the model object creates.
 The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
-Finally the model must have a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
+Finally the flow solver has a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
 The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.
@@
 For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \citep{albrecht2014fracture}.
-The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge Kutta method would be more appropriate.
+The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge-Kutta method would be more appropriate.
 

### 2020-11-03 icepack.tex
Some minor fixes and formatting

--- 2020-11-03 icepack.tex
@@
 The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
-The current mass transport implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
-Implementing a solver that also works for discontinuous basis functions involves is more challenging but still possible.
+The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.
 

### 2020-11-04 icepack.tex
Assorted random corrections

--- 2020-11-03 icepack.tex
@@
 In this paper we introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
-Icepack is built on the finite element modeling library firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
+Icepack is built on the finite element modeling library Firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
 The diagnostic models in icepack are formulated through action principles that are specified in UFL.
@@
 Icepack improves upon the groundwork laid in VarGlaS in three main respects.
-First, icepack includes a simple 3D flow model that uses several features only available in firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
+First, icepack includes a simple 3D flow model that uses several features only available in Firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
 This model will be described in section \S\ref{sec:physics-hybrid-model}.
@@
 For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
-In C++ one could avoid modifying the diagnostic solve routine by several different means.
-The function signature could be expressed using variadic templates.
+In C++ one could avoid changing the signature of the diagnostic solve routine in several ways.
+First, the function signature could be expressed using variadic templates.
 This is an advanced language feature and relying on it would present a steep learning curve for novices.
-The inputs could also be wrapped up in a class.
+Second, the inputs could be wrapped up in a class.
 Users can then define a subclass if they need to change the input fields.
-This would be a fairly idiomatic approach for the language but requires more knowledge on the part of users of object-oriented programming and its conventions than the approach we'll describe below.
+This would be a fairly idiomatic approach for the language but requires users to know more about object-oriented programming than the approach we'll describe below.
 Finally, the inputs could be wrapped in a dictionary (a \texttt{std::map} from the C++ standard library).
 This approach sacrifices the static type checking guarantees of C++, one of the main advantages of the language.
-At this rate there is no marginal benefit over using a dynamically-typed language such as Python.
+At this rate there is little marginal gain over using a dynamically-typed language such as Python.
 
@@
 In adopting this approach, we are restricted to using keyword arguments instead of positional arguments.
-Using only keyword arguments enhances readability and comprehensibility for this particular case of invoking a physics solver.
+We argue that employing only keyword arguments enhances readability and comprehensibility for the particular user case of calling a physics solver.
 The user only needs to know the argument names, which are chosen to agree with the English name most commonly used in the literature -- ``friction'', ``rheology'', ``velocity'', etc.
@@
 For example, in defining a bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance.
-In this case, positional arguments correspond better with the semantics of the problem.
+Positional arguments correspond better with the semantics of the problem in this case than keyword arguments.
 
@@
 
-The hybrid flow uses several features that are only available in firedrake to better exploit the special structure of the problem.
+The hybrid flow uses several features that are only available in Firedrake to better exploit the special structure of the problem.
 Implementing this model also required some mathematical sleight-of-hand related to the terminus boundary condition that has not appeared in the literature before.
@@
 
-For problems defined on extruded geometries, firedrake includes support for tensor product elements, which includes using different bases in the horizontal and vertical directions \citep{mcrae2016automated}.
+For problems defined on extruded geometries, Firedrake includes support for tensor product elements, which includes using different bases in the horizontal and vertical directions \citep{mcrae2016automated}.
 Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
@@
 
-Extruded meshes and tensor product elements are available in firedrake but not in FEniCS.
+Extruded meshes and tensor product elements are available in Firedrake but not in FEniCS.
 Other general-purpose finite element modeling packages that support tensor product elements include deal.II and nektar++ \citep{bangerth2007deal, cantwell2015nektar++}.
@@
 where the critical stress $\tau_c$ is a certain specified fraction of the water pressure in the subglacial hydrological system.
-In icepack sliding laws are not expressed directly but rather as the the derivative of an acction functional.
+Sliding laws in icepack are not expressed directly, but rather as the the derivative of an action functional.
 To implement the Schoof sliding law we need to know the antiderivative of equation \eqref{eq:schoof-sliding}.

### 2020-11-09 icepack.tex
Some reorganizing

@@
 
-\subsection{Data assimilation}
+\subsection{Heat transport} \label{sec:heat-transport}
+
+\subsection{Damage transport} \label{sec:damage-transport}
+
+\section{Data assimilation} \label{sec:data-assimilation}
 
 Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
-The class \texttt{InverseProblem} represents the specification of the inverse problem.
-To specify an inverse problem, the user must provide:
+The class \texttt{InverseProblem} represents the specification of an inverse problem, which requires:
 \begin{itemize}
@@
 \end{itemize}
-At present, this class assumes that the observed state is always the ice velocity, but in principle the same design would suffice for more complicated inverse problems.
-More importantly, the inverse problem is flexible enough to account for the fact that we might have substituted in our own parameterization for the rheology or friction coefficient.
-For example, nearly all studies in the literature use an alternate parameterization of the friction coefficient in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
-This information is passed to the inverse problem by specifying the names of the keyword arguments for the observed state and the field we wish to observe.
+This class currently assumes that the observed state is always the ice velocity.
+In principle the same design would suffice for more complicated inverse problems and this is an area of active development.
+
+The inverse problem class is flexible enough to account for users defining their own parameterization for the rheology or friction coefficient as described in \S\ref{sec:physics-substitution}.
+This flexibility with respect to parameterization is not just convenient but essential for common data assimilation workflows.
+Nearly all studies in the literature introduce a parameterization of the field to be estimated in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
+For example, one could define the friction coefficient $C$ in terms of an auxiliary variable $\beta$ as $C = \beta^2$ to guarantee positivity.
+One could also just as easily use $C \propto \exp(\beta)$.
+The data assimilation routines in icepack can work the same for any parameterization of the physics because Firedrake provides a rich set of routines for symbolically calculating functional derivatives.
+The inverse problem class only needs to know which functional needs to be differentiated with respect to which field.
 
@@
 This class will be described further in section \S\ref{sec:numerics-inverse-solvers}.
-The inverse solver is completely agnostic to the particular parameterization of the physics in terms of the inferred field as long as it guarantees positivity.
-This feature interacts seamlessly with the ability to modify individual model physics components as described in \S\ref{sec:physics-substitution}.
-
-\subsection{Damage transport}
-
-\subsection{Heat transport}
 
@@
 
-In this section we describe how various parts of the model physics were implemented in icepack, including both software design and the numerical solution procedures.
+In the previous sections we described \emph{what} problems icepack can solve, i.e. various physics models and data assimilation problems.
+In this section, we'll describe \emph{how} these problems are solved.
+This separation between the two questions parallels the broader design of the software package.
+
 The key classes that users interact with are flow models and solvers.

### 2020-11-09 icepack.tex
Add reference on Picard vs Newton

@@
 
-Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation \textcolor{red}{cite -- I think Mathieu and Dan did this}, but few if any have considered the problem of when to stop iterating.
+Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation, for example \citet{perego2012parallel}.
+Few have considered the problem of when to stop iterating.
 The most common stopping criteria are when (1) the 2-norm of the residual is sufficiently small or (2) the relative change in the iterates is sufficiently small.
@@
 
-As shown in equation \eqref{action-functional}, the action has units of power and is the sum of the dissipation due to viscosity, friction, gravitational driving, and ocean back-pressure at the terminus.
+As shown in equation \eqref{action-functional}, the action for most models has units of power and is the sum of dissipation due to viscosity, friction, gravitational driving, and ocean back-pressure at the terminus.
 The viscous and frictional terms are convex, positive functions of the velocity.

### 2020-11-09 icepack.tex
Add text on damage model + stub out demo

@@
 
+Other physical fields besides temperature can influence ice fluidity.
+At large spatial scales (> 5 km), crevasse fields affect ice flow by reducing the lateral area over which stress can be transmitted.
+Modeling individual fractures is not computationally feasible for large-scale simulations.
+Instead, we have implemented the phenomenological model described in \citet{albrecht2014fracture}, which is based on the theory of continuum damage mechanics.
+This model is defined in the class \texttt{DamageTransport}.
+Prognostic damage models can be broken down into three parts: (1) evolve the damage field based on the membrane stress of the glacier, (2) advect the damage field with flow, and (3) feed the damage field back into the fluidity of the glacier.
+Changes in fluidity in turn affect the membrane stress; the coupling between bulk damage and ice flow goes both ways.
+The model from \citet{albrecht2014fracture} adds sources of damage where the membrane stress exceeds a critical value and sinks of damage when the principal strain rate is less than a critical value.
+These kinds of simplified models ignore complications such as inertial or 3D effects.
+They are intended more to capture some important features of the real physical system ecnomically than to describe it exactly.
+
 \section{Data assimilation} \label{sec:data-assimilation}
@@
 The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.
-Other problems within glaciology have more of a hyperbolic character and there the argument for using shock-capturing methods is much stronger.
-For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \citep{albrecht2014fracture}.
-The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge-Kutta method would be more appropriate.
+
+Other problems in glaciology have more of a hyperbolic character.
+For example, the thresholding behavior of the source terms for the damage model (see \S\ref{sec:damage-transport}) can create sharp discontinuities.
+The implicit Euler scheme would obscure this important feature.
+For the damage solver, we have instead used a strong stability-preserving Runge-Kutta method in time and a discontinuous Galerkin basis in space to best capture these sharp interfaces \citep{shu1988efficient}.
 
@@
 
+\subsection{Synthetic ice shelf with damage transport}
+
+We simulated the evolution of a synthetic ice shelf towards steady state and coupled it with the damage transport model described in \S\ref{sec:damage-transport}.
+The geometry of the ice shelf consists of the intersection of two circles, one with a larger radius and offset center, designed to roughly mimic the shape and size of real ice shelves.
+The radius of the whole shelf is 200km.
+Four ice streams flow into the shelf with varying speeds and a prescribed inflow thickness.
+In the first phase of the experiment, the ice shelf is propagated to steady state without damage for 200 years.
+After this time interval, the flux imbalance is less than 1\% of the influx.
+In the second phase, damage transport is turned on and coupled to the ice velocity.
+An interesting feature to observed in the approximate equilibrium damage field is that the highest values occur between the streams and not within them.
+Additionally, the spacing between the streams changes as a result of adding damage.
+The final ice thickness, velocity, strain rate, and damage are shown in figure \textcolor{red}{ZZZ}.
+
 \subsection{Larsen C Ice Shelf}

### 2020-11-09 icepack.tex
Add reference on SSA/Stokes comparison

@@
 
-Several studies have compared the shallow stream approximation to higher-order models, such as Blatter-Pattyn or full Stokes \textcolor{red}{cite the MIPs}.
-The most appreciable difference between lower- and higher-order models occurs near the glacier grounding line, where the full Stokes equations can represent bridging stresses \textcolor{red}{cite van der Veen}.
+Several studies have compared the shallow stream approximation to 3D models such as Blatter-Pattyn or full Stokes \citep{pattyn2013grounding}.
+The most appreciable difference between lower- and higher-order models occurs near the glacier grounding line, where the full Stokes equations can represent bridging stresses \citep{van2013fundamentals}.
+The lack of vertical strain rates in the ice viscosity in 2D models can also lead to different equilibrium grounding line positions under the same external forcing.
 The shallow stream approximation is warranted for studying individual fast-flowing glaciers where the concern is more with studying processes and responses than with complete accuracy.

### 2020-11-10 icepack.tex
Add description of heat transport

@@
 \subsection{Heat transport} \label{sec:heat-transport}
+
+We implemented the enthalpy transport model described in \citet{aschwanden2012enthalpy}.
+The enthalpy describes the heat content of the material in a way that incorporates both temperature and latent heat stored in meltwater.
+The temperature and meltwater fraction can be uniquely calculated at any point from the value of the enthalpy, so nothing is sacrificed in describing heat content one way or the other.
+Using the enthalpy has the advantage of circumventing many of the difficulties associated with tracking the interface between cold and temperate ice.
+Our implementation of the model uses all of the simplifying assumptions described in \citet{aschwanden2012enthalpy}, for example that horizontal diffusion is negligible and that heat capacity and conductivity are not temperature-dependent within each phase.
+
+The fluidity factor $A$ in Glen's law is roughly a known function of both temperature and melt fraction and we have included this function in the package.
+Having some description of heat transport is thus necessary for describing realistic glacier flows, but it is not sufficient by itself.
+For one, while the dependence of fluidity on temperature is known fairly well from laboratory experiments, the dependence on melt fraction is known with much less certainty.
+It is likely that users of this module will want to substitute in their own parameterization for melt fraction dependence.
+Second, other processes such as damage, fabric development, and impurities can affect the fluidity as well.
+For these reasons, the diagnostic model takes in the fluidity as an argument and users must calculate the fluidity field themselves, either using the default parameterization that icepack supplies or one of their own.
+Our general design principle is that icepack will solve differential equations for the various prognostic fields but it's up to users to decide how these fields are coupled.
+
+We have not implemented a model for the surface or englacial transport of meltwater, although it would be possible to include this process.
+Instead, we rely on an external scheme to act as a sink of enthalpy for meltwater fraction values above some user-defined critical value, which defaults to 1\%.
 

### 2020-11-10 icepack.tex
Remove hybrid multigrid section

@@
 
-\subsubsection{Multigrid-type algorithm}
-
-One of the main advantages of this hybrid model is the ease with which a more accurate solution can be bootstrapped from a less accurate solution, since the two fields are defined on exactly the same geometry.
-For example, if we only have a very crude initial guess for the solution, we can obtain a degree-1 solution relatively cheaply.
-The resulting velocity field is then used as a much more informed initial guess for a degree-5 solution.
-\textcolor{red}{Benchmark.}
-Since the vertical basis functions are orthogonal, the projection of the degree-1 solution into the degree-5 function space can be calculated by simply copying array elements.
-By contrast, if we tried to use the same strategy to bootstrap a solution with a small number of vertical layers to a solution with a larger number of vertical layers, we would also have to interpolate values to the fine layers.
-
-Other hybrid models in the literature use only a fixed number of vertical basis functions.
-For example, \citet{bassis2010hamilton} use only the two vertical basis functions $\psi_1(\zeta) = 1$, $\psi_2(\zeta) = 1 - (1 - \zeta)^{n + 1}$, the idea being that the vertical profile can be written as a direct superposition of the plug flow and shallow ice approximation modes.
-By contrast, we leave the number of vertical basis functions up to the user.
-This flexibility enables the multigrid-type approach described above.
-
-The bootstrapping approach described here can be used both for the forward model and for inverse problems.
-First, we estimate the unknown parameter using only degree-1 vertical basis functions for the velocity field, which captures most of the spatial variability of the unknown parameter.
-The resulting coefficient is then used as the initial guess with higher degree vertical basis functions for the velocity \textcolor{red}{benchmark}.
-
 \section{Demonstrations}

### 2020-11-11 icepack.tex
Add synthetic ice shelf figure

@@
 
-\subsection{Synthetic ice shelf with damage transport}
+\subsection{Synthetic ice shelf}
+
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/gibbous/damage.png}
+    \caption{Results of ice shelf damage simulation: (a) ice velocity, (b) change in ice speed with and without damage, (c) thickness, and (d) damage field.}
+    \label{fig:damage}
+\end{figure}
 
@@
 Additionally, the spacing between the streams changes as a result of adding damage.
-The final ice thickness, velocity, strain rate, and damage are shown in figure \textcolor{red}{ZZZ}.
+The final ice thickness, velocity, strain rate, and damage are shown in figure \ref{fig:damage}.
 

### 2020-11-11 demos/gibbous/ice-shelf.py

@@ -2,8 +2,9 @@ import os
 import argparse
 import tqdm
 import firedrake
-from firedrake import as_vector, inner, ds
+from firedrake import sym, grad, as_vector, inner, ds
 import icepack
+from icepack.models.viscosity import membrane_stress
 
 parser = argparse.ArgumentParser()
 parser.add_argument('--mesh')
@@ -19,6 +20,7 @@ mesh = firedrake.Mesh(args.mesh)
 Q = firedrake.FunctionSpace(mesh, family='CG', degree=2)
 V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=2)
 Δ = firedrake.FunctionSpace(mesh, family='DG', degree=1)
+S = firedrake.TensorFunctionSpace(mesh, family='DG', degree=1)
 
 # Load the initial data from an HDF5 file if available
 if args.input:
@@ -105,8 +107,15 @@ for step in tqdm.trange(num_timesteps):
         dt, thickness=h, accumulation=a, velocity=u, thickness_inflow=h0
     )
     if args.damage:
+        ε = firedrake.project(sym(grad(u)), S)
+        M = firedrake.project((1 - D) * membrane_stress(ε, A), S)
         D = damage_solver.solve(
-            dt, damage=D, velocity=u, fluidity=A, damage_inflow=D_inflow
+            dt,
+            damage=D,
+            velocity=u,
+            strain_rate=ε,
+            membrane_stress=M,
+            damage_inflow=D_inflow
         )
 
     u = flow_solver.diagnostic_solve(

### 2020-11-11 demos/gibbous/make_plots.py

@@ -1,28 +1,34 @@
+import argparse
 import numpy as np
 from mpl_toolkits.axes_grid1 import make_axes_locatable
 import firedrake
 from firedrake import sqrt, inner
 import icepack.plot
 
+parser = argparse.ArgumentParser()
+parser.add_argument('--mesh')
+parser.add_argument('--undamaged')
+parser.add_argument('--damaged')
+parser.add_argument('--output')
+args = parser.parse_args()
+
 def colorbar(figure, axes, mappable, *args, **kwargs):
     divider = make_axes_locatable(axes)
     cax = divider.append_axes('right', size='5%', pad=0.05)
     return figure.colorbar(mappable, *args, cax=cax, **kwargs)
 
-mesh = firedrake.Mesh('ice-shelf.msh')
+mesh = firedrake.Mesh(args.mesh)
 Q = firedrake.FunctionSpace(mesh, family='CG', degree=2)
 V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=2)
 Δ = firedrake.FunctionSpace(mesh, family='DG', degree=1)
 
-filename = 'steady-state-undamaged'
-with firedrake.DumbCheckpoint(filename, mode=firedrake.FILE_READ) as chk:
+with firedrake.DumbCheckpoint(args.undamaged, mode=firedrake.FILE_READ) as chk:
     h_undamaged = firedrake.Function(Q)
     u_undamaged = firedrake.Function(V)
     chk.load(h_undamaged, name='h')
     chk.load(u_undamaged, name='u')
 
-filename = 'steady-state-damaged'
-with firedrake.DumbCheckpoint(filename, mode=firedrake.FILE_READ) as chk:
+with firedrake.DumbCheckpoint(args.damaged, mode=firedrake.FILE_READ) as chk:
     h_damaged = firedrake.Function(Q)
     u_damaged = firedrake.Function(V)
     D = firedrake.Function(Δ)
@@ -36,37 +42,30 @@ speed = sqrt(inner(u_undamaged, u_undamaged))
 v = u_undamaged / speed
 δu = firedrake.interpolate(inner(u_damaged - u_undamaged, v), Q)
 
-fig, axes = icepack.plot.subplots(nrows=1, ncols=2, sharex=True, sharey=True)
-axes[0].set_ylabel('distance (m)')
[…]

### 2020-11-11 icepack.tex
Add velocity error estimate from measures
Describe Glen flow law

@@
 
-\textcolor{red}{Should Glen's flow law be mentioned or written somewhere?}
+All the momentum transport models inherit a fundamental nonlinearity in the mechanics of ice flow.
+For a Newtonian viscous fluid, the stress tensor $\tau$ and the strain rate tensor $\dot\varepsilon$ are linearly proportional to each other.
+Glaciers, however, have a nonlinear constitutive relation:
+\begin{equation}
+    \dot\varepsilon = A|\tau|^{n - 1}\tau
+\end{equation}
+where $A$ is the \emph{fluidity} and $n$ is the \emph{Glen flow law exponent} \citep{cuffey2010physics}.
+The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.
 
@@
 We used the InSAR phase-based ice velocity map from \citet{mouginot2019continent}.
-This dataset has nominal errors over the Larsen Ice Shelf between \textcolor{red}{find it}.
+This dataset has nominal errors over the Larsen Ice Shelf on the order 1 -- 7 meters per year.
 We used the recently-released BedMachine map of the thickness of Antarctica \citep{morlighem2019deep}.

### 2020-11-20 icepack.tex
Describe UFL more

@@
 A third advantage of variational principles is that the action functional itself can be used to define a solver convergence criterion that is independent of the mesh and requires little tuning on the part of the user.
-Icepack features a 3D diagnostic model based on terrain-following coordinates and vertical spectral discretization.
-This model resolves both plug- and shear-flow components of horizontal ice flow with a minimum of extra computational expense over 2D, depth-averaged models.
+Icepack includes the 2D shallow ice and shallow stream models and a novel implementation of the 3D Blatter-Pattyn model.
 Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
@@
 The glacier flow modeling package closest in spirit to icepack is VarGlaS \citep{brinkerhoff2013data}.
-VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}.
-The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for describing weak forms of PDE.
+VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}, whereas icepack is built on top of Firedrake.
+The Firedrake project began as an outgrowth of FEniCS.
+Both packages share a similar goal of saving users from manually writing low-level code for assembling the systems of equations that discretize their physics model.
+Instead, users describe the weak form of the partial differential equations they wish to solve using a high-level domain-specific language (DSL) called the \emph{Unified Form Language} or UFL \citep{alnaes2014unified}.
+This DSL is embedded entirely into Python, i.e. the complete syntax of UFL can be mapped directly onto overloaded operators in the Python programming language.
+Both FEniCS and Firedrake then generate optimized C or C++ code to assemble the discretized system of equations from this symbolic description of the problem \citep{kirby2006compiler, rathgeber2016firedrake}.
+Combining a DSL and a code generator frees users from the very error-prone process of writing these assembly kernels themselves, and makes the syntax of the code align closer to the syntax of written mathematical expressions.
+Additionally, having a high-level symbolic description of the problem makes it possible to automatically derive tangent linear models and adjoints \citep{mitusch2019dolfin}.
+
 Icepack improves upon the groundwork laid in VarGlaS in three main respects.

### 2020-11-20 icepack.tex
Corrections (thanks Jess)

@@
 where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
-The thickness is represented using continuous Galerkin basis functions.
+The thickness is represented using continuous, piecewise polynomial basis functions in each cell of the mesh.
 The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
 The current mass transport implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
-Implementing a solver that also works for discontinuous basis functions involves is more challenging but still possible.
+We have not yet implemented a solver that works for discontinuous basis functions, but this extension is completely feasible within our framework and may come in a future release of icepack.
 
@@
 Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.
+We have not implemented this feature but support for non-hydrostatic ice shelves may come in a future release of icepack.
 
@@
 
-There are three diagnostic models implemented in icepack.
-The \emph{shallow ice approximation} (SIA) and \emph{shallow stream approximation} (SSA) are 2D models describing the depth-averaged velocity of ice flow, while the \emph{first-order} or \emph{Blatter-Pattyn} approximation is a 3D model describing the horizontal velocity (termed the \emph{hybrid model} in icepack) \citep{blatter1995velocity, pattyn2003new}.
+There are four diagnostic models implemented in icepack.
+See \citet{dukowicz2010consistent} for a thorough explanation of the various approximations and action principles velow.
+
+The most complex and most physically accurate model is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
+(In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
+The only approximation that the first-order model makes is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
+This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
+Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.
+
+From the first-order model, two approximations are possible.
+First, the \emph{shallow ice approximation} (SIA) comes from assuming that vertical shear is the dominant mode of ice flow.
 The SIA model \citep{hutter1981effect} describes the interior of ice sheets well, where flow can be simply described as bed-parallel shear and surface and basal slopes are small.
 This approximation breaks down near ice divides, where vertical velocities dominate, and near ice margins, where basal sliding can be a large fraction of the surface speed.
+Second, one could assume that ice flow is purely by horizontal extension and that the surface and bed velocities are practically the same.
+This is called the \emph{shallow stream approximation} (SSA).
 The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.
-This model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
-Plug flow is a good approximation in fast-flowing ice streams and outlet glaciers near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
-The only approximation in the hybrid model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
-This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
-Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.
+Finally, for floating ice, the shallow stream approximation simplifies further to the \emph{shallow shelf approximation}.
+The shallow shelf model has no bed friction and the ice surface elevation can be eliminated from the system.
 
@@
 The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.
-
-\subsubsection{Variational principles}
 
@@
 
+Every momentum transport model in icepack is encapsulated in its own Python class.
+The key responsibility of the model classes is to take in the input fields -- the ice thickness, velocity, etc. -- and return a symbolic description of the action functional in UFL.
+
 The action principle is especially useful for designing a robust numerical solver.
@@
 The result is a 2D system of equations for depth-averaged horizontal velocity.
+The class in icepack that represents this physics model is called \texttt{ShallowIce}.
 The terms in the action functional are:
@@
 The shallow ice approximation applies well in ice-sheet interiors.
-For this reason, and because the equations are particularly simple to solve, this approximation has been a common choice for ice-sheet modeling \textcolor{red}{cite}.
+For this reason, and because the equations are particularly simple to solve, this approximation has been a common choice for ice-sheet modeling \citep{kirchner2016shallow}.
 Caution must be taken, however, because areas of the ice sheet that experience high vertical velocities (e.g., ice divides) or plug flow (e.g., ice streams) are not well represented by this approximation.
 
-\subsubsection{Shallow stream approximation}
+\subsubsection{Shallow stream and shelf approximations}
 
@@
 In this case, the momentum equation can be vertically integrated to obtain a 2D system of equations.
+The class in icepack that represents this physics model is called \texttt{IceStream}.
 The terms in the action functional are:
@@
 \end{equation}
-Using integration by parts, the terminus component can be eliminated completely.
+Additionally, the terminal stress term of the action disappears by applying integration by parts to the gravity term.
+Since the action functional for ice shelves has fewer terms than for grounded ice streams, we have defined a separate \texttt{IceShelf} model class.
+The ice shelf and ice stream models share common components, i.e. the viscosity and side wall stress.
 
@@
 
-\subsection{Hybrid model} \label{sec:physics-hybrid-model}
+\subsubsection{Hybrid model} \label{sec:physics-hybrid-model}
 
-The shallow ice and shallow stream models in icepack follow directly from the variational principles described above.
+The first-order model in icepack is described in the class \texttt{HybridModel}.
+The shallow ice and shallow stream models follow directly from the variational principles described above.
 The hybrid flow model, while also based on a variational principle, uses two more tricks: terrain-following coordinates and vertical spectral methods.
@@
 
-\subsubsection{Terrain-following coordinates}
+\paragraph{Terrain-following coordinates}
 
@@
 
-\subsubsection{Spectral discretization}
+\paragraph{Spectral discretization}
 
@@
 These users then have a steep learning curve to overcome when they need to modify functionality that lives in the core library, as they now need to learn an entirely new programming language.
-Some low-level kernel code is always necessary for runtime efficiency but with Firedrake this responsibility is deferred to a code generator.
+Some low-level kernel code is always necessary for runtime efficiency but with Firedrake this responsibility is deferred to a code generator \citep{rathgeber2016firedrake}.
 The average user can remain largely unaware of how the code generator works even when solving very advanced problems.
@@
 In order to use terrain-following coordinates, the hybrid model assumes that the geometry of the domain is an extruded mesh, where a 2D footprint mesh is lifted into 3D.
-Firedrake includes support for creating extruded meshes by calling the function \texttt{ExtrudedMesh} on the 2D footprint \citep{bercea2016structure}.
+Firedrake includes support for creating extruded meshes by calling the function \texttt{ExtrudedMesh} on the 2D footprint \citep{bercea2016structure, mcrae2016automated}.
 The cells of an extruded mesh are triangular prisms instead of the more common tetrahedra used for general 3D meshes.
@@
 \subsection{Synthetic ice sheet}
+
 As a second test case for icepack, we ran an experiment inspired by \cite{kessler2008fjord}.
@@
 
-Similar to \cite{kessler2008fjord}, we use the shallow ice model; however, we do not couple it to bed-erosion, calving, and GIA models.
-We initialize the experiment with a simple, unrealistic ice sheet placed on the bed topography \textcolor{red}{(figure X)}.
-We let the ice sheet evolve without climate forcing for 500 years until it is near equilibrium.
+Similar to \cite{kessler2008fjord}, we use the shallow ice model; however, we do not couple it to bed-erosion, calving, or GIA.
+We initialized the experiment with a simple but unrealistic ice thickness \textcolor{red}{(figure X)}.
+We then evolve the ice sheet without climate forcing for 500 years until it is near equilibrium.
 The time it takes to reach equilibrium is much shorter than the \cite{kessler2008fjord} experiment because the bedrock is static.
@@
 The ice-flow model chosen for this simulation is computationally cheap compared to the other ice-flow models, such that it can be used to simulate an ice sheet over thousands of years.
-By simply switching between ice-flow models, icepack is able to simulate a range of glaciological conditions over a range of timescales and spatial scales.
+By simply switching between ice-flow models, icepack is able to simulate many different glaciological conditions over a wide range of spatiotemporal scales.
 For example, icepack can be used to simulate a small domain dominated by an ice stream, an ice catchment that includes slow-moving interior ice and a floating ice shelf, or, as in this example, an entire ice sheet.

### 2020-11-23 icepack.tex
Change use of diagnostic + prognostic model vs mass + momentum transport

@@
 
-\subsection{Mass transport}
+\subsection{Prognostic model}
 
-The PDE describing how the ice thickness changes in time is a conservative advection equation:
+The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
+The prognostic model is a conservative advection equation:
 \begin{equation}
@@
 The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
-The current mass transport implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
+The current prognostic model implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
 We have not yet implemented a solver that works for discontinuous basis functions, but this extension is completely feasible within our framework and may come in a future release of icepack.
@@
 
-\subsection{Momentum transport}
+\subsection{Diagnostic model}
 
-There are four diagnostic models implemented in icepack.
+There are four \emph{diagnostic} or \emph{momentum transport} models implemented in icepack.
 See \citet{dukowicz2010consistent} for a thorough explanation of the various approximations and action principles velow.
@@
 
-All the momentum transport models inherit a fundamental nonlinearity in the mechanics of ice flow.
+All the diagnostic models inherit a fundamental nonlinearity in the mechanics of ice flow.
 For a Newtonian viscous fluid, the stress tensor $\tau$ and the strain rate tensor $\dot\varepsilon$ are linearly proportional to each other.
@@
 
-Every momentum transport model in icepack is encapsulated in its own Python class.
+Every diagnostic model in icepack is encapsulated in its own Python class.
 The key responsibility of the model classes is to take in the input fields -- the ice thickness, velocity, etc. -- and return a symbolic description of the action functional in UFL.
@@
 The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
-In this case, the momentum equation can be vertically integrated to obtain a 2D system of equations.
+The momentum equations can, again, be vertically integrated to obtain a 2D system of equations.
 The class in icepack that represents this physics model is called \texttt{IceStream}.
@@
 The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
-To solve the mass transport equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
+To solve the advection equation equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
 The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
@@
 The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
-The extra cost of using an implicit time discretization for the mass transport equation is dwarfed by the cost of the diagnostic solve in any case.
-Advanced users who are interested in maximizing performance can subclass the mass transport solver to implement a faster explicit scheme.
+The extra cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
+Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.
 

### 2020-11-23 icepack.tex
Minor corrections

@@
 In adopting this approach, we are restricted to using keyword arguments instead of positional arguments.
-We argue that employing only keyword arguments enhances readability and comprehensibility for the particular user case of calling a physics solver.
+We argue that employing only keyword arguments is a strength rather than weakness because it enhances readability and comprehensibility for the particular use case of calling a physics solver.
 The user only needs to know the argument names, which are chosen to agree with the English name most commonly used in the literature -- ``friction'', ``rheology'', ``velocity'', etc.
@@
 Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
-The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
+The use of an unconditionally stable scheme guarantees that they will get an answer, rather than a runtime error, should they try to use a large timestep.
 The extra cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
@@
     \includegraphics[width=0.95\linewidth]{demos/gibbous/damage.png}
-    \caption{Results of ice shelf damage simulation: (a) ice velocity, (b) change in ice speed with and without damage, (c) thickness, and (d) damage field.}
+    \caption{Results of ice shelf damage simulation: (a) ice velocity without damage, (b) change in ice speed with and without damage, (c) thickness, and (d) damage field.}
     \label{fig:damage}
@@
 The solver classes, as opposed to the model classes, provide the interface for making these kinds of algorithmic choices.
-While alternative solvers might offer faster runtime performance than direct factorization, they also requires making additional choices choices -- how often to restart GMRES?
+While alternative solvers might offer faster runtime performance than direct factorization, they also requires making additional choices -- how often to restart GMRES?
 How much fill-in to allow in the incomplete factorization?

### 2020-11-23 icepack.tex
Rework explanation of substituting physics components

@@
 
-One of our goals with icepack is to facilitate experimentation with the model physics.
+One of our goals with icepack is to facilitate experimentation with the model physics, even for novice users.
+Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would meet these needs.
 To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
@@
 In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
-For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
-In C++ one could avoid changing the signature of the diagnostic solve routine in several ways.
-First, the function signature could be expressed using variadic templates.
-This is an advanced language feature and relying on it would present a steep learning curve for novices.
-Second, the inputs could be wrapped up in a class.
-Users can then define a subclass if they need to change the input fields.
-This would be a fairly idiomatic approach for the language but requires users to know more about object-oriented programming than the approach we'll describe below.
-Finally, the inputs could be wrapped in a dictionary (a \texttt{std::map} from the C++ standard library).
-This approach sacrifices the static type checking guarantees of C++, one of the main advantages of the language.
-At this rate there is little marginal gain over using a dynamically-typed language such as Python.
+For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function, which is undesirable.
+In C++, one could avoid changing the signature of the diagnostic solve routine by (1) using variadic templates, (2) wrapping the inputs in a class, or (3) passing all arguments in a dictionary.
+Using variadic templates or wrapping the inputs in a class would require users to know more about generic or object-oriented programming than a novice might.
+Using a dictionary data structure to pass arguments is relatively easier but would be more idiomatic in Python than in C++.
 
-Any model physics component in icepack can be substituted for a parameterization of the user's choosing.
-Each flow model object is initialized with a set of functions to calculate the various terms of the action functional.
-For example, the ice stream model is initialized with functions to calculate the contributions due to viscosity, basal friction, side wall drag, and gravity.
-The default parameterizations are set to the most common choices in the literature, but if the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
-When the diagnostic solve procedure is invoked, all of its arguments are then passed on by keyword to the method that calculates the action functional.
-Any fields that are unnecessary for calculating a given component are ignored.
-For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function makes no use of these arguments.
-The user can completely change the functional form and add new arguments, with the restriction that arguments are passed by keyword and not positionally.
+In icepack, users can substitute any diagnostic model component for the parameterization of their choosing, including adding new fields.
+From the user's perspective, substituting model physics components does not require any advanced language features beyond keyword arguments.
+To understand how this is possible, we'll briefly describe the path that the input fields take through the program.
+\begin{enumerate}
+    \item The user passes all arguments to the diagnostic solve procedure by keyword, essentially in a dictionary mapping names to fields.
+    \item The diagnostic solve procedure creates a symbolic representation of the action functional by summing up several terms, like the viscosity, basal friction, etc.
+        Calculating these terms is delegated to specialized procedures for each term.
+        Each term procedure gets the entire collection of fields.
+    \item The routine that calculates the terms of the action selects which fields it actually needs from the argument dictionary, and then creates the symbolic representation of that term.
+    \item Finally, once the symbolic representation of the action functional has been created, all the responsibility passes to the nonlinear solver.
+\end{enumerate}
+To substitute different model components, the user intervenes at step 3.
+Each model object -- shallow ice, shallow stream, etc. -- is initialized with a default set of routines to calculate the terms of that model's action functional.
+The user can replace these default routines with one of their own choosing by passing the function of their choice when that model object is initialized.
+
+In step 3, any fields that were unnecessary for calculating a given component are ignored.
+For example, the gravitational driving power routine will pull out the velocity, thickness, and surface elevation.
+It will have access to other fields, for example the ice fluidity.
+These extra arguments are ignored.
 
 In adopting this approach, we are restricted to using keyword arguments instead of positional arguments.
-We argue that employing only keyword arguments is a strength rather than weakness because it enhances readability and comprehensibility for the particular use case of calling a physics solver.
+We argue that employing only keyword arguments is a strength rather than a weakness because it enhances readability and comprehensibility for the particular use case of calling a physics solver.
 The user only needs to know the argument names, which are chosen to agree with the English name most commonly used in the literature -- ``friction'', ``rheology'', ``velocity'', etc.
 The order of the arguments is arbitrary and immaterial.
-The preference for argument passing by name is specific to this use case and not universal.
-For example, in defining a bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance.
-Positional arguments correspond better with the semantics of the problem in this case than keyword arguments.
+The preference for argument passing by name is specific to this use case, however, and is not universal.
+
+One key limitation of our approach is that users can change individual terms of the action functional but they cannot add or remove terms.
+This calls for creating an entirely new diagnostic model.
 

### 2020-11-24 icepack.tex
Cut some redundant text

@@
 
-One key limitation of our approach is that users can change individual terms of the action functional but they cannot add or remove terms.
-This calls for creating an entirely new diagnostic model.
-
-One of our main design goals for icepack is to enable users to ``plug and play'' their own physics.
-Since many aspects of glacier flow are not understood completely, we view extensibility as an essential requirement for any software package in this domain.
-Writing the entire package in Python was necessary to meet this goal.
-Many scientific software packages consist of a highly-optimized core library in C, C++, or Fortran, together with a wrapper that exposes this library to an interpreted language like Python.
-This approach in principle offers the best of both worlds -- the runtime efficiency of a compiled language with the convenience and interactivity of an interpreted one.
-Many practitioners, however, will use such a library entirely through the interpreted surface layer and have no familiarity with the language in which the fast core was written.
-These users then have a steep learning curve to overcome when they need to modify functionality that lives in the core library, as they now need to learn an entirely new programming language.
-Some low-level kernel code is always necessary for runtime efficiency but with Firedrake this responsibility is deferred to a code generator \citep{rathgeber2016firedrake}.
-The average user can remain largely unaware of how the code generator works even when solving very advanced problems.
+One key limitation of our approach is that, while users can change what goes into one or more terms of the action functional, they cannot add or remove terms.
+This case would call for writing a new diagnostic model.
 

### 2020-11-24 icepack.tex
Add reference on Larsen B breakup

@@
 Several recent studies have focused on Larsen C because it may be unstable in the warmer climate of the coming decades.
-From January to March of 2002, the neighboring Larsen B Ice Shelf disintegrated due to surface melt pond-induced fracture \textcolor{red}{check that this is the currently-accepted explanation and cite}, and this mechanism could also lead to the breakup of Larsen C.
+From January to March of 2002, the neighboring Larsen B Ice Shelf disintegrated due to surface melt pond-induced fracture \citep{banwell2013breakup}.
+This mechanism might also lead to the breakup of Larsen C in the future.
 One of the key factors affecting the stability of ice shelves is the presence of marine ice -- seawater that freezes onto the base of an ice shelf -- in the suture zones where two flow units meet \citep{kulessa2014marine}.

### 2020-11-24 icepack.tex
Remove obsolete reference to ice shelf time series

@@
 
-\citet{adusumilli2018variable} used surface elevation measurements from multiple satellite altimeters, together with firn density estimates from a climate model and tide height estimates, to compile a time series of ice thickness change over the Larsen Ice Shelf for the past 23 years.
-This same approach can be used to derive an improved estimate the thickness of the ice shelf from, e.g. the REMA surface elevation dataset.
-Future work will focus on this and other methods to assimilate more accurate thickness data and to explicitly account for the statistical errors in this field.
-
 \subsubsection{Parameterization}

### 2020-11-24 icepack.tex
Adjust wording in usability section

@@
 Previous work on numerical modeling of glacier flow has focused largely on technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory, does it scale to large numbers of processors, can models accurately predict grounding line retreat, etc.
-The fallible human learning to use the model is largely absent from the discussion.
-For graduate students or other researchers who are not experts in computational physics, the difficulty of learning to use a particular modeling tool may be more of a rate-limiting factor than the speed or efficiency of the model.
+The human learning to use an unfamiliar software tool is largely absent from the discussion.
+For graduate students or other researchers who are not experts in computational physics, the difficulty of learning to use a particular software package may be more of a rate-limiting factor than the speed or efficiency of that software.
 
@@
 Users already familiar with, say, the \texttt{IceStream} class can then use the \texttt{HeatTransport} class under the assumption that the input fields -- the current temperature $T$, ice velocity $u$, and basal melt rate $m$ -- are passed as keyword arguments with the same name as the fields themselves.
-This obviates the need for consulting the documentation or examples.
+Consistency obviates the need for repeatedly consulting the documentation or examples once users are already familiar with the software.
 
 \textbf{Progressive evaluation: How easily can users get feedback during their use of the software?}
-Progressive evaluation is the main advantage of having a user interface in Python or another interpreted language, as opposed to using compiled programs that must run all at once.
-Icepack includes a suite of example programs that demonstrate all of the main functions of the package.
-These demos are formatted as Jupyter notebooks, a document format that includes code, figures, and explanatory text with typeset mathematics that runs interactively in a web browser \citep{kluyver2016jupyter}.
-A notebook is a sequence of cells, each of which consists of either code or text.
-The code cells can be executed one-at-a-time and this enables users to insert diagnostics in each cell as they go.
-Most importantly, the notebook format also integrates the plotting library matplotlib so that figures created in each cell appear directly beneath that cell as it is executed.
-The ability to visualize the current state of the simulation at every step is a powerful tool for debugging and sanity-checking.
-\textcolor{red}{Picture of an executed jupyter notebook.}
-The interface for icepack also supports inspection of partial results in other ways.
-Running a simulation amounts to writing a loop that alternates prognostic and diagnostic solves.
-Users can add arbitrary code to this loop for sanity checking or feedback on progress, or they can manually step through the simulation one iteration at a time.
-Other packages only support a more coarse-grained view.
-The user specifies the starting and ending time of a simulation, and the degree of feedback is hard-coded into the package itself.
+Progressive evaluation is the main advantage of having a user interface in an interpreted programming language such as Python as opposed to a compiled language where programs run in batch mode.
+All non-trivial simulations of physical systems begin life broken.
+The ability to manually step through a simulation and examine the entire state in an interpreter is critical to finding errors faster.
+Icepack was designed to give fine-grained control over how simulations work in order to support this mode of debugging.
+The API provides routines for solving the diagnostic and prognostic equations.
+It is the user's responsibility for making the repeated calls to these routines, either in a loop or by manually iterating through one step at a time.
+With this responsibility comes the freedom to add in arbitrary code.
+This capability might be used to add sanity checks, such as printing minimum and maximum thickness or velocity values.
+It can also be used to get feedback on how long the simulation will take, or to save results for later visualization.
+Other packages support a more coarse-grained view where the user only specifies the start and end time of a simulation and has more limited options for inspecting the state of the running program.
+
+Icepack makes extensive use of Jupyter notebooks as a form of executable documentation.
+Jupyter notebooks are a document format that includes code and explanatory text with typeset mathematics that runs interactively in a web browser \citep{kluyver2016jupyter}.
+The contents of a notebook can executed incrementally much like running code in the Python interpreter.
+Most importantly, Jupyter notebooks will render and display visualizations on the fly.
+This enables a workflow where plotting all intermediate results serves for sanity-checking an experimental simulation.
 
@@
 A user interested in achieving greater runtime performance can pass additional keyword arguments instead specifying, say, the preconditioned conjugate gradient method.
-This choice is of interest mostly to advanced users so we keep the linear solver method as a default argument.
+This choice is of interest mostly to advanced users so we keep the linear solver algorithm as a default argument.
 In so doing, we avoid confronting novice users with options that they might not understand.
 
-Advanced users who wish to tune solver performance for large simulations will need some way to make choices about algorithms.
-For example, a more advanced user might choose the GMRES iterative solver together with an incomplete LU preconditioner to solve linear systems.
-The solver classes, as opposed to the model classes, provide the interface for making these kinds of algorithmic choices.
+Advanced users who do wish to tune solver performance for large simulations will need some way to make choices about algorithms.
+For example, one might choose the GMRES iterative solver together with an incomplete LU preconditioner to solve linear systems.
+The solver classes, as opposed to the model classes, provide the interface for making these choices.
 While alternative solvers might offer faster runtime performance than direct factorization, they also requires making additional choices -- how often to restart GMRES?
@@
 Many glaciologists do not have the background in numerical linear algebra to know that adjusting these parameters could make the difference between solver convergence or breakdown.
-Similarly, users might want to select between different discretization strategies for the Stokes equations, which must be chosen carefully in order to satisfy the Ladyzhenskaya-Babu\v{s}ka-Brezzi (LBB) conditions \citep{boffi2013mixed}.
-When using Galerkin least-squares \textcolor{red}{check that's actually what it is} stabilization of the weak form, the user has to pick a value of the stabilization parameter.
+As another example, users might want to select between different discretization strategies for the Stokes equations.
+The discretization of the Stokes equations has to be chosen carefully in order to satisfy the Ladyzhenskaya-Babu\v{s}ka-Brezzi (LBB) conditions \citep{boffi2013mixed}.
+When using Galerkin least-squares stabilization of the weak form, the user has to pick a value of the stabilization parameter.
 Determining exactly what value of this parameter is necessary to guarantee stability is a subtle problem, even more so for the kinds of highly anisotropic meshes that are commonly encountered in 3D glacier flow modeling.

### 2020-11-25 icepack.tex
Add ice sheet figure

@@
 
-As a second test case for icepack, we ran an experiment inspired by \cite{kessler2008fjord}.
-In the original work, the authors couple an ice-flow model with simple models for bed-erosion, calving, and \textcolor{red}{glacial isostatic adjustment (GIA)} to investigate the formation of deep fjords that are characteristic of coastlines on Baffin Island, Greenland, and British Columbia, among others.
-Their first experiment is set in an idealized circular domain that contains a bedrock plateau rimmed by a ridge that is studded with four valleys.
-They find that fjords become well-developed after about one million years and that ice velocities are highest in these fjords.
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/ice-sheet/ice-sheet.png}
+    \caption{Synthetic ice sheet simulation: (a) the bed elevation profile consists of a ring of mountains with valleys of varying depth surrounding a flat plateau where the ice is initialized to (b) the change in thickness over the simulation period}
+    \label{fig:ice-sheet}
+\end{figure}
 
-Similar to \cite{kessler2008fjord}, we use the shallow ice model; however, we do not couple it to bed-erosion, calving, or GIA.
-We initialized the experiment with a simple but unrealistic ice thickness \textcolor{red}{(figure X)}.
-We then evolve the ice sheet without climate forcing for 500 years until it is near equilibrium.
-The time it takes to reach equilibrium is much shorter than the \cite{kessler2008fjord} experiment because the bedrock is static.
-The first part of our simulation is an exponential relaxation towards equilibrium.
-A velocity pattern emerges that is similar to results from \cite{kessler2008fjord}, with the highest velocities where ice is constricted through the deepest valleys \textcolor{red}{(figure X)}.
-In and upstream of the valleys, the surface is drawn down due to the elevated export of ice \textcolor{red}{as shown in figure X}.
+As a demonstration of the shallow ice approximation model, we ran an experiment inspired by \cite{kessler2008fjord}.
+In this work, the authors coupled an ice flow model to simple models for bed-erosion, calving, and glacial isostatic adjustment (GIA) to investigate the formation of deep fjords that are characteristic of coastlines on Baffin Island, Greenland, and British Columbia, among others.
+We emulated their domain and bedrock geometry -- a plateau surrounded by a ridge punctuated by four valleys -- without also simulating erosion or GIA.
+These effects will be the subject of future work.
 
-The ice-flow model chosen for this simulation is computationally cheap compared to the other ice-flow models, such that it can be used to simulate an ice sheet over thousands of years.
-By simply switching between ice-flow models, icepack is able to simulate many different glaciological conditions over a wide range of spatiotemporal scales.
-For example, icepack can be used to simulate a small domain dominated by an ice stream, an ice catchment that includes slow-moving interior ice and a floating ice shelf, or, as in this example, an entire ice sheet.
+Figure \ref{fig:ice-sheet} shows the results of this computational experiment.
+We initialized the experiment with a simple but unrealistic ice thickness.
+We then evolved the ice sheet without climate forcing for 500 years.
+The ice sheet relaxes very rapidly in the first 200 years, but after this changes are much slower as ice must be funneled through one of the four narrow valleys.
+The resulting velocity pattern is similar to that of \citet{kessler2008fjord}, with the highest velocities where ice flows out from the valleys.
+In and upstream of the valleys, the surface is drawn down due to the elevated export of ice.
+
+The diagnostic model used in this demonstration is cheap enough that it can be used to simulate ice sheets over several millenia in a matter of minutes on a desktop.
+A key feature of icepack is the ability to choose between many different diagnostic models.
+Users are free to decide whichever model is most appropriate for the spatial and temporal scales of their problem.
 

### 2020-11-25 demos/ice-sheet/ice-sheet.py

@@ -1,0 +1,144 @@
+import os
+import argparse
+import tqdm
+import numpy as np
+import firedrake
+from firedrake import (
+    max_value,
+    min_value,
+    Constant,
+    interpolate,
+    sqrt,
+    inner,
+    exp
+)
+import icepack
+from icepack.constants import (
+    ice_density as ρ_I,
+    glen_flow_law as n,
+    gravity as g
+)
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--timestep', type=float)
+parser.add_argument('--num-steps', type=int)
+parser.add_argument('--output')
+args = parser.parse_args()
+
+# Create the mesh and function spaces
+mesh = firedrake.UnitDiskMesh(5)
+R = 250e3
+mesh.coordinates.dat.data[:] *= R
+
+Q = firedrake.FunctionSpace(mesh, family='CG', degree=2)
+V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=2)
+
+# Create the bed elevation
+x, y = firedrake.SpatialCoordinate(mesh)
+r = sqrt(x**2 + y**2)
+
+b_base = firedrake.Constant(400)
+b_max = firedrake.Constant(1400)
+
+# Radius of the plateau interior and ridge
+ro = 125e3
+Ro = firedrake.Constant(200e3)
+
+def tanh(z):
+    return (exp(z) - exp(-z)) / (exp(z) + exp(-z))
+
+def θ(z):
+    return (tanh(z) + 1) / 2
+
+def sech(z):
+    return 2 / (exp(z) + exp(-z))
+
+a = firedrake.Constant(50e3)
+ξ = (sqrt(x**2 + y**2) - ro) / a
+
+b_expr_plateau = b_base * (1 - θ(3 * ξ))
+
+ζ = (r - Ro) / Ro
+b_expr_ridge = (b_max - b_base) * sech(3 * ξ)
+
+ρ1 = firedrake.Constant(1/4)
+μ1 = 1 - ρ1 * θ(3 * (x - ro/4) / a) * sech(2 * y / a)
+
+ρ2 = firedrake.Constant(3/8)
+μ2 = 1 - ρ2 * θ(3 * (y - ro/4) / a) * sech(2 * x / a)
+
+ρ3 = firedrake.Constant(1/2)
[…]

### 2020-11-25 demos/ice-sheet/make_plots.py

@@ -1,0 +1,50 @@
+import os
+import argparse
+import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+import firedrake
+import icepack.plot
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--input')
+parser.add_argument('--output')
+args = parser.parse_args()
+
+mesh = firedrake.UnitDiskMesh(5)
+R = 250e3
+mesh.coordinates.dat.data[:] *= R
+
+Q = firedrake.FunctionSpace(mesh, family='CG', degree=2)
+V = firedrake.VectorFunctionSpace(mesh, family='CG', degree=2)
+
+b = firedrake.Function(Q)
+h0 = firedrake.Function(Q)
+h = firedrake.Function(Q)
+u = firedrake.Function(V)
+
+input_name = os.path.splitext(args.input)[0]
+with firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ) as chk:
+    chk.load(b, name='bed')
+    chk.load(h, name='thickness')
+    chk.load(h0, name='thickness-initial')
+    chk.load(u, name='velocity')
+
+δh = firedrake.interpolate(h - h0, Q)
+
+fig, axes = icepack.plot.subplots(ncols=2, sharex=True, sharey=True)
+
+axes[0].set_ylabel('distance (meters)')
+for ax in axes:
+    ax.get_xaxis().set_visible(False)
+
+axes[0].set_title('Bed elevation')
+colors = icepack.plot.tripcolor(b, cmap='magma', axes=axes[0])
+fig.colorbar(colors, ax=axes[0], orientation='horizontal', pad=0.02)
+
+axes[1].set_title('Thickness change')
+colors = icepack.plot.tripcolor(
+    δh, vmin=-200, vmax=200, axes=axes[1], cmap='twilight'
+)
+fig.colorbar(colors, ax=axes[1], orientation='horizontal', pad=0.02)
+
+fig.savefig(args.output, bbox_inches='tight')

### 2020-12-02 icepack.tex
Wrote discussion section

@@
 A key feature of icepack is the ability to choose between many different diagnostic models.
-Users are free to decide whichever model is most appropriate for the spatial and temporal scales of their problem.
+Users are free to decide what model works best for the spatial and temporal scales of their problem, how accurately they need to solve this problem to produce useful results, and what computational resources they are willing to devote.
 
@@
 
+We have presented a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
+This package advances the state of the art in this field by providing a platform for easily experimenting with the model physics.
+In this paper, we have presented three demonstrations of this feature:
+\begin{enumerate}
+    \item coupling a model of ice shelf flow to a phenomenological model of damage,
+    \item changing the sliding law in a simulation of a marine-terminating glacier, and
+    \item inferring the fluidity of a floating ice shelf in a way that guarantees positivity.
+\end{enumerate}
+The phyics both of how glaciers flow and of how they interact with their environment are not completely understood.
+Consequently, the ability to change components of the model is an essential feature for any tool aimed at researchers in this field.
+
+We have also paid special attention to how we can design this software package to be most usable for its intended audience.
+Usability is not often discussed in the literature on computational science.
+We believe that this is because of two difficulties: (1) quantifying the degree to which usability is a rate-limiting factor and (2) concretely assessing what features make software tools more or less usable.
+By contrast, measuring computational performance is much more feasible although still fraught with difficulties of its own.
+(This is not to say that performance optimization is easy by any means.)
+In working with several graduate students and postdoctoral researchers in glaciology, we have observed that usability is a substantial bottleneck for researchers at these career stages.
+For this reason, we have chosen to focus explicitly on usability by drawing on the research literature in HCI.
+We argue that the same design features that enhance usability for relative novices to the subject area will also enhance the productivity of expert users.
+
+This paper has presented several physics models currently implemented in icepack along with demonstrations.
+Future developments will include:
+\begin{enumerate}
+    \item an implementation of full Stokes flow
+    \item dynamic glacier termini to enable calving models
+    \item adaptivity through moving mesh methods
+\end{enumerate}
+New features will be guided by feedback from icepack users and from the glaciological community at large.
+By providing implementations of several glacier flow models from less to more complex and by enabling users to experiment with the physics, this tool both lowers the barrier to entry for novices to glacier flow modeling and provides a pathway for these users to progress towards ever more sophisticated and advanced simulations.
+
 \bibliographystyle{plainnat}

### 2020-12-02 icepack.tex
Fix up ice sheet plot

@@
 
-A sample of the source code implementing both of these approaches is shown in \textcolor{red}{Figure X}.
+The source code implementing both of these approaches is contained in the supplementary material.
 To change the sliding law, users only need to pass one function or the other to the model object at initialization.
@@
     \includegraphics[width=0.95\linewidth]{demos/ice-sheet/ice-sheet.png}
-    \caption{Synthetic ice sheet simulation: (a) the bed elevation profile consists of a ring of mountains with valleys of varying depth surrounding a flat plateau where the ice is initialized to (b) the change in thickness over the simulation period}
+    \caption{Synthetic ice sheet simulation. (a) The bed elevation profile consists of a ring of mountains with valleys of varying depth surrounding a flat plateau where the ice sheet is initialized.
+    The black square outlines the domain of subplot (b) which shows the velocity of the ice as it passes through the southernmost mountain pass, with contours of the bed elevation shown in greyscale.}
     \label{fig:ice-sheet}

### 2020-12-02 demos/ice-sheet/make_plots.py

@@ -31,20 +31,41 @@ with firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ) as chk:
 
 δh = firedrake.interpolate(h - h0, Q)
 
-fig, axes = icepack.plot.subplots(ncols=2, sharex=True, sharey=True)
+fig, axes = icepack.plot.subplots(ncols=2)
 
 axes[0].set_ylabel('distance (meters)')
 for ax in axes:
     ax.get_xaxis().set_visible(False)
 
+xmin, xmax = -50e3, +50e3
+ymin, ymax = -175e3, -75e3
+
 axes[0].set_title('Bed elevation')
 colors = icepack.plot.tripcolor(b, cmap='magma', axes=axes[0])
-fig.colorbar(colors, ax=axes[0], orientation='horizontal', pad=0.02)
-
-axes[1].set_title('Thickness change')
-colors = icepack.plot.tripcolor(
-    δh, vmin=-200, vmax=200, axes=axes[1], cmap='twilight'
+fig.colorbar(
+    colors, label='meters', ax=axes[0], orientation='horizontal', pad=0.02
 )
-fig.colorbar(colors, ax=axes[1], orientation='horizontal', pad=0.02)
+axes[0].plot(
+    [xmin, xmax, xmax, xmin, xmin],
+    [ymin, ymin, ymax, ymax, ymin],
+    color='k'
+)
+
+axes[1].set_title('Velocity')
+axes[1].set_xlim(xmin, xmax)
+axes[1].set_ylim(ymin, ymax)
+axes[1].yaxis.set_label_position('right')
+axes[1].yaxis.tick_right()
+firedrake.tricontour(b, 20, cmap='Greys', axes=axes[1])
+streamlines = firedrake.streamplot(
+    u, resolution=5e3, cmap='magma', seed=1, axes=axes[1]
+)
+fig.colorbar(
+    streamlines,
+    label='meters / year',
+    ax=axes[1],
+    orientation='horizontal',
+    pad=0.02
+)
 
 fig.savefig(args.output, bbox_inches='tight')

### 2020-12-03 icepack.tex
Coauthors
Corrections from Ian

@@
 \title{\emph{icepack}: a novel glacier flow modeling package}
-\author{Daniel Shapero}
+\author{Daniel Shapero, Jessica Badgeley, Andrew Hoffmann}
 \date{}
@@
 There are four \emph{diagnostic} or \emph{momentum transport} models implemented in icepack.
-See \citet{dukowicz2010consistent} for a thorough explanation of the various approximations and action principles velow.
+For each of the diagnostic models, we use a formulation of the physics based on a \emph{minimization} or \emph{action principle} \citep{dukowicz2010consistent}.
+Action principles are completely equivalent to the usual weak form of a partial differential equation but have certain numerical advantages that we will describe below.
 
@@
 \textbf{Progressive evaluation: How easily can users get feedback during their use of the software?}
-Progressive evaluation is the main advantage of having a user interface in an interpreted programming language such as Python as opposed to a compiled language where programs run in batch mode.
-All non-trivial simulations of physical systems begin life broken.
+Progressive evaluation is the main advantage of having a user interface in an interpreted programming language such as Python as opposed to a compiled language where programs can only run in batch mode.
+In the early stages of the development process, any non-trivial simulation of a physical system exists as a prototype which may be non-functional or even broken.
 The ability to manually step through a simulation and examine the entire state in an interpreter is critical to finding errors faster.
@@
 The contents of a notebook can executed incrementally much like running code in the Python interpreter.
-Most importantly, Jupyter notebooks will render and display visualizations on the fly.
+Most importantly, Jupyter notebooks can render and display visualizations on the fly.
 This enables a workflow where plotting all intermediate results serves for sanity-checking an experimental simulation.

### 2020-12-09 demos/mismip/mismip+.py
Use trust region in MISMIP+

@@ -51,8 +51,10 @@ f_c = 4e3
 d_c = 500
 w_c = 24e3
 
-B_y = d_c * (1 / (1 + exp(-2 * (y - Ly/2 - w_c) / f_c)) +
-             1 / (1 + exp(2 * (y - Ly / 2 + w_c) / f_c)))
+B_y = d_c * (
+    1 / (1 + exp(-2 * (y - Ly / 2 - w_c) / f_c)) +
+    1 / (1 + exp(+2 * (y - Ly / 2 + w_c) / f_c))
+)
 
 z_deep = -720
 z_b = interpolate(max_value(B_x + B_y, z_deep), Q)
@@ -72,7 +74,8 @@ from icepack.constants import (
 )
 
 def friction(**kwargs):
-    u, h, s, C = map(kwargs.get, ('velocity', 'thickness', 'surface', 'friction'))
+    keys = ('velocity', 'thickness', 'surface', 'friction')
+    u, h, s, C = map(kwargs.get, keys)
 
     p_W = ρ_W * g * max_value(0, -(s - h))
     p_I = ρ_I * g * h
@@ -82,11 +85,21 @@ def friction(**kwargs):
     u_c = (τ_c / C)**m
     u_b = sqrt(inner(u, u))
 
-    return τ_c * ((u_c**(1/m + 1) + u_b**(1/m + 1))**(m / (m + 1)) - u_c)
+    return τ_c * ((u_c**(1 / m + 1) + u_b**(1 / m + 1))**(m / (m + 1)) - u_c)
 
 # Create the model object and extra options
 model = icepack.models.IceStream(friction=friction)
-opts = {'dirichlet_ids': [1], 'side_wall_ids': [3, 4], 'tolerance': 1e-8}
+opts = {
+    'dirichlet_ids': [1],
+    'side_wall_ids': [3, 4],
+    'diagnostic_solver_type': 'petsc',
+    'diagnostic_solver_parameters': {
+        'snes_type': 'newtontr',
+        'ksp_type': 'preonly',
+        'pc_type': 'lu',
+        'pc_factor_mat_solver_type': 'mumps'
+    }
+}
 solver = icepack.solvers.FlowSolver(model, **opts)
 
 # Read the input data if there is any

### 2020-12-09 demos/mismip/mismip+.py
Output entire time series in mismip

@@ -117,6 +117,9 @@ if args.input:
 
     input_name = os.path.splitext(args.input)[0]
     with firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ) as chk:
+        timesteps, indices = chk.get_timesteps()
+        chk.set_timestep(timesteps[-1], idx=indices[-1])
+
         chk.load(h_in, name='h')
         chk.load(u_in, name='u')
 
@@ -162,33 +165,38 @@ h_c0 = Constant(75.0)   # cut-off water column thickness
 δt = args.timestep
 num_steps = int(args.time / δt)
 
-trange = tqdm.trange(num_steps)
-for step in trange:
-    z_d = s - h             # elevation of the ice base
-    h_c = z_d - z_b         # water column thickness
-    melt = Ω * tanh(h_c / h_c0) * max_value(z_0 - z_d, 0)
-    a.interpolate(accumulation - melt)
-
-    h = solver.prognostic_solve(
-        δt, thickness=h, velocity=u, accumulation=a, thickness_inflow=h0
-    )
-    s = icepack.compute_surface(thickness=h, bed=z_b)
-
-    u = solver.diagnostic_solve(
-        velocity=u,
-        thickness=h,
-        surface=s,
-        fluidity=A,
-        friction=C
-    )
-
-    min_thickness = firedrake.COMM_WORLD.allreduce(h.dat.data_ro.min(), MPI.MIN)
-    avg_thickness = firedrake.assemble(h * dx) / area
-    msg = f"avg/min thickness: ({avg_thickness:4.2f}, {min_thickness:4.2f})"
-    trange.set_description(msg)
-
-# Write out the results
 output_name = os.path.splitext(args.output)[0]
 with firedrake.DumbCheckpoint(output_name, mode=firedrake.FILE_CREATE) as chk:
+    trange = tqdm.trange(num_steps)
+    for step in trange:
+        chk.set_timestep(step * δt)
+        chk.store(h, name='h')
+        chk.store(u, name='u')
+
+        z_d = s - h             # elevation of the ice base
+        h_c = z_d - z_b         # water column thickness
+        melt = Ω * tanh(h_c / h_c0) * max_value(z_0 - z_d, 0)
+        a.interpolate(accumulation - melt)
+
+        h = solver.prognostic_solve(
+            δt, thickness=h, velocity=u, accumulation=a, thickness_inflow=h0
+        )
[…]

### 2020-12-09 demos/mismip/plot.py

@@ -24,9 +24,12 @@ h = firedrake.Function(Q)
 u = firedrake.Function(V)
 
 input_name = os.path.splitext(args.input)[0]
-chk = firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ)
-chk.load(h, name='h')
-chk.load(u, name='u')
+with firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ) as chk:
+    timesteps, indices = chk.get_timesteps()
+    chk.set_timestep(timesteps[-1], idx=indices[-1])
+
+    chk.load(h, name='h')
+    chk.load(u, name='u')
 
 fig, axes = icepack.plot.subplots(
     nrows=2, sharex=True, sharey=True, figsize=(6.4, 2.8)

### 2020-12-09 icepack.tex
Add line numbers
Show ice volume from MISMIP

@@
 %\usepackage[titletoc,toc,title]{appendix}
-
-%\linenumbers
+\usepackage{lineno}
+\linenumbers
 %\doublespacing
@@
 
-Figure \ref{fig:mismip} shows the outcome of the MISMIP+ experiment performed using icepack.
+Figure \ref{fig:mismip-result} shows the thickness and ice speed after spinning up the MISMIP+ geometry and input data to steady state with the sliding law described above.
+The spin-up used used degree-1 basis functions for both thickness and velocity.
+To get a high-resolution estimate for the steady state, the spin-up started from a relatively coarse resolution to propagate out most of the transient signal.
+Then the mesh was successively refined and spun up again to propagate out the remaining high-wavenumber transients.
+This process was repeated three times.
+The net result is that most of the spin-up is done at relatively little computational cost.
+Figure \ref{fig:mismip-volumes} shows the total ice mass during the retreat and readvance phases of the experiment.
+The initial response to turning on melting is very rapid but then becomes roughly linear with time.
+When the high melt is turned off at the 100-year mark, the ice readvances, but the rate is much slower than the rate of decrease when melt was on.
+This asymmetric response is typical and expected from ice physics.
 
 \begin{figure}[h]
-    \includegraphics[width=0.95\linewidth]{demos/mismip/steady-state.png}
+    \includegraphics[width=0.95\linewidth]{demos/mismip/steady-state-result.png}
     \caption{The steady state thickness and velocity of the MISMIP+ experimental setup with the Weertman sliding law.}
-    \label{fig:mismip}
+    \label{fig:mismip-result}
+\end{figure}
+
+\begin{figure}[h]
+    \includegraphics[width=0.5\linewidth]{demos/mismip/volumes.png}
+    \caption{Total ice mass history during retreat (first 100 years) and readvance (second 100 years) of MISMIP+ experiment.}
+    \label{fig:mismip-volumes}
 \end{figure}

### 2020-12-09 demos/mismip/plot-volumes.py

@@ -1,0 +1,46 @@
+import os
+import argparse
+import numpy as np
+import matplotlib.pyplot as plt
+import firedrake
+from firedrake import dx
+import icepack.plot
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--output')
+parser.add_argument('--retreated')
+parser.add_argument('--readvanced')
+args = parser.parse_args()
+
+Lx, Ly = 640e3, 80e3
+ny = 20
+nx = int(Lx/Ly) * ny
+coarse_mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
+level = 3
+mesh_hierarchy = firedrake.MeshHierarchy(coarse_mesh, level)
+mesh = mesh_hierarchy[level]
+
+Q = firedrake.FunctionSpace(mesh, family='CG', degree=1)
+h = firedrake.Function(Q)
+
+def get_volumes(name):
+    with firedrake.DumbCheckpoint(name, mode=firedrake.FILE_READ) as chk:
+        timesteps, indices = chk.get_timesteps()
+        volumes = np.zeros_like(timesteps)
+
+        for timestep, index in zip(timesteps, indices.astype(int)):
+            chk.set_timestep(timestep, idx=index)
+            chk.load(h, name='h')
+            volumes[index] = firedrake.assemble(h * dx)
+
+    return volumes
+
+retreating_volumes = get_volumes(os.path.splitext(args.retreated)[0])
+readvancing_volumes = get_volumes(os.path.splitext(args.readvanced)[0])
+volumes = np.concatenate((retreating_volumes, readvancing_volumes[1:]))
+
+fig, axes = plt.subplots()
+axes.plot(1/24 * np.array(list(range(len(volumes)))), volumes / 1e9 / 917.)
+axes.set_xlabel('Simulation time (years)')
+axes.set_ylabel('Ice mass (gigatons)')
+fig.savefig(args.output, dpi=300, bbox_inches='tight')

### 2020-12-09 icepack.tex
Use GMD template

@@
-\documentclass{article}
-
-\usepackage{amsmath}
-%\usepackage{amsfonts}
-\usepackage{amsthm}
-%\usepackage{amssymb}
-%\usepackage{mathrsfs}
-%\usepackage{fullpage}
-%\usepackage{mathptmx}
-%\usepackage[varg]{txfonts}
-\usepackage{natbib}
-\usepackage{color}
-\usepackage[charter]{mathdesign}
-\usepackage[pdftex]{graphicx}
+%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
+%% ---------------------------------
+%% This template should be used for copernicus.cls
+%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
+%% For further assistance please contact Copernicus Publications at: production@copernicus.org
+%% https://publications.copernicus.org/for_authors/manuscript_preparation.html
+
+%% Please use the following documentclass and journal abbreviations for preprints and final revised papers.
+
+%% 2-column papers and preprints
+\documentclass[journal abbreviation, manuscript]{copernicus}
+
+%% Journal abbreviations (please use the same for preprints and final revised papers)
+
+% Advances in Geosciences (adgeo)
+% Advances in Radio Science (ars)
+% Advances in Science and Research (asr)
+% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
+% Annales Geophysicae (angeo)
+% Archives Animal Breeding (aab)
+% ASTRA Proceedings (ap)
+% Atmospheric Chemistry and Physics (acp)
+% Atmospheric Measurement Techniques (amt)
+% Biogeosciences (bg)
+% Climate of the Past (cp)
+% DEUQUA Special Publications (deuquasp)
+% Drinking Water Engineering and Science (dwes)
+% Earth Surface Dynamics (esurf)
+% Earth System Dynamics (esd)
+% Earth System Science Data (essd)
+% E&G Quaternary Science Journal (egqsj)
+% European Journal of Mineralogy (ejm)
+% Fossil Record (fr)
+% Geochronology (gchron)
+% Geographica Helvetica (gh)
+% Geoscience Communication (gc)
+% Geoscientific Instrumentation, Methods and Data Systems (gi)
+% Geoscientific Model Development (gmd)
+% History of Geo- and Space Sciences (hgss)
+% Hydrology and Earth System Sciences (hess)
+% Journal of Bone and Joint Infection (jbji)
+% Journal of Micropalaeontology (jm)
+% Journal of Sensors and Sensor Systems (jsss)
+% Magnetic Resonance (mr)
+% Mechanical Sciences (ms)
+% Natural Hazards and Earth System Sciences (nhess)
+% Nonlinear Processes in Geophysics (npg)
+% Ocean Science (os)
+% Primate Biology (pb)
+% Proceedings of the International Association of Hydrological Sciences (piahs)
+% Scientific Drilling (sd)
+% SOIL (soil)
+% Solid Earth (se)
+% The Cryosphere (tc)
+% Weather and Climate Dynamics (wcd)
+% Web Ecology (we)
+% Wind Energy Science (wes)
+
+%% \usepackage commands included in the copernicus.cls:
+%\usepackage[german, english]{babel}
+%\usepackage{tabularx}
+%\usepackage{cancel}
+%\usepackage{multirow}
+%\usepackage{supertabular}
+%\usepackage{algorithmic}
+%\usepackage{algorithm}
+%\usepackage{amsthm}
 %\usepackage{float}
-%\usepackage{hyperref}
-%\usepackage[modulo, displaymath, mathlines]{lineno}
-%\usepackage{setspace}
-%\usepackage[titletoc,toc,title]{appendix}
-\usepackage{lineno}
-\linenumbers
-%\doublespacing
-
-\theoremstyle{definition}
-\newtheorem*{defn}{Definition}
-\newtheorem*{exm}{Example}
-
-\theoremstyle{plain}
-\newtheorem*{thm}{Theorem}
-\newtheorem*{lem}{Lemma}
-\newtheorem*{prop}{Proposition}
-\newtheorem*{cor}{Corollary}
-
-\newcommand{\argmin}{\text{argmin}}
-\newcommand{\ud}{\hspace{2pt}\mathrm{d}}
-\newcommand{\bs}{\boldsymbol}
-\newcommand{\PP}{\mathsf{P}}
-
-\title{\emph{icepack}: a novel glacier flow modeling package}
-\author{Daniel Shapero, Jessica Badgeley, Andrew Hoffmann}
-\date{}
+%\usepackage{subfig}
+%\usepackage{rotating}
 
@@
 
-\tableofcontents
-\newpage
+\title{\emph{icepack}: a new glacier flow modeling package in Python, version 1.0}
+
+% \Author[affil]{given_name}{surname}
+
+\Author[1]{Daniel}{Shapero}
+\Author[2]{Jessica}{Badgeley}
+\Author[2]{Andrew}{Hoffmann}
+
+\affil[1]{Polar Science Center, Applied Physics Laboratory, University of Washington, Seattle, WA, USA}
+\affil[2]{Department of Earth and Space Sciences, University of Washington, Seattle, WA, USA}
+
+%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
+
+%% If an author is deceased, please mark the respective author name(s) with a dagger, e.g. "\Author[2,$\dag$]{Anton}{Aman}", and add a further "\affil[$\dag$]{deceased, 1 July 2019}".
+
+%% If authors contributed equally, please mark the respective author names with an asterisk, e.g.
+"\Author[2,*]{Anton}{Aman}" and "\Author[3,*]{Bradley}{Bman}" and add a further affiliation: "\affil[*]{These authors contributed equally to this work.}".
+
+\correspondence{Daniel Shapero (shapero@uw.edu)}
+
+\runningtitle{icepack}
+
+\runningauthor{Shapero, Badgeley, Hoffmann}
+
+\received{}
+\pubdiscuss{} %% only important for two-stage journals
+\revised{}
+\accepted{}
+\published{}
+
+%% These dates will be inserted by Copernicus Publications during the typesetting process.
+
+\firstpage{1}
 
@@
 \begin{abstract}
-In this paper we introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
+We introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
 Icepack is built on the finite element modeling library Firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
@@
 The components of each action functional can be substituted for different forms of the user's choosing, which makes it easy to experiment with the model physics.
-Many post-processing and analysis tasks on simulation results also amount to the evaluation of some functional.
-By using a variational formulation of the model physics, the specification of a problem and the analysis of the solution employ the same conceptual vocabulary.
-A third advantage of variational principles is that the action functional itself can be used to define a solver convergence criterion that is independent of the mesh and requires little tuning on the part of the user.
-Icepack includes the 2D shallow ice and shallow stream models and a novel implementation of the 3D Blatter-Pattyn model.
+The action functional itself can be used to define a solver convergence criterion that is independent of the mesh and requires little tuning on the part of the user.
+Icepack includes the 2D shallow ice and shallow stream models.
+We have also defined a 3D hybrid model based on spectral semi-discretization of the Blatter-Pattyn equations.
 Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
@@
 
-\section{Introduction}
+% \copyrightstatement{TEXT} %% This section is optional and can be used for copyright transfers.
+
+\introduction  %% \introduction[modified heading if necessary]
 
@@
 First, icepack includes a simple 3D flow model that uses several features only available in Firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
-This model will be described in section \S\ref{sec:physics-hybrid-model}.
+This model will be described in section \ref{sec:physics-hybrid-model}.
 Second, the architecture of icepack is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of the models included.
@@
 \begin{align}
-    & \text{action} = \iint\text{stress} \times \text{strain rate}\ud z\ud x - \int\text{basal friction} \times \text{sliding velocity}\ud x \nonumber \\
-    & \quad - \iint\text{surface slope}\times\text{velocity}\ud z\ud x - \iint\text{ocean pressure}\times\text{velocity}\ud z\ud \gamma
+    & \text{action} = \iint\text{stress} \times \text{strain rate}\,dz\,dx - \int\text{basal friction} \times \text{sliding velocity}\,dx \nonumber \\
+    & \quad - \iint\text{surface slope}\times\text{velocity}\,dz\,dx - \iint\text{ocean pressure}\times\text{velocity}\,dz\,d\gamma
     \label{action-functional}
 \end{align}
-where $\ud z\ud x$ denotes integration over the entire glacier, $\ud x$ denotes integration over the glacier footprint, and $\ud z\ud\gamma$ over the side wall boundary.
+where $dz\,dx$ denotes integration over the entire glacier, $dx$ denotes integration over the glacier footprint, and $dz\,d\gamma$ over the side wall boundary.
 The action has units of power (energy/time) and can be related to the rate of decrease of the thermodynamic free energy.
@@
 \begin{align}
-    \text{mass} & =\frac{1}{2}\int_\Omega u\cdot u\ud x\\\
-    \text{gravity} & = \int_\Omega\frac{2A(\rho_I g)^n}{n+2} h^{n + 1}|\nabla s|^{n - 1}\nabla s\cdot u\ud x\\
-    \text{penalty} & = \frac{1}{2}\int_\Omega \ell^2\nabla u\cdot \nabla u\ud x
+    \text{mass} & =\frac{1}{2}\int_\Omega u\cdot u\,dx\\\
+    \text{gravity} & = \int_\Omega\frac{2A(\rho_I g)^n}{n+2} h^{n + 1}|\nabla s|^{n - 1}\nabla s\cdot u\,dx\\
+    \text{penalty} & = \frac{1}{2}\int_\Omega \ell^2\nabla u\cdot \nabla u\,dx
 \end{align}
@@
 \begin{align}
-    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}\ud x \\
-    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}\ud x \\
-    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\ud x \\
-    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\ud\gamma
+    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}dx \\
+    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}dx \\
+    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\,dx \\
+    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\,d\gamma
 \end{align}
@@
 
-\section{Data assimilation} \label{sec:data-assimilation}
+\section{Data assimilation}
 
@@
 \begin{equation}
-    \ud^2J(u_k)\cdot v_k = -\ud J(u_k).
+    d^2J(u_k)\cdot v_k = -dJ(u_k).
     \label{eq:newton-search-direction}
@@
 
-For a convex problem, $\ud^2J$ is a symmetric and positive-definite matrix.
+For a convex problem, $d^2J$ is a symmetric and positive-definite matrix.
 This has two advantages.
@@
 A convergence criterion that works equally well independent of the mesh, finite element discretization, and the quality of the initial guess can be defined based on the idea of the \emph{Newton decrement} \citep{nocedal2006numerical}.
-Since the second derivative operator $\ud^2J(u_k)$ is positive-definite, the Newton search direction $v_k$ computed from equation \eqref{eq:newton-search-direction} is a descent direction for $J$:
-\begin{equation}
-    \ud J(u_k)\cdot v_k < 0.
+Since the second derivative operator $d^2J(u_k)$ is positive-definite, the Newton search direction $v_k$ computed from equation \eqref{eq:newton-search-direction} is a descent direction for $J$:
+\begin{equation}
+    dJ(u_k)\cdot v_k < 0.
 \end{equation}
@@
 \begin{equation}
-    J(u_k) - J(u) \approx \frac{1}{2}|\ud J(u_k)\cdot v_k|.
+    J(u_k) - J(u) \approx \frac{1}{2}|dJ(u_k)\cdot v_k|.
 \end{equation}
@@
 \begin{equation}
-    |\ud J(u_k)\cdot v_k| < \epsilon K(u_k)
+    |dJ(u_k)\cdot v_k| < \epsilon K(u_k)
 \end{equation}
@@
 \begin{equation}
-    \phi_k = -M^{-1}\ud J(\theta_k)
+    \phi_k = -M^{-1}dJ(\theta_k)
 \end{equation}
@@
 \begin{equation}
-    \text{backpressure power} = \int_\Gamma\int_0^1 \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+\ud\zeta\ud\gamma,
+    \text{backpressure power} = \int_\Gamma\int_0^1 \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+\,d\zeta\,d\gamma,
     \label{backpressure}
@@
 \begin{equation}
-    E(u - u^o) = \int_\Omega\left(\sqrt{\frac{|u - u^o|^2}{\sigma^2} + \gamma^2} - \gamma\right)\ud x.
+    E(u - u^o) = \int_\Omega\left(\sqrt{\frac{|u - u^o|^2}{\sigma^2} + \gamma^2} - \gamma\right)dx.
     \label{eq:l1-error-metric}
@@
 
-\section{Discussion}
+\conclusions  %% \conclusions[modified heading if necessary]
 
@@
 
-\bibliographystyle{plainnat}
+%% The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
+%% It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
+
+\codedataavailability{
+All code used in this repository is free and open source and all data sets used in the demonstrations are publicly available.
+The icepack source repository is at https://github.com/icepack/icepack and the git commit hash of the version of the code used for this publication is 250490a528c4abe473825bebbc823b2de0c43000.
+The demonstrations used glacier outlines that were hand-digitized from satellite imagery in order to generate the model domains.
+These outlines are kept in version control and hosted at https://github.com/icepack/glacier-meshes and the git commit hash for the version used in this publication is c98a8b7536b1891611566257d944e5ea024f2cdf.
+Additional observational data are hosted at the US National Snow and Ice Data Center (https://www.nsidc.org).
+}
+
+\appendix
+\section{}    %% Appendix A
+
+\subsection{}     %% Appendix A1, A2, etc.
+
+\noappendix       %% use this to mark the end of the appendix section. Otherwise the figures might be numbered incorrectly (e.g. 10 instead of 1).
+
+%% Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
+
+%% Option 1: If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
+%% They will be correctly named automatically.
+
+%% Option 2: If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
+%% To rename them correctly to A1, A2, etc., please add the following commands in front of them:
+
+\appendixfigures  %% needs to be added in front of appendix figures
+
+\appendixtables   %% needs to be added in front of appendix tables
+
+%% Please add \clearpage between each table and/or figure. Further guidelines on figures and tables can be found below.
+
+\authorcontribution{
+DS designed and implemented icepack with contributions from JB and AH.
+All authors contributed to the demonstration codes and to writing this paper.
+} %% this section is mandatory
+
+\competinginterests{
+All authors declare that they have no competing interests.
+} %% this section is mandatory even if you declare that no competing interests are present
+
+\begin{acknowledgements}
+We would like thank Ian Joughin for testing the model and for many valuable discussions.
+\end{acknowledgements}
+
+%% REFERENCES
+
+%% The reference list is compiled as follows:
+
+%\begin{thebibliography}{}
+
+%\bibitem[AUTHOR(YEAR)]{LABEL1}
+%REFERENCE 1
+
+%\bibitem[AUTHOR(YEAR)]{LABEL2}
+%REFERENCE 2
+
+%\end{thebibliography}
+
+%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
+%% authors experienced with BibTeX only have to include the following two lines:
+%%
+\bibliographystyle{copernicus}
 \bibliography{icepack.bib}
+%%
+%% URLs and DOIs can be entered in your BibTeX file as:
+%%
+%% URL = {http://www.xyz.org/~jones/idx_g.htm}
+%% DOI = {10.5194/xyz}
+
+%% LITERATURE CITATIONS
+%%
+%% command                        & example result
+%% \citet{jones90}|               & Jones et al. (1990)
+%% \citep{jones90}|               & (Jones et al., 1990)
+%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
+%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
+%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
+%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
+%% \citeauthor{jones90}|          & Jones et al.
+%% \citeyear{jones90}|            & 1990
+
+%% FIGURES
+
+%% When figures and tables are placed at the end of the MS (article in one-column style), please add \clearpage
+%% between bibliography and first table and/or figure as well as between each table and/or figure.
+
+% The figure files should be labelled correctly with Arabic numerals (e.g. fig01.jpg, fig02.png).
+
+%% ONE-COLUMN FIGURES
+
+%%f
+%\begin{figure}[t]
+%\includegraphics[width=8.3cm]{FILE NAME}
+%\caption{TEXT}
+%\end{figure}
+%
+%%% TWO-COLUMN FIGURES
+%
+%%f
+%\begin{figure*}[t]
+%\includegraphics[width=12cm]{FILE NAME}
+%\caption{TEXT}
+%\end{figure*}
+%
+%
+%%% TABLES
+%%%
+%%% The different columns must be seperated with a & command and should
+%%% end with \\ to identify the column brake.
+%
+%%% ONE-COLUMN TABLE
+%
+%%t
+%\begin{table}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table}
+%
+%%% TWO-COLUMN TABLE
+%
+%%t
+%\begin{table*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table*}
+%
+%%% LANDSCAPE TABLE
+%
+%%t
+%\begin{sidewaystable*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{sidewaystable*}
+%
+%
+%%% MATHEMATICAL EXPRESSIONS
+%
+%%% All papers typeset by Copernicus Publications follow the math typesetting regulations
+%%% given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry,
+%%% 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
+%%%
+%%% Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
+%%% Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
+%%% Items/objects which are defined are typeset in roman font (Car A, Car B)
+%%% Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
+%%% Abbreviations from 2 letters are typeset in roman font (RH, LAI)
+%%% Vectors are identified in bold italic font using \vec{x}
+%%% Matrices are identified in bold roman font
+%%% Multiplication signs are typeset using the LaTeX commands \times (for vector products, grids, and exponential notations) or \cdot
+%%% The character * should not be applied as mutliplication sign
+%
+%
+%%% EQUATIONS
+%
+%%% Single-row equation
+%
+%\begin{equation}
+%
+%\end{equation}
+%
+%%% Multiline equation
+%
+%\begin{align}
+%& 3 + 5 = 8\\
+%& 3 + 5 = 8\\
+%& 3 + 5 = 8
+%\end{align}
+%
+%
+%%% MATRICES
+%
+%\begin{matrix}
+%x & y & z\\
+%x & y & z\\
+%x & y & z\\
+%\end{matrix}
+%
+%
+%%% ALGORITHM
+%
+%\begin{algorithm}
+%\caption{...}
+%\label{a1}
+%\begin{algorithmic}
+%...
+%\end{algorithmic}
+%\end{algorithm}
+%
+%
+%%% CHEMICAL FORMULAS AND REACTIONS
+%
+%%% For formulas embedded in the text, please use \chem{}
+%
+%%% The reaction environment creates labels including the letter R, i.e. (R1), (R2), etc.
+%
+%\begin{reaction}
+%%% \rightarrow should be used for normal (one-way) chemical reactions
+%%% \rightleftharpoons should be used for equilibria
+%%% \leftrightarrow should be used for resonance structures
+%\end{reaction}
+%
+%
+%%% PHYSICAL UNITS
+%%%
+%%% Please use \unit{} and apply the exponential notation
 

### 2020-12-11 icepack.tex
Add funding info
More grants acknowledgment
Add corrections from Ian

@@
 \Author[2]{Andrew}{Hoffmann}
+\Author[1]{Ian}{Joughin}
 
@@
 
-\runningauthor{Shapero, Badgeley, Hoffmann}
+\runningauthor{Shapero, Badgeley, Hoffmann, Joughin}
 
@@
 We have also defined a 3D hybrid model based on spectral semi-discretization of the Blatter-Pattyn equations.
-Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
-The overall design philosophy of icepack is to be as usable as possible for a wide a swathe of the glaciological community, including both experts and novices in computational science.
+Finally, icepack includes a Gauss-Newton solver for inverse problems that runs substantially faster than the BFGS method often used in the glaciological literature.
+The overall design philosophy of icepack is to be as usable as possible for a wide a swath of the glaciological community, including both experts and novices in computational science.
 \end{abstract}
@@
 
-Numerical modeling has become an essential part of the workflow of glaciologists across all disciplines.
+Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
+Several software packages for glacier flow modeling already exist and are effective in the hands of experts.
 We highlight four main uses of glacier models in the literature:
 \begin{enumerate}
-    \item predicting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
+    \item projecting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
     \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood,
@@
 \end{enumerate}
-Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
-Several glacier flow models already exist and are effective in the hands of experts.
-These models are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
-Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
+To accomplish these tasks, these tools are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
+Many glaciologists receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
 The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
-Our goal is to make a tool that will be both accessible to non-experts and useful for experts.
+In this paper, we introduce a new Python software package for glacier flow modeling called \emph{iceapck}.
+Our goal is to make a tool that will be both accessible to novices and more productive for experts.
 
 The glacier flow modeling package closest in spirit to icepack is VarGlaS \citep{brinkerhoff2013data}.
-VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}, whereas icepack is built on top of Firedrake.
-The Firedrake project began as an outgrowth of FEniCS.
-Both packages share a similar goal of saving users from manually writing low-level code for assembling the systems of equations that discretize their physics model.
+VarGlaS is implemented using the finite element modeling (FEM) package FEniCS \citep{logg2012automated}, whereas icepack is built on top of Firedrake, which began as an outgrowth of FEniCS.
+Both packages share a similar goal of saving users from manually writing low-level code for assembling the systems of equations that discretize their model physics.
 Instead, users describe the weak form of the partial differential equations they wish to solve using a high-level domain-specific language (DSL) called the \emph{Unified Form Language} or UFL \citep{alnaes2014unified}.
@@
 Both FEniCS and Firedrake then generate optimized C or C++ code to assemble the discretized system of equations from this symbolic description of the problem \citep{kirby2006compiler, rathgeber2016firedrake}.
-Combining a DSL and a code generator frees users from the very error-prone process of writing these assembly kernels themselves, and makes the syntax of the code align closer to the syntax of written mathematical expressions.
+Combining a DSL and a code generator frees users from the very error-prone process of writing these assembly kernels themselves, and makes the syntax of the code align more closely to the syntax of written mathematical expressions.
 Additionally, having a high-level symbolic description of the problem makes it possible to automatically derive tangent linear models and adjoints \citep{mitusch2019dolfin}.
@@
 Icepack improves upon the groundwork laid in VarGlaS in three main respects.
-First, icepack includes a simple 3D flow model that uses several features only available in Firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
-This model will be described in section \ref{sec:physics-hybrid-model}.
-Second, the architecture of icepack is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of the models included.
-The software design that made this possible is discussed in section \S\ref{sec:physics-substitution}.
-Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.
+First, icepack includes a simple 3D flow model that uses several features only available in Firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated} (see \S\ref{sec:physics-hybrid-model}).
+Second, icepack's architecture is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of its constituent models (see \S\ref{sec:physics-substitution}).
+Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS (see section \S\ref{sec:numerics-inverse-solvers}).
+
+\section{Physics}
 
 The two main components of a glacier flow model are a \emph{diagnostic} and a \emph{prognostic} equation.
-The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation.
-The inputs to the diagnostic equation are the ice thickness, surface elevation, velocity at the inflow boundary, rheology, and bed friction coefficient.
-The output of the diagnostic equation is the ice velocity throughout the entire domain.
+The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation (PDE).
 The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
-The inputs to the prognostic equation are the current value of the ice thickness, velocity, and the surface and basal mass balance.
-The output is the ice thickness at the subsequent timestep.
 Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation.
@@
 For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
-Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock.
+Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock \citep{cuffey2010physics}.
 The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.
-
-\section{Physics}
 
@@
 where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
-The thickness is represented using continuous, piecewise polynomial basis functions in each cell of the mesh.
-The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
-The current prognostic model implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
-We have not yet implemented a solver that works for discontinuous basis functions, but this extension is completely feasible within our framework and may come in a future release of icepack.
-
-The surface elevation is calculated at each timestep of a simulation as
+Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
+We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.
+
+The surface elevation is calculated as
 \begin{equation}
@@
 For each of the diagnostic models, we use a formulation of the physics based on a \emph{minimization} or \emph{action principle} \citep{dukowicz2010consistent}.
-Action principles are completely equivalent to the usual weak form of a partial differential equation but have certain numerical advantages that we will describe below.
-
-The most complex and most physically accurate model is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
+Action principles are completely equivalent to the usual weak form of a partial differential equation but have certain numerical advantages, as described below.
+
+The most complex and most physically accurate model that we implement is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
 (In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
@@
 The SIA model \citep{hutter1981effect} describes the interior of ice sheets well, where flow can be simply described as bed-parallel shear and surface and basal slopes are small.
-This approximation breaks down near ice divides, where vertical velocities dominate, and near ice margins, where basal sliding can be a large fraction of the surface speed.
-Second, one could assume that ice flow is purely by horizontal extension and that the surface and bed velocities are practically the same.
-This is called the \emph{shallow stream approximation} (SSA).
+This approximation breaks down near ice margins, where basal sliding can be a large fraction of the surface speed and where membrane stresses are substantial.
+Second, one could assume that ice flow is purely by horizontal extension and that the surface and bed velocities are practically the same, which is called the \emph{shallow stream approximation} (SSA).
 The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.
-Finally, for floating ice, the shallow stream approximation simplifies further to the \emph{shallow shelf approximation}.
-The shallow shelf model has no bed friction and the ice surface elevation can be eliminated from the system.
 
@@
 The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.
+Strictly speaking, the fluidity should be a tensor field, but is almost universally treated as a scalar.
 
@@
 Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge from any initial guess.
-Minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation, although the two formulations are equivalent.
+Minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation while having no additional computational cost.
+Both formulations are mathematically equivalent.
 
@@
 
-The shallow ice model can be derived from the Blatter-Pattyn approximation by assuming the ice flow is dominated by bed-parallel shear and that surface and basal slopes are small, i.e.
+The shallow ice model can be derived from the Blatter-Pattyn approximation by assuming ice flow is dominated by bed-parallel shear and that surface and basal slopes are small, i.e.
 $\partial u/\partial x \ll \partial u/\partial z$.
@@
 \end{align}
-and the action fuctional is
+and the action functional is
 \begin{equation}
@@
 The shallow ice approximation applies well in ice-sheet interiors.
-For this reason, and because the equations are particularly simple to solve, this approximation has been a common choice for ice-sheet modeling \citep{kirchner2016shallow}.
-Caution must be taken, however, because areas of the ice sheet that experience high vertical velocities (e.g., ice divides) or plug flow (e.g., ice streams) are not well represented by this approximation.
+For this reason, and because the equations are particularly simple to solve, this approximation has been a common choice for ice-sheet modeling \citep{cuffey2010physics, kirchner2016shallow}.
+This approximation does not work well in areas of the ice sheet where the flow has substantial membrane stresses, like fast outlet glaciers.
 
@@
 The lack of vertical strain rates in the ice viscosity in 2D models can also lead to different equilibrium grounding line positions under the same external forcing.
-The shallow stream approximation is warranted for studying individual fast-flowing glaciers where the concern is more with studying processes and responses than with complete accuracy.
+The advantage of the SSA is that that it can capture most of the overall flow features of an individual fast-flowing glacier but at much lower computational cost than the Stokes equations.
 
@@
 The shallow ice and shallow stream models follow directly from the variational principles described above.
-The hybrid flow model, while also based on a variational principle, uses two more tricks: terrain-following coordinates and vertical spectral methods.
+The hybrid flow model, while also based on a variational principle, uses two more advanced mathematical techniques: terrain-following coordinates and spectral discretization in the vertical dimension.
 Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
-\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
+\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband models.
 \citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
@@
 
-\paragraph{Terrain-following coordinates}
-
-Rather than use the usual Cartesian coordinate system, the hybrid flow model uses terrain-following coordinates.
+\textbf{Terrain-following coordinates}.
+Rather than the usual Cartesian coordinate system, the hybrid flow model uses terrain-following coordinates.
 The terrain-following vertical coordinate $\zeta$ is
@@
 
-\paragraph{Spectral discretization}
-
+\textbf{Spectral discretization}.
 Terrain-following coordinates open up several choices for how to describe the vertical variation of the velocity field.
@@
 
-The combination of using extruded meshes and tensor product elements in the vertical direction is referred to as \emph{semi-discretization}.
-This approach can be thought of merely as a way to discretize a PDE that has special structure.
-Alternatively, we can view vertical semi-discretization as defining a family of models indexed by the number of vertical basis functions.
+The combination of using extruded meshes and tensor product elements in the vertical direction can be thought of merely as a way to discretize a PDE that has special structure.
+Alternatively, we can view discretization in the vertical as defining a family of models indexed by the number of vertical basis functions.
 The order-$d$ model defines a coupled system of PDEs for $d$ vector fields.
@@
 
-The user then has to decide how many vertical modes are enough.
+The user then has to decide how many vertical modes are sufficient.
 Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
@@
 Many aspects of glacier physics are not completely understood.
-For example, the most common description of basal friction assumes that glacier sliding occurs through regelation, in which case the basal shear stress can be written
+For example, the most commonly used sliding law is the power law
 \begin{equation}
@@
 \end{equation}
-with $m = 3$ \citep{weertman1957sliding}.
-Many authors have argued that, in fast-flowing regions of the ice sheet, glacier sliding occurs instead by plastic deformation within the subglacial sediments \citep{tulaczyk2000basal}.
-For plastic sliding, the shear stress is dependent on the yield strength of the subglacial sediments and not on the sliding speed, in which case $m = \infty$ would be most appropriate.
-Finally, the Schoof sliding law aims to reconcile the two modes of sliding \citep{schoof2005effect}.
-Below a critical sliding speed, the shear stress is aymptotic to $|u|^{1/m}$ as in the Weertman sliding law, but above the critical speed the shear stress is independent of the sliding speed.
+for some exponent $m$.
+Older research assumed $m = 3$ based on the theory of regelation \citep{weertman1957sliding}, which has since been referred to as the Weertman sliding law.
+When $m = \infty$, the basal shear stress is independing of the sliding speed; this is referred to as perfect plasticity.
+More recently, \citet{schoof2005effect} proposed the functional form
+\begin{equation}
+    \tau_b = -C\left(\frac{|u|}{|u| + u_c}\right)^{\frac{1}{m}}\frac{u}{|u|}
+\end{equation}
+where $u_c$ is some critical speed.
+This sliding law acts like the Weertman law when $|u| \ll u_c$ and like the plastic law when $|u| \gg u_c$.
+This latter equation has been found to agree best with laboratory experiments \citep{zoet2020slip}.
 
@@
 One of our goals with icepack is to facilitate experimentation with the model physics, even for novice users.
-Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would meet these needs.
+Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would appear meet these needs.
 To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
 For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
-In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
+In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress, which Python accomodates easily.
 For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function, which is undesirable.
@@
 For example, the gravitational driving power routine will pull out the velocity, thickness, and surface elevation.
-It will have access to other fields, for example the ice fluidity.
-These extra arguments are ignored.
+While the routine will not use them, it also has access to other fields, for example the ice fluidity.
 
@@
 The preference for argument passing by name is specific to this use case, however, and is not universal.
-
-One key limitation of our approach is that, while users can change what goes into one or more terms of the action functional, they cannot add or remove terms.
-This case would call for writing a new diagnostic model.
 
@@
 We have not implemented a model for the surface or englacial transport of meltwater, although it would be possible to include this process.
-Instead, we rely on an external scheme to act as a sink of enthalpy for meltwater fraction values above some user-defined critical value, which defaults to 1\%.
+Instead, we rely on an external scheme to act as a sink of enthalpy for meltwater fraction values above some user-defined critical value.
+A common choice for this critical value is 1\% \citep{aschwanden2012enthalpy}.
 
@@
     \item the observed field and the name of the argument to the diagnostic solver,
-    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver,
+    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver, and
     \item extra data passed to the diagnostic solver such as boundary conditions.
@@
 One could also just as easily use $C \propto \exp(\beta)$.
-The data assimilation routines in icepack can work the same for any parameterization of the physics because Firedrake provides a rich set of routines for symbolically calculating functional derivatives.
+The data assimilation routines in icepack can work the same way for any parameterization of the physics because Firedrake provides a rich set of routines for symbolically calculating functional derivatives.
 The inverse problem class only needs to know which functional needs to be differentiated with respect to which field.
@@
 The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the model object creates.
-The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
-Finally the flow solver has a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
-The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.
+The Newton solver itself is completely standard but the convergence criterion is not (see section \S\ref{sec:convergence-criteria}).
+Finally, the flow solver has a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
 
@@
 To solve the advection equation equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
-The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
-This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
+The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}, but this scheme introduces a tuneable stabilization parameter.
+In the interest of simplicity, we instead default to the implicit Euler scheme, which has no such parameters.
 The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.
 
-Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
+Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy (CFL) condition.
 The use of an unconditionally stable scheme guarantees that they will get an answer, rather than a runtime error, should they try to use a large timestep.
-The extra cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
+Users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
+The extra computational cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
 Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.
@@
 This has two advantages.
-First, the search direction is always a descent direction for $J$.
-For non-convex problems this is not always the case and some adaptation strategy must be employed in the vicinity of saddle points.
+First, the search direction is always a descent direction for $J$, which is not always the case for non-convex problems.
 Second, symmetry and positivity enable the use of specialized linear solvers, such as the Cholesky decomposition or the conjugate gradient algorithm, that are superior to their more general counterparts in many respects.
@@
 Some packages combat this problem by using a few iterations of the more robust but slower Picard method first \citep{gagliardini2013capabilities}.
-While this approach can be effective it requires tuning the number of Picard iterations and there is no guarantee that an adequate amount for one problem will work well on another problem.
-This issue rarely appears on realistic input data.
-When solving inverse problems, however, the intermediate guesses for the inferred field can be wildly unrealistic before converging.
+While this approach can be effective, it requires tuning the number of Picard iterations.
+There is no guarantee that an adequate amount for one problem will work well on another.
+This issue rarely appears on realistic input data, but when solving inverse problems, the intermediate guesses for the inferred field can be wildly unrealistic before converging.
 A forward model solver that is not sufficiently robust can crash in these extreme scenarios.
@@
 
-Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation, for example \citet{perego2012parallel}.
+Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation \citep{perego2012parallel}.
 Few have considered the problem of when to stop iterating.
@@
 
-A convergence criterion that works equally well independent of the mesh, finite element discretization, and the quality of the initial guess can be defined based on the idea of the \emph{Newton decrement} \citep{nocedal2006numerical}.
+We can devise a convergence criterion that works equally well, independent of the discretization and the quality of the initial guess, based on the \emph{Newton decrement} \citep{nocedal2006numerical}.
 Since the second derivative operator $d^2J(u_k)$ is positive-definite, the Newton search direction $v_k$ computed from equation \eqref{eq:newton-search-direction} is a descent direction for $J$:
@@
 \end{equation}
-We can then use the Newton decrement to decide when to stop the iteration.
+We can then use the Newton decrement to decide when to stop the iteration, as described below.
 
@@
 
-The derivative of the objective functional with respect to the unknown parameter is calculated using the symbolic differentiation features of Firedrake.
+The derivative of the objective functional with respect to the unknown parameter is calculated using the adjoint method and the symbolic differentiation features of Firedrake.
 The user does not need to provide any routines for the derivatives, only the symbolic form of the error metric and the regularization functional.
@@
 
-The hybrid flow uses several features that are only available in Firedrake to better exploit the special structure of the problem.
+The hybrid flow model uses several features that are available in Firedrake to better exploit the special structure of the problem.
 Implementing this model also required some mathematical sleight-of-hand related to the terminus boundary condition that has not appeared in the literature before.
@@
     \caption{The normalized ocean pressure ($p_w / \rho_Wg$) and Legendre polynomial approximations of several degrees (left), and the residuals of the approximation (right).
-    For this particular example, the waterline is at $\zeta = 1/3$.
+    For this particular example, the waterline is at $\zeta = 1/3$, which would be representative of a glacier grounded on a higher moraine.
     The moments of each of the residuals up to the approximation degree are all zero.}
@@
 The relative difference in basal shear stress between the two sliding laws is less than 10\% throughout the entire range.
+Although figure \ref{fig:sliding-laws} shows a comparison of both sliding laws with the same value of the critical speed, by using different values, the agreement between our sliding law can be brought into much closer agreement with the Schoof law.
 
@@
 \citet{asay2016experimental} also suggest using a sliding law that transitions sharply to exact plasticity above the critical speed.
-Expressing this sliding law in UFL requires a conditional in the velocity and is thus no longer differentiable.
-This lack of differentiability makes the forward solver crash.
+Expressing this sliding law in UFL requires a conditional in the velocity and is thus no longer differentiable, causing the forward solver to crash.
 The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.
@@
 We emulated their domain and bedrock geometry -- a plateau surrounded by a ridge punctuated by four valleys -- without also simulating erosion or GIA.
-These effects will be the subject of future work.
 
@@
 
-The diagnostic model used in this demonstration is cheap enough that it can be used to simulate ice sheets over several millenia in a matter of minutes on a desktop.
+The diagnostic model used in this demonstration is computationally cheap enough that it can be used to simulate ice sheets over several millenia in a matter of minutes on a desktop.
 A key feature of icepack is the ability to choose between many different diagnostic models.
@@
 In the second phase, damage transport is turned on and coupled to the ice velocity.
-An interesting feature to observed in the approximate equilibrium damage field is that the highest values occur between the streams and not within them.
+An interesting feature to observe in the approximate equilibrium damage field is that the highest values occur between the streams and not within them.
 Additionally, the spacing between the streams changes as a result of adding damage.
@@
 Marine ice is warmer than meteoric ice and usually includes brine pockets, which are discernible in radio echo sounding measurements as the absence of reflection from the ice shelf base \citep{holland2009marine}.
-This warmer and impurity-laden ice is more ductile and thus less prone to fracture than cold and brittle meteoric ice.
+This warmer and impurity-laden ice is more ductile and thus should be less prone to fracture than cold and brittle meteoric ice.
 Ocean models predict that marine ice forms under Larsen C as well \citep{holland2009marine}.
@@
 This dataset has nominal errors over the Larsen Ice Shelf on the order 1 -- 7 meters per year.
-We used the recently-released BedMachine map of the thickness of Antarctica \citep{morlighem2019deep}.
-This data product improves on existing thickness maps for the continent by using the mass conservation law, together with measurements of the velocity, surface elevation, and surface mass balance, to improve thickness estimates where few measurements are available.
+We used the recently-released BedMachine map of the thickness of Antarctica, which takes advantage of newly available remote sensing data \citep{morlighem2019deep}.
 
@@
 The rheology parameter of an ice shelf is strictly positive.
-The optimization algorithm, however, can explore unphysical regions of parameter space without some a priori constraint on what values are reasonable.
+The optimization algorithm, however, can explore unphysical regions of parameter space without some a priori constraints.
 Rather than try to solve an inequality-constrained problem, most studies in glaciology instead re-parameterize the problem in terms of some auxiliary field in such a way that the rheology is manifestly positive.
-In this case we will use the parameterization
+In this case we use the parameterization
 \begin{equation}
@@
 
-The final value of the model-data misfit \eqref{eq:l1-error-metric} matches the value we would expect if the velocity errors actually came from this probability distribution.
-In a separate computational experiment, we used the older bedmap2 ice thickness map instead of BedMachine \citep{fretwell2013bedmap2}.
+The final value of the model-data misfit from equation \eqref{eq:l1-error-metric} matches the value we would expect if the velocity errors actually came from this probability distribution.
+In a separate computational experiment, we used the older bedmap2 ice thickness \citep{fretwell2013bedmap2}.
 We were unable to achieve the same model-data misfit using bedmap2 at the same grid resolution with any regularization parameter.
@@
 One of the main goals for icepack is to create a tool that is accessible to researchers who might not be experts in scientific computing.
-Previous work on numerical modeling of glacier flow has focused largely on technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory, does it scale to large numbers of processors, can models accurately predict grounding line retreat, etc.
-The human learning to use an unfamiliar software tool is largely absent from the discussion.
+Previous work on numerical modeling of glacier flow has focused largely on the technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory; does it scale to large numbers of processors; can models accurately predict grounding line retreat, etc.
+The person learning to use an unfamiliar software tool is largely absent from the discussion.
 For graduate students or other researchers who are not experts in computational physics, the difficulty of learning to use a particular software package may be more of a rate-limiting factor than the speed or efficiency of that software.
 
-The field of \emph{human-computer interaction} (HCI) asks how we can design software that is easiest to learn and use effectively.
+The field of \emph{human-computer interaction} (HCI) asks how we can design software that is easier to learn and use effectively.
 In the following, we will describe some of the design choices in icepack and how they relate to what HCI researchers call the \emph{cognitive dimensions of notations}.
-\citet{green1996usability} introduced this concept to assess the usability of visual programming languages, but the criteria they laid out in this paper have been used to analyze software systems across many disciplines.
+\citet{green1996usability} introduced this concept to assess the usability of visual programming languages, but the criteria they laid out in their study have been used to analyze software systems across many disciplines.
 
@@
 \end{enumerate}
-The phyics both of how glaciers flow and of how they interact with their environment are not completely understood.
+The phyics of how glaciers interact with their environment are not completely understood.
 Consequently, the ability to change components of the model is an essential feature for any tool aimed at researchers in this field.
@@
 \begin{enumerate}
-    \item an implementation of full Stokes flow
-    \item dynamic glacier termini to enable calving models
-    \item adaptivity through moving mesh methods
+    \item an implementation of full Stokes flow,
+    \item dynamic glacier termini to enable calving models, and
+    \item adaptivity through moving mesh methods.
 \end{enumerate}
@@
 \appendix
-\section{}    %% Appendix A
-
-\subsection{}     %% Appendix A1, A2, etc.
+% \section{}    %% Appendix A
+
+% \subsection{}     %% Appendix A1, A2, etc.
 
@@
 DS designed and implemented icepack with contributions from JB and AH.
+IJ tested the model and assisted with design and debugging.
 All authors contributed to the demonstration codes and to writing this paper.
@@
 \end{acknowledgements}
+
+\textit{Financial support}.
+DS was supported by US National Science Foundation grant \#1835321 and National Aeronautics and Space Administration grant \#80NSSC20K0954.
+JB was supported by NSF grant \#1256082.
+AH was supported by NASA grant \#80NSSC20K1627.
+IJ was supported by NSF grants \#1835321 and \#1643285.
 

### 2020-12-11 icepack.tex
Cite Doug B more

@@
 \end{enumerate}
-To accomplish these tasks, these tools are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
+To accomplish these tasks, modeling tools are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
 Many glaciologists receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
 The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
-In this paper, we introduce a new Python software package for glacier flow modeling called \emph{iceapck}.
+In this paper, we introduce a new Python software package for glacier flow modeling called \emph{icepack}.
 Our goal is to make a tool that will be both accessible to novices and more productive for experts.
@@
 The hybrid flow model, while also based on a variational principle, uses two more advanced mathematical techniques: terrain-following coordinates and spectral discretization in the vertical dimension.
-Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
+
+Each of these techniques has appeared in the literature on glacier modeling before but rarely all in the same place for a 3D model.
 \citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband models.
@@
 \citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem, but this model used a finite difference discretization in the vertical direction.
+The model used in \citet{brinkerhoff2015dynamics} is the closest to the one we present below.
+This work used both terrain-following coordinates and a tensor product basis of Lagrange finite elements in the vertical and higher-degree polynomials in a single vertical layer.
+For the vertical basis functions, they used a plug flow mode and one shear mode.
 
@@
 Rather than divide the spatial domain into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
-This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element.
+This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element \citep{mcrae2016automated}.
 Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
@@
 The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.
+Our main reason for choosing to build icepack using the Firedrake package was because it natively supports tensor product elements.
 
@@
 Going to higher degree gives a more accurate approximation at the expense of greater computational effort.
+
+\citet{brinkerhoff2015dynamics} used a similar approach to the one described above, with a one vertical basis function for plug flow and one for shear flow.
+The shear flow basis function was chosen to be exact assuming the SIA balance with cold ice, together with a heuristic approximation for polythermal ice.
+By contrast, our approach allows for an arbitrary number of shear modes, although using a high-degree basis in the presence of a cold-temperate transition surface will introduce ringing artifacts.
+In principle, one could confront this problem by using two layers -- one cold and one temperate -- instead of a single layer.
+The extruded mesh capabilities in Firedrake provide the needed functionality for implementing such a model.
+We have not yet implemented a multilayer model but this will be the subject of future work.
 
@@
 
+The following demonstrations aim to show the capabilities of icepack on both synthetic and real problems.
+The key features of icepack that these demonstrations aim to highlight are the variety of different physics models implemented and the flexibility of the components of these physics models.
+
 \subsection{MISMIP+}

### 2020-12-11 icepack.tex
Add use case for 3D model

@@
 
-The source code implementing both of these approaches is contained in the supplementary material.
 To change the sliding law, users only need to pass one function or the other to the model object at initialization.
+(See \emph{Code and data availability} below for the source code.)
 Users do not need to implement a subclass that overrides some parent method.
-This approach would be idiomatic in C++, but it requires some facility with object-oriented programming that a non-expert might lack.
+This approach would be idiomatic in C++, but it requires knowledge of object-oriented programming that a non-expert might lack.
 
@@
 When the high melt is turned off at the 100-year mark, the ice readvances, but the rate is much slower than the rate of decrease when melt was on.
-This asymmetric response is typical and expected from ice physics.
+This asymmetric response is typical and expected from ice physics; the rates broadly agree with the reference results computed with BISICLES in the original experimental specification \citep{asay2016experimental}.
+
+We also ran the experimental setup to steady state using the hybrid flow model with vertical basis functions up to degree 2.
+This is the most minimal set of vertical basis functions that can resolve plug flow and the stress boundary conditions at the ice base and surface.
+The ratio of basal velocity to surface velocity is shown in figure \ref{fig:mismip-sliding-ratio}.
+The areas with the most significant vertical deformation are at the inflow boundary and where the troughs at the side walls are steepest.
+Otherwise, the sliding ratio is above 0.8 throughout almost the entire domain.
 
@@
     \includegraphics[width=0.95\linewidth]{demos/mismip/steady-state-result.png}
-    \caption{The steady state thickness and velocity of the MISMIP+ experimental setup with the Weertman sliding law.}
+    \caption{The steady state thickness and velocity of the MISMIP+ experimental setup with the modified Schoof sliding law of equation \eqref{eq:power-dissipation-modified-schoof-sliding}.}
     \label{fig:mismip-result}
@@
     \label{fig:mismip-volumes}
+\end{figure}
+
+\begin{figure}[h]
+    \includegraphics[width=0.95\linewidth]{demos/mismip/sliding-ratio.png}
+    \caption{Ratio of basal velocity to surface velocity in steady state of MISMIP+ scenario computed with hybrid model.}
+    \label{fig:mismip-sliding-ratio}
 \end{figure}

### 2020-12-11 demos/mismip/mismip+3d.py

@@ -1,0 +1,171 @@
+import os
+import argparse
+import numpy as np
+import matplotlib.pyplot as plt
+import tqdm
+from mpi4py import MPI
+import firedrake
+from firedrake import (
+    inner, dx, Constant, interpolate, as_vector, exp, sqrt, max_value
+)
+import icepack, icepack.plot
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--input')
+parser.add_argument('--output')
+parser.add_argument('--level', type=int, default=0)
+parser.add_argument('--time', type=float)
+parser.add_argument('--timestep', type=float)
+args = parser.parse_args()
+
+# Create the 2D mesh and function spaces
+Lx, Ly = 640e3, 80e3
+ny = 20
+nx = int(Lx/Ly) * ny
+area = Lx * Ly
+
+coarse_mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
+mesh_hierarchy = firedrake.MeshHierarchy(coarse_mesh, args.level)
+mesh2d = mesh_hierarchy[args.level]
+Q2d = firedrake.FunctionSpace(mesh2d, 'CG', 1)
+V2d = firedrake.VectorFunctionSpace(mesh2d, 'CG', 1)
+
+# Load in the results from the previous stage
+h2d = firedrake.Function(Q2d)
+u2d = firedrake.Function(V2d)
+input_name = os.path.splitext(args.input)[0]
+with firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ) as chk:
+    timesteps, indices = chk.get_timesteps()
+    chk.set_timestep(timesteps[-1], idx=indices[-1])
+
+    chk.load(h2d, name='h')
+    chk.load(u2d, name='u')
+
+# Create the 3D mesh and function spaces and lift the previous stage results
+# into 3D
+mesh = firedrake.ExtrudedMesh(mesh2d, layers=1)
+Q = firedrake.FunctionSpace(mesh, 'CG', 1, vfamily='R', vdegree=0)
+V_0 = firedrake.VectorFunctionSpace(
+    mesh, 'CG', 1, vfamily='R', vdegree=0, dim=2
+)
+V = firedrake.VectorFunctionSpace(
+    mesh, 'CG', 1, vfamily='GL', vdegree=2, dim=2
+)
+
+h0 = icepack.lift3d(h2d, Q)
+u0 = firedrake.interpolate(icepack.lift3d(u2d, V_0), V)
+
+# Create the bed topography
[…]

### 2020-12-11 demos/mismip/plot-3d.py

@@ -1,0 +1,60 @@
+import os
+import argparse
+import numpy as np
+import matplotlib.pyplot as plt
+import firedrake
+from firedrake import inner, sqrt
+import icepack.plot
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--input')
+parser.add_argument('--level', type=int)
+parser.add_argument('--output')
+args = parser.parse_args()
+
+Lx, Ly = 640e3, 80e3
+ny = 20
+nx = int(Lx/Ly) * ny
+coarse_mesh = firedrake.RectangleMesh(nx, ny, Lx, Ly)
+mesh_hierarchy = firedrake.MeshHierarchy(coarse_mesh, args.level)
+mesh2d = mesh_hierarchy[args.level]
+mesh = firedrake.ExtrudedMesh(mesh2d, 1)
+
+Q = firedrake.FunctionSpace(mesh, 'CG', 1, vfamily='R', vdegree=0)
+V = firedrake.VectorFunctionSpace(mesh, 'CG', 1, vfamily='GL', vdegree=2, dim=2)
+
+h = firedrake.Function(Q)
+u = firedrake.Function(V)
+
+input_name = os.path.splitext(args.input)[0]
+with firedrake.DumbCheckpoint(input_name, mode=firedrake.FILE_READ) as chk:
+    timesteps, indices = chk.get_timesteps()
+    chk.set_timestep(timesteps[-1], idx=indices[-1])
+
+    chk.load(h, name='h')
+    chk.load(u, name='u')
+
+x, y, ζ = firedrake.SpatialCoordinate(mesh)
+
+P_0 = 1
+P_1 = np.sqrt(3) * (2 * ζ - 1)
+P_2 = np.sqrt(5) * (6 * ζ**2 - 6 * ζ + 1)
+
+weight = P_0 - np.sqrt(3) * P_1 + np.sqrt(5) * P_2
+u_b = icepack.depth_average(u, weight=weight)
+
+weight = P_0 + np.sqrt(3) * P_1 + np.sqrt(5) * P_2
+u_s = icepack.depth_average(u, weight=weight)
+
+U_b = sqrt(inner(u_b, u_b))
+U_s = sqrt(inner(u_s, u_s))
+Q2d = firedrake.FunctionSpace(mesh2d, 'CG', 1)
+ratio = firedrake.project(U_b / U_s, Q2d)
+
+fig, axes = icepack.plot.subplots()
[…]

### 2020-12-12 icepack.tex
Final corrections

@@
 
+We verified the correctness of our implementation by check that numerical results converge at the expected rate to the analytical Beuler profile for a circular symmetric ice sheet \citep{greve2009dynamics}.
+
 The shallow ice approximation applies well in ice-sheet interiors.
@@
 
+We verified the correctness of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
+When there is basal friction, the shallow stream equations do not have a simple analytical solution.
+To get an exact solution, we chose the ice velocity and thickness a priori and used the computer algebra system SymPy \citep{sympy} to generate a friction coefficient that makes these fields an exact solution.
+We then checked that numerical results converge to this manufactured solution at the expected rate.
+
 Several studies have compared the shallow stream approximation to 3D models such as Blatter-Pattyn or full Stokes \citep{pattyn2013grounding}.
@@
 When $m = \infty$, the basal shear stress is independing of the sliding speed; this is referred to as perfect plasticity.
-More recently, \citet{schoof2005effect} proposed the functional form
+More recently, \citet{schoof2005effect} proposed an alternative form that acts like the Weertman law when $|u| \ll u_c$ and like the plastic law when $|u| \gg u_c$.
+A simplified form of this model can be expressed as
 \begin{equation}
@@
 where $u_c$ is some critical speed.
-This sliding law acts like the Weertman law when $|u| \ll u_c$ and like the plastic law when $|u| \gg u_c$.
-This latter equation has been found to agree best with laboratory experiments \citep{zoet2020slip}.
+This latter equation has been found to agree best with laboratory experiments on till \citep{zoet2020slip} and in reproducing observed velocity variations \citep{joughin2019regularized}.
 
@@
 One of our goals with icepack is to facilitate experimentation with the model physics, even for novice users.
-Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would appear meet these needs.
-To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
-For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
+Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would appear to meet these needs.
+To support use cases like implementing the Schoof sliding law, it must be possible not just to change the value of a single parameter but to completely alter the functional form of a given model component.
+For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model component.
 In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress, which Python accomodates easily.
@@
 \begin{enumerate}
-    \item The user passes all arguments to the diagnostic solve procedure by keyword, essentially in a dictionary mapping names to fields.
+    \item The user passes all arguments to the diagnostic solve procedure by keyword.
     \item The diagnostic solve procedure creates a symbolic representation of the action functional by summing up several terms, like the viscosity, basal friction, etc.
@@
 
+\subsection{Performance}
+
+Icepack largely inherits the performance capabilities of the Firedrake package, for which we refer to the benchmarks in \citet{rathgeber2016firedrake}.
+Firedrake is in turn built on PETSc, which includes a rich suite of nonlinear solvers and preconditioners that have been demonstrated to scale up to hundreds of processors \citep{balay2019petsc}.
+Firedrake also includes special features for defining sophisticated solvers and preconditioners that take advantage of problem-specific structure \citep{kirby2018solver}.
+Icepack exposes these tools through the interface of the flow solver objects, so users can select any solver and preconditioner from PETSc.
+
+We have mainly developed icepack for process-scale studies of individual glaciers or drainage basins.
+For the demonstrations presented below, nearly all simulations run in a matter of minutes to hours on a single core.
+Larger problems, such as continental-scale modeling, will require more sophisticated and possibly problem-specific approaches.
+For example, a rudimentary preconditioner for the hybrid model system could use the degree-0 model as the coarse space in a multigrid-type approach.
+These optimizations will be the subject of future work.
+
 \section{Demonstrations}
@@
 Using variational principles to express all constitutive laws is less flexible than specifying the sliding law directly and this is a distinct disadvantage.
-We were able to implement a sliding law that exhibits many of the characteristics of the Schoof sliding law.
+We were nonetheless able to implement a sliding law that exhibits the important characteristics of the Schoof sliding law.
 \citet{asay2016experimental} also suggest using a sliding law that transitions sharply to exact plasticity above the critical speed.
@@
 The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.
+Moreover, there are few physical reasons to prefer one form over another, as long as the asymptotic behavior is correct above and below the critical speed.
 

### 2020-12-12 icepack.tex
Add Zenodo release

@@
 All code used in this repository is free and open source and all data sets used in the demonstrations are publicly available.
-The icepack source repository is at https://github.com/icepack/icepack and the git commit hash of the version of the code used for this publication is 250490a528c4abe473825bebbc823b2de0c43000.
+The icepack source repository is at https://github.com/icepack/icepack.
+Icepack is released under the GPLv3 license.
+The git commit hash of the version of the code used for this publication is b78b0ee571590cf5af242101c159154c6872b7b9, see also the Zenodo release at https://doi.org/10.5281/zenodo.1205640.
+The icepack documentation and user manual is hosted at https://icepack.github.io.
+The source code for the demonstrations used in this paper is hosted at https://github.com/icepack/icepack-paper, commit hash 4db3dc281e83b4304751ded3b77f166942b0efc6.
 The demonstrations used glacier outlines that were hand-digitized from satellite imagery in order to generate the model domains.
@@
 \begin{acknowledgements}
-We would like thank Ian Joughin for testing the model and for many valuable discussions.
 \end{acknowledgements}

### 2020-12-23 icepack.tex
Acknowledgments
Citation for MOA

@@
     Lower values indicate more deformable ice.
-    Right: stream plot of ice velocity computed from this fluidity parameter.}
+    Right: stream plot of ice velocity computed from this fluidity parameter.
+    Background image is the MODIS Mosaic of Antarctica \citep{scambos2007modis, haran2014modis}, courtesy of NASA NSIDC DAAC.}
     \label{fig:larsen}
@@
 \begin{acknowledgements}
+We would like to thank the Firedrake development team for their support and for many helpful discussions.
 \end{acknowledgements}

### 2021-01-12 icepack.tex
Use short commit hashes

@@
 Icepack is released under the GPLv3 license.
-The git commit hash of the version of the code used for this publication is b78b0ee571590cf5af242101c159154c6872b7b9, see also the Zenodo release at https://doi.org/10.5281/zenodo.1205640.
+The git commit hash of the version of the code used for this publication is b78b0ee5, see also the Zenodo release at https://doi.org/10.5281/zenodo.1205640.
 The icepack documentation and user manual is hosted at https://icepack.github.io.
-The source code for the demonstrations used in this paper is hosted at https://github.com/icepack/icepack-paper, commit hash 4db3dc281e83b4304751ded3b77f166942b0efc6.
+The source code for the demonstrations used in this paper is hosted at https://github.com/icepack/icepack-paper, commit hash 4db3dc28.
 The demonstrations used glacier outlines that were hand-digitized from satellite imagery in order to generate the model domains.
-These outlines are kept in version control and hosted at https://github.com/icepack/glacier-meshes and the git commit hash for the version used in this publication is c98a8b7536b1891611566257d944e5ea024f2cdf.
+These outlines are kept in version control and hosted at https://github.com/icepack/glacier-meshes and the git commit hash for the version used in this publication is c98a8b75.
 Additional observational data are hosted at the US National Snow and Ice Data Center (https://www.nsidc.org).

### 2021-03-16 reviews/response-1.tex
Stub out responses to reviewers

\documentclass{article}

\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amsthm}
%\usepackage{amssymb}
%\usepackage{mathrsfs}
%\usepackage{fullpage}
%\usepackage{mathptmx}
%\usepackage[varg]{txfonts}
\usepackage{color}
\usepackage[charter]{mathdesign}
\usepackage[pdftex]{graphicx}
%\usepackage{float}
%\usepackage{hyperref}
%\usepackage[modulo, displaymath, mathlines]{lineno}
%\usepackage{setspace}
%\usepackage[titletoc,toc,title]{appendix}
\usepackage{natbib}

%\linenumbers
%\doublespacing

\theoremstyle{definition}
\newtheorem*{defn}{Definition}
\newtheorem*{exm}{Example}

\theoremstyle{plain}
\newtheorem*{thm}{Theorem}
\newtheorem*{lem}{Lemma}
\newtheorem*{prop}{Proposition}
\newtheorem*{cor}{Corollary}

\newcommand{\argmin}{\text{argmin}}
\newcommand{\ud}{\hspace{2pt}\mathrm{d}}
\newcommand{\bs}{\boldsymbol}
\newcommand{\PP}{\mathsf{P}}
\let\divsymb=\div % rename builtin command \div to \divsymb
\renewcommand{\div}[1]{\operatorname{div} #1} % for divergence
\newcommand{\Id}[1]{\operatorname{Id} #1}

\begin{document}

\title{Response to review \#1}
\author{}
\date{}

\maketitle

Thank you Doug for this helpful review!
Original comments are in blue, our responses in black.

\section*{Summary}

\textcolor{blue}{In ‘icepack: a new glacier flow modeling package in Python, version 1.0’, Shapero and co-authors present a promising new ice sheet modeling framework. The
framework contains mechanisms for solving both the prognostic mass balance
equations for updating ice sheet geometry, as well as diagnostic solvers for approximations to the non-linear Stokes’ equations. Throughout both the software
and the manuscript describing it, the authors focus on ensuring usability (readability), a trait that is bound to make this software (and paper) frequently used.
Despite its accessibility, the capabilities of icepack are already impressive, made
all the more so by its explicit design prioritization of easy extensibility.
I find the manuscript to be exceptionally well-crafted, and I think that it
could be published as is. That said, I offer a few suggestions, comments, and
points of clarification below.}

\section*{Minor points}

\begin{description}
\item[L15-18] \textcolor{blue}{It would be nice to have a cited example or three for each of these
suggested use cases. This would help the reader identify the kinds of
practical problems where icepack might fill a need.}

\item[L74] \textcolor{blue}{A low aspect ratio isn’t really an approximation; it’s an existential fact.
The approximation that the first order approximation makes is that vertical resistive stresses (or bridging stresses as they are referred to later
in this manuscript) are small, pressure is hydrostatic, and bed slopes are
small.}

\item[L92] \textcolor{blue}{It would be useful to offer a reference regarding an anisotropic fluidity.}

\item[L100] \textcolor{blue}{Not clear where the Legendre transform enters: the viscous and frictional
dissipation can be read off from Eq. 4.
Eqs. 6 and 12 While I understand that it is convenient to manipulate the
action to reflect the algebraic manipulations to yield the analytical SIA
solution, the break in symmetry between Eq. 6 and Eq. 12 is frustrating,
1given that they both are name ‘gravity’, and that they should in some
sense be the same regardless of which strain rates are assumed to be zero.}

\item[L144] \textcolor{blue}{That the terminal potential term, Eq. 13 disappears is not obvious. This
should probably be clarified, since many readers will be surprised by this.}

\item[L152] \textcolor{blue}{Cite the method of manufactured solutions.}

\item[L159] \textcolor{blue}{The benefit to avoiding complicated 3D meshing should not be under-
stated, in addition to the reduction in the cost of computational solution.}

\item[L188] \textcolor{blue}{This is a bit of a red herring, given that impenetrability is a natural
boundary condition on the incompressibility equation. It can go right
in the action principle, no extra Lagrange multiplier (besides pressure)
needed.}

\item[L367] \textcolor{blue}{I’m surprised by this. For challenging geometries, I’ve always needed to
stabilize even when using implicit Euler. Are you sure that implicit Euler
is unconditionally stable even for non-linear advection like this?}

\item[L453] \textcolor{blue}{Many advances have occurred in the last 5 years regarding gradient de-
scent due to its necessity for optimizing neural networks. These may yet
be useful in this context if you have to deal with a large scale optimization
problem where forming the Hessian becomes prohibitive.}
\item[L459] \textcolor{blue}{Gauss-Newton needs a reference if you’re not going to describe it here.
Eq. 34 It’s worth noting that the Schoof law is phenomenological, and was
selected because it has the right shape and obeys Iken’s bound. As such,
if your sliding law obeys Iken’s bound and looks right, then it’s not any
less valid.}
\item[L580] \textcolor{blue}{I think that having to specify a variational principle for the sliding law
is useful because it guarantees a law that is positive (semi-)definite, as it
must be to be physical.}
\item[L701–713] \textcolor{blue}{I enjoyed reading this paragraph, but I wonder if stabilizing the
Stokes’ equations is the best illustration of the point, given that icepack
does not in fact have Stokes’ equations implemented (although I imagine
it could be done in short order). Maybe stabilizing a transport equation
would be a more appropriate case?}
\end{description}

\end{document}

### 2021-03-16 reviews/response-2.tex
Stub out responses to reviewers

\documentclass{article}

\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amsthm}
%\usepackage{amssymb}
%\usepackage{mathrsfs}
%\usepackage{fullpage}
%\usepackage{mathptmx}
%\usepackage[varg]{txfonts}
\usepackage{color}
\usepackage[charter]{mathdesign}
\usepackage[pdftex]{graphicx}
%\usepackage{float}
%\usepackage{hyperref}
%\usepackage[modulo, displaymath, mathlines]{lineno}
%\usepackage{setspace}
%\usepackage[titletoc,toc,title]{appendix}
\usepackage{natbib}

%\linenumbers
%\doublespacing

\theoremstyle{definition}
\newtheorem*{defn}{Definition}
\newtheorem*{exm}{Example}

\theoremstyle{plain}
\newtheorem*{thm}{Theorem}
\newtheorem*{lem}{Lemma}
\newtheorem*{prop}{Proposition}
\newtheorem*{cor}{Corollary}

\newcommand{\argmin}{\text{argmin}}
\newcommand{\ud}{\hspace{2pt}\mathrm{d}}
\newcommand{\bs}{\boldsymbol}
\newcommand{\PP}{\mathsf{P}}
\let\divsymb=\div % rename builtin command \div to \divsymb
\renewcommand{\div}[1]{\operatorname{div} #1} % for divergence
\newcommand{\Id}[1]{\operatorname{Id} #1}

\begin{document}

\title{Response to review \#2}
\author{}
\date{}

\maketitle

\section*{Summary}

\textcolor{blue}{In this manuscript, the authors introduce a new land-ice modeling software package known as icepack.
icepack is written in Python on top of the Firedrake library, which uses the domain-specific Unified Form
Language (UFL) and provides a high-level symbolic description of the problem to facilitate the add of new
physics and/or equations. The intended user base of icepack is the glaciological community, in particular,
glaciologists who may not have an extensive background/training in computational science. The idea is to
provide a code that would be easy to use/develop by this class of prospective users. Following a description
of the code, some numerical examples are presented to demonstrate the method’s capabilities and accuracy.
The manuscript in question is well-written and interesting to read. Addressing usability is noteworthy
and something that not enough authors in glaciology/climate science address. I personally have some
qualms with some of the philosophy described by the authors, namely I worry about folks who are not
familiar with numerical methods developing an application code in which a lot of what is under the
code is hidden from them in some sense. In an ideal world, one would have glaciologists working with
computational scientists to help them pick the right solvers, discretizations, etc., for their problem. The
authors are correct that some solver options are for optimizing performance, which is secondary to getting
the code/model running; but there are also solver/algorithm choices that depend very much on the physics
(e.g., CG is only valid for symmetric problems) - is icepack designed so as to prevent the naive user from
inadvertently using the incorrect default setting for their problem? I will assume it is to the extent it can
be, and that the authors’ argument is that the default hidden settings are likely to do less damage than
some arbitrary settings a user might put in his/her input file without knowing what they are doing. Also,
I realize that the reality is that many glaciologists do not have strong ties to computational scientists, and
still wish to make progress in numerical modeling of land-ice; therefore, I will not focus too much on much
“philosophical” perspective described above.
Another qualm I have about the paper has to do with performance - I am skeptical whether icepack
can really be performant if one tries to run it on continental scale problems, and it is not clear to me if
the code is even parallel. I think the intention may be to use icepack as more of a sandbox for prototyping
small problems (similar to FeniCS), in which case, this is not a huge deal.
Despite the above concerns, I like this paper and see it published in GMD. I like in particular the
idea of describing all the equations using the variational principle/action functional and having everything
else propagate from there - not enough people do this. I do, however, feel that there is a lot of missing
information in various parts of the paper, which should be filled in in the revision prior to the paper being
suitable for publication. Please see below my enumerated list of questions/comments to address in the
revision.}

\section*{Specific comments}

\begin{enumerate}
\item \textcolor{blue}{The authors suggest that C++ makes it inherently difficult to add new physics/PDEs (e.g., on p. 10
1and p. 20), which I somewhat disagree with. One advancement that can make a C++-based code
easy to add to is Automatic Differentiation (AD) - with AD, one can effectively code the weak form
of the residual within a C++ code and AD will handle the rest, making it very easy to add new
physics. An example is the Albany/Land Ice model (previously known as Albany/FELIX) of Tezaur
et al. (https://doi.org/10.5194/gmd-8-1197-2015). I think it would be worth mentioning that
there have been efforts like Albany/Land Ice out there to make C++-based codes more accessible to
users of varied backgrounds. I agree that even an “easy-to-use” C++ code will be more difficult and
more intimidating than a Python code, so I am not trying to minimize the authors’ efforts at all.}
\item \textcolor{blue}{Can the authors comment on the overhead of the symbolic descriptions/manipulations done by their
framework? This sounds potentially like it would be very expensive. How does the cost compare to
automatic differentiation, for example? A broader question is: is computational performance/cost a
concern for users of icepack, or is it intended to be a “sandbox” in which performance is secondary
to being able to code up something “quick-and-dirty” for initial prototyping?}
\item \textcolor{blue}{I was a little bit confused about the reference of the FO Stokes-based model in this paper as a “hybrid
model”. I see that it is hybrid in the sense that you have a different discretization in the horizontal and
vertical direction, but there are also hybrid ice models that use different PDEs in different domains,
e.g., the ISCAL model of Ahlkrona et al. (https://doi.org/10.1016/j.quascirev.2016.01.032).
Is the term “hybrid model” a common name for the approach in Section 2.2.3? Perhaps it is and I
am not aware of it.}
\item \textcolor{blue}{Does the hybrid model described in the paper have the same applicability as say the First Order
Stokes model? Can it be used for both Greenland and Antarctica at continental scales?}
\item \textcolor{blue}{Section 2.2.1: in my opinion, the authors do not provide sufficient justification for the penalty term,
equation (7). They describe this as something that is added to smooth over artifacts - this would be
needed based on the discretization, which there is little discussion of. The authors should state what
order finite elements they are using - I presume they are linear, and that this is why the stabilization
is needed? Why is stabilization needed only for the SIA? I think these things should be made clear.}
\item \textcolor{blue}{Section 2.2.2: there is some imprecision here in equation (13) - you have not defined anywhere that $\Gamma$
is the boundary, and which boundary you are referring to. One can figure it out, but it is not precise.
$\Omega$ is not defined either though one will assume invariably that this is an open bounded domain in 2D
or 3D depending on which approximation one is looking at.}
\item \textcolor{blue}{The boundary conditions are not discussed very rigorously systematically - the authors seem to
sprinkle in some boundary conditions here and there. I think the boundary conditions need to be
given for each of the models at the time the models are presented - boundary conditions are needed
to complete the definition of each models.}
\item \textcolor{blue}{Certain terms in the equations I do not believe are defined anywhere, for instance, in equation (10),
there is no expression given for the strains $\dot\epsilon(u)$.
This is one of the things I cam across that need to
be made more precise.}
\item \textcolor{blue}{Section 2.2.3: the authors comment that higher degree polynomials can be used in the vertical layer
in the hybrid approach. What order is typically used?}
\item \textcolor{blue}{Section 2.2.3: this might be a naive question, but does Glen’s flow law come into the hybrid model?
I was expecting to see it there, but maybe I’m missing something.}
\item \textcolor{blue}{P. 10: the discussion here about substituting model components suggests it may be possible to use
different models in different regions and couple them (a la the ISCAL method). Is this possible, or
something that the authors are thinking to add to their model/code?}
\item \textcolor{blue}{Section 2.4: This section is very incomplete. You need to give the enthalpy equation and given the
Glen’s law expression as well, since it is mentioned.}
\item \textcolor{blue}{In my opinion, there is not enough discussion of the thickness equation (Section 2.1) and how it
is discretized. In typical ice sheet models, this equation is used to change the ice extent - one
meshes up a region of “potential” ice, and then uses the thickness to dynamically determine a mask
for ice-covered regions. Do you do something like this in your model? It should be discussed for
completeness. I think you maybe start to do this in Section 4.1, but it is very confusing and hard to
make the connection.}
\item \textcolor{blue}{What sort of meshes do you use in your model (in the horizontal dimension, for the hybrid one)?
Structured/unstructured? Hex/tet (quad/tri)?}
\item \textcolor{blue}{Section 3: It is not clear from the description what the inverse problem you are describing is for.
Is it to obtain parameters in the model like the basal friction using observational data of e.g. sur-
face velocity? There really needs to be more discussion here, and I personally would like to see
a mathematical statement of a representative inverse problem you are solving. It would be worth
citing the work Perego et al. on optimization-based inversion, if what you are doing is similar:
https://doi.org/10.1002/2014JF003181. BTW, the basal friction has not been defined, yet it is
discussed - it needs to be defined earlier, when talking about boundary conditions (which needs to
be added).}
\item \textcolor{blue}{Section 4.1: I find this section confusing. I assume you are talking about discretizing the thickness
equation here - that should be made clear. I disagree with several statements in this section as
well. “The simplest explicit timestepping schemes are unstable with CG finite elements” - if you are
talking about CFL stability, this is not true. You need to satisfy a CFL condition which could give
rise to very small time-steps but you can get the scheme to be stable. I’m also confused about the
notion of SUPG as a time-stepping scheme - I think of SUPG as a finite element approach to deal
with advection-dominated flow problems, for example, that does not have anything to do with time-
stepping. Maybe you are referring to upwinding? In any case, I think SUPG has nothing to do with
forward Euler, so the discussion about forward Euler requiring parameters is erroneous. Additionally,
I don’t understand the comment about implicit Euler smoothing out sharp discontinuities... I believe
explicit and implicit Euler have effectively the same diffusion and dispersion properties, so there
should not really be a difference between the schemes. Did the authors verify their time-stepper on
a manufactured problem to ensure that it was implemented correctly?}
\item \textcolor{blue}{Section 4.2: there is an approach discussed in Tezaur et al. (https://doi.org/10.5194/gmd-8-1197-2015)
for dealing with bad initial conditions in a Newton solver that relies on homotopy continuation that
would be worth citing. It is an alternate to the approach you describe that lets you get away with not
doing a line search for Newton. By the way, it should be no surprise that Newton is not converging
3without a line search - in general Newton is not guaranteed to converge from an arbitrary initial
guess without the line search.}
\item \textcolor{blue}{Section 4.3, lines 411-412: there are actually ways to construct weighted norms to deal with the issue
of DOFs having different orders of magnitude for the purpose of convergence.}
\item \textcolor{blue}{Section 4.4: Again, it is not clear to me what is your inverse problem. You need to state this explicitly
so it is clear.}
\item \textcolor{blue}{Section 4.5.1: you talk about problems defined on “extruded geometries” - do you ever use non-
extruded geometries? It has been shown in various references that there can be numerical problems
for land-ice solvers that do not use extruded geometries, e.g. Tezaur et al.}
\item \textcolor{blue}{Section 4.5.2: I think this section needs to be made earlier, and other BCs need to be added to that
discussion.}
\item \textcolor{blue}{Section 4.6: The authors mention running their code on 1 core. Is the code parallel - can it be run
on multiple cores? Are there any hope for performance portability of the code to take advantage of
emerging HPC architectures, e.g., GPUs?}
\item \textcolor{blue}{Section 4.6: Can you please clarify what you mean by the following statement? “Large problems,
such as continental-scale modeling, will require more sophisticated and possibly problem-specific
approaches”. I’m wondering in particular about the problem-specific approach part. There are
models like first order Stokes that can be used at the continental scale and they are not really
problem specific.}
\item \textcolor{blue}{p. 20: I don’t understand why you need to create an analytical expression of (32) using special
functions. Is this something specific to your framework, which requires expressions in a certain form
for the symbolic representation?}
\item \textcolor{blue}{It’s great that you have executable documentation in something easy-to-use such as Jupyter note-
books! (no need to address this comment)}
\item \textcolor{blue}{p. 29: are you considering putting in the first-order Stokes//Blatter Pattyn model into your code
framework?}
\item \textcolor{blue}{The methods do not discuss their code development/testing stance on icepack. How do users con-
tribute to the code - through pull requests? Is there regression/performance testing? Continuous
integration testing? These are all really important, especially if you have non-experts contributing
to the code!}
\end{enumerate}

\section*{Minor comments}

\begin{itemize}
\item \textcolor{blue}{“UFL” is not defined in the abstract.}
\item \textcolor{blue}{p. 4, line 90: A is also called the “flow factor”. I would mention here that it is usually a function of
the temperature, which comes from a different equation.}
\item \textcolor{blue}{p. 5, line 125: should be “checking”, not “check”.}
\item \textcolor{blue}{p. 5, line 126: I think it should be “Bueler” not “Beuler”, if I’m thinking of the right person.}
\item \textcolor{blue}{p. 5, line 147: change “we verified the correctness of the ice shelf model” to “we verified the correctness
of our implementation of the ice shelf model”.}
\item \textcolor{blue}{p. 10, line 270: change “we’ll” to “we will”.}
\end{itemize}

\end{document}

### 2021-03-16 reviews/response-3.tex
Stub out responses to reviewers

\documentclass{article}

\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amsthm}
%\usepackage{amssymb}
%\usepackage{mathrsfs}
%\usepackage{fullpage}
%\usepackage{mathptmx}
%\usepackage[varg]{txfonts}
\usepackage{color}
\usepackage[charter]{mathdesign}
\usepackage[pdftex]{graphicx}
%\usepackage{float}
%\usepackage{hyperref}
%\usepackage[modulo, displaymath, mathlines]{lineno}
%\usepackage{setspace}
%\usepackage[titletoc,toc,title]{appendix}
\usepackage{natbib}

%\linenumbers
%\doublespacing

\theoremstyle{definition}
\newtheorem*{defn}{Definition}
\newtheorem*{exm}{Example}

\theoremstyle{plain}
\newtheorem*{thm}{Theorem}
\newtheorem*{lem}{Lemma}
\newtheorem*{prop}{Proposition}
\newtheorem*{cor}{Corollary}

\newcommand{\argmin}{\text{argmin}}
\newcommand{\ud}{\hspace{2pt}\mathrm{d}}
\newcommand{\bs}{\boldsymbol}
\newcommand{\PP}{\mathsf{P}}
\let\divsymb=\div % rename builtin command \div to \divsymb
\renewcommand{\div}[1]{\operatorname{div} #1} % for divergence
\newcommand{\Id}[1]{\operatorname{Id} #1}

\begin{document}

\title{Response to review \#3}
\author{}
\date{}

\maketitle

\section*{Summary}

\textcolor{blue}{Icepack is an important new model, and a description paper is appropriate.  Four features of Icepack stand out: its use of Firedrake/Python, the flexible action-principle design of its stress-balance solver module, its from-the-start attention to data assimilation and inverse modeling, and its design as a modeling environment and language instead of a ready-to-run model.  All of these choices are addressed appropriately in this manuscript.  Readers are very likely to try the model, which fullfills a major purpose of a description paper at this early-ish development stage.}

\textcolor{blue}{However, this description paper can be improved in three significant ways:}

\textcolor{blue}{A.  Greater attention to the meaning of Icepack as a *time-dependent and geometry-evolving model*, and to related defaults.}

\textcolor{blue}{B.  Avoidance of *bad linguistic habits* inherited (mostly) from the ice sheet modeling literature.}

\textcolor{blue}{C.  As currently laid out, the paper treats the reader mostly as a potential co-developer of the model, while Icepack's *effectiveness as a simulation tool for science* is muddled.}

\textcolor{blue}{Fully addressing concern A would require major code extensions, which are not my intention in commenting on this.  Rather, the manuscript should make the reader aware of which evolution aspects are well-handled by current Icepack and which are in future development.}

\textcolor{blue}{The above leading concerns will be addressed in more detail below, in a list which addresses specific line numbers.  The associated potential improvement "A","B","C" is listed when appropriate.}

\section*{Line-by-line comments}

\textcolor{blue}{13, AC:  Presumably all software packages can be effective in the hands of experts, and here it is suggested that Icepack stands-out as better for non-experts.  My question is whether it will be effective for non-experts interested in announced goals 1 and 4 (among those listed on lines 15--18).  This is hard to believe given how Icepack seems to be designed around, and the manuscript focused on, goals 2 and 3.  Consider the reader who wants to simulate glacier extent for some years into the future for a mountain glacier or Greenland, who can supply a simulated climate (atmosphere/ocean), a bed topography, and a current geometry in some data files.  Does this paper convince me that they need *less* expert knowledge to use Icepack for that purpose than other existing ice sheet models?  Not yet.  Of course a full usage answer occurs in online tutorials and examples, not just the manuscript.  Nonetheless the absence of attention to transport equation boundary conditions, and to how mass/energy surface inputs are handled, gives me the impression that such a reader is on a long co-development path with the Icepack authors, requiring the development of much expert knowledge before first useful results.  Said a different way, expert knowledge is required for any software that does not have an aggressive scheme for putting reasonable defaults into the hands of novices.  To paraphrase a recent Firedrake paper [Farrell et al 2020], it is a mechanism-vs-policy concern.  Icepack has a library of mechanisms, but an absence of apparent policy means expert knowledge is needed to recover usability for real-science applications.  I rolled my eyes at the implications of the sentence on line 13, and these concerns remained after reading the whole paper.}

\textcolor{blue}{21--39:  I also think Python+Firedrake+PETSc is the most promising environment to build a new ice flow simulation library/model.  I'm on board!}

\textcolor{blue}{44--46, B:  Let me vote to *not* maintain the "diagnostic"/"prognostic" linguistic tribalism.  The world calls these equations "conservation of momentum" and "conservation of mass".  (Indeed one should remind the reader that the latter describes thickness evolution because of how glacier models normally parameterize fluid geometry.)  An "also known as the 'prognostic equation' [cite]" is appropriate, but another paper using this tribal language will cause yet more students to need to unlearn silly language in order to read the mainstream fluids and numerical PDE literature.}

\textcolor{blue}{46--47:  Here!  Here!  It is a infinite-dimensional DAE!  Good.  Many readers will be unfamiliar with the concept; I cite [Ascher \& Petzold 1998] for that but there may be better references.}

\textcolor{blue}{56, equation (1) and nearby, AB:  Two concerns.  First, it is later acknowledged that this is not really an advection equation (lines 376--377), so one does not need to call it that here either.  The SIA is not some weird alternative universe of weak-willed modelers, it is what *all models should produce* in the large.  That model, the only clearly-understood coupled model, makes equation (1) a diffusion, as noted.  Surely calling this a "transport equation" or even "thickness transport equation" is adequate.  One then points-out that $q=hu$ is one way to parameterize flux, and that $u$ comes from a coupled equation \emph{driven by $\nabla s$} (even in the Stokes case).  It might be acknowledged that (1) does not have a PDE "type" in the classical sense.  (It is a DAE, after all.)  Second, equation (1) holds with what boundary conditions?  This manuscript maintains the tradition of pretending not to notice.  To quote [Schoof \& Hewitt 2013], "A sometimes weakly perceived point in glaciology is that the model above is in fact a free-boundary problem ...", and *this applies to (1) regardless of the stress balance model*.  Is this paper just going to pretend boundary conditions for the main, and first, equation don't exist?  (Or pretend that all glaciers end in cliffs of sufficient height so that the singular change of coordinates (16) is no problem?)  This manuscript could even cite existing wellposedness literature for (1), especially [Calvo et al 2003] and [Jouvet \& Bueler 2012], which confirm the meaningfulness of the free-boundary problem formulation and the resultant simultaneous Dirichlet (h=0) and Neumann (q=0) boundary conditions at grounded margins.  And then say what Icepack will when solving (1).}

\textcolor{blue}{68, B:  It is not clear why the conservation of momentum model for a slow fluid should ever be called a "momentum transport" model.  There is no d/dt for momentum!  Surely "conservation of momentum equation" or "stress balance equation" suffice.}

\textcolor{blue}{74, B:  No, the Blatter-Pattyn (BP) equation is not a "hybrid".  This word is perfectly descriptive when categorizing ice sheet momentum models, and I am not sure who decided to mangle the language in this way.  (It precedes this manuscript.)  There are principled models, of which SIA, SSA, and BP are three, arising from small parameter arguments.  There were horizontal hybrids (Ritz) and there still are vertical hybrids (Pollard, Bueler, Winkelmann).  Hybrids, by definition, combine more than one principled model in some (less principled!) manner.  If you are trying to say that BP is a shallow model, but less-shallow than SIA and SSA ... that's true!  And if you are saying that BP can balance stresses in a plug or a shear flow, that's true too.  Say those things.}

\textcolor{blue}{101--102, B:  Regarding the sentence "The action principle can be viewed as a consequence of the Onsager reciprocity relations for systems near to equilibrium ...":  I am trying to imagine a reader for whom reading that would be a useful learning experience regarding the action principle, as opposed to an irritating one.}

\textcolor{blue}{108:  I do not understand the intended meaning of "vastly more convenient numerically" here.  The cost of evaluating the objective, versus a merit function like the square of the residual norm, is the same.  The cost of evaluating a residual is the same whether or not it happens to be a gradient of an objective.  Perhaps what is meant is more like: "Minimizing a convex functional provides better convergence guarantees than solving a general system of nonlinear equations ...", which is certainly true.  At least a forward reference to section 4.2 would help explain the phrase.}

\textcolor{blue}{115, equation (5), B:  Now the bad linguistic habits are being inherited from the FE world.  This is not "mass" in any sense other than that any matrix with entries $a_ij = \int psi^i psi^j dx$ is always called a "mass matrix" in the FE world.  And the reader may not have that FE bad habit already, in which case confusion ensues because (5) has nothing to do with mass conservation!  My suggestion for this term, implicitly acknowledging how it arises in the SIA, is to call it "localization".}

\textcolor{blue}{115--124:  The action combining (5) and (6) is already convex and coercive \emph{in $L^2$}.  The reason to add the penalty term is (presumably) because the FE space choices want to work in $H^1$.  (Adding (7) with $\ell > 0$ makes $J$ coercive in $H^1$.)  This penalized form is a perfectly reasonable idea, and it makes the SIA behave more like other stress balances, *and* it is one of those unprincipled things one does to get it all to work properly ... if you add another kludge you could even call yourself a "hybrid".}

\textcolor{blue}{126:  This author's name is Bueler not Beuler.  (The latter is a PETSc \texttt{-ts\_type}, so it is easy to get confused.)}

\textcolor{blue}{137--140:  There is a double negative in equations (12) and (14) which is not used in (6) and (8), respectively.  Does this reflect anything important?}

\textcolor{blue}{148--149:  It is of course true that for general boundary conditions the "shallow stream equations do not have a simple analytical solution."  What is probably meant here, however, is that there is not a well-known, exactly-solvable, basal-friction-included boundary value problem suitable for testing.  But that's not true, and indeed the *very first theoretical paper in glaciology* provides one, namely [Böðvarsson 1955]!  See the full story, and a derived marine flow-line exact solution suitable for testing, in [Bueler 2014].  The method of manufactured solutions applies, of course, but at loss of clarity on the meaning and magnitude of the resulting numerical errors, and with great danger of testing the wrong parts of a (nonlinear) system phase space, and with loss of the history of mathematical glaciology.}

\textcolor{blue}{158:  I guess the phrase "individual fast-flowing glacier" arises because of the sense that the SSA does not handle slow flow very well, which is true, and that fast ice is separated by slow flow.  Nonetheless, I would replace "features of an individual fast-flowing glacier" by "features of fast-flowing glaciers".}

\textcolor{blue}{170--171:  I am not clear on "a tensor product basis of Lagrange finite elements in the vertical and higher-order polynomials in a single vertical layer".  Should the first "vertical" actually be "horizontal"?  The novice reader not already used to FE stuff could use a figure or more words here, I suspect.}

\textcolor{blue}{175, equation (16), A:  From here forward, in the manuscript, the lateral boundary condition for equation (1) is even less clear because this change of variables takes the lateral free margin and blows it up.  (In the grounded ice case, and where "blow up" is used in the mathematical sense.)  With what consequences for simulating moving margins?  Section 4.1 does not address this (i.e. how the boundary condition for the transport equation behaves as h->0; a boundary condition is not mentioned).  I believe, on the other hand, that the SIA grounded-margin example in section 5.2 does not use (16), which as described applies only applying to "the hybrid flow model", but section 5.2 does not mention the margin anyway.  I am concerned that positive consequences of using (16) are emphasized while negative consequences are not mentioned, and that Icepack has no path around these negative consequences.}

\textcolor{blue}{215:  For clarity to readers not already thinking about Firedrake's extruded meshes, I suggest adding a parenthetical: "linear or quadratic elements on (horizontal) triangles".}

\textcolor{blue}{216:  Instead of "Lagrange interpolating polynomial basis", which readers accustomed to numerical analysis terminology (outside FE) might associate to the Lagrange method of finding interpolating polynomials, it is probably clearer to just say "Lagrange elements".  Perhaps emphasize the nodal placement?  (Legendre nodes, concentrated near endpoints, versus the equally-spaced nodes of Lagrange elements.)  Orthogonality is not the only reason why Legendre is better; node location implies better approximation properties.}

\textcolor{blue}{227:  "Smoke test" is an unfamiliar idiom to me.  As in testing machinery or testing electronics or smoking-out bugs or what-am-I-smoking?  Pity the non-native speaker.}

\textcolor{blue}{227:  "most minimalistic" --> "minimal"}

\textcolor{blue}{234--235:  I have received the advice "write about what you have done, not what you haven't", and I think it applies here.  Using two layers in this way assumes one polythermal structure and will not allow transition to the other.  No need to consider or mention it.}

\textcolor{blue}{253--289:  This chunk of text is important, and it describes one of the best aspects of Icepack's use of Python.  I think it should be put at the start of subsection 2.3, which will give clarity that when equations (22) and (23) appear, they are primarily an example of submodel plug-ability.}

\textcolor{blue}{283--289:  I'm on board with keyword arguments!}

\textcolor{blue}{298--299:  Suggested replacement for the mushy phrase: "describing realistic glacier flows, but it is not sufficient by itself" --> "parameterizing fluidity".}

\textcolor{blue}{300--301, C:  I seriously doubt that more than one user in 100 will want to substitute their own melt-fraction dependence, because no one has the data for it.  So: "likely" --> "possible".  (This is a clear case where good defaults are more important than modularity.)}

\textcolor{blue}{297--305, C:  This paragraph reads, to me, like an argument that expert knowledge *will* be required for the Icepack user.  ("users must calculate the fluidity field themselves")  That is, unless Icepack includes both defaults and higher-level simulation drivers which allow the novice user to never know about these possibilities.  More broadly, what does Icepack design offer, when it comes to avoiding explicit choices of all submodels, on the way to a first scientifically-useful model for the beginner?}

\textcolor{blue}{318--320:  Mushy sentences.  Needed?}

\textcolor{blue}{345, C:  The sentence "The key classes that users interact with are flow models and solvers" is right at the heart of my concern C.  Indeed, after developing PISM with much (developer group) concern about flow models and solvers, we spent the next 10 years talking to users about their data formats, parameter-study schema, and surface process models (supra, sub, and calving front).  Barely had time to work on flow models and solvers ... until I quit day-to-day PISM development.  The by-far largest population of scientists care about how flowing ice interacts with their climates.  They use ice flow models on the assumption that ice flow modelers know how to model ice flow!  This "key classes" sentence is not describing users, it is describing co-developers.}

\textcolor{blue}{368:  "Practicing ... may be unfamiliar with" is unnecessary.  Just say what you have to say.}

\textcolor{blue}{362--381, A:  Three concerns about section 4.1.  First, as already noted, transport equation (1) has boundary values and they are again silent here.  Second, there are multiple transport problems in glacier modeling: equation (1), the advection-dominated enthalpy equation, the ice shelf damage transport equation, the age equation, ... etc.  Is this paragraph covering them all, or just equation (1)?}

\textcolor{blue}{368--373, AC:  The third concern, regarding this paragraph, is that Icepack apparently has not yet set-up effective adaptive time-stepping.  (I have not gotten a runtime error from a modern explicit ODE solver for a long time!  Have you?  Some runs are slow cause they are stiff, indeed.)  This should be addressed/acknowledged.  In particular, assuming a perfectly-implemented \emph{implicit} solver, how does $\Delta t$ scale with $\Delta x$?  If the problem were really advective, so that the goal is to model influences as they travel at the characteristic speed, then the answer would be $\Delta t = O(\Delta x)$ for accuracy, even with your implicit scheme.  (Better than $O(\Delta x^2)$, yes indeed ... but then explicit steppers would be fine ... but (1) is not an advection ... we must think more.)  Though the problem is actually some diffusive/advective mix, some defined scaling of $\Delta t$ with $\Delta x$ is still needed for accuracy.  This can come from an adaptive time-stepper, or be designed from scratch.  Then there is the matter of the user's data time scale for surface mass balance, etc., which implies "events" in the adaptive time-stepper (when data is read).  In fact, the actual questions a glacier model must answer, regarding time steps, are these:  How does time-stepping change under spatial grid refinement?  Does it converge in space-time refinement?  Is it robust over realistic geometries and inputs?  What is the user interface?  This paragraph convinces me that Icepack is not yet there, which should not be completely buried.}

\textcolor{blue}{363:  "equation equation"}

\textcolor{blue}{365, A:  The "In the interest of simplicity" phrase here tells me the authors simply have not thought-through the time-evolution of glacier geometry and the needed boundary conditions.  (You have a scheme for unconditional-stably generating glacier surfaces which conserves mass in the presence of ablation?  Then don't hide it!)  Please take the problem seriously:  Address how you maintain reasonable margin shape and positive thickness in an implicit scheme.  It is o.k. to admit that time steps cannot be arbitrarily long in your scheme, to get convergence; the "conditional/unconditional stable" language is an artifact of linear PDE theory, and your problem, taken seriously, is super-duper-nonlinear.}

\textcolor{blue}{376--377, B:  This sentence is a very good argument for *not* calling (1) an advection equation.}

\textcolor{blue}{383--444, A:  Sections 4.2 and 4.3 convince me that when the Icepack team takes on aspects of model design they care about then it comes out very well!  These sections suggest the momentum balance ("diagnostic" ... grrr) solver has great defaults.  Likewise with sections 4.4 and 4.5.  (Now for serious attention to time-stepping, mass conservation, and the user's surface mass- and energy-flux data, and clarity on Icepack's TODO list.)}

\textcolor{blue}{459--462:  Gauss-Newton is a good choice for this purpose, as pointed out for exactly this purpose, inversion of glacier models, by [Habermann et al 2012].  Which should be cited.}

\textcolor{blue}{463--466:  Symbolic derivatives.  Again I am on board with the benefits of a Firedrake-based tool chain.}

\textcolor{blue}{469, A:  Here "terminus boundary condition" refers to the momentum balance equation.  There is, as far as I can tell, no regard for the "terminus boundary condition" needed when solving equation (1), i.e. the mass conservation terminus boundary condition, even in the case of a cliff, much less a grounded margin.}

\textcolor{blue}{474--487:  Extruded meshes.  On board with Firedrake.  But if the model claims to have a solution of (1) then there would be some mechanism for addressing regions which become ice-free within a time step?  No mention thereof.}

\textcolor{blue}{493--515, A:  Section 4.5 describes a nice solution to a real problem.  In fact, can one consider this kind of fit to the glacier profile at a grounded margin?  Not obvious how to proceed, I agree, but note that the thickness field h of a glacier has the similar continuous-but-non-differentiable character as the pressure on the ocean calving front.}

\textcolor{blue}{517--522:  Rich problem-aware preconditioners and solvers.  On board with Firedrake.}

\textcolor{blue}{523, A:  This honest, clear sentence is not at all expected by the reader who remembers lines 15-19 in the Introduction.  Please let the reader know earlier.}

\textcolor{blue}{533--586:  The MISMIP+ experiment is part of the 2008-onward tradition of running ice sheet models in rectangular boxes (MISMIP, EISMINT-HOM, ...).  This is fine in a field where one also has some laboratory fluid models, i.e. actual stuff, which fit in boxes.  But this tradition has a distressing impact on new model development, which is to modularize around the choices one makes between intercomparison-specified boxes and their associated boundary conditions.  (Yes, that is what Icepack looks like right now.)}

\textcolor{blue}{588--597, A:  This example would be more impressive with hybrid physics, right?, but Icepack is not there yet.  (Or else it would have been applied here.)  Readers of GMD should be taken to be serious people.  Tell them the score, and I don't mean relative to what other models are capable of.  What needs to improve?}

\textcolor{blue}{598:  Regarding "computationally cheap enough ... in a matter of minutes on a desktop":  I suppose the excuse for this sentence is that other people get away with writing such stuff?  The question is how run time scales with mesh resolution.  The reader can handle a plot, if you want to generate one.}

\textcolor{blue}{604--647:  Nice examples!  Ice shelf physics is a scope where Icepack is a convincing choice for a research project.}

\textcolor{blue}{662--663:  Regarding "The person learning ... is largely absent from the discussion", have you looked at the PISM User's Manual?  (Start from page 1.)  This opinion ranges from disputable to insulting, but mostly reflects not paying too much attention to unpublished prior literature (i.e. online manuals).  On exactly that topic, the online Icepack documentation is excellent.}

\textcolor{blue}{666--713:  Saying these HCI principles out loud is a very worthwhile aspect of the manuscript.}

\textcolor{blue}{677--678, C:  There is a *big* gap between "user interface in an interpreted language" and a "program that can only run in batch mode".  C-based PETSc programs like the PISM ice sheet model, and many other command-line programs, live in the interior of this gap.  For example, please consider these three HCI principles as they apply to command-line git.  It is neither "a user interface in an interpreted program" nor a "batch mode" only thing.  (Most command-line tools are not!)  Git has a steep learning curve, because it is a DAG modeling language, but such command-line design can hit your principles too.  Indeed, a good antidote to your false dichotomy is [Brown et al 2014], and addressing UI points made there would increase the credibility of section 6.  (You'll see that Icepack is definitely a Brown-approved design library-wise, but doing science is not equal to designing a library API either.)  Ultimately any science application of Icepack will be a map from inputs (data) to outputs (simulations or inversions), and the usability of that map (e.g. in ensemble simulations) is different from the develop-ability of it in an interactive environment.  Python is a great environment for experimentation, and for ice sheet modeling, and progressive evaluation is a desirable principle, but an interactive Python session is not the only alternative to 1990s climate-model design.  Presumably Icepack usage is intended to progress from all-interactive mode to parallel production/ensemble runs in batch systems anyway?  Address that?}

\textcolor{blue}{681--683, C:  Regarding "The API ... one step at a time":  There is a *huge* amount of expert knowledge required to do science with an "ice sheet model" which does not have a policy for doing things "one step at a time".  Imagine a paper that proposed a new-and-better WRF model but said something like this about the demands on the user!}

\textcolor{blue}{682--683, C:  Also, this "user" is a co-developer.}

\textcolor{blue}{693--713, C:  This basic point about abstraction gradient is good.  But solver/discretization components/choices is not the only such gradient.  From talking to a lot of novice glacier modelers, I assert the key abstraction gradient, which an ice sheet model must finesse, regards ice flow physics.  (What aspects of the full, coupled, nonlinear dynamics are the next ones that the user's constructed Icepack model should/can handle for the science goal?  How to build-in that physics without unnecessary parameters?  How to generate intermediate results which reveal which processes are missing, re the science goal?)  Thinking on this stuff is where Icepack developers could make their next real progress.  That means de-emphasizing solver components/discretizations in the user's view, primarily by setting aggressive defaults, even as the developers must get solvers right (which is the strength of Icepack).}

\textcolor{blue}{721:  "phyics"}

\textcolor{blue}{724:  Having reviewed quite a few ice sheet modeling papers, every single one claimed something about its usability.  Less mush, more answers please.}

\textcolor{blue}{725:  I'm on board with (2), and wish for more careful analysis of Icepack's contribution here, but "quantifying" in (1) is not what you are doing in this manuscript.  (Nor, probably, do you have the ability to do it.)}

\textcolor{blue}{732--740, AC:  If I saw "4. robust time-evolution and climate interaction tools", or similar, on this list then I would be more of a believer in Icepack's future.  Will you be able to break out of stress-balance-and-inverse-model-solver-playground mode, and start answering some of the science questions enumerated in the Introduction?}

\section*{References}

Ascher, U. and Petzold, L. (1998). Computer Methods for Ordinary Differential Equations and Differential-Algebraic Equations, SIAM, Philadelphia.

Böðvarsson G (1955) On the flow of ice-sheets and glaciers. Jökull,

5, 1–8 (find this too-little-read paper at: https://github.com/bueler/bod-marine/blob/master/Bodvardsson1955.pdf)

Brown, J., Knepley, M. G., \& Smith, B. F. (2014). Run-time extensibility and librarization of simulation software. Computing in Science \& Engineering, 17(1), 38-45.

Bueler, E. (2014). An exact solution for a steady, flowline marine ice sheet. Journal of Glaciology, 60(224), 1117-1125.

Calvo, N., Díaz, J. I., Durany, J., Schiavi, E., \& Vázquez, C. (2003). On a doubly nonlinear parabolic obstacle problem modelling ice sheet dynamics. SIAM Journal on applied mathematics, 63(2), 683-707.

Farrell, P. E., Kirby, R. C., \& Marchena-Menendez, J. (2020). Irksome: Automating Runge--Kutta time-stepping for finite element methods. arXiv preprint arXiv:2006.16282.

Habermann, M., Maxwell, D., \& Truffer, M. (2012). Reconstruction of basal properties in ice sheets using iterative inverse methods. Journal of Glaciology, 58(210), 795-808.

Jouvet, G., \& Bueler, E. (2012). Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation. SIAM Journal on Applied Mathematics, 72(4), 1292-1314.

Schoof, C., \& Hewitt, I. (2013). Ice-sheet dynamics. Annual Review of Fluid Mechanics, 45, 217-239.

\end{document}

### 2021-03-17 reviews/response-1.tex
Most of reviewer 1 corrections

@@
 
+Changed to: ``The first-order model is based on an asymptotic expansion of the Stokes equations in the ratio of the ice thickness to a typical horizontal length scale.
+The aspect ratio of most glacier flows is on the order of 1/20 or less, although there are some exceptions.
+For example, the main trunk of Jakobshavn Isbrae in Greenland flows through a very deep and narrow trough with an aspect ratio closer to 1/5.
+''
+
 \item[L92] \textcolor{blue}{It would be useful to offer a reference regarding an anisotropic fluidity.}
 
+Added a reference to Gillet-Chaulet et al. 2006.
+
 \item[L100] \textcolor{blue}{Not clear where the Legendre transform enters: the viscous and frictional
-dissipation can be read off from Eq. 4.
-Eqs. 6 and 12 While I understand that it is convenient to manipulate the
+dissipation can be read off from Eq. 4.}
+
+Other reviewers didn't find this digression useful so we've removed it.
+
+\item[Eqs. 6 and 12]\textcolor{blue}{While I understand that it is convenient to manipulate the
 action to reflect the algebraic manipulations to yield the analytical SIA
@@
 
+Expanded the explanation to: ``Additionally, the terminal stress term of the action disappears after applying integration by parts to the gravity term to shift the gradient of the surface elevation over onto the velocity.''
+
 \item[L152] \textcolor{blue}{Cite the method of manufactured solutions.}
+
+Added a citation to Roache 2002, Code verification by the method of manufactured solutions.
 
@@
 stated, in addition to the reduction in the cost of computational solution.}
+
+3D meshing is something to be avoided, but we make a big deal elsewhere in the paper about using extruded meshes.
+With extruded meshes you only need to do 2D meshing, for which there are algorithms with much better guarantees, but solving 3D problems like the Stokes equations is still possible.
+We chose not to add any text here as it would then contradict or obscure some of the points we make later.
 
@@
 
+I don't think this is entirely correct.
+Imposing zero velocity at the boundary is easy, but imposing no normal flow or a fixed normal flow together with friction along the opposite directions is much more challenging.
+When the boundary is flat, you can set one component of the velocity to 0.
+But if the finite element basis for the velocity has degrees of freedom that are located at the mesh vertices, you have to confront the fact that the unit outward normal vector isn't uniquely defined at the mesh vertices.
+In the next sentence we cited a paper from the Elmer/Ice group where they devised an ad-hoc (but still very effective) scheme for defining the normal vector at mesh vertices and thus imposing this boundary condition.
+You can also do it with Nitsche's method for linear problems but to my knowledge no one has figured that out for power-law fluids.
+I think that the natural boundary condition you're thinking of in most weak formulations of the Stokes equations is that the average of the pressure is zero.
+
 \item[L367] \textcolor{blue}{I’m surprised by this. For challenging geometries, I’ve always needed to
@@
 is unconditionally stable even for non-linear advection like this?}
+
+See Donea and Huerta, \emph{Finite element methods for flow problems}, chapter 3, sections 4-6.
+We did have similar experiences to what you describe when preparing the MISMIP+ test case for this paper.
+The melt phase produces very high ablation rates concentrated in a small region right near the grounded line.
+In our initial setup, the mesh we used didn't adequately resolve this feature.
+The finite element interpolation errors can then have an oscillatory pattern that, while not directly amplified by the implicit Euler timestepping scheme as such, still persists and gives a nonsense solution.
+This behavior manifests even at timesteps substantially below the CFL timestep.
+High-amplitude oscillatory garbage in the thickness field can then result in unrealistically large driving stresses and crash the nonlinear solver for the velocity.
+The extra terms from the Lax-Wendroff method help to diffuse out these oscillatory features.
+Both schemes give good results when the mesh is sufficiently refined; the Lax-Wendroff scheme just require less refinement.
+So this could be more an issue of spatial resolution than stability of the timestepping scheme.
+You're right that saving users from having to think about the CFL condition doesn't completely alleviate all the difficulties and we state as much in the next sentence.
 
@@
 problem where forming the Hessian becomes prohibitive.}
-\item[L459] \textcolor{blue}{Gauss-Newton needs a reference if you’re not going to describe it here.
-Eq. 34 It’s worth noting that the Schoof law is phenomenological, and was
+
+One hard criterion we have is that the method needs to naturally map from the dual of the parameter space (where the gradient lives) back to the parameter space itself.
+Neglecting to do so often results in unspeakable horrors like the vertex degrees of freedom converging at a different rate than the edge degrees of freedom under mesh refinement.
+See Schwedes et al. 2017, \emph{Mesh Dependence in PDE-Constrained Optimisation}.
+Using the Hessian or an approximation does that.
+The acceleration tricks for first-order optimization methods are really amazing but I've yet to find a nice way to adapt them to problems posed over Sobolev spaces.
+By contrast, I know that I can scale second-order methods to larger problems than we're solving now by using matrix-free application of the Hessian and coming up with better preconditioners.
+
+\item[L459] \textcolor{blue}{Gauss-Newton needs a reference if you’re not going to describe it here.}
+
+We added a reference to Pratt et al. 1998 which, although focused on applications in seismology, I think does a better job describing it than any other reference.
+
+\item[Eq. 34] \textcolor{blue}{It’s worth noting that the Schoof law is phenomenological, and was
 selected because it has the right shape and obeys Iken’s bound. As such,
@@
 less valid.}
+
+This was stated around line 586 but it makes more sense to say that earlier in the text.
+We've moved the statement accordingly.
+
 \item[L580] \textcolor{blue}{I think that having to specify a variational principle for the sliding law
@@
 must be to be physical.}
+
+I'm not completely sure what you mean.
+If the basal shear stress has the wrong sign this is arguably just as easy to check by looking at the action functional as it is a nonlinear system of equations.
+If you mean that the action has to be convex then this isn't true, the sliding law could be rate-weakening, but then there might be multiple steady states.
+
 \item[L701–713] \textcolor{blue}{I enjoyed reading this paragraph, but I wonder if stabilizing the
@@
 would be a more appropriate case?}
+
+This would be a stronger point if we actually had a Stokes solver (it's in the works now) but the example was just to be illustrative.
 \end{description}

### 2021-03-17 icepack.tex
Most of reviewer 1 corrections

@@
 (In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
-The only approximation that the first-order model makes is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
-This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
-Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.
+The first-order model is based on an asymptotic expansion of the Stokes equations in the ratio of the ice thickness to a typical horizontal length scale.
+The aspect ratio of most glacier flows is on the order of 1/20 or less, although there are some exceptions.
+For example, the main trunk of Jakobshavn Isbrae in Greenland flows through a very deep and narrow trough with an aspect ratio closer to 1/5.
 
@@
 The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.
-Strictly speaking, the fluidity should be a tensor field, but is almost universally treated as a scalar.
+Strictly speaking, the fluidity should be a tensor field, but is almost universally treated as a scalar \citep{gillet2006flow}.
 
@@
 where $dz\,dx$ denotes integration over the entire glacier, $dx$ denotes integration over the glacier footprint, and $dz\,d\gamma$ over the side wall boundary.
-The action has units of power (energy/time) and can be related to the rate of decrease of the thermodynamic free energy.
-Moreover, the energy lost to viscous and frictional heating can be calculated from the action and its Legendre transform.
-The action principle can be viewed as a consequence of the Onsager reciprocity relations for systems near to equilibrium \citep{de2013non}.
+The action has units of power (energy/time) and can be related to the rate of decrease of the thermodynamic free energy \citep{de2013non}.
 
@@
 \end{equation}
-Additionally, the terminal stress term of the action disappears by applying integration by parts to the gravity term.
+Additionally, the terminal stress term of the action disappears after applying integration by parts to the gravity term to shift the gradient of the surface elevation over onto the velocity.
 Since the action functional for ice shelves has fewer terms than for grounded ice streams, we have defined a separate \texttt{IceShelf} model class.
@@
 To get an exact solution, we chose the ice velocity and thickness a priori and used the computer algebra system SymPy \citep{sympy} to generate a friction coefficient that makes these fields an exact solution.
-We then checked that numerical results converge to this manufactured solution at the expected rate.
+We then checked that numerical results converge to this manufactured solution at the expected rate \citep{roache2002code}.
 
@@
 
-Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian.
+Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian \citep{pratt1998gauss}.
 Each iteration of Gauss-Newton is more expensive than that of BFGS or gradient descent because it requires the solution of a more complex linear system than just the mass matrix.
@@
 To implement the Schoof sliding law we need to know the antiderivative of equation \eqref{eq:schoof-sliding}.
-Using a computer algebra system, we found that the antiderivative of this expression in terms of hypergeometric functions.
-The Unified Form Language has several transcendental functions (sine, cosine, exponential, etc.), but it does not currently support hypergeometric functions.
+We found the antiderivative of this expression using a computer algebra system, but the result includes hypergeometric functions.
+The Unified Form Language has several transcendental functions (sine, cosine, exponential, etc.) but it does not currently support hypergeometric functions.
 Instead, we implemented a sliding relation that exhibits the important features of the Schoof law, i.e.
@@
 Although figure \ref{fig:sliding-laws} shows a comparison of both sliding laws with the same value of the critical speed, by using different values, the agreement between our sliding law can be brought into much closer agreement with the Schoof law.
+Finally, the Schoof law itself is phenomenological.
+A different sliding law with the right asymptotic behavior is no more or less valid.
 
@@
 The numerical advantages of using variational principles are so great, however, that we view this tradeoff as acceptable.
-Moreover, there are few physical reasons to prefer one form over another, as long as the asymptotic behavior is correct above and below the critical speed.
 

### 2021-03-17 reviews/response-2.tex
Start on response to review 2

@@
 \item \textcolor{blue}{The authors suggest that C++ makes it inherently difficult to add new physics/PDEs (e.g., on p. 10
-1and p. 20), which I somewhat disagree with. One advancement that can make a C++-based code
+and p. 20), which I somewhat disagree with. One advancement that can make a C++-based code
 easy to add to is Automatic Differentiation (AD) - with AD, one can effectively code the weak form
@@
 more intimidating than a Python code, so I am not trying to minimize the authors’ efforts at all.}
+
+I've added a citation to Tezaur et al. and to another paper from the ISSM group about AD.
+I agree with you about AD tools: they definitely relieve the burden of having to rewrite the code to calculate variational derivatives of functionals of the solutions of the model with respect to input parameters upon changing the model or parameters.
+The experiences I had rolling my own adjoints are what pushed me to tools that would either have AD or a more symbolic approach like what FEniCS, Firedrake, and Devito offer.
+In idiomatic Python it's nonetheless possible to be much more flexible about function signatures than in C++ by virtue of being able to throw arbitrary data into \texttt{kwargs}.
+Now of course you pay for this in that all input validation is done at runtime rather than at compile time.
+
 \item \textcolor{blue}{Can the authors comment on the overhead of the symbolic descriptions/manipulations done by their
@@
 to being able to code up something “quick-and-dirty” for initial prototyping?}
+
+See Rathgeber et al. 2016 for more information about the architecture of Firedrake and for performance benchmarks.
+Loosely speaking, the path through the toolchain goes like this.
+A user creates a symbolic description of the weak form of the PDE they want to solve.
+Firedrake then computes a hash of this expression and looks to see if it has encountered this problem before.
+If not, it does a long, complicated, and expensive series of transformations to generate highly optimized C code that fills the relevant matrices and vectors.
+This C code is then compiled into a dynamic library for later reuse.
+If the user has solved this problem before, Firedrake simply looks up the dynamic library that it already generated.
+(Crucially, the only thing that matters is the symbolic shape of the problem, not the actual data that goes into it -- you don't have to do codegen all over again just because you changed the boundary conditions or forcing.)
+In either case, Firedrake then calls into PETSc's scalable nonlinear equation solvers (SNES) to solve the resulting system of equations.
+As with any just-in-time compiled language, performance is slow the first time the code is run, and faster ever after.
+Most importantly, \textbf{the performance-critical parts are all written in C}.
+Firedrake has been shown to scale up to large problems on thousands of processors.
+
+It is possible to ruin the performance of the application by accidentally hard-coding a floating point value into a symbolic expression of a PDE and then changing that value in a loop.
+For example you could easily make this mistake if you were doing adaptive timestepping on an evolutionary problem discretized via the method of lines.
+The remedy is to wrap this value in a \texttt{firedrake.Constant} object.
+But this kind of performance regression is easily caught using \texttt{htop} and are more a result of programmer error.
+
+For the problems that we have used icepack for so far, we have focused more on individual glaciers or catchments, and thus performance has been a secondary concern.
+We aim to move towards larger continental-scale problems in the future.
+The rate-limiting factor there is more our ability to find the right incantation of PETSc solver options and preconditioners than it is any inherent limitation in our tools.
+
 \item \textcolor{blue}{I was a little bit confused about the reference of the FO Stokes-based model in this paper as a “hybrid
@@
 Is the term “hybrid model” a common name for the approach in Section 2.2.3? Perhaps it is and I
-am not aware of it.}
-\item \textcolor{blue}{Does the hybrid model described in the paper have the same applicability as say the First Order
+am not aware of it.
+Does the hybrid model described in the paper have the same applicability as say the First Order
 Stokes model? Can it be used for both Greenland and Antarctica at continental scales?}
+
+The fundamental physics are the first-order model obtained by asymptotic expansion of the Stokes equations in the aspect ratio, also known as the Blatter-Pattyn equations.
+What I had imagined is that using only vertical basis functions up to degree 2 essentially defines its own semi-discrete physics model, similar to two- or three-layer ocean models.
+You can view these as very coarse discretizations of the primitive equations, or you can view them as simplified models in their own right.
+But I made this naming choice before there were many other collaborators on the project.
+The ensuing confusion has shown that this was a bad choice of terminology that we'll change in a future version.
+This model can be used for both regions at continental scales -- it can capture but plug and shear flow.
+
 \item \textcolor{blue}{Section 2.2.1: in my opinion, the authors do not provide sufficient justification for the penalty term,
@@
 something that the authors are thinking to add to their model/code?}
+
+Implementing this idea will require some new developments to Firedrake, namely first-class support for subdomains, defining different PDEs on different subdomains, and defining matching conditions for the solutions at the interfaces between subdomains.
+The Firedrake developers are working on this feature right now as it's very much in demand.
+We are very much interested in, for example, using SIA in the interior of the ice sheet and SSA in the ice streams and margins, as this would give a much less computationally-intensive way to do some form of whole-ice sheet modeling than using, say, the first order model.
 \item \textcolor{blue}{Section 2.4: This section is very incomplete. You need to give the enthalpy equation and given the
@@
 doing a line search for Newton. By the way, it should be no surprise that Newton is not converging
-3without a line search - in general Newton is not guaranteed to converge from an arbitrary initial
+without a line search - in general Newton is not guaranteed to converge from an arbitrary initial
 guess without the line search.}
@@
 emerging HPC architectures, e.g., GPUs?}
+
+Firedrake relies on the package loo.py for code generation, which can target C and OpenCL.
+There is ongoing work with the developers of loo.py to target GPUs and other accelerators by generating OpenCL instead.
+Firedrake is built on PETSc and thus can run on parallel machines.
+See Rathgeber et al.
+2016 for performance and scaling benchmarks; Firedrake has been run on problems with millions of degrees of freedom on supercomputers, for example the UK national supercomputer ARCHER.
+
 \item \textcolor{blue}{Section 4.6: Can you please clarify what you mean by the following statement? “Large problems,
@@
 books! (no need to address this comment)}
-\item \textcolor{blue}{p. 29: are you considering putting in the first-order Stokes//Blatter Pattyn model into your code
+\item \textcolor{blue}{p. 29: are you considering putting in the first-order Stokes/Blatter Pattyn model into your code
 framework?}
+
+The thing that we mistakenly called the ``hybrid'' model is really the first order / Blatter-Pattyn model.
+We have rewritten some of the text to try and make this clearer.
 \item \textcolor{blue}{The methods do not discuss their code development/testing stance on icepack. How do users con-
@@
 \item \textcolor{blue}{“UFL” is not defined in the abstract.}
+
+Changed the sentence to: ``Icepack is built on the finite element modeling library Firedrake, which uses the Unified Form Language (UFL), a domain-specific language embedded into Python for describing weak forms of partial differential equations.''
+
 \item \textcolor{blue}{p. 4, line 90: A is also called the “flow factor”. I would mention here that it is usually a function of
 the temperature, which comes from a different equation.}
-\item \textcolor{blue}{p. 5, line 125: should be “checking”, not “check”.}
-\item \textcolor{blue}{p. 5, line 126: I think it should be “Bueler” not “Beuler”, if I’m thinking of the right person.}
+\item \textcolor{blue}{p. 5, line 125: should be “checking”, not “check”.} \checkmark
+\item \textcolor{blue}{p. 5, line 126: I think it should be “Bueler” not “Beuler”, if I’m thinking of the right person.} \checkmark
 \item \textcolor{blue}{p. 5, line 147: change “we verified the correctness of the ice shelf model” to “we verified the correctness
-of our implementation of the ice shelf model”.}
-\item \textcolor{blue}{p. 10, line 270: change “we’ll” to “we will”.}
+of our implementation of the ice shelf model”.} \checkmark
+\item \textcolor{blue}{p. 10, line 270: change “we’ll” to “we will”.} \checkmark
 \end{itemize}

### 2021-03-17 icepack.tex
Start on response to review 2

@@
 We introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
-Icepack is built on the finite element modeling library Firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
+Icepack is built on the finite element modeling library Firedrake, which uses the Unified Form Language (UFL), a domain-specific language embedded into Python for describing weak forms of partial differential equations.
 The diagnostic models in icepack are formulated through action principles that are specified in UFL.
@@
 
-We verified the correctness of our implementation by check that numerical results converge at the expected rate to the analytical Beuler profile for a circular symmetric ice sheet \citep{greve2009dynamics}.
+We verified the correctness of our implementation by checking that numerical results converge at the expected rate to the analytical Bueler profile for a circular symmetric ice sheet \citep{greve2009dynamics}.
 
@@
 
-We verified the correctness of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
+We verified the correctness of our implementation of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
 When there is basal friction, the shallow stream equations do not have a simple analytical solution.
@@
 Using a dictionary data structure to pass arguments is relatively easier but would be more idiomatic in Python than in C++.
+Some of the difficulty associated with changing the model physics in a C++ code can be alleviated with automatic differentiation tools, which have been used successfully in Albany/FELIX \citep{tezaur2015albany} and ISSM \citep{huck2018usability}.
+Nonetheless, C++ or any language with a stronger type discipline will be much more restrictive about changing function signatures.
 
@@
 From the user's perspective, substituting model physics components does not require any advanced language features beyond keyword arguments.
-To understand how this is possible, we'll briefly describe the path that the input fields take through the program.
+To understand how this is possible, we will briefly describe the path that the input fields take through the program.
 \begin{enumerate}

### 2021-03-24 reviews/response-1.tex
Finish response to review 1

@@
 
-\textcolor{blue}{In ‘icepack: a new glacier flow modeling package in Python, version 1.0’, Shapero and co-authors present a promising new ice sheet modeling framework. The
+\textcolor{blue}{In `icepack: a new glacier flow modeling package in Python, version 1.0', Shapero and co-authors present a promising new ice sheet modeling framework. The
 framework contains mechanisms for solving both the prognostic mass balance
-equations for updating ice sheet geometry, as well as diagnostic solvers for approximations to the non-linear Stokes’ equations. Throughout both the software
+equations for updating ice sheet geometry, as well as diagnostic solvers for approximations to the non-linear Stokes equations. Throughout both the software
 and the manuscript describing it, the authors focus on ensuring usability (readability), a trait that is bound to make this software (and paper) frequently used.
@@
 
-\item[L74] \textcolor{blue}{A low aspect ratio isn’t really an approximation; it’s an existential fact.
+We've added one or two references for each.
+
+\item[L74] \textcolor{blue}{A low aspect ratio isn't really an approximation; it’s an existential fact.
 The approximation that the first order approximation makes is that vertical resistive stresses (or bridging stresses as they are referred to later
@@
 solution, the break in symmetry between Eq. 6 and Eq. 12 is frustrating,
-1given that they both are name ‘gravity’, and that they should in some
+given that they both are name `gravity', and that they should in some
 sense be the same regardless of which strain rates are assumed to be zero.}
@@
 
-\item[L367] \textcolor{blue}{I’m surprised by this. For challenging geometries, I’ve always needed to
+\item[L367] \textcolor{blue}{I'm surprised by this. For challenging geometries, I've always needed to
 stabilize even when using implicit Euler. Are you sure that implicit Euler
@@
 
-\item[L453] \textcolor{blue}{Many advances have occurred in the last 5 years regarding gradient de-
-scent due to its necessity for optimizing neural networks. These may yet
+\item[L453] \textcolor{blue}{Many advances have occurred in the last 5 years regarding gradient descent due to its necessity for optimizing neural networks. These may yet
 be useful in this context if you have to deal with a large scale optimization
@@
 
-\item[L459] \textcolor{blue}{Gauss-Newton needs a reference if you’re not going to describe it here.}
+\item[L459] \textcolor{blue}{Gauss-Newton needs a reference if you're not going to describe it here.}
 
@@
 
-\item[Eq. 34] \textcolor{blue}{It’s worth noting that the Schoof law is phenomenological, and was
-selected because it has the right shape and obeys Iken’s bound. As such,
-if your sliding law obeys Iken’s bound and looks right, then it’s not any
+\item[Eq. 34] \textcolor{blue}{It's worth noting that the Schoof law is phenomenological, and was
+selected because it has the right shape and obeys Iken's bound. As such,
+if your sliding law obeys Iken's bound and looks right, then it's not any
 less valid.}
@@
 
-\item[L701–713] \textcolor{blue}{I enjoyed reading this paragraph, but I wonder if stabilizing the
-Stokes’ equations is the best illustration of the point, given that icepack
-does not in fact have Stokes’ equations implemented (although I imagine
+\item[L701-713] \textcolor{blue}{I enjoyed reading this paragraph, but I wonder if stabilizing the
+Stokes equations is the best illustration of the point, given that icepack
+does not in fact have Stokes equations implemented (although I imagine
 it could be done in short order). Maybe stabilizing a transport equation

### 2021-03-24 icepack.tex
Finish response to review 1

@@
 \begin{enumerate}
-    \item projecting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
-    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood,
-    \item estimating unobservable quantities, such as bed friction or rheology, from observational data, and
-    \item reconstructing what glaciers of the near- or distant-past may have looked like.
+    \item projecting future glacier extent and estimating the sea-level rise contribution from glacier dynamics \citep{joughin2014marine},
+    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood \citep{nick2010physically, werder2013modeling},
+    \item estimating unobservable quantities, such as bed friction or rheology, from observational data \citep{vieli2007causes, shapero2016basal}, and
+    \item reconstructing what glaciers of the near- or distant-past may have looked like \citep{marshall2002north}.
 \end{enumerate}

### 2021-03-24 reviews/response-2.tex
More responses to reviewer 2

@@
 
+I hope this paragraph made the point more that we had specific goals that could better be achieved in Python and not as some blanket condemnation of C++ -- I don't like being dogmatic about language choice.
 I've added a citation to Tezaur et al. and to another paper from the ISSM group about AD.
@@
 In idiomatic Python it's nonetheless possible to be much more flexible about function signatures than in C++ by virtue of being able to throw arbitrary data into \texttt{kwargs}.
-Now of course you pay for this in that all input validation is done at runtime rather than at compile time.
+Now of course you pay for this in that all input validation is done at runtime rather than at compile time, but it's a tradeoff we had to make.
 
@@
 The ensuing confusion has shown that this was a bad choice of terminology that we'll change in a future version.
-This model can be used for both regions at continental scales -- it can capture but plug and shear flow.
+This model can be used for both regions at continental scales -- it can capture both plug and shear flow.
 
@@
 or 3D depending on which approximation one is looking at.}
+
+These were not stated explicitly anywhere.
+We've added them to table 1.
 \item \textcolor{blue}{The boundary conditions are not discussed very rigorously systematically - the authors seem to
@@
 be made more precise.}
+
+This was stated in table 1 but we've added the definition as $\frac{1}{2}(\nabla u + \nabla u^\top)$ and added a sentence to the text referring readers to table 1.
 \item \textcolor{blue}{Section 2.2.3: the authors comment that higher degree polynomials can be used in the vertical layer
 in the hybrid approach. What order is typically used?}
-\item \textcolor{blue}{Section 2.2.3: this might be a naive question, but does Glen’s flow law come into the hybrid model?
-I was expecting to see it there, but maybe I’m missing something.}
+
+Added the following text: ``Going up to a degree-4 model is sufficient to capture the exact solution for the shallow ice approximation.
+In the tutorial notebooks for icepack, we use up to degrees 2 and 4, but the test suite checks up to degree 8.''
+\item \textcolor{blue}{Section 2.2.3: this might be a naive question, but does Glen's flow law come into the hybrid model?
+I was expecting to see it there, but maybe I'm missing something.}
 \item \textcolor{blue}{P. 10: the discussion here about substituting model components suggests it may be possible to use
@@
 \item \textcolor{blue}{Section 2.4: This section is very incomplete. You need to give the enthalpy equation and given the
-Glen’s law expression as well, since it is mentioned.}
+Glen's law expression as well, since it is mentioned.}
 \item \textcolor{blue}{In my opinion, there is not enough discussion of the thickness equation (Section 2.1) and how it
 is discretized. In typical ice sheet models, this equation is used to change the ice extent - one
-meshes up a region of “potential” ice, and then uses the thickness to dynamically determine a mask
+meshes up a region of ``potential'' ice, and then uses the thickness to dynamically determine a mask
 for ice-covered regions. Do you do something like this in your model? It should be discussed for
@@
 Structured/unstructured? Hex/tet (quad/tri)?}
+
+We use unstructured triangular meshes although Firedrake in principle can use unstructured quad meshes.
 \item \textcolor{blue}{Section 3: It is not clear from the description what the inverse problem you are describing is for.
-Is it to obtain parameters in the model like the basal friction using observational data of e.g. sur-
-face velocity? There really needs to be more discussion here, and I personally would like to see
+Is it to obtain parameters in the model like the basal friction using observational data of e.g. surface velocity? There really needs to be more discussion here, and I personally would like to see
 a mathematical statement of a representative inverse problem you are solving. It would be worth
@@
 equation here - that should be made clear. I disagree with several statements in this section as
-well. “The simplest explicit timestepping schemes are unstable with CG finite elements” - if you are
+well. ``The simplest explicit timestepping schemes are unstable with CG finite elements'' - if you are
 talking about CFL stability, this is not true. You need to satisfy a CFL condition which could give
-rise to very small time-steps but you can get the scheme to be stable. I’m also confused about the
+rise to very small time-steps but you can get the scheme to be stable. I'm also confused about the
 notion of SUPG as a time-stepping scheme - I think of SUPG as a finite element approach to deal
@@
 forward Euler, so the discussion about forward Euler requiring parameters is erroneous. Additionally,
-I don’t understand the comment about implicit Euler smoothing out sharp discontinuities... I believe
+I don't understand the comment about implicit Euler smoothing out sharp discontinuities... I believe
 explicit and implicit Euler have effectively the same diffusion and dispersion properties, so there
@@
 guess without the line search.}
+
+Added a reference to the Albany paper as well as another one on trust region methods.
+We described a fairly rudimentary line search method in more detail than perhaps is necessary for a reader who's a seasoned modeler.
+It's a bit of a pet peeve of mine when papers just say ``We used Newton!'' without any attention to the globalization strategy, which can make a huge difference to the solver robustness.
 \item \textcolor{blue}{Section 4.3, lines 411-412: there are actually ways to construct weighted norms to deal with the issue
 of DOFs having different orders of magnitude for the purpose of convergence.}
+
+I think you can use a lumped mass matrix as the $H_0$ in BFGS too, but I find it to be far preferable to use something like Gauss-Newton which gives mesh-independent convergence and which achieves close to the second-order rate of full Newton on many problems.
+No change to the text.
 \item \textcolor{blue}{Section 4.4: Again, it is not clear to me what is your inverse problem. You need to state this explicitly
@@
 
-\item \textcolor{blue}{Section 4.6: Can you please clarify what you mean by the following statement? “Large problems,
+\item \textcolor{blue}{Section 4.6: Can you please clarify what you mean by the following statement? ``Large problems,
 such as continental-scale modeling, will require more sophisticated and possibly problem-specific
-approaches”. I’m wondering in particular about the problem-specific approach part. There are
+approaches''. I’m wondering in particular about the problem-specific approach part. There are
 models like first order Stokes that can be used at the continental scale and they are not really
 problem specific.}
-\item \textcolor{blue}{p. 20: I don’t understand why you need to create an analytical expression of (32) using special
+
+We have clarified the text to state that ``problem-specific'' refers more to the strategies we use to solve the resulting nonlinear systems of equations rather than to what equations are being solved, e.g.
+Stokes vs first-order Stokes:
+``For the demonstrations presented below, nearly all simulations run in a matter of minutes to hours on a single core.
+We have used sparse LU factorization to solve linear systems for many problem instances in order to eliminate the linear solver as a possible failure mode.
+Defaulting to a robust solution method is especially important for onboarding novice users who may not be familiar with different iterative linear solvers and preconditioners.
+Larger problems, such as continental-scale modeling, will require solving the diagnostic equations using the conjugate gradient method with an appropriate preconditioner to achieve parallel scalability.
+The particular structure of the problems we solve may be useful in choosing a preconditioner.
+For example, a rudimentary preconditioner for the hybrid model system could use the degree-0 model as the coarse space in a multigrid-type approach.
+These optimizations will be the subject of future work.''
+
+Since we use a modal basis in the vertical to discretize solutions of the 3D model, we can devise a p-type multigrid scheme along this axis.
+This is an example of using problem-specific knowledge to choose a solution strategy.
+Just using LU or throwing a black-box algebraic multigrid preconditioner at it would be failing to use this special structure.
+
+\item \textcolor{blue}{p. 20: I don't understand why you need to create an analytical expression of (32) using special
 functions. Is this something specific to your framework, which requires expressions in a certain form
 for the symbolic representation?}
-\item \textcolor{blue}{It’s great that you have executable documentation in something easy-to-use such as Jupyter note-
-books! (no need to address this comment)}
+
+We don't need an analytical expression of equation 32, but rather of the antiderivative of that function with respect to $u$.
+This problem is specific to icepack because we have made the choice to use action principles to describe all of the diagnostic models.
+A package that also made the choice to use action principles but which was built on a different finite element modeling library or coded in an entirely different language would also need the antiderivative of this function.
+We are hampered by the fact UFL does not include support for hypergeometric functions.
+If we were instead writing everything from scratch in C++, we could call into Boost or GSL to evaluate hypergeometric functions.
+We believe that the benefits outweight the costs but this is a definite drawback of our approach.
+\item \textcolor{blue}{It's great that you have executable documentation in something easy-to-use such as Jupyter notebooks! (no need to address this comment)}
 \item \textcolor{blue}{p. 29: are you considering putting in the first-order Stokes/Blatter Pattyn model into your code
@@
 \begin{itemize}
-\item \textcolor{blue}{“UFL” is not defined in the abstract.}
+\item \textcolor{blue}{``UFL'' is not defined in the abstract.}
 
@@
 
-\item \textcolor{blue}{p. 4, line 90: A is also called the “flow factor”. I would mention here that it is usually a function of
+\item \textcolor{blue}{p. 4, line 90: A is also called the ``flow factor''. I would mention here that it is usually a function of
 the temperature, which comes from a different equation.}
-\item \textcolor{blue}{p. 5, line 125: should be “checking”, not “check”.} \checkmark
-\item \textcolor{blue}{p. 5, line 126: I think it should be “Bueler” not “Beuler”, if I’m thinking of the right person.} \checkmark
-\item \textcolor{blue}{p. 5, line 147: change “we verified the correctness of the ice shelf model” to “we verified the correctness
-of our implementation of the ice shelf model”.} \checkmark
-\item \textcolor{blue}{p. 10, line 270: change “we’ll” to “we will”.} \checkmark
+\item \textcolor{blue}{p. 5, line 125: should be ``checking'', not ``check''.} \checkmark
+\item \textcolor{blue}{p. 5, line 126: I think it should be ``Bueler'' not ``Beuler'', if I'm thinking of the right person.} \checkmark
+\item \textcolor{blue}{p. 5, line 147: change ``we verified the correctness of the ice shelf model'' to ``we verified the correctness
+of our implementation of the ice shelf model''.} \checkmark
+\item \textcolor{blue}{p. 10, line 270: change ``we'll'' to ``we will''.} \checkmark
 \end{itemize}

### 2021-03-24 icepack.tex
More responses to reviewer 2

@@
         $\dot a_b$ & basal mass balance \\
-        $\dot\varepsilon$ & strain rate \\
+        $\dot\varepsilon$ & strain rate, $\frac{1}{2}(\nabla u + \nabla u^\top)$\\
         $C$ & bed friction coefficient \\
@@
         $\nu$ & unit outward normal \\
-        $J$ & action functional
+        $J$ & action functional \\
+        $\Omega$ & spatial domain \\
+        $\Gamma$ & calving terminus
     \end{tabular}
     \caption{Mathematical symbols}
+    \label{tab:mathematical-symbols}
 \end{table}
@@
     \caption{Physical constants}
+    \label{tab:physical-constants}
 \end{table}
+
+In this section, we'll describe the mathematical formulation of each of the models implemented in icepack.
+See table \ref{tab:mathematical-symbols} for definitions of all mathematical symbols and table \ref{tab:physical-constants} for physical constants.
 
@@
 Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
-The degree 2 model is the most minimalistic model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
-Going to higher degree gives a more accurate approximation at the expense of greater computational effort.
-
-\citet{brinkerhoff2015dynamics} used a similar approach to the one described above, with a one vertical basis function for plug flow and one for shear flow.
+The degree-2 model is the most minimalistic model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
+Going up to a degree-4 model is sufficient to capture the exact solution for the shallow ice approximation.
+In the tutorial notebooks for icepack, we use up to degrees 2 and 4, but the test suite checks up to degree 8.
+
+\citet{brinkerhoff2015dynamics} used a similar approach to the one described above, with one vertical basis function for plug flow and one for shear flow.
 The shear flow basis function was chosen to be exact assuming the SIA balance with cold ice, together with a heuristic approximation for polythermal ice.
@@
 By contrast, a damped Newton procedure using a line search that satisfies the Armijo-Wolfe criteria is guaranteed to converge on non-degenerate, if unrealistic, input data.
+The line search method that we used here is one way to achieve global convergence of Newton's method, but there are other approaches to achieve global convergence, for example homotopy continuation \citep{tezaur2015albany} or the trust region method \citep{bellavia2007globalization}.
 
@@
 For the demonstrations presented below, nearly all simulations run in a matter of minutes to hours on a single core.
-Larger problems, such as continental-scale modeling, will require more sophisticated and possibly problem-specific approaches.
+We have used sparse LU factorization to solve linear systems for many problem instances in order to eliminate the linear solver as a possible failure mode.
+Defaulting to a robust solution method is especially important for onboarding novice users who may not be familiar with different iterative linear solvers and preconditioners.
+Larger problems, such as continental-scale modeling, will require solving the diagnostic equations using the conjugate gradient method with an appropriate preconditioner to achieve parallel scalability.
+The particular structure of the problems we solve may be useful in choosing a preconditioner.
 For example, a rudimentary preconditioner for the hybrid model system could use the degree-0 model as the coarse space in a multigrid-type approach.

### 2021-03-25 reviews/response-2.tex
More responses to reviewer 2

@@
 Glen's law expression as well, since it is mentioned.}
+
+We have added some text and equations describing the model we used, the boundary conditions, and the shear heating rate, the latter of which implicitly includes Glen's flow law.
 \item \textcolor{blue}{In my opinion, there is not enough discussion of the thickness equation (Section 2.1) and how it
@@
 be added).}
+
+We added a brief description of the mathematics of the inverse problem to be solved.
+We also added: ``The state to be estimated can be any single input field to the diagnostic model -- basal friction, rheology, or another field that the user has added by customizing the model.''
+
 \item \textcolor{blue}{Section 4.1: I find this section confusing. I assume you are talking about discretizing the thickness
@@
 so it is clear.}
-\item \textcolor{blue}{Section 4.5.1: you talk about problems defined on “extruded geometries” - do you ever use non-
+
+See correction to section 3.
+The point of the inverse solver class is that it is very general with respect to what field is being inferred and what diagnostic model is being used.
+Since the solver was designed to work for many different inverse problems
+\item \textcolor{blue}{Section 4.5.1: you talk about problems defined on ``extruded geometries'' - do you ever use non-
 extruded geometries? It has been shown in various references that there can be numerical problems
 for land-ice solvers that do not use extruded geometries, e.g. Tezaur et al.}
+
+We have restricted our implementation to use only extruded geometries.
+Having made this choice, we might not ever be able to solve really geometrically complex problems like what the Elmer/Ice crowd did with the drainage of the lake underneath T\^ete Rousse glacier.
+But the simplifications that this results in for the vast majority of glaciological applications are so advantageous that it's a sacrifice we're willing to accept.
 \item \textcolor{blue}{Section 4.5.2: I think this section needs to be made earlier, and other BCs need to be added to that

### 2021-03-25 icepack.tex
Change Marshall ref to Huybrechts 2002
More responses to reviewer 2

@@
     \item estimating unobservable quantities, such as bed friction or rheology, from observational data \citep{vieli2007causes, shapero2016basal}, and
-    \item reconstructing what glaciers of the near- or distant-past may have looked like \citep{marshall2002north}.
+    \item reconstructing what glaciers of the near- or distant-past may have looked like \citep{huybrechts2002sea}.
 \end{enumerate}
@@
         $A$ & rheology coefficient \\
+        $E$ & enthalpy density \\
         $\nu$ & unit outward normal \\
@@
         $\rho_W$ & seawater density \\
+        $c$ & specific heat capacity of ice \\
+        $L$ & latent heat of melting of ice \\
         $g$ & gravitational acceleration
@@
 We implemented the enthalpy transport model described in \citet{aschwanden2012enthalpy}.
-The enthalpy describes the heat content of the material in a way that incorporates both temperature and latent heat stored in meltwater.
+The enthalpy density $E$ describes the heat content of the material in a way that incorporates both temperature $T$ and latent heat stored in meltwater:
+\begin{equation}
+    E = \rho_I(cT + Lf)
+\end{equation}
+where $L$ is the latent heat of melting of ice and $f$ is the melt fraction.
+The melt fraction can only be positive when the ice as at the pressure-melting point.
 The temperature and meltwater fraction can be uniquely calculated at any point from the value of the enthalpy, so nothing is sacrificed in describing heat content one way or the other.
@@
 Our implementation of the model uses all of the simplifying assumptions described in \citet{aschwanden2012enthalpy}, for example that horizontal diffusion is negligible and that heat capacity and conductivity are not temperature-dependent within each phase.
-
+The resulting PDE is
+\begin{equation}
+    \left(\frac{\partial}{\partial t} + \nabla\cdot u\right)E - \frac{\partial}{\partial z}\alpha\frac{\partial E}{\partial z} = q
+\end{equation}
+where $u$ is the full 3D velocity, $\alpha$ the (temperature-dependent) thermal diffusivity, and $q$ the volumetric heat sources.
+This form assumes some amount of vertical diffusion of water at the melting point.
+A realistic treatment would require treating the ice as a porous medium and modeling englacial water transport explicitly, which we have not done.
+The boundary condition at the ice base is a fixed flux of heat from geothermal sources or from the oceans.
+We depart from \citet{aschwandend2012enthalpy} in one respect.
+Rather than fix the ice surface temperature to the atmospheric temperature, we instead use a Robin boundary condition that makes the ice temperature adjust to the external temperature:
+\begin{equation}
+    -\alpha\frac{\partial E}{\partial z}\Big|_{z = s} = \kappa(E - E_s)
+\end{equation}
+where $\kappa$ is a surface exchange coefficient and $E_s$ is the temperature-equivalent enthalpy that forces the surface.
+Using a Robin instead of a Dirichlet boundary condition at the surface allows for more gradual adjustment to above-freezing temperatures that might otherwise create areas of unrealistically high melt fraction.
+Except in blue ice areas, the thermal contact between ice and the atmosphere is mediated by firn, which we do not model explicitly.
+The Robin boundary condition is a compromise to work around this missing component of the system.
+Finally, if there is an upstream boundary the user must provide the inflow values of the enthalpy with depth, while outflow boundary conditions are applied at the glacier terminus.
+
+The heat transport model assumes that the user will supply a volumetric heating rate, but the model itself does not calculate the heating rate from other fields.
+The simplest description for the shear heating rate is
+\begin{equation}
+    q = \tau : \dot\varepsilon = A^{\frac{-1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}.
+\end{equation}
 The fluidity factor $A$ in Glen's law is roughly a known function of both temperature and melt fraction and we have included this function in the package.
 Having some description of heat transport is thus necessary for describing realistic glacier flows, but it is not sufficient by itself.
-For one, while the dependence of fluidity on temperature is known fairly well from laboratory experiments, the dependence on melt fraction is known with much less certainty.
+While the dependence of fluidity on temperature is known fairly well from laboratory experiments, the dependence on melt fraction is known with much less certainty \citep{cuffey2010physics}.
 It is likely that users of this module will want to substitute in their own parameterization for melt fraction dependence.
-Second, other processes such as damage, fabric development, and impurities can affect the fluidity as well.
-For these reasons, the diagnostic model takes in the fluidity as an argument and users must calculate the fluidity field themselves, either using the default parameterization that icepack supplies or one of their own.
+Other processes such as damage, fabric development, and impurities can affect the fluidity as well.
+Moreover, there may be other volumetric heat sources that users wish to account for, such as cryo-hydrologic warming \citep{phillips2010cryo}.
+These considerations defy any attempt to have one function that calculates the volumetric heating rate from the other state variables.
+Instead, we opted for an interface where users calculate this rate themselves, with tutorials and examples that show how to do so for simple scenarios, leaving the freedom to alter the thermal feedbacks as they see fit.
 Our general design principle is that icepack will solve differential equations for the various prognostic fields but it's up to users to decide how these fields are coupled.
@@
 Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
+Mathematically, this inverse problem amounts to finding a critical point of the functional
+\begin{equation}
+    L(u, \theta, \lambda) = J(u) + R(\theta) + \langle F(u, \theta), \lambda\rangle
+\end{equation}
+where $J$ is the misfit between the computed velocity $u$ and observations, $\theta$ is the field to be inferred and $R$ a regularization functional that measures the spatial variability of this field, $F$ is the diagnostic physics, and $\lambda$ a Lagrange multiplier to enforce the physics \citep{macayeal1992basal, joughin2004basal, larour2005rheology, shapero2016basal}.
+The optimal value $\theta$ gives the best fit to observations subject to the constraints of the physics and that it should not overfit to noise in the data.
 The class \texttt{InverseProblem} represents the specification of an inverse problem, which requires:
@@
 This class currently assumes that the observed state is always the ice velocity.
+The state to be estimated can be any single input field to the diagnostic model -- basal friction, rheology, or another field that the user has added by customizing the model.
 In principle the same design would suffice for more complicated inverse problems and this is an area of active development.
@@
 \begin{equation}
-    E(u - u^o) = \int_\Omega\left(\sqrt{\frac{|u - u^o|^2}{\sigma^2} + \gamma^2} - \gamma\right)dx.
+    J(u) = \int_\Omega\left(\sqrt{\frac{|u - u^o|^2}{\sigma^2} + \gamma^2} - \gamma\right)dx.
     \label{eq:l1-error-metric}

### 2021-03-31 reviews/response-2.tex
Start on responses to reviewer 3

@@
 
-\item \textcolor{blue}{I was a little bit confused about the reference of the FO Stokes-based model in this paper as a “hybrid
-model”. I see that it is hybrid in the sense that you have a different discretization in the horizontal and
+\item \textcolor{blue}{I was a little bit confused about the reference of the FO Stokes-based model in this paper as a ``hybrid
+model''. I see that it is hybrid in the sense that you have a different discretization in the horizontal and
 vertical direction, but there are also hybrid ice models that use different PDEs in different domains,
 e.g., the ISCAL model of Ahlkrona et al. (https://doi.org/10.1016/j.quascirev.2016.01.032).
-Is the term “hybrid model” a common name for the approach in Section 2.2.3? Perhaps it is and I
+Is the term ``hybrid model'' a common name for the approach in Section 2.2.3? Perhaps it is and I
 am not aware of it.

### 2021-03-31 reviews/response-3.tex
Start on responses to reviewer 3

@@
 
+This depends on what kind of expertise you mean.
+We believe that more work can be done to relieve glaciologists \emph{who are not experts in scientific computing} from having to understand, e.g., the details of what preconditioners are used in solving linear systems, what globalization strategy is used in a Newton solver, etc.
+By way of an ``aggressive scheme for putting reasonable defaults into the hands of novices'', we describe this policy at the end of section 6, beginning in line 695.
+We do not claim to have reduced the need for expertise in glaciology itself, i.e. understanding mass and energy balance, etc.
+We also consider having some understanding of the different boundary conditions that might apply to the system to be part of this knowledge.
+The text has been expanded to describe these boundary conditions in more detail (see comments that follow).
+Granted, (1) there is more demonstration work that we need to do in order to cover use cases like modeling a small mountain glacier and (2) there are technical hurdles we still need to overcome to do a better job handling the margins.
+Very likely your recent work on variational inequalities will factor into improvement on the second front.
+Addressing these concerns within this manuscript would result in a dramatic increase in scope.
+
 \textcolor{blue}{21--39:  I also think Python+Firedrake+PETSc is the most promising environment to build a new ice flow simulation library/model.  I'm on board!}
@@
 (Indeed one should remind the reader that the latter describes thickness evolution because of how glacier models normally parameterize fluid geometry.)  An "also known as the 'prognostic equation' [cite]" is appropriate, but another paper using this tribal language will cause yet more students to need to unlearn silly language in order to read the mainstream fluids and numerical PDE literature.}
+
+I agree with you that this isn't great terminology.
+But I didn't think it was a fight I could win so I went with the rest of the glaciological literature and used these names.
+By this point the nomenclature is used in so many places throughout the package that switching would be a breaking change for existing users.
 

### 2021-04-01 reviews/response-3.tex
More responses to reviewer 3

@@
 \usepackage{natbib}
+\usepackage{hyperref}
+\hypersetup{
+    colorlinks=true,
+    linkcolor=blue,
+    filecolor=magenta,
+    urlcolor=magenta,
+}
 
@@
 We believe that more work can be done to relieve glaciologists \emph{who are not experts in scientific computing} from having to understand, e.g., the details of what preconditioners are used in solving linear systems, what globalization strategy is used in a Newton solver, etc.
-By way of an ``aggressive scheme for putting reasonable defaults into the hands of novices'', we describe this policy at the end of section 6, beginning in line 695.
+By way of an ``aggressive scheme for putting reasonable defaults into the hands of novices'', we describe our policy at the end of section 6, beginning in line 695.
+Other software packages have (exactly as you say) all the mechanisms, but do not display a consistent policy.
+For example, several of the demonstration codes in Elmer default to choices of iterative solvers and preconditioners that work on the particular problem in the example but which are likely to require hand-tuning when generalized to real data.
+See for example \href{https://github.com/ElmerCSC/elmerfem/blob/aaa6c4a345c73ba248c475d27975c5e3a9fa7758/elmerice/examples/Inverse_Methods/MacAyeal_Stokes/SIF/OPTIM_CONT.sif#L161}{this line} from Elmer's demonstration of an inverse solver, which uses the GCR scheme and an ILU0 preconditioner to solve the momentum equation.
+A graduate student could easily try to take this code and use it on real data only to have convergence failures to diagnose.
+Other examples from Elmer do use direct solvers, for example their \href{https://github.com/ElmerCSC/elmerfem/blob/aaa6c4a345c73ba248c475d27975c5e3a9fa7758/elmerice/examples/Test_SSA/ismip_SSA_2D.sif#L105}{demonstration code} to run the ISMIP test case with the SSA model uses a direct solver from UMFPACK for the momentum equation.
+The mechanisms are there, but a consistent policy is not.
+Other packages show a similar pattern.
+For example, BISICLES has hand-tuned numbers of Picard iterations and multigrid smoothing iterations in their demonstration codes.
+A new user either must (1) know enough about numerical methods to diagnose the problem from bad input data, bad linear solver options, or bad nonlinear solver options, or (2) have a mentor who can.
+Either way this creates unnecessary friction for people who may be getting their first experience of modeling by using these examples.
+This is not to put down or undermine the incredible technical work that the developers of Elmer, BISICLES, and other packages have done over the years by any means; it's just a reflection of what knowledge the users need to have going in.
+
 We do not claim to have reduced the need for expertise in glaciology itself, i.e. understanding mass and energy balance, etc.
@@
 Very likely your recent work on variational inequalities will factor into improvement on the second front.
-Addressing these concerns within this manuscript would result in a dramatic increase in scope.
 
@@
 
-I agree with you that this isn't great terminology.
-But I didn't think it was a fight I could win so I went with the rest of the glaciological literature and used these names.
-By this point the nomenclature is used in so many places throughout the package that switching would be a breaking change for existing users.
+I agree with you that this is not great terminology, but this isn't a fight we could win.
+Many other software packages for glacier flow modeling, including PISM and ISSM, use this terminology.
+By now the nomenclature is used in so many places throughout icepack that switching would be a breaking change for existing users.
 
@@
 
+Changed from ``momentum transport'' to ``stress balance''.
+
 \textcolor{blue}{74, B:  No, the Blatter-Pattyn (BP) equation is not a "hybrid".
@@
 
+This was bad terminology.
+I adopted it before there were many other contributors to the project and everyone who has come along afterwards has told me that it's confusing.
+What I had imagined is that a discretization of BP using only a single vertical layer and relatively low-degree (up to 4) polynomials in the vertical direction could be considered a kind of approximate model in its own right, similar to how oceanographers will use semi-discrete two- or three-layer ocean models.
+In the next release we will change the name to ``Blatter-Pattyn'' or ``First Order'' and deprecate the name ``Hybrid'' for removal in a future release or until we develop an actual hybrid model which is less computationally intensive than this Thing That I Should Not Have Called a Hybrid Model.
+
 \textcolor{blue}{101--102, B:  Regarding the sentence "The action principle can be viewed as a consequence of the Onsager reciprocity relations for systems near to equilibrium ...":  I am trying to imagine a reader for whom reading that would be a useful learning experience regarding the action principle, as opposed to an irritating one.}
+
+The previous two sentences have been removed from the text but we've kept the reference to the book by De Groot and Mazur.
+When I've spoken to students about action principles and why they're useful, I usually highlight the numerical advantages of solving convex optimization problems as opposed to the more general problem of solving large nonlinear systems of equations.
+But the question invariably comes up as to why a given physics problem should have a minimization principle at all while others do not.
+Non-equilibrium thermodynamics provides some insight and I think we would be doing a disservice by not mentioning this at all.
 
@@
 
+Added a forward reference to section 4.2 and changed this sentence to ``Algorithms for minimizing convex functionals have better convergence guarantees than algorithms for solving general nonlinear systems of equations while having no additional computational cost'' with a reference to Nocedal and Wright.
+
 \textcolor{blue}{115, equation (5), B:  Now the bad linguistic habits are being inherited from the FE world.
@@
 My suggestion for this term, implicitly acknowledging how it arises in the SIA, is to call it "localization".}
+
+Changed to ``localization''.
 
@@
 
+Apologies, the text has been corrected.
+
 \textcolor{blue}{137--140:  There is a double negative in equations (12) and (14) which is not used in (6) and (8), respectively.  Does this reflect anything important?}
@@
 
+Added a reference to Bueler 2014 and Böðvarsson 1955.
+You make a great point about testing the wrong parts of a nonlinear phase space and this was actually a bit of a struggle when creating the tests for this solver.
+You could put in some thickness and velocity fields and a PDE and SymPy will spit out a manufactured friction, but there's no guarantee that the result will be physically reasonable.
+I had to manually adjust the input parameters so that the basal shear stress values would come out to something sensible.
+A sentence has been added to the text to describe this part of the testing process.
+
 \textcolor{blue}{158:  I guess the phrase "individual fast-flowing glacier" arises because of the sense that the SSA does not handle slow flow very well, which is true, and that fast ice is separated by slow flow.
 Nonetheless, I would replace "features of an individual fast-flowing glacier" by "features of fast-flowing glaciers".}
+
+Done.
 
@@
 The novice reader not already used to FE stuff could use a figure or more words here, I suspect.}
+
+That was a typo, fixed.
 
@@
 
+Done.
+
 \textcolor{blue}{216:  Instead of "Lagrange interpolating polynomial basis", which readers accustomed to numerical analysis terminology (outside FE) might associate to the Lagrange method of finding interpolating polynomials, it is probably clearer to just say "Lagrange elements".
@@
 
+Done and added a reference to Szab\'o et al 2004.
+
 \textcolor{blue}{227:  "Smoke test" is an unfamiliar idiom to me.  As in testing machinery or testing electronics or smoking-out bugs or what-am-I-smoking?  Pity the non-native speaker.}
 
+The term comes from electronics.
+``Plug the circuit board in.
+If you see smoke coming out, unplug it.
+You don't need to do any more testing.''
+Changed to ``sanity test''.
+
 \textcolor{blue}{227:  "most minimalistic" --> "minimal"}
+
+Done.
 
@@
 No need to consider or mention it.}
+
+I was anticipating criticism from Doug Brinkerhoff.
+These sentences have been removed.
 
@@
 
+Done.
+
 \textcolor{blue}{300--301, C:  I seriously doubt that more than one user in 100 will want to substitute their own melt-fraction dependence, because no one has the data for it.
@@
 (This is a clear case where good defaults are more important than modularity.)}
+
+Changed to ``Some users of this module may want to substitute in their own parameterization for melt fraction dependence.''
+I think the fact that melt fraction dependence is not constrained very well means users are \emph{more} likely to change this in their code than if we did have good data.
+After all, if the data were good then we could just use whatever parameterization the experimentalists told us to and no one would ever have to touch it.
 
@@
 
+I think this falls under the heading of expert knowledge of glaciology and not of numerics.
+To give an extreme example, early on in grad school I was making a simulation of several Greenland glaciers and my officemate gave me the helpful advice to increase the fluidity of ice that would have accumulated during the last glacial period because of the dustier atmosphere.
+To use your phrase from before, we provide policies about numerics and mechanisms for physics.
+
 \textcolor{blue}{318--320:  Mushy sentences.  Needed?}
+
+We wanted to emphasize that this model is very ad hoc and that we're not passing it off as more accurate than it is.
+These sentences have been removed.
 
@@
 \textcolor{blue}{368:  "Practicing ... may be unfamiliar with" is unnecessary.  Just say what you have to say.}
+
+Changed to ``Some glaciologists may be unfamiliar with the Courant-Friedrichs-Lewy (CFL) condition.''
 
@@
 
+Text has been slightly reorganized, and we added ``Icepack currently lacks an adaptive timestepping scheme.
+Unconditionally stable schemes allow taking long timesteps, but taking very long timesteps will give inaccurate solutions.
+At present, users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
+Adaptive timestepping will be added in a future release.''
+
+This is one of the areas where we don't (yet) live up to our goal of keeping glaciologists from having to worry about numerics.
+Our plan is to use the package firedrake\_ts, which offers an interface to the PETSc timestepping schemes.
+
 \textcolor{blue}{363:  "equation equation"}
+
+Fixed.
 
@@
 \textcolor{blue}{459--462:  Gauss-Newton is a good choice for this purpose, as pointed out for exactly this purpose, inversion of glacier models, by [Habermann et al 2012].  Which should be cited.}
+
+Added a citation to Pratt et al 1998 and Habermann et al 2012.
 
@@
 
+This sentence has been removed.
+
 \textcolor{blue}{666--713:  Saying these HCI principles out loud is a very worthwhile aspect of the manuscript.}
@@
 \textcolor{blue}{721:  "phyics"}
+
+Fixed.
 

### 2021-04-01 icepack.tex
More responses to reviewer 3

@@
 
-There are four \emph{diagnostic} or \emph{momentum transport} models implemented in icepack.
+There are four \emph{diagnostic} or \emph{stress balance} models implemented in icepack.
 For each of the diagnostic models, we use a formulation of the physics based on a \emph{minimization} or \emph{action principle} \citep{dukowicz2010consistent}.
@@
 
-The action principle is especially useful for designing a robust numerical solver.
+The action principle is especially useful for designing a robust numerical solver, the implementation of which will be described in \S\ref{subsec:convex-optimization}.
 For viscous flow problems near to steady-state, the action is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
 Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge from any initial guess.
-Minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation while having no additional computational cost.
+Algorithms for minimizing convex functionals have better convergence guarantees than algorithms for solving general nonlinear systems of equations while having no additional computational cost \citep{nocedal2006numerical}.
 Both formulations are mathematically equivalent.
@@
 \begin{align}
-    \text{mass} & =\frac{1}{2}\int_\Omega u\cdot u\,dx\\\
+    \text{localization} & =\frac{1}{2}\int_\Omega u\cdot u\,dx\\\
     \text{gravity} & = \int_\Omega\frac{2A(\rho_I g)^n}{n+2} h^{n + 1}|\nabla s|^{n - 1}\nabla s\cdot u\,dx\\
@@
 \begin{equation}
-    J = \text{mass} + \text{gravity} + \text{penalty}.
+    J = \text{localization} + \text{gravity} + \text{penalty}.
 \end{equation}
@@
 We verified the correctness of our implementation of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
-When there is basal friction, the shallow stream equations do not have a simple analytical solution.
+There are analytical solutions to the shallow stream equations with basal friction \citep{bodhvarsson1955flow, bueler2014exact}; these solutions are more complex algebraically than those for the case without basal friction.
 To get an exact solution, we chose the ice velocity and thickness a priori and used the computer algebra system SymPy \citep{sympy} to generate a friction coefficient that makes these fields an exact solution.
+We hand-tuned the input parameters so that the values of the manufactured friction coefficient were within reasonable bounds.
 We then checked that numerical results converge to this manufactured solution at the expected rate \citep{roache2002code}.
@@
 The lack of vertical strain rates in the ice viscosity in 2D models can also lead to different equilibrium grounding line positions under the same external forcing.
-The advantage of the SSA is that that it can capture most of the overall flow features of an individual fast-flowing glacier but at much lower computational cost than the Stokes equations.
+The advantage of the SSA is that that it can capture most of the overall flow features of fast-flowing glaciers but at much lower computational cost than the Stokes equations.
 
@@
 The model used in \citet{brinkerhoff2015dynamics} is the closest to the one we present below.
-This work used both terrain-following coordinates and a tensor product basis of Lagrange finite elements in the vertical and higher-degree polynomials in a single vertical layer.
+This work used both terrain-following coordinates and a tensor product basis of Lagrange finite elements in the horizontal and higher-degree polynomials in a single vertical layer.
 For the vertical basis functions, they used a plug flow mode and one shear mode.
@@
 \end{equation}
-For example, we can use piecewise linear or quadratic elements on triangles and use quintic or higher degree polynomials in the vertical.
-Rather than use the usual Lagrange interpolating polynomial basis in the vertical dimension, we can instead use the Legendre polynomial basis.
-The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.
+For example, we can use piecewise linear or quadratic elements on (horizontal) triangles and use quintic or higher degree polynomials in the vertical.
+Rather than use the usual Lagrange elements in the vertical dimension, we can instead use the Legendre polynomial basis.
+The Legendre polynomials are mutually orthogonal, which makes the mass matrix block-diagonal, and provide better approximation properties \citep{szabo2004p}.
 Our main reason for choosing to build icepack using the Firedrake package was because it natively supports tensor product elements.
@@
 The user then has to decide how many vertical modes are sufficient.
-Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
-The degree-2 model is the most minimalistic model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
+Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a sanity test for the hybrid model.
+The degree-2 model is the minimal model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
 Going up to a degree-4 model is sufficient to capture the exact solution for the shallow ice approximation.
@@
 The shear flow basis function was chosen to be exact assuming the SIA balance with cold ice, together with a heuristic approximation for polythermal ice.
-By contrast, our approach allows for an arbitrary number of shear modes, although using a high-degree basis in the presence of a cold-temperate transition surface will introduce ringing artifacts.
-In principle, one could confront this problem by using two layers -- one cold and one temperate -- instead of a single layer.
-The extruded mesh capabilities in Firedrake provide the needed functionality for implementing such a model.
-We have not yet implemented a multilayer model but this will be the subject of future work.
+By contrast, our approach allows for an arbitrary number of shear modes.
 
@@
 The fluidity factor $A$ in Glen's law is roughly a known function of both temperature and melt fraction and we have included this function in the package.
-Having some description of heat transport is thus necessary for describing realistic glacier flows, but it is not sufficient by itself.
+Parameterizing fluidity is not sufficient by itself however.
 While the dependence of fluidity on temperature is known fairly well from laboratory experiments, the dependence on melt fraction is known with much less certainty \citep{cuffey2010physics}.
-It is likely that users of this module will want to substitute in their own parameterization for melt fraction dependence.
+Some users of this module may want to substitute in their own parameterization for melt fraction dependence.
 Other processes such as damage, fabric development, and impurities can affect the fluidity as well.
@@
 The model from \citet{albrecht2014fracture} adds sources of damage where the membrane stress exceeds a critical value and sinks of damage when the principal strain rate is less than a critical value.
-These kinds of simplified models ignore complications such as inertial or 3D effects.
-They are intended more to capture some important features of the real physical system ecnomically than to describe it exactly.
 
@@
 The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
-To solve the advection equation equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
+To solve the advection equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
 The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}, but this scheme introduces a tuneable stabilization parameter.
@@
 
-Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy (CFL) condition.
+Some glaciologists may be unfamiliar with the Courant-Friedrichs-Lewy (CFL) condition.
 The use of an unconditionally stable scheme guarantees that they will get an answer, rather than a runtime error, should they try to use a large timestep.
-Users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
 The extra computational cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
 Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.
+
+Icepack currently lacks an adaptive timestepping scheme.
+Unconditionally stable schemes allow taking long timesteps, but taking very long timesteps will give inaccurate solutions.
+At present, users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
+Adaptive timestepping will be added in a future release.
 
@@
 
-\subsection{Convex optimization}
+\subsection{Convex optimization}\label{subsec:convex-optimization}
 
@@
 
-Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian \citep{pratt1998gauss}.
+Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian \citep{pratt1998gauss, habermann2012reconstruction}.
 Each iteration of Gauss-Newton is more expensive than that of BFGS or gradient descent because it requires the solution of a more complex linear system than just the mass matrix.
@@
 Previous work on numerical modeling of glacier flow has focused largely on the technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory; does it scale to large numbers of processors; can models accurately predict grounding line retreat, etc.
-The person learning to use an unfamiliar software tool is largely absent from the discussion.
 For graduate students or other researchers who are not experts in computational physics, the difficulty of learning to use a particular software package may be more of a rate-limiting factor than the speed or efficiency of that software.
@@
 \end{enumerate}
-The phyics of how glaciers interact with their environment are not completely understood.
+The physics of how glaciers interact with their environment are not completely understood.
 Consequently, the ability to change components of the model is an essential feature for any tool aimed at researchers in this field.

### 2021-04-02 reviews/response-3.tex
More responses to reviewer 3

@@
 
+Added a reference to Ascher and Petzold.
+
 \textcolor{blue}{56, equation (1) and nearby, AB:  Two concerns.
@@
 And then say what Icepack will when solving (1).}
+
+We removed the statement that this is an advection equation and added the following sentences: ``This problem has the apparent form of a conservative advection equation.
+The velocity $u$ is coupled to the thickness and surface slope, however, in such a way that the whole problem is not hyperbolic and in many cases does not have a PDE `type' in the usual sense.''
 
@@
 
-I was anticipating criticism from Doug Brinkerhoff.
 These sentences have been removed.
@@
 Changed to ``Some users of this module may want to substitute in their own parameterization for melt fraction dependence.''
-I think the fact that melt fraction dependence is not constrained very well means users are \emph{more} likely to change this in their code than if we did have good data.
-After all, if the data were good then we could just use whatever parameterization the experimentalists told us to and no one would ever have to touch it.
 
@@
 
-I think this falls under the heading of expert knowledge of glaciology and not of numerics.
-To give an extreme example, early on in grad school I was making a simulation of several Greenland glaciers and my officemate gave me the helpful advice to increase the fluidity of ice that would have accumulated during the last glacial period because of the dustier atmosphere.
-To use your phrase from before, we provide policies about numerics and mechanisms for physics.
+What you're describing is knowledge of glaciology, not scientific computing.
+We have stated explicitly in the first sentence of section 6 that we aim to reduce the need for users to be experts in the latter but not the former.
+If we were to define how some fields are coupled, we would be faced with a difficult design choice of where this is or is not appropriate.
+Including the effect of temperature on rheology seems fairly obvious.
+Some modeling papers also account for much higher fluidity for ice that accumulated during the last glacial period because of the greater dust content.
+Should it also be the responsibility of the package to include this effect?
+We did not feel that, at this stage of the development, we could also implement high-level simulation drivers that hide these details while also satisfying our other design goals.
 
@@
 This "key classes" sentence is not describing users, it is describing co-developers.}
+
+Again I think this seems to be a disagreement about what the software package should do and what the users should be responsible for.
+Icepack lacks any of the high-level simulation drivers that PISM has.
+Giving users a relatively lower level interface to the physics solvers was a concession to the fact that we can't predict what they will want to do with an ice flow model by way of coupling to climate, oceans, solid earth, etc.
+Obviously you and the developers of PISM have had far more experience.
+We may well have misjudged what glaciologists do and don't find difficult about modeling and what type of interface best serves the needs of the most people.
 
@@
 
+Changed to ``Relatively few works in the computational science literature draw directly from relevant work in HCI when discussing usability; see for example Hannay et al. 2009, Harris et al. 2020.''
+
 \textcolor{blue}{725:  I'm on board with (2), and wish for more careful analysis of Icepack's contribution here, but "quantifying" in (1) is not what you are doing in this manuscript.
 (Nor, probably, do you have the ability to do it.)}
+
+Changed from ``quantifying'' to ``evaluating''.
+We're making an unsubstantiated claim that usability is indeed a rate-limiting factor.
 

### 2021-04-02 icepack.tex
More responses to reviewer 3

@@
 The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
-Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation.
+Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation; see \citet{ascher1998computer}.
 
@@
 The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
-The prognostic model is a conservative advection equation:
+The prognostic model is:
 \begin{equation}
@@
 where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
+This problem has the apparent form of a conservative advection equation.
+The velocity $u$ is coupled to the thickness and surface slope, however, in such a way that the whole problem is not hyperbolic and in many cases does not have a PDE ``type'' in the usual sense.
+
 Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
@@
 We have also paid special attention to how we can design this software package to be most usable for its intended audience.
-Usability is not often discussed in the literature on computational science.
-We believe that this is because of two difficulties: (1) quantifying the degree to which usability is a rate-limiting factor and (2) concretely assessing what features make software tools more or less usable.
+Relatively few works in the computational science literature draw directly from relevant work in HCI when discussing usability; see for example \citet{hannay2009scientists, harris2020array}.
+We believe that this is because of two difficulties: (1) evaluating the degree to which usability is a rate-limiting factor and (2) concretely assessing what features make software tools more or less usable.
 By contrast, measuring computational performance is much more feasible although still fraught with difficulties of its own.

### 2021-04-03 reviews/response-2.tex
More responses and describe BCs

@@
 to complete the definition of each models.}
+
+We added this statement to section 2.1: ``We implement two types of boundary conditions for the prognostic equation.
+Users can specify an inflow flux value and this value becomes a source of thickness at any point along the domain boundary where the ice velocity is pointing in to the domain.
+The flux at the inflow boundary can change in time.
+Second, we impose outflow boundary conditions on any part of the domain where the ice velocity is pointing outwards.
+Which segments of the boundary are inflow or outflow are diagnosed automatically by calculating the sign of the dot product between the velocity and the unit outward normal vector.''
+
+We also added an entirely new section which is now \S2.3 in the text just on the boundary conditions for the different diagnostic models.
+
 \item \textcolor{blue}{Certain terms in the equations I do not believe are defined anywhere, for instance, in equation (10),
@@
 make the connection.}
+
+We added this statement to section 2.1: ``Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
+In the examples we use up to degree 2 and the unit tests use up to degree 4.
+We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.''
+See also previous comment on boundary conditions for the thickness equation.
+We also added a longer description of our treatment of ice-free regions (which is very ad hoc for now) at the end of section 2.1.
+This is a weak point at present and we plan to improve this in future versions.
+
 \item \textcolor{blue}{What sort of meshes do you use in your model (in the horizontal dimension, for the hybrid one)?
@@
 a manufactured problem to ensure that it was implemented correctly?}
+
+This was sloppily written.
+What we should have said was that SUPG confers some of the benefits of upwind finite difference stencils when using continuous Galerkin basis functions.
+I was basing the statement about stability on the expression for the numerical amplification factor for the $\theta$-scheme with piecewise linear finite elements from section 3.5.2 in Donea and Huerta, Finite Element Methods for Flow Problems.
+In any case, we've cut much of this section to focus more on the implementation in icepack rather than what other packages use.
+The text was also out of date; the actual default now is an implicit scheme with a Lax-Wendroff correction that gives higher order accuracy in time.
+Our statement that implicit Euler has predominantly diffusive errors was not to imply that explicit Euler doesn't share the same property.
+We were trying to draw a contrast between what types of errors are tolerable for the prognostic model as opposed to other problems like damage transport, which is described in the next paragraph in the text.
+
 \item \textcolor{blue}{Section 4.2: there is an approach discussed in Tezaur et al. (https://doi.org/10.5194/gmd-8-1197-2015)
@@
 discussion.}
+
+See response to previous comment and the additional section we added on boundary conditions.
+This section describes some extra care that we had to do in our implementation which was purely a consequence of our choice of basis functions and which does not appear in the idealized mathematical form of the model.
+In keeping with our overall goal of splitting the paper up into a section on what we're solving and a different section on how we're solving it, we've kept this section where it is.
+
 \item \textcolor{blue}{Section 4.6: The authors mention running their code on 1 core. Is the code parallel - can it be run
@@
 We believe that the benefits outweight the costs but this is a definite drawback of our approach.
+
 \item \textcolor{blue}{It's great that you have executable documentation in something easy-to-use such as Jupyter notebooks! (no need to address this comment)}
@@
 We have rewritten some of the text to try and make this clearer.
+
 \item \textcolor{blue}{The methods do not discuss their code development/testing stance on icepack. How do users con-

### 2021-04-03 reviews/response-3.tex
More responses and describe BCs

@@
 By way of an ``aggressive scheme for putting reasonable defaults into the hands of novices'', we describe our policy at the end of section 6, beginning in line 695.
-Other software packages have (exactly as you say) all the mechanisms, but do not display a consistent policy.
+Other software packages have (exactly as you say) all the mechanisms, but the policies on display suggest the priority is for speed above all else.
 For example, several of the demonstration codes in Elmer default to choices of iterative solvers and preconditioners that work on the particular problem in the example but which are likely to require hand-tuning when generalized to real data.
@@
 Other examples from Elmer do use direct solvers, for example their \href{https://github.com/ElmerCSC/elmerfem/blob/aaa6c4a345c73ba248c475d27975c5e3a9fa7758/elmerice/examples/Test_SSA/ismip_SSA_2D.sif#L105}{demonstration code} to run the ISMIP test case with the SSA model uses a direct solver from UMFPACK for the momentum equation.
-The mechanisms are there, but a consistent policy is not.
-Other packages show a similar pattern.
-For example, BISICLES has hand-tuned numbers of Picard iterations and multigrid smoothing iterations in their demonstration codes.
+Other packages show a similar pattern -- BISICLES has hand-tuned numbers of Picard iterations and multigrid smoothing iterations in their demonstration codes.
 A new user either must (1) know enough about numerical methods to diagnose the problem from bad input data, bad linear solver options, or bad nonlinear solver options, or (2) have a mentor who can.
-Either way this creates unnecessary friction for people who may be getting their first experience of modeling by using these examples.
-This is not to put down or undermine the incredible technical work that the developers of Elmer, BISICLES, and other packages have done over the years by any means; it's just a reflection of what knowledge the users need to have going in.
+Experienced users have all the tools at hand that they need to create large simulations that will run on supercomputers, but at the same time these choices can create friction for people who may be getting their first experience of modeling by using these examples.
 
@@
 Many other software packages for glacier flow modeling, including PISM and ISSM, use this terminology.
-By now the nomenclature is used in so many places throughout icepack that switching would be a breaking change for existing users.
 
@@
 
-We removed the statement that this is an advection equation and added the following sentences: ``This problem has the apparent form of a conservative advection equation.
-The velocity $u$ is coupled to the thickness and surface slope, however, in such a way that the whole problem is not hyperbolic and in many cases does not have a PDE `type' in the usual sense.''
+We removed the statement that this is an advection equation and added the following sentences:
+``This problem has the apparent form of a conservative advection equation, but the velocity $u$ is coupled to the thickness and surface slope in such a way that the whole problem is not hyperbolic.
+For the specific case of the shallow ice approximation (see section \S2.2.1), the coupled system is parabolic.
+In all other cases, the problem does not have a PDE `type' in the usual sense because the velocity is found through solving an elliptic PDE where the thickness and surface slope are coefficients.''
+
+Our discussion of boundary conditions was mostly lacking.
+We added the following:
+
+``We implement two types of boundary conditions for the prognostic equation.
+Users can specify an inflow flux value and this value becomes a source of thickness at any point along the domain boundary where the ice velocity is pointing in to the domain.
+The flux at the inflow boundary can change in time.
+Second, we impose outflow boundary conditions on any part of the domain where the ice velocity is pointing outwards.
+Which segments of the boundary are inflow or outflow are diagnosed automatically by calculating the sign of the dot product between the velocity and the unit outward normal vector.
+
+The mass transport equation for ice thickness is a free boundary problem, where the free boundary is the contour between ice-covered and ice-free regions (Schoof and Hewitt 2013).
+A naively-implemented prognostic solver could erroneously compute negative thickness values in subsequent timesteps when there is ablation in ice-free regions.
+A common and ad-hoc approach to work around this issue is to truncate the thickness at zero at every timestep.
+The principled approach is to instead treat the free boundary problem directly as a variational inequality (Jouvet and Bueler 2012).
+Icepack currently lacks a principled scheme for tracking this free boundary and we instead rely on truncation.
+Treating this problem as a variational inequality in icepack will be the subject of future development.
+PETSc includes scalable solvers for variational inequalities (Bueler 2020) that are also available through Firedrake.''
+
+We also added a new section which is now section \S2.3 in the text on the boundary conditions for the diagnostic models.
 
@@
 
+These modules were developed by different authors.
+We will get rid of the discrepancy in a future version.
+
 \textcolor{blue}{148--149:  It is of course true that for general boundary conditions the "shallow stream equations do not have a simple analytical solution."  What is probably meant here, however, is that there is not a well-known, exactly-solvable, basal-friction-included boundary value problem suitable for testing.
@@
 I am concerned that positive consequences of using (16) are emphasized while negative consequences are not mentioned, and that Icepack has no path around these negative consequences.}
+
+Section 2.3 in the text describes the lateral boundary conditions for the diagnostic models and an admission that we assume the entire spatial domain is ice-covered.
+It's possible that I've painted us into a corner with this choice.
+In the $h \to 0$ limit, the only singularity in the equations is in the vertical shear stress, which -- including a factor of $h$ from the determinant of the coordinate transformation -- scales like $h^{-1}$.
+A dirty hack would be to replace this term with \texttt{1 / max\_value(h\_min, h)}, effectively clamping it at the reciprocal of this minimum thickness.
+This would still permit ice-free areas at the risk of underestimating vertical shear stresses in areas where the ice is thinner than whatever threshold value was used.
+Alternatively, we may have no choice but to switch to a $z$-coordinate model in future versions.
 
@@
 
+See response to previous comment on boundary conditions.
+We cut some of the text in this section in response to comments from reviewer \#2 and clarified that we were referring to the mass transport equation.
+
 \textcolor{blue}{368--373, AC:  The third concern, regarding this paragraph, is that Icepack apparently has not yet set-up effective adaptive time-stepping.
@@
 
-Text has been slightly reorganized, and we added ``Icepack currently lacks an adaptive timestepping scheme.
+Text has been reorganized, and we added ``Icepack currently lacks an adaptive timestepping scheme.
 Unconditionally stable schemes allow taking long timesteps, but taking very long timesteps will give inaccurate solutions.
@@
 to admit that time steps cannot be arbitrarily long in your scheme, to get convergence; the "conditional/unconditional stable" language is an artifact of linear PDE theory, and your problem, taken seriously, is super-duper-nonlinear.}
+
+See response to previous comment.
+We've changed the text to say that the scheme is unconditionally stable for the advection equation with a note that the coupled system is not linear nor is it hyperbolic.
+We also changed a later sentence to: ``Implicit schemes allow taking longer timesteps than explicit ones, but taking very long timesteps will give inaccurate solutions and, in the presence of ablation, may yield negative thickness values.''
 
@@
 
+See response to previous comments.
+
 \textcolor{blue}{474--487:  Extruded meshes.
@@
 No mention thereof.}
+
+You can use extruded meshes of variable thickness and that thickness can become 0.
+So the choice of extruded meshes does not force one to use terrain-following coordinates as we did, whether or not that was a good idea.
 
@@
 
+See response to previous comments.
+The point of the exercise was to reproduce at least some of the results from a prior paper that used the shallow ice model exclusively.
+
 \textcolor{blue}{598:  Regarding "computationally cheap enough ...
@@
 The reader can handle a plot, if you want to generate one.}
+
+We avoided discussing performance or scalability considerations in this paper.
+See our responses to reviewer \#2.
+These will be the subject of future work.
 
@@
 
+We added the statement: ``Simulations that have been debugged can then easily be transformed into a single Python script, for example using the tool \texttt{nbconvert}, for production runs and parallel execution.
+In other words, while there is an interactive interface, there is also a faster batch mode interface as well.''
+
 \textcolor{blue}{681--683, C:  Regarding "The API ...
@@
 \textcolor{blue}{682--683, C:  Also, this "user" is a co-developer.}
+
+Several authors have noted the necessity of providing a path for users to become developers for the long-term health of an open-source software project.
+See for example Turk (2013), Scaling a code in the human dimension or Bangerth and Heister (2013), What makes computational open source software libraries successful?
 
@@
 That means de-emphasizing solver components/discretizations in the user's view, primarily by setting aggressive defaults, even as the developers must get solvers right (which is the strength of Icepack).}
+
+It's possible that icepack will evolve to have a more end-to-end simulation driver as the main interface.
+No change to the text.
 
@@
 
-Changed from ``quantifying'' to ``evaluating''.
+Expanded to: ``We believe that this is because of two difficulties.
+First, the degree to which usability is a rate-limiting factor for scientists is hard to quantify and likely differs widely across disciplines.
+Second, concretely assessing what features make software tools more or less usable is highly subjective.''
 We're making an unsubstantiated claim that usability is indeed a rate-limiting factor.
+We could be wrong about this.
+Even if we're right, the steps that we took to address this problem could have been misguided.
 
@@
 Will you be able to break out of stress-balance-and-inverse-model-solver-playground mode, and start answering some of the science questions enumerated in the Introduction?}
+
+We have changed the list to include ``improved physics formulations and solvers that work in ice-free areas''.
 

### 2021-04-03 icepack.tex
More responses and describe BCs

@@
 
-The rheology and friction coefficient are functions of other fields that have their own evolution equations.
+The evolution equations determining for ice thickness and velocity are necessary for any realistic glacier simulation, but other fields with their own evolution equations can be a part of the problem as well.
+The rheology and friction are the main coefficients in the diagnostic equation.
+These coefficients are functions of other fields that have their own evolution equations.
 For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
@@
     \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
+    \label{eq:prognostic-eqn}
 \end{equation}
 where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
-This problem has the apparent form of a conservative advection equation.
-The velocity $u$ is coupled to the thickness and surface slope, however, in such a way that the whole problem is not hyperbolic and in many cases does not have a PDE ``type'' in the usual sense.
+This problem has the apparent form of a conservative advection equation, but the velocity $u$ is coupled to the thickness and surface slope in such a way that the whole problem is not hyperbolic.
+For the specific case of the shallow ice approximation (see section \S\ref{subsubsec:shallow-ice}), the coupled system is parabolic.
+In all other cases, the problem does not have a PDE ``type'' in the usual sense because the velocity is found through solving an elliptic PDE where the thickness and surface slope are coefficients.
 
 Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
+In the examples we use up to degree 2 and the unit tests use up to degree 4.
 We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.
@@
 Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.
-We have not implemented this feature but support for non-hydrostatic ice shelves may come in a future release of icepack.
+
+We implement two types of boundary conditions for the prognostic equation.
+Users can specify an inflow flux value and this value becomes a source of thickness at any point along the domain boundary where the ice velocity is pointing in to the domain.
+The flux at the inflow boundary can change in time.
+Second, we impose outflow boundary conditions on any part of the domain where the ice velocity is pointing outwards.
+Which segments of the boundary are inflow or outflow are diagnosed automatically by calculating the sign of the dot product between the velocity and the unit outward normal vector.
+
+The mass transport equation for ice thickness is a free boundary problem, where the free boundary is the contour between ice-covered and ice-free regions \citep{schoof2013ice}.
+A naively-implemented prognostic solver could erroneously compute negative thickness values in subsequent timesteps when there is ablation in ice-free regions.
+A common and ad-hoc approach to work around this issue is to truncate the thickness at zero at every timestep.
+The principled approach is to instead treat the free boundary problem directly as a variational inequality \citep{jouvet2012steady}.
+Icepack currently lacks a principled scheme for tracking this free boundary and we instead rely on truncation.
+Treating this problem as a variational inequality in icepack will be the subject of future development.
+PETSc includes scalable solvers for variational inequalities \citep{bueler2020petsc} that are also available through Firedrake.
 
@@
 (In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
-The first-order model is based on an asymptotic expansion of the Stokes equations in the ratio of the ice thickness to a typical horizontal length scale.
+The first-order model is based on an asymptotic expansion of the Stokes equations with respect to the ratio of the ice thickness to a typical horizontal length scale.
 The aspect ratio of most glacier flows is on the order of 1/20 or less, although there are some exceptions.
@@
 
-\subsubsection{Shallow ice approximation}
+\subsubsection{Shallow ice approximation}\label{subsubsec:shallow-ice}
 
@@
     \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\,dx \\
-    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\,d\gamma
+    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\,d\gamma \label{eq:terminus-bc}
 \end{align}
@@
 By contrast, our approach allows for an arbitrary number of shear modes.
+
+\subsection{Boundary conditions}
+
+The shallow shelf, shallow stream, and hybrid models require boundary conditions to be well-posed.
+We implement three different types which are largely the same for each physics model.
+\begin{enumerate}
+    \item Users can specify the velocity on segments of the domain boundary where ice is flowing in; this is a Dirichlet condition which is eliminated from the system at the level of the solver.
+Supplying inflow boundary conditions for the hybrid model necessarily requires more information, namely the variation of the horizontal velocity with depth.
+    \item The Neumann condition at the ice terminus (equation \eqref{eq:terminus-bc}) is a natural boundary condition which is imposed by adding the relevant term to the action functional.
+    \item We allow for side wall boundary conditions where the ice has non-zero thickness but abuts, say, a fjord wall or an ice shelf embayment that exerts resistive stresses.
+In the normal direction, the boundary condition is that $u \cdot \nu = 0$ where $\nu$ is the unit outward normal vector.
+In the tangential direction, the boundary condition is $hM\cdot\nu = -C|u|^{\frac{1}{m} - 1}u$ where $C$ is a side wall friction coefficient and $M$ is the membrane stress tensor.
+Putting these together gives a mixed Dirichlet-Robin boundary condition.
+The side wall friction (Robin) boundary condition is natural and thus can be added directly to the action functional.
+The constraint of no normal flow is harder to impose directly for a curved boundary; we used the penalty method.
+\end{enumerate}
+Future versions of the code will use Nitsche's method \citep{nitsche1971variationsprinzip} instead of the penalty method to enforce the no normal flow constraint.
+Nitsche's method dramatically improves the conditioning of the associated optimization problem.
+Our implementation of the hybrid model requires special care to handle the terminus boundary condition as a consequence of our choice of vertical basis functions; see section \S\ref{subsubsec:ocean-bc}.
+
+The shallow ice model alone is able to cope with ice-free regions in the domain, albeit with truncation at 0 thickness.
+All of the remaining models and their boundary conditions assume that the entire spatial domain is ice-covered.
+We will improve the solvers to address ice-free areas in future versions.
 
@@
 The boundary condition at the ice base is a fixed flux of heat from geothermal sources or from the oceans.
-We depart from \citet{aschwandend2012enthalpy} in one respect.
+We depart from \citet{aschwanden2012enthalpy} in one respect.
 Rather than fix the ice surface temperature to the atmospheric temperature, we instead use a Robin boundary condition that makes the ice temperature adjust to the external temperature:
@@
 
-The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
-To solve the advection equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
-The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}, but this scheme introduces a tuneable stabilization parameter.
-In the interest of simplicity, we instead default to the implicit Euler scheme, which has no such parameters.
-The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.
-
-Some glaciologists may be unfamiliar with the Courant-Friedrichs-Lewy (CFL) condition.
-The use of an unconditionally stable scheme guarantees that they will get an answer, rather than a runtime error, should they try to use a large timestep.
-The extra computational cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
+Icepack offers two timestepping schemes for solving the prognostic model (equation \eqref{eq:prognostic-eqn}).
+For testing purposes we include the implicit Euler scheme, which is first-order accurate and unconditionally stable for the advection equation.
+The default is a second-order accurate scheme based on an implicit version of the Lax-Wendroff method, which is also unconditionally stable for the advection equation \citep{donea2003finite}.
+Explicit schemes all require a timestep that satisfies the Courant-Friedrichs-Lewy (CFL) stability condition, which many glaciologists may be unfamiliar with.
+With an unconditionally stable scheme, users will get an answer, rather than a runtime error, should they try to use a larger timestep.
+The extra computational cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve.
 Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.
-
-Icepack currently lacks an adaptive timestepping scheme.
-Unconditionally stable schemes allow taking long timesteps, but taking very long timesteps will give inaccurate solutions.
-At present, users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
-Adaptive timestepping will be added in a future release.
-
-The implicit Euler scheme tends to diffuse out sharp discontinuities that may be present in the true solution \citep{donea2003finite}.
+We note that, while we the stability properties of different schemes for the linear advection equation guided our choice of method, the coupled system for both thickness and velocity is not linear and not hyperbolic.
+
+The implicit Euler and Lax-Wendroff schemes tend to diffuse out sharp discontinuities that may be present in the true solution \citep{donea2003finite}.
 Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
@@
 The implicit Euler scheme would obscure this important feature.
-For the damage solver, we have instead used a strong stability-preserving Runge-Kutta method in time and a discontinuous Galerkin basis in space to best capture these sharp interfaces \citep{shu1988efficient}.
+For the damage solver, we have instead used a strong stability-preserving Runge-Kutta method in time, a discontinuous Galerkin basis in space, and a flux limiter to best capture these sharp interfaces \citep{shu1988efficient}.
+
+Icepack currently lacks an adaptive timestepping scheme.
+Implicit schemes allow taking longer timesteps than explicit ones, but taking very long timesteps will give inaccurate solutions and, in the presence of ablation, may yield negative thickness values.
+At present, users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
+Adaptive timestepping will be added in a future release.
 
@@
 
-\subsubsection{Ocean boundary condition}
+\subsubsection{Ocean boundary condition}\label{subsubsec:ocean-bc}
 
@@
 
+Simulations that have been debugged can then easily be transformed into a single Python script, for example using the tool \texttt{nbconvert}, for production runs and parallel execution.
+In other words, while there is an interactive interface, there is also a faster batch mode interface as well.
+
 \textbf{Abstraction gradient: What are the levels of abstraction exposed by the library?
@@
 Relatively few works in the computational science literature draw directly from relevant work in HCI when discussing usability; see for example \citet{hannay2009scientists, harris2020array}.
-We believe that this is because of two difficulties: (1) evaluating the degree to which usability is a rate-limiting factor and (2) concretely assessing what features make software tools more or less usable.
+We believe that this is because of two difficulties.
+First, the degree to which usability is a rate-limiting factor for scientists is hard to quantify and likely differs widely across disciplines.
+Second, concretely assessing what features make software tools more or less usable is highly subjective.
 By contrast, measuring computational performance is much more feasible although still fraught with difficulties of its own.
@@
     \item an implementation of full Stokes flow,
-    \item dynamic glacier termini to enable calving models, and
-    \item adaptivity through moving mesh methods.
+    \item improved physics formulations and solvers that work in ice-free areas,
+    \item adaptivity in time and space,
 \end{enumerate}

### 2021-04-05 reviews/response-1.tex
More responses and describe BCs
Expand discussion of hybrid model, final responses

@@
 
+These modules were implemented by two different authors, the SSA / hybrid models by myself (DRS) and the SIA model by Jess Badgeley (second author) as a project for her PhD thesis and to learn more about finite element analysis.
+We recognize that there's a discrepancy here and this will be corrected in future releases of the package.
+
 \item[L144] \textcolor{blue}{That the terminal potential term, Eq. 13 disappears is not obvious. This
@@
 See Donea and Huerta, \emph{Finite element methods for flow problems}, chapter 3, sections 4-6.
+This section was also written very early on but became out of date by the time we submitted the paper.
+Since then, we added an implicit version of the Lax-Wendroff scheme, which has better stability properties and higher order accuracy.
+The text has been amended to reflect this change.
+
 We did have similar experiences to what you describe when preparing the MISMIP+ test case for this paper.

### 2021-04-05 reviews/response-2.tex
Expand discussion of hybrid model, final responses

@@
 The remedy is to wrap this value in a \texttt{firedrake.Constant} object.
-But this kind of performance regression is easily caught using \texttt{htop} and are more a result of programmer error.
+This kind of performance regression is easily caught using \texttt{htop}.
+While this is more a result of programmer error, it's an easy mistake to make and the Firedrake team are working on ways to diagnose it and issue appropriate warnings.
 
@@
 But I made this naming choice before there were many other collaborators on the project.
-The ensuing confusion has shown that this was a bad choice of terminology that we'll change in a future version.
+The ensuing confusion has shown that this was a bad choice of terminology and we intend to change it in a future version.
 This model can be used for both regions at continental scales -- it can capture both plug and shear flow.
@@
 is needed? Why is stabilization needed only for the SIA? I think these things should be made clear.}
+
+See comments by reviewer \#3.
+The technical answer is that this makes the solution live in the Sobolev space $H^1(\Omega)$.
+A more heuristic answer is that the shallow ice approximation is usually assumed to hold only over distances greater than a few ice thicknesses.
+The penalty term is meant to filter out variability at length scales where the model doesn't even apply.
 \item \textcolor{blue}{Section 2.2.2: there is some imprecision here in equation (13) - you have not defined anywhere that $\Gamma$
@@
 I was expecting to see it there, but maybe I'm missing something.}
+
+This was a bad oversight on our part -- the hybrid model does use Glen's flow law and we've added more detail at the end of this section describing the terms in the action functional (equations 22 through 26 in the revised text).
+The main difference is the term for viscous power dissipation.
 \item \textcolor{blue}{P. 10: the discussion here about substituting model components suggests it may be possible to use

### 2021-04-05 icepack.tex
Expand discussion of hybrid model, final responses

@@
 
+\textbf{Action functional.}
+We can now describe all of the terms in the action functional for this model:
+\begin{align}
+    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega\int_0^1hA^{-\frac{1}{n}}\sqrt{|\dot\varepsilon(u)|^2 + h^{-2}|\partial_\zeta u|^2}^{\frac{1}{n} + 1}\,d\zeta\,dx \\
+    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u(\zeta = 0)|^{\frac{1}{m} + 1}dx \\
+    \text{gravity} & = -\int_\Omega\int_0^1\rho_Igh\nabla s\cdot u\,d\zeta\,dx \\
+    \text{terminus} & = \int_\Gamma\int_0^1(p_I - p_W)u\cdot\nu\;d\zeta\,d\gamma
+\end{align}
+where $p_I$ is the ice overburden pressure and $p_W$ is the water pressure (if any) at the calving terminus from the ocean or a pro-glacial lake.
+The full action functional is exactly analogous to that of the shallow stream model:
+\begin{equation}
+    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}
+\end{equation}
+Again, we note that the expression for horizontal the strain rate in terrain-following coordinates comes from equation \eqref{eq:geometric-correction}, which includes the additional geometric correction factor.
+
+In terrain-following coordinates, the ice overburden pressure at relative depth $\zeta$ is $p_I = \rho_Igh(1 - \zeta)$.
+The water pressure is zero above the waterline and increases linearly below it.
+Letting $\zeta_{\text{sl}}$ denote the relative depth to the waterline, the water pressure is
+\begin{equation}
+    p_W = \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+
+\end{equation}
+where $w_+$ denotes the positive part of a real number.
+This quantity is continuous but only piecewise linear.
+Integrating it correctly requires more work above and beyond the usual symbolic approach; see section \S\ref{subsubsec:ocean-bc} for details.
+
 \subsection{Boundary conditions}

### 2021-04-06 reviews/response-1.tex
Correct names

@@
 
-These modules were implemented by two different authors, the SSA / hybrid models by myself (DRS) and the SIA model by Jess Badgeley (second author) as a project for her PhD thesis and to learn more about finite element analysis.
+These modules were implemented by two different authors, the SSA / hybrid models by myself (DRS) and the SIA model by Jessica Badgeley (second author) as a project for her PhD thesis and to learn more about finite element analysis.
 We recognize that there's a discrepancy here and this will be corrected in future releases of the package.

### 2021-04-06 icepack.tex
Correct names
Truly brutal correction

@@
 
-\Author[1]{Daniel}{Shapero}
-\Author[2]{Jessica}{Badgeley}
-\Author[2]{Andrew}{Hoffmann}
-\Author[1]{Ian}{Joughin}
+\Author[1]{Daniel R}{Shapero}
+\Author[2]{Jessica A}{Badgeley}
+\Author[2]{Andrew O}{Hoffmann}
+\Author[1]{Ian R}{Joughin}
 
@@
 
-The evolution equations determining for ice thickness and velocity are necessary for any realistic glacier simulation, but other fields with their own evolution equations can be a part of the problem as well.
+Other fields with their own evolution equations can be a part of the problem as well as thickness and velocity.
 The rheology and friction are the main coefficients in the diagnostic equation.

### 2021-04-06 icepack.tex
More wordsmithing

@@
 
-Other fields with their own evolution equations can be a part of the problem as well as thickness and velocity.
-The rheology and friction are the main coefficients in the diagnostic equation.
+The evolution equations for ice thickness and velocity are necessary for any simulation, both other fields with their own dynamics can be a part of the problem as well.
+For example, the rheology and friction are the main coefficients in the diagnostic equation.
 These coefficients are functions of other fields that have their own evolution equations.
-For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
+The rheology is a function of temperature, englacial water content, and damage from crevassing.
 Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock \citep{cuffey2010physics}.

### 2021-04-06 icepack.tex
Name spellings

@@
 \Author[2]{Jessica A}{Badgeley}
-\Author[2]{Andrew O}{Hoffmann}
+\Author[2]{Andrew O}{Hoffman}
 \Author[1]{Ian R}{Joughin}
@@
 
-\runningauthor{Shapero, Badgeley, Hoffmann, Joughin}
+\runningauthor{Shapero, Badgeley, Hoffman, Joughin}
 
@@
 \textit{Financial support}.
-DS was supported by US National Science Foundation grant \#1835321 and National Aeronautics and Space Administration grant \#80NSSC20K0954.
-JB was supported by NSF grant \#1256082.
-AH was supported by NASA grant \#80NSSC20K1627.
-IJ was supported by NSF grants \#1835321 and \#1643285.
+DRS was supported by US National Science Foundation grant \#1835321 and National Aeronautics and Space Administration grant \#80NSSC20K0954.
+JAB was supported by NSF grant \#1256082.
+AOH was supported by NASA grant \#80NSSC20K1627.
+IRJ was supported by NSF grants \#1835321 and \#1643285.
 

### 2021-04-06 reviews/response-2.tex
Responses to 2

@@
 provide a code that would be easy to use/develop by this class of prospective users. Following a description
-of the code, some numerical examples are presented to demonstrate the method’s capabilities and accuracy.
+of the code, some numerical examples are presented to demonstrate the method's capabilities and accuracy.
 The manuscript in question is well-written and interesting to read. Addressing usability is noteworthy
@@
 inadvertently using the incorrect default setting for their problem? I will assume it is to the extent it can
-be, and that the authors’ argument is that the default hidden settings are likely to do less damage than
+be, and that the authors' argument is that the default hidden settings are likely to do less damage than
 some arbitrary settings a user might put in his/her input file without knowing what they are doing. Also,
@@
 still wish to make progress in numerical modeling of land-ice; therefore, I will not focus too much on much
-“philosophical” perspective described above.
+``philosophical'' perspective described above.
 Another qualm I have about the paper has to do with performance - I am skeptical whether icepack
@@
 there have been efforts like Albany/Land Ice out there to make C++-based codes more accessible to
-users of varied backgrounds. I agree that even an “easy-to-use” C++ code will be more difficult and
-more intimidating than a Python code, so I am not trying to minimize the authors’ efforts at all.}
+users of varied backgrounds. I agree that even an ``easy-to-use'' C++ code will be more difficult and
+more intimidating than a Python code, so I am not trying to minimize the authors' efforts at all.}
 
@@
 automatic differentiation, for example? A broader question is: is computational performance/cost a
-concern for users of icepack, or is it intended to be a “sandbox” in which performance is secondary
-to being able to code up something “quick-and-dirty” for initial prototyping?}
+concern for users of icepack, or is it intended to be a ``sandbox'' in which performance is secondary
+to being able to code up something ``quick-and-dirty'' for initial prototyping?}
 
 See Rathgeber et al. 2016 for more information about the architecture of Firedrake and for performance benchmarks.
+We are constrained only by what Firedrake can do, so we refer to this paper for performance with no change to the text.
+
 Loosely speaking, the path through the toolchain goes like this.
@@
 The fundamental physics are the first-order model obtained by asymptotic expansion of the Stokes equations in the aspect ratio, also known as the Blatter-Pattyn equations.
+We have amended the text to make this clear.
 What I had imagined is that using only vertical basis functions up to degree 2 essentially defines its own semi-discrete physics model, similar to two- or three-layer ocean models.
@@
 The penalty term is meant to filter out variability at length scales where the model doesn't even apply.
+No change to the text.
+
 \item \textcolor{blue}{Section 2.2.2: there is some imprecision here in equation (13) - you have not defined anywhere that $\Gamma$
@@
 We are very much interested in, for example, using SIA in the interior of the ice sheet and SSA in the ice streams and margins, as this would give a much less computationally-intensive way to do some form of whole-ice sheet modeling than using, say, the first order model.
+No change to the text.
+
 \item \textcolor{blue}{Section 2.4: This section is very incomplete. You need to give the enthalpy equation and given the
@@
 We use unstructured triangular meshes although Firedrake in principle can use unstructured quad meshes.
+We've added the word ``unstructured'' to clarify this.
+
 \item \textcolor{blue}{Section 3: It is not clear from the description what the inverse problem you are describing is for.
@@
 But the simplifications that this results in for the vast majority of glaciological applications are so advantageous that it's a sacrifice we're willing to accept.
+No change to the text.
+
 \item \textcolor{blue}{Section 4.5.2: I think this section needs to be made earlier, and other BCs need to be added to that
@@
 Firedrake is built on PETSc and thus can run on parallel machines.
-See Rathgeber et al.
+See again Rathgeber et al.
 2016 for performance and scaling benchmarks; Firedrake has been run on problems with millions of degrees of freedom on supercomputers, for example the UK national supercomputer ARCHER.
+No change to the text.
 
@@
 such as continental-scale modeling, will require more sophisticated and possibly problem-specific
-approaches''. I’m wondering in particular about the problem-specific approach part. There are
+approaches''. I'm wondering in particular about the problem-specific approach part. There are
 models like first order Stokes that can be used at the continental scale and they are not really
@@
 Just using LU or throwing a black-box algebraic multigrid preconditioner at it would be failing to use this special structure.
+That said, we did not want to speculate too much on approaches that we haven't implemented yet.
 
@@
 We believe that the benefits outweight the costs but this is a definite drawback of our approach.
+No change to the text.
 
@@
 
-\item \textcolor{blue}{The methods do not discuss their code development/testing stance on icepack. How do users con-
-tribute to the code - through pull requests? Is there regression/performance testing? Continuous
+\item \textcolor{blue}{The methods do not discuss their code development/testing stance on icepack.
+How do users contribute to the code - through pull requests?
+Is there regression/performance testing?
+Continuous
 integration testing? These are all really important, especially if you have non-experts contributing
 to the code!}
+
+This information is on the icepack website (https://icepack.github.io/developers).
+Users contribute through pull requests which are automatically checked against a regression testing suite.
+We try to keep the test coverage at 95\% or higher.
+There is at present no automated performance testing short of looking at the timings from our CI service.
+Our development practices have changed appreciably even during the process of writing this manuscript; several of the contributors are students who are learning more about version control in tandem with learning to implement new or modify existing models.
+We have added references in the code and data availability section about where to find this information.
+
 \end{enumerate}

### 2021-04-06 icepack.tex
Responses to 2

@@
 All code used in this repository is free and open source and all data sets used in the demonstrations are publicly available.
-The icepack source repository is at https://github.com/icepack/icepack.
+The icepack source repository and regression testing results are at https://github.com/icepack/icepack.
 Icepack is released under the GPLv3 license.
 The git commit hash of the version of the code used for this publication is b78b0ee5, see also the Zenodo release at https://doi.org/10.5281/zenodo.1205640.
-The icepack documentation and user manual is hosted at https://icepack.github.io.
+The icepack documentation, user manual, and contribution guidelines are hosted at https://icepack.github.io.
+
 The source code for the demonstrations used in this paper is hosted at https://github.com/icepack/icepack-paper, commit hash 4db3dc28.

### 2021-04-06 reviews/response-3.tex
Responses to 3

@@
 
+We have altered the text to state more clearly that our goal is to help users who are not experts in scientific computing.
+We also added the following sentence to the conclusion: ``Exactly how to apply principles from HCI to maximize usability is, nonetheless, not an exact science.
+Our implementation may have failed to meet this goal and changes in future versions will be guided by what users find most difficult.''
+
 \textcolor{blue}{21--39:  I also think Python+Firedrake+PETSc is the most promising environment to build a new ice flow simulation library/model.  I'm on board!}
@@
 Many other software packages for glacier flow modeling, including PISM and ISSM, use this terminology.
+No change to the text.
 
@@
 
-Added a reference to Bueler 2014 and Böðvarsson 1955.
+Added a reference to Bueler 2014 and B\"o\dh varsson 1955.
 You make a great point about testing the wrong parts of a nonlinear phase space and this was actually a bit of a struggle when creating the tests for this solver.
@@
 We did not feel that, at this stage of the development, we could also implement high-level simulation drivers that hide these details while also satisfying our other design goals.
+See also response to the first comment and the text added to the conclusion.
+We believe that our design choices will ultimately enhance usability for the intended audience and we did our best by applying principles from HCI.
+But we may well have missed the mark and the interface may change to respond to evolving understanding of our target audience's needs.
 
@@
 We may well have misjudged what glaciologists do and don't find difficult about modeling and what type of interface best serves the needs of the most people.
+See response to first comment.
 
@@
 So the choice of extruded meshes does not force one to use terrain-following coordinates as we did, whether or not that was a good idea.
+No change to the text.
 
@@
 Not obvious how to proceed, I agree, but note that the thickness field h of a glacier has the similar continuous-but-non-differentiable character as the pressure on the ocean calving front.}
+
+The technique might be applicable in other areas but we haven't had the chance to try it yet.
+No change to the text.
 

### 2021-04-06 icepack.tex
Responses to 3

@@
 For this reason, we have chosen to focus explicitly on usability by drawing on the research literature in HCI.
-We argue that the same design features that enhance usability for relative novices to the subject area will also enhance the productivity of expert users.
+Exactly how to apply principles from HCI to maximize usability is, nonetheless, not an exact science.
+Our implementation may have failed to meet this goal and changes in future versions will be guided by what users find most difficult.
+Finally, we argue that the same design features that enhance usability for relative novices to the subject area will also enhance the productivity of expert users.
 
@@
 New features will be guided by feedback from icepack users and from the glaciological community at large.
-By providing implementations of several glacier flow models from less to more complex and by enabling users to experiment with the physics, this tool both lowers the barrier to entry for novices to glacier flow modeling and provides a pathway for these users to progress towards ever more sophisticated and advanced simulations.
+By providing implementations of several glacier flow models from less to more complex and by enabling users to experiment with the physics, this tool both lowers the barrier to entry for novices to numerical modeling and provides a pathway for these users to progress towards ever more sophisticated and advanced simulations.
 

### 2021-04-28 reviews/response-3.tex
Final changes

@@
 
+Thank you for this thorough review!
+Original comments are in blue, our responses are in black, and hyperlinks in magenta.
+
 \section*{Summary}
@@
 We have altered the text to state more clearly that our goal is to help users who are not experts in scientific computing.
+We added the following to the introduction: ``We have focused efforts thus far on process studies of individual glaciers or drainage basins (use cases 2 and 3 of the list above).
+Development of icepack is ongoing and we will broaden our efforts to encompass more use cases in future.''
 We also added the following sentence to the conclusion: ``Exactly how to apply principles from HCI to maximize usability is, nonetheless, not an exact science.
@@
 
+We added the following to the introduction: ``We have focused efforts thus far on process studies of individual glaciers or drainage basins (use cases 2 and 3 of the list above).
+Development of icepack is ongoing and we will broaden our efforts to encompass more use cases in future.''
+
 \textcolor{blue}{533--586:  The MISMIP+ experiment is part of the 2008-onward tradition of running ice sheet models in rectangular boxes (MISMIP, EISMINT-HOM, ...).
@@
 (Yes, that is what Icepack looks like right now.)}
+
+We included this assuming that it's a necessity to repeat at least some of the standard benchmarks from the literature in order to publish a new model.
+No change to the text.
 
@@
 It's possible that icepack will evolve to have a more end-to-end simulation driver as the main interface.
-No change to the text.
+See changes to intro and conclusion.
 

### 2021-06-11 icepack.tex
Final changes
Final technical corrections

@@
 Our goal is to make a tool that will be both accessible to novices and more productive for experts.
+We have focused efforts thus far on process studies of individual glaciers or drainage basins (use cases 2 and 3 of the list above).
+Development of icepack is ongoing and we will broaden our efforts to encompass more use cases in future.
 
@@
 
-The evolution equations for ice thickness and velocity are necessary for any simulation, both other fields with their own dynamics can be a part of the problem as well.
+The evolution equations for ice thickness and velocity are necessary for any simulation, but other fields with their own dynamics can be a part of the problem as well.
 For example, the rheology and friction are the main coefficients in the diagnostic equation.
@@
 The mass transport equation for ice thickness is a free boundary problem, where the free boundary is the contour between ice-covered and ice-free regions \citep{schoof2013ice}.
-A naively-implemented prognostic solver could erroneously compute negative thickness values in subsequent timesteps when there is ablation in ice-free regions.
-A common and ad-hoc approach to work around this issue is to truncate the thickness at zero at every timestep.
-The principled approach is to instead treat the free boundary problem directly as a variational inequality \citep{jouvet2012steady}.
-Icepack currently lacks a principled scheme for tracking this free boundary and we instead rely on truncation.
-Treating this problem as a variational inequality in icepack will be the subject of future development.
+Where there is ablation in ice-free regions, a correctly-implemented prognostic solver will compute a negative thickness, which is unphysical and requires correction.
+The prognostic solver in icepack truncates the thickness to zero, which approximates the free boundary, where there is ablation in ice-free regions.
+An implicit approach would instead treat the free boundary problem directly as a variational inequality \citep{jouvet2012steady}.
+Icepack currently lacks such a scheme and this will be the subject of future development.
 PETSc includes scalable solvers for variational inequalities \citep{bueler2020petsc} that are also available through Firedrake.
